### 2025-01-13 Figures/PaperFiguresextended.Rmd
Added code to create figures

---
title: "Figures"
author: "LD"
date: '2023-06-06'
output: html_document
editor_options: 
  chunk_output_type: console
chunk_output_type: console
---
```{r setup}
library(Seurat)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(cowplot)
library(ggbreak) 
library(tidyr)
library(patchwork)
library(ComplexUpset)
library(ggrepel)
library(Polychrome)
library(ComplexHeatmap)
library(gplots)
library(RColorBrewer)
library(randomcoloR)

dir_out <- "/home/common/data/output/projects/ENDO/E044/A033/"
dir.create(dir_out, recursive = TRUE)
setwd(dir_out)
```

In script
Replaced ProlifPeriSecFinal with MenstrualCyclePhase_main

```{r save downsampled data}
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")

# Downsample the number of cells per identity class
Idents(data) <- data$sample
data_ds <- subset(x = data, downsample = 200)
saveRDS(data_ds, "/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")

#Some plotting
DimPlot(data, group.by = "AnnotationMain", raster = FALSE)
DimPlot(data, group.by = "Symphony_Refined_A029", raster = FALSE)
DimPlot(data, group.by = "Symphony_Refined_A029_United", raster = FALSE)

#Save object subsetted for ScaiNet samples
ScaiNETsamples <- read.csv("/home/common/data/output/projects/ENDO/E041/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")$names
Idents(data) <- data$sample
subset <- subset(data, idents = ScaiNETsamples)
rm(data)
gc()
meta <- subset@meta.data
saveRDS(subset, "/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_ScaiNETsamples.rds")
rm(subset)
gc()

meta$AnnotationMain

```

Remove? TODO
```{r order of samples by menstrual cycle phase and pseudotime}
meta_final <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")

MenstrualCyclePhase_sampleOrder <- meta_final %>%
  arrange(MenstrualCyclePhase,epithelial.pseudotime) %>%
  distinct(sample,epithelial.pseudotime, MenstrualCyclePhase,CycleDay) %>%
  pull(sample)
saveRDS(MenstrualCyclePhase_sampleOrder, paste0(dir_out,"sampleorder_MenstrualCyclePhase_greenCirclemedian.RDS"))
```

Add additional data to the Seurat metadata and reformat
```{r plotting of cell types per sample}
#Set directory links and create directories
dir_data <- "L:/GitHubDesktop/endometriosis_endometrium_scRNA_atlas/_Data/"
dir_out <- "L:/GitHubDesktop/endometriosis_endometrium_scRNA_atlas/Figures_out/"
if (!dir.exists(dir_out)) {
  dir.create(dir_out)
}

#Load data
#A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_metadata_ptime_snn_Final.rds")
meta <- readRDS(paste0(dir_data, "EndoAtlas_GEO_meta_complete.rds"))

##Modify menstrual cycle information
# Create column MenstrualCyclePhase_main based on MenstrualCyclePhase by merging the secretory sub-phases from 
meta$MenstrualCyclePhase_main <- as.character(meta$MenstrualCyclePhase)
meta$MenstrualCyclePhase_main[meta$MenstrualCyclePhase_main %in% c("secretory_early", "secretory_mid", "secretory_late")] <- "secretory"
#Factor samples by MenstrualCyclePhase_main for plotting
meta$MenstrualCyclePhase_main <- factor(meta$MenstrualCyclePhase_main, levels = c("proliferative","periovulatory","secretory"))

##Add additional metadata from manuscript Supplementary Table 1
# Read the tab-delimited text file
SupplementaryTable1 <- read.delim(paste0(dir_data, "SupplementaryTable1.txt"), header = TRUE, stringsAsFactors = FALSE)
# Perform a left join of the `meta` dataframe with `SupplementaryTable1`
meta <- dplyr::left_join(
  meta,
  SupplementaryTable1 %>% 
    dplyr::select(
      Patient.sample.ID,
      epithelial.pseudotime,
      sequencing.batch,
      Minor.exclusion.criteria.logical,
      Menstrual.cycle.phase.marker.analysis,
      Cell.type.frequency.analysis..Figure.2e.f.,
      DEG.analysis..Figure.3a.b.,
      Ligand.receptor.analysis..Figure.3c.h.
    ),
  by = c("sample" = "Patient.sample.ID")
)

#Double-check, TODO
table(meta$sample, meta$Minor.exclusion.criteria.logical)
prolifsamples <- meta%>%
  dplyr::filter(MenstrualCyclePhase_main == "proliferative") %>%
  select(sample) %>%
  distinct() %>%
  arrange(sample) %>%
  pull(sample)
DEGsamples <- unique(read.csv("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/celltype_freq_sample.csv")$Var2)
DEGsamples <- DEGsamples[order(DEGsamples)]
setdiff(prolifsamples,DEGsamples) #MinorExclusion samples overlap with samples excluded in DEG analysis, sample87 is late secretory

##Determine order of samples by MenstrualCyclePhase and epithelial.pseudotime
MenstrualCyclePhase_sampleOrder <- meta%>%
  select(sample, MenstrualCyclePhase,epithelial.pseudotime) %>%
  distinct() %>%
  arrange(MenstrualCyclePhase, epithelial.pseudotime) %>%
  mutate(MenstrualCyclePhase_sampleOrder = row_number()) %>%
  ungroup()
#Factor samples by MenstrualCyclePhase and epithelial.pseudotime for plotting
meta$sample <- factor(meta$sample, levels = c(MenstrualCyclePhase_sampleOrder$sample))

#Save color pals
qual_col_pals <- brewer.pal.info[brewer.pal.info$category == "qual", ]
col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
```

Supplementary Data Figure 4 a
```{r plotting of cell types per sample}
#########################################################################
# bar plot of the number of AnnotationRefined cells in each sample
df <- meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(sample, AnnotationMain, AnnotationRefined, EndometriosisGrade, MenstrualCyclePhase, epithelial.pseudotime, EndometriosisStatus) %>%
  summarise(row_count = n()) %>%
  ungroup()

#Bar plot
#####use this for paper supplement
#Bar plot split by main5 
#MenstrualCyclePhase_sampleOrder <- readRDS(paste0(dir_out,"sampleorder_MenstrualCyclePhase_greenCirclemedian.RDS"))
#df$sample <- factor(df$sample, levels = MenstrualCyclePhase_sampleOrder)
P63 = createPalette(63, c("#ff0000", "#00ff00", "#0000ff")) 
swatch(P63)
names(P63) <- NULL
#Make barplots with different seeds
seed <- 1 #find in dir.create(paste0(dir_out, "/seedopt/"))
set.seed(seed)  
p <- ggplot(df, aes(x = sample, y = row_count, fill = AnnotationRefined)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = sample(P63)) +
  xlab('Sample ID') +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
  ) +
  facet_grid(AnnotationMain ~ ., scales = "free_y")
ggsave(paste0(dir_out, "/seedopt/Barplot_sample_AnnotationRefined_main5_", seed, ".pdf"),
       plot = p,
       width = 15, height = 6)

##Consider leaving out
a1 <- ggplot(df %>% select(sample, EndometriosisStatus) %>% distinct (), aes(x = sample, y = EndometriosisStatus, fill = EndometriosisStatus)) +
  geom_bar(stat = 'identity', position = 'fill') +
  xlab('Sample ID') +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
  ) 
a2 <- ggplot(df %>% select(sample, MenstrualCyclePhase) %>% distinct (), aes(x = sample, y = MenstrualCyclePhase, fill = MenstrualCyclePhase)) +
  geom_bar(stat = 'identity', position = 'fill') +
  xlab('Sample ID') +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
  )
a1 + a2 + plot_layout(nrow = 2)
ggsave(paste0(dir_out, "Barplot_sample_AnnotationRefined_main5_annotation.pdf"),
  plot = last_plot(),
  width = 15, height = 6)

```

Figure 2 e
```{r plotting of cell types per sample}
##########################################################################
#BoxPlots sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main
df2 <- meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main, EndometriosisStatus) %>% #no sample, AnnotationMain combinations are dropped
  summarise(row_count = n()) %>%
  ungroup() %>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(sample) %>%
  mutate(AnnotationMain_perc = row_count / sum(row_count) * 100) %>%
  ungroup()

#BoxPlots percentage, EndometriosisStatus #for publication
ggplot(df2, aes(x = MenstrualCyclePhase_main, y = AnnotationMain_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge()) +
  RotatedAxis() +
  xlab('Sample ID') +
  facet_wrap(~AnnotationMain,scales = "free", ncol = 6)
ggsave(paste0(dir_out, "Boxplot_MenstrualCyclePhase_main_main5_sample_EndometriosisStatus_perc_publication.pdf"),
  plot = last_plot(),
  width = 12, height = 4.4)

###Significance testing CTL vs Endo
# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
df2_ttest_EndometriosisStatus <- df2 %>%
  group_by(AnnotationMain, MenstrualCyclePhase_main) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#epithelial, proliferative just not significant (p.value = 0.0898)

###Significance testing prolif vs periov vs sec, CTL + Endo
## Perform t-test Prolif vs Peri for each cell type
df2_ttest_ProlifPeri <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#none significant
## Perform t-test Prolif vs Sec for each cell type
df2_ttest_ProlifSec <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#epi (0.00977) and str (0.0255) significant

## Perform t-test periovulatory vs sec for each cell type
df2_ttest_PeriSec <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#epi (0.00794) and str (0.0137) significant
```

Supplementary Data Figure 4b, delete?
```{r plotting of cell types per sample}
##########################################################################################
##Plot differences for DEG analysis samples
#Extract sample names of samples used for DEG analysis
#DEGsamples <- unique(read.csv("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/celltype_freq_sample.csv")$Var2)

DEGsamples <- SupplementaryTable1 %>%
  filter(DEG.analysis..Figure.3a.b. == TRUE) %>%
  pull(Patient.sample.ID)

#Make df 
#Introduce 0 for the ones with 0 count!
df3 <- meta%>%
  group_by(sample, AnnotationRefined) %>%
  summarise(AnnotationRefined_counts = n()) %>%
  ungroup() %>%
  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
  filter(!is.na(AnnotationRefined)) %>%
  group_by(sample) %>%
  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
  ungroup()  %>% #60 samples * 63 cell types = 3780 rows, correct
  filter(sample %in% DEGsamples) #23 samples * 63 cell types = 1449 rows, correct
df3 <- left_join(df3, dplyr::distinct(meta[,c("sample","EndometriosisStatus")]), by = "sample")

# Perform t-test for each cell type
df3_ttest <- df3 %>%
  group_by(AnnotationRefined) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#No significant difference between the EndometriosisStatus in any cell type

###Plot
#BoxPlots percentage, EndometriosisStatus, percentage
ggplot(df3, aes(x = AnnotationRefined, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge(jitter.width = 0.2)) +
  facet_wrap(~AnnotationRefined, scales = "free", ncol = 7)
ggsave(paste0(dir_out, "Boxplot_DEGsamples_SymphonyRefinedFinal_EndometriosisStatus_perc.pdf"),
  plot = last_plot(),
  width = 17, height = 12)
```

Supplementary Data Figure 4b
```{r plotting of cell types per sample}
##########################################################################################
##Plot differences for samples without minor exclusion criteria

#Make df 
#Introduce 0 for the ones with 0 count!
df4 <- meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationRefined) %>%
  summarise(AnnotationRefined_counts = n()) %>%
  ungroup() %>%
  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
  filter(!is.na(AnnotationRefined)) %>%
  group_by(sample) %>%
  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
  ungroup() %>% #60 samples * 63 cell types = 3780 rows, correct
  filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
df4 <- left_join(df4, dplyr::distinct(meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main")]), by = "sample")
write.csv(df4, paste0(dir_out,"df4_AnnotationRefined_frequency.csv"))

# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
df4_ttest_EndometriosisStatus <- df4 %>%
  group_by(AnnotationRefined, MenstrualCyclePhase_main) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#Prv_VSMC secretory, secretory is the only significant one (pval 0.009075467)!!
df4_ttest_EndometriosisStatus$p.value_adjBH <- p.adjust(df4_ttest_EndometriosisStatus$p.value, method = "BH", n = length(df4_ttest_EndometriosisStatus$p.value))
#After correction for multiple testing Prv_VSMC secretory is not significant anymore

df4_ttest_EndometriosisStatus_secretory <- df4_ttest_EndometriosisStatus %>%
                      filter(MenstrualCyclePhase_main == "secretory")
df4_ttest_EndometriosisStatus_secretory$p.value_adjBH <- p.adjust(df4_ttest_EndometriosisStatus_secretory$p.value, method = "BH", n = length(df4_ttest_EndometriosisStatus_secretory$p.value))
#Even when subsetting secretory phase, after correction for multiple testing Prv_VSMC secretory is not significant anymore

## Perform t-test Prolif vs Peri for each cell type
df4_ttest_ProlifPeri <- df4 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
  group_by(AnnotationRefined) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#Few significant
## Perform t-test Prolif vs Sec for each cell type
df4_ttest_ProlifSec <- df4 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
  group_by(AnnotationRefined) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
write.csv(df4_ttest_ProlifSec, paste0(dir_out,"df4_ttest_ProlifSec.csv"))
df4_ttest_ProlifSec <- read.csv(paste0(dir_out,"df4_ttest_ProlifSec.csv"))
df4_ttest_ProlifSec$p.value_adjBH <- p.adjust(df4_ttest_ProlifSec$p.value, method = "BH", n = length(df4_ttest_ProlifSec$p.value))
write.csv(df4_ttest_ProlifSec, paste0(dir_out,"df4_ttest_ProlifSec_adjpval.csv"))

#About half are significant, show in paper supplement
## Perform t-test periovulatory vs sec for each cell type
df4_ttest_PeriSec <- df4 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
  group_by(AnnotationRefined) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#Less than half are significant, don't show in paper supplement

#BoxPlots percentage, EndometriosisStatus, percentage ###for paper Extended Data Fig. S4
ggplot(df4, aes(x = MenstrualCyclePhase_main, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge(jitter.width = 0.2)) +
  facet_wrap(~AnnotationRefined, scales = "free", ncol = 7) +
  theme_bw() +
  expand_limits(y = 0)

ggsave(paste0(dir_out, "Boxplot_DEGsamples_SymphonyRefinedFinal_EndometriosisStatus_perc.pdf"),
  plot = last_plot(),
  width = 20, height = 12)
```

Figure 2f
```{r plotting of cell types per sample}
#Early and Mid Secretory Prv_VSMCs 
df4 <- meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationRefined) %>%
  summarise(AnnotationRefined_counts = n()) %>%
  ungroup() %>%
  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
  filter(!is.na(AnnotationRefined)) %>%
  group_by(sample) %>%
  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
  ungroup() %>% #60 samples * 63 cell types = 3780 rows, correct
  filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
df4 <- left_join(df4, dplyr::distinct(meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main","MenstrualCyclePhase")]), by = "sample")

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main <- df4 %>%
  filter(AnnotationRefined %in% c("Prv_VSMC secretory", "Prv_VSMC secretory_late")) %>% #same result with and without "Prv_VSMC secretory_late"
  dplyr::filter(MenstrualCyclePhase %in% c("proliferative","periovulatory","secretory_mid","secretory_early")) %>%
  group_by(sample, EndometriosisStatus, MenstrualCyclePhase_main) %>%
  summarize(
    total_counts_MenstrualCyclePhase_main = sum(AnnotationRefined_counts, na.rm = TRUE),
    total_perc_MenstrualCyclePhase_main = sum(AnnotationRefined_perc, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(secEarylMid_Prv_VSMC_MenstrualCyclePhase_main, aes(x = MenstrualCyclePhase_main, y = total_perc_MenstrualCyclePhase_main, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge(jitter.width = 0.2))
ggsave(paste0(dir_out, "Boxplot_sec_Prv_VSMC_NOsecretory_late_total_perc_MenstrualCyclePhase_main_EndometriosisStatus.pdf"),
  plot = last_plot(),
  width = 17, height = 12)

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("secretory")) %>% 
  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#significant, p.value 0.0261 -> paper

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval_prolif <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative")) %>% 
  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#significant, p.value NA

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval_peri <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory")) %>% 
  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#significant, p.value 0.7639853

p.adjust(c(0.0261), method = "BH", n = length(3)) #Value does not change
p.adjust(c(0.7639853), method = "BH", n = length(3)) #Value does not change

#Calculate fold change ENDO vs nonENDO secretory
# Assuming secEarylMid_Prv_VSMC_MenstrualCyclePhase_main is your dataframe

library(dplyr)

# Filter the dataframe for 'secretory' in MenstrualCyclePhase_main and then group by EndometriosisStatus
mean_total_perc <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
  filter(MenstrualCyclePhase_main == "secretory") %>%
  group_by(EndometriosisStatus) %>%
  summarize(mean_total_perc = mean(total_perc_MenstrualCyclePhase_main))
```

Figure 1d
```{r plotting of cell types per sample}
p5 <- ggplot(df5, aes(x = AnnotationMain, y = row_count)) +
  geom_col() +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip()
p5
ggsave(paste0(dir_out, "Barplot_EndometriosisGrade_main5.pdf"),
  plot = last_plot(),
  width = 15, height = 6)

#main5 boxplot 
df6 <- meta%>%
  group_by(sample, AnnotationMain, EndometriosisGrade) %>%
  summarise(row_count = n()) %>%
  ungroup()
p6 <- ggplot(df6, aes(x = AnnotationMain, y = row_count)) +
  geom_boxplot() +
  #geom_jitter(position = position_jitter(width = 0.2, height = 0), size = 0.5) +
  scale_y_log10() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
p6
ggsave(paste0(dir_out, "Boxplot_EndometriosisGrade_main5_colEndometriosisGrade.pdf"),
         plot = last_plot(),
         width = 15, height = 6)

# bar plot of the number of AnnotationRefined cells in each sample
df7 <- meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(AnnotationMain, EndometriosisGrade) %>%
  summarise(row_count = n()) %>%
  ungroup() 
ggplot(df7, aes(x = AnnotationMain, y = row_count, fill = EndometriosisGrade)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()
ggsave(paste0(dir_out, "Barplot_AnnotationMain_EndometriosisGrade.pdf"),
  plot = last_plot(),
  width = 6, height = 4)

```

UMAPs

epithelial UMAP
2b,2c,3b,3e,

atlas UMAP
S3f, S7a

Figure 1c
```{r}
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
table(A003A033$AnnotationRefined)
data <- AddMetaData(data,A003A033)
E031A004meta <- read.csv("/home/common/data/output/projects/ENDO/E044/A033/E031A004_proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated_meta.csv")
E031A004meta$cell_id <- paste(E031A004meta$sample, E031A004meta$cell.names, sep = "_")
rownames(E031A004meta) <- E031A004meta$cell_id 
data <- AddMetaData(data, E031A004meta[, "Annotation", drop = FALSE])

order <- A003A033 %>%
  select(AnnotationMain, AnnotationRefined) %>%
  filter(!is.na(AnnotationMain)) %>%
  distinct() %>%
  arrange(AnnotationRefined) %>%
  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
  arrange(AnnotationMain)
data$AnnotationRefined <- factor(x = data$AnnotationRefined, levels = order$AnnotationRefined)
p24 <- DimPlot(data, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
p24

```

Figure 2b
```{r}
#Figure 2b entire atlas
Fig2b1 <- DimPlot(data, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
Fig2b1
ggsave(filename = paste0(dir_out, 'Fig2b1.pdf'),
       plot = Fig2b1,
       width =20, height=9)

#Figure 2b epithelial cells
Fig2b2 <- DimPlot(data, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
Fig2b2
ggsave(filename = paste0(dir_out, 'Fig2b2.pdf'),
       plot = Fig2b2,
       width =20, height=9)

```

Supplementary Data Figure 1c
```{r}
SFig1c <- FeaturePlot(data, features = "CycleDay", raster = FALSE)
SFig1c
ggsave(filename = paste0(dir_out, 'SFig1c.pdf'),
       plot = SFig1c,
       width =20, height=9)
```

Supplementary Data Figure 1d
```{r}
SFig1d <- FeaturePlot(data, features = "ProgesteroneSerum", raster = FALSE)
SFig1d
ggsave(filename = paste0(dir_out, 'SFig1d.pdf'),
       plot = SFig1d,
       width =20, height=9)

```

Supplementary Data Figure 3e
/home/common/data/output/projects/ENDO/E044/A066/phase_sc_Fig3_Wang_epi_pseudot/evaluation/ #S3c/d
```{r plott umaps of epithelial cells}
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A062/E044A050_seed621_filteredWO_ciliatedMesothelialMUC5B.rds")
A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
table(A003A033$AnnotationRefined)
data <- AddMetaData(data,A003A033)
data <- AddMetaData(data, meta_ptime)
data$epithelial.pseudotime <- as.integer(data$epithelial.pseudotime)
data$Fig3_Wang_epi_pseudot[is.na(data$Fig3_Wang_epi_pseudot)] <- 1 #checked samples

#General Plotting
p1 <- FeaturePlot(data, features = "ProgesteroneSerum", split.by = "EndometriosisStatus")
p2 <- DimPlot(data, group.by = "EndometriosisStatus", split.by = "CyclePhase")
p3 <- DimPlot(data, split.by = "Clusters", group.by = "EndometriosisStatus")
#plot_grid(p1,p2,p3, ncol=1)
ggsave(filename = paste0(dir_out, 'EpiUMAP_general.pdf'),
       plot = plot_grid(p1,p2,p3, ncol=1),
       width =8 ,height=10)

#Plotting of mens cycle info
p1 <- DimPlot(data, group.by = "Fig3_Wang_epi_pseudot")
p2 <- DimPlot(data, group.by = "MenstrualCyclePhase")

#FeaturePlot(seurat_obj, features = "trajectory_day567_epithelial.pseudotime")
#plot_grid(p1,p2,p3,p4,p5,p6, ncol=3)

ggsave(filename = paste0(dir_out, 'EpiUMAP_PhasePrediction_w11.pdf'),
       plot = plot_grid(p1,p2, ncol=2),
       width =11 ,height=5)

```

```{r plotting umaps of entire dataset}
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
table(A003A033$AnnotationRefined)
data <- AddMetaData(data,A003A033)
E031A004meta <- read.csv("/home/common/data/output/projects/ENDO/E044/A033/E031A004_proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated_meta.csv")
E031A004meta$cell_id <- paste(E031A004meta$sample, E031A004meta$cell.names, sep = "_")
rownames(E031A004meta) <- E031A004meta$cell_id 
data <- AddMetaData(data, E031A004meta[, "Annotation", drop = FALSE])

#Plot clusters, batch,sample EndometriosisStatus, cycle phase
p1 <- DimPlot(data, reduction = "umap", group.by = "EndometriosisStatus", raster = FALSE)
p2 <- DimPlot(data, reduction = "umap", group.by = "batch", raster = FALSE)
p3 <- DimPlot(data, reduction = "umap", group.by = "sample", raster = FALSE) +
  theme(legend.text = element_text(size=4))+
  theme(legend.key.size = unit(0.2, 'cm'))+  guides(color = guide_legend(ncol=1) )
p4 <- DimPlot(data, reduction = "umap", group.by = "celline", raster = FALSE)
p5 <- DimPlot(data, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
p6 <- DimPlot(data, reduction = "umap", group.by = "AnnotationMain", raster = FALSE)
p7 <- DimPlot(data, reduction = "umap", group.by = "Symphony_Refined_A029", raster = FALSE)
p8 <- DimPlot(data, reduction = "umap", group.by = "CyclePhase", raster = FALSE)
p9 <- FeaturePlot(data, features = "ProgesteroneSerum", raster = FALSE)
ggsave(filename = paste0(dir_out, 'UMAPs_p1p9.pdf'),
       plot = plot_grid(p1,p2,p3,p4,p6,p7,p5,p8,p9, ncol=3),
       width =28 ,height=24)
ggsave(filename = paste0(dir_out, 'UMAPs_p1p9.png'),
       plot = plot_grid(p1,p2,p3,p4,p6,p7,p5,p8,p9, ncol=3),
       width =28 ,height=24)

p10 <- DimPlot(data, reduction = "umap", group.by = "AnnotationMain", raster = FALSE)
p11 <- FeaturePlot(data, reduction = "umap", features = "Symphony_Global_prob", raster = FALSE)
p12 <- DimPlot(data, reduction = "umap", group.by = "Symphony_Refined", raster = FALSE)
p13 <- FeaturePlot(data, reduction = "umap", features = "Symphony_Refined_prob", raster = FALSE)
p14 <- DimPlot(data, reduction = "umap", group.by = "AnnotationRefined", raster = FALSE)
ggsave(filename = paste0(dir_out, 'UMAPs_p10p14.pdf'),
       plot = plot_grid(p10,p11,p12,p13,p14, ncol=3),
       width =28 ,height=16)

p20 <- DimPlot(data, reduction = "umap", group.by = "batch", raster = FALSE)
p21 <- FeaturePlot(data, reduction = "umap", features = "ProgesteroneSerum", raster = FALSE)
p22 <- FeaturePlot(data, reduction = "umap", features = "CycleDay", raster = FALSE)
p23 <- DimPlot(data, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)

#Order data by main5
###In progress
order <- A003A033 %>%
  select(AnnotationMain, AnnotationRefined) %>%
  filter(!is.na(AnnotationMain)) %>%
  distinct() %>%
  arrange(AnnotationRefined) %>%
  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
  arrange(AnnotationMain)
data$AnnotationRefined <- factor(x = data$AnnotationRefined, levels = order$AnnotationRefined)
p24 <- DimPlot(data, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
p24
data$Fig3_Wang_epi_pseudot[is.na(data$Fig3_Wang_epi_pseudot)] <- 1 #checked samples
p25 <- DimPlot(data, reduction = "umap", group.by = "Fig3_Wang_epi_pseudot",na.value = "",  raster = FALSE)
p25
#Highlight cycling fibroblasts
Idents(data) <- data$Annotation
table(data$Annotation)
cyclingfibs <- WhichCells(data, idents = c("Cycling Fibroblasts"))
p26 <- DimPlot(data, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
p26

plot_grid(p20,p21,p22,p23, ncol=2)
ggsave(filename = paste0(dir_out, 'UMAPs_p20p23.pdf'),
       plot = plot_grid(p20,p21,p22,p23, ncol=2),
       width =24, height=18)
ggsave(filename = paste0(dir_out, 'UMAPs_p20p23.png'),
       plot = plot_grid(p20,p21,p22,p23, ncol=2),
       width =24, height=18)
ggsave(filename = paste0(dir_out, 'UMAPs_p24.pdf'),
       plot = p24,
       width =20, height=9)
ggsave(filename = paste0(dir_out, 'UMAPs_p24.png'),
       plot = p24,
       width =20, height=9)
ggsave(filename = paste0(dir_out, 'UMAPs_p25.pdf'),
       plot = p25,
       width =13, height=9)
ggsave(filename = paste0(dir_out, 'UMAPs_p25.png'),
       plot = p25,
       width =13, height=9)
ggsave(filename = paste0(dir_out, 'UMAPs_p26.pdf'),
       plot = p26,
       width =13, height=9)
ggsave(filename = paste0(dir_out, 'UMAPs_p26.png'),
       plot = p26,
       width =13, height=9)

#Plot objects with split.by sample
p15 <- DimPlot(data, split.by = "sample", raster = FALSE)
ggsave(plot = p15, filename = paste0(dir_out,'A003_samples.pdf'),width =45 ,height=5)

#Plot objects with split.by EndometriosisStatus
p16 <- DimPlot(data, split.by = "EndometriosisStatus", raster = FALSE)
ggsave(plot = p16, filename = paste0(dir_out,'A003_EndometriosisStatus.pdf'),width =15 ,height=5)

rm(data)
gc()
```

Supplementary Data Figure 3 g
```{r}
####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad

####### Convert .h5ad files to .h5seurat files
library(SeuratDisk)
Convert("endo-2022_epithelial.h5ad", dest = "h5seurat", assay = "RNA", overwrite = TRUE)

Tan_epithelial <- LoadH5Seurat(paste0("endo-2022_epithelial.h5seurat"), assays = "RNA")

Tan_epithelial <- ScaleData(object = Tan_epithelial, features = data@assays[["RNA"]]@data@Dimnames[[1]])

#Subset cells from the endometrium (exclude cells from endometriosis lesions)
Idents(Tan_epithelial) <- Tan_epithelial$sample_type_rename
Tan_epithelial <- subset(Tan_epithelial, idents = c("Ctrl", "EuE"))

# factor levels in better order (E10 and E11 don't have endometrium cells)
Tan_epithelial$sample <- factor(Tan_epithelial$sample, levels = c("C01", "C02", "C03", "E01", "E02", "E03", "E04", "E05", "E06", "E07", "E08", "E09"))

# Generate heatmap of the endometrium cells from Tan et al. 2022 with the same epithelial menstrual marker genes as in our manuscript Figure 2d, for comparison
feature_list_figure2 <- c("SFRP4", "LAMC2", "MMP11", "MMP7", "SPRY1", "TFPI2","STEAP4","ADAMTS8","CAPN6","SLC37A2","SCGB1D2", "MT1F", "MT1X","RIMKLB", "S100P", "DEFB1","G0S2","NNMT","GPX3", "PAEP")

heatmap <- DoHeatmap(
  object = Tan_epithelial,
  features = intersect(feature_list_figure2,names),
  cells = NULL,
  group.by = "sample",
  assay = "RNA",
  group.bar = TRUE) + 
  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
heatmap
ggsave(
    filename = paste0(dir_out, "Tan_epithelial_Fig2EpiMarkers.pdf"),
    plot = last_plot(),
    width = 12.3,
    height = 5)

#As control also plot menstrual marker genes from Wang et al. 2020 (DOI: 10.1038/s41591-020-1040-z)
feature_list_Wang <- c("PLAU", "MMP7", "THBS1", "CADM1", "NPAS3", "ATP1A1", "ANK3", "ALPL", "TRAK1", "SCGB1D2", "MT1F", "MT1X", "MT1E", "MT1G", "CXCL14", "MAOA", "DPP4", "NUPR1", "GPX3", "PAEP")

heatmap <- DoHeatmap(
  object = Tan_epithelial,
  features = intersect(feature_list_Wang, names),
  cells = NULL,
  group.by = "sample",
  assay = "RNA",
  group.bar = TRUE) + 
  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
heatmap
ggsave(
    filename = paste0(dir_out, "Tan_epithelial_WangEpiMarkers.pdf"),
    plot = last_plot(),
    width = 15,
    height = 15)

#The menstrual marker gene expression distribution of each sample looks comparable to our manuscript figure 2d, confirming that the different samples in Tan et al. 2022 express menstrual marker genes from different menstrual cycle phases

```

Supplementary Data Figure 3 h
```{r}
####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_stromal.h5ad

####### Convert .h5ad files to .h5seurat files
library(SeuratDisk)
Convert("endo-2022_stromal.h5ad", dest = "h5seurat", assay = "RNA", overwrite = TRUE)
Tan_stromal <- LoadH5Seurat(paste0("endo-2022_stromal.h5seurat"), assays = "RNA")

# 
FEATURES <- data@assays[["RNA"]]@data@Dimnames[[1]]
Tan_stromal <- ScaleData(object = Tan_stromal, features = FEATURES)

#Subset cells from the endometrium (exclude cells from endometriosis lesions)
Idents(Tan_stromal) <- Tan_stromal$sample_type_rename
Tan_stromal <- subset(Tan_stromal, idents = c("Ctrl", "EuE"))

# factor levels in better order (E10 and E11 don't have endometrium cells)
Tan_stromal$sample <- factor(Tan_stromal$sample, levels = c("C01", "C02", "C03", "E01", "E02", "E03", "E04", "E05", "E06", "E07", "E08", "E09"))

# Generate heatmap of the endometrium cells from Tan et al. 2022 with the same stromal menstrual marker genes as in our manuscript Figure 2d, for comparison
feature_list_figure2_stromal <- c("SFRP1", "H1-1", "H3C2", "WNT5A","CILP","BRINP1", "CFD","PTGDS","SLIT3","RPRM", "TIMP3", "IGFBP1","LEFTY2","LRRC15","LUM")

#Generate heatmap

heatmap <- DoHeatmap(
  object = Tan_stromal,
  features = intersect(feature_list_figure2_stromal,FEATURES),
  cells = NULL,
  group.by = "sample",
  assay = "RNA",
  group.bar = TRUE) + 
  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
heatmap
ggsave(
    filename = paste0(dir_out, "Tan_stromal_Fig2stromalMarkers.pdf"),
    plot = last_plot(),
    width = 12.3,
    height = 5)

#Wang genes
feature_list_Wang_stromal <- c('STC1', 'NFATC2', 'BMP2' ,'PMAIP1' ,'MMP11' ,'SFRP1', 'WNT5A' ,'ZFYVE21' ,'CILP' ,'SLF2' ,'MATN2', 'S100A4' ,'DKK1' ,'CRYAB', 'FOXO1', 'IL15', 'FGF7', 'LMCD1')

heatmap <- DoHeatmap(
  object = Tan_stromal,
  features = intersect(feature_list_Wang_stromal, FEATURES),
  cells = NULL,
  group.by = "sample",
  assay = "RNA",
  group.bar = TRUE) + 
  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
heatmap
ggsave(
    filename = paste0(dir_out, "Tan_stromal_WangstromalMarkers.pdf"),
    plot = last_plot(),
    width = 15,
    height = 15)

#The menstrual marker gene expression distribution of each sample looks comparable to our manuscript figure 2d, confirming that the different samples in Tan et al. 2022 express menstrual marker genes from different menstrual cycle phases

```

Supplementary Data Figure 1a and 1b
```{r VlnPlots of nCounts and nFeatures}
##VlnPlots for "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT","percent.mt"
VlnPlot(data,
  features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT","percent.mt"),
  group.by='EndometriosisStatus', ncol = 1, pt.size=0)
ggsave(paste0(dir_out, "VlnPlot_GroupByEndometriosisStatus_nFeature_nCounts_RNA_SCT_percMT.pdf"),
  plot = last_plot(),
  width = 4, height = 13)

##VlnPlots grouped by sample of "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT","percent.mt" 
# default level sort order is alphabetical, therefore refactor before plotting
data[["EndometriosisStatus"]] <- factor(unlist(data[["EndometriosisStatus"]])) 
#VlnPlot
VlnPlot(data,
  features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT","percent.mt"),
  group.by='sample', ncol = 1, pt.size=0)
ggsave(paste0(dir_out, "VlnPlot_GroupBySample_nFeature_nCounts_RNA_SCT_percMT.pdf"),
  plot = last_plot(),
  width = 25, height = 15)
#Plot VlnPlots with boxplots and save them
VlnPlot(data, features = "nCount_RNA", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
VlnPlot(data, features = "nFeature_RNA", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
VlnPlot(data, features = "nCount_SCT", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
VlnPlot(data, features = "nFeature_SCT", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
VlnPlot(data, features = "percent.mt", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)

```

PCA 
A072 1e, 2a, S1e, S3a
A082 S1f

```{r different counts and means/medians}
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_metadata.rds")

#Percentage of main 5
table <- as.data.frame(table(data$AnnotationMain))
table$Percentage <- round((table$Freq / sum(table$Freq)) * 100, 1)

#Number of cells for CTL and Endo
table(data$EndometriosisStatus)

#Number of patients for CTL and Endo
data %>% select(sample,EndometriosisStatus) %>%
  distinct() %>%
  group_by(EndometriosisStatus) %>%
  summarize(number_of_rows = n())

#mean number of cells per sample
mean(table(data$sample))

#median count/feature
median_summary <- data %>%
  select(nCount_RNA, nCount_SCT, nFeature_RNA, nFeature_SCT) %>%
  summarize(
    median_nCount_RNA = median(nCount_RNA),
    median_nCount_SCT = median(nCount_SCT),
    median_nFeature_RNA = median(nFeature_RNA),
    median_nFeature_SCT = median(nFeature_SCT)
  )

```

Supplementary Data Figure 2
```{refined cluster markers}
#Load data
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
meta <- data@meta.data

#Set SCT as default and remove RNA to save space
DefaultAssay(data) <- "SCT"
data@assays$RNA <- NULL
gc()

#Remove data without annotation (no NAs!)
Idents(data) <- data$Symphony_Refined
data <- subset(data, subset = Symphony_Refined %in% levels(as.factor(data$Symphony_Refined)))
gc()

#setdiff(pops_to_plot,unique(data$Symphony_Refined))
######################################################
#Tan2022 Sup Fig. 7c, epithelial
genes_to_plot <- c("TPPP3", "PIFO", "FOXJ1", "TP63", "KRT5", "SOX9", "LGR5", "MUC5B", "TFF3", "RUNX3", "SAA1", "SIX1", "PROM1", "PROM2", "ANPEP", "KIAA1324", "SPDEF", "SMAD9", "CD36", "HSD17B2", "SCGB2A2", "GDA", "FGFR2", "WNT7A", "FGF9", "PTGS1", "MSLN", "UPK3B", "CALB2", "PGR", "ESR1", "MMP7", "TIMP1", "IDO1", "PSAT1", "ENPP3", "GNG11", "CREB3L1", "IHH", "TOP2A", "MT1F", "MT1E", "MT1X", "SERPINA5", "SPP1", "PAEP", "CXCL14", "DPP4")
pops_to_plot <- c("mesothelial", "ciliated", "mid-secretory", "lumenal 2", "lumenal 1", "lumenal", "glandular early-secretory", "glandular", "TP63+/KRT5+", "MUC5B+")

# reorder clusters
pops_seurat <- subset(data, idents = pops_to_plot)
pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
                            levels=pops_to_plot)
p <- Seurat::DotPlot(
  object = pops_seurat,
  scale = T,
  features = genes_to_plot,
  idents = pops_to_plot,
) + theme(axis.text.x = element_text(angle = 90))
ggsave(plot = p, "./dotplots/Tan2022Fig3b_epithelial_SCT.pdf", width = 17, height = 7)
######################################################
#DotPlot of features from Tan2022 Fig. 4b, Lymphocytes
genes_to_plot <- c("FCGR3A","KLRD1","ITGA1","LILRB1","ITGAX","KLRC1","KLRC2","EPAS1","KIR2DL4","IGFBP2","CD160","GZMH","KLRF1","FGFBP2","TBX21","CD8A","CD8B","CRTAM","CCR5","NCR3","RORC","IL7R","TCF7","CD4","CCR7","SELL","LEF1","FOXP3","CTLA4","IL2RA","CXCR6","ZNF683","ITGAE","XCL1","SLC4A10","DPP4","GZMK","IGKC", "MZB1", "DERL3", "CD79A", "CD79B", "MS4A1", "BANK1")
pops_to_plot <- c(meta %>% filter(AnnotationMain == "lymphocyte") %>% pull(Symphony_Refined) %>% unique())

# reorder clusters
pops_seurat <- subset(data, idents = pops_to_plot)
pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
                           levels=c("B cell", "plasma", "CD8 T$_{RM}$","CD4 T$_{RM}$","T$_{Reg}$", "T$_N$/T$_{CM}$" , "ILC" , "T$_{EM}$","CTL", "pNK", "NK3","NK2","NK1"))

p <- Seurat::DotPlot(
  object = pops_seurat,
  scale = T,
  features = genes_to_plot,
  idents = pops_to_plot,
) + theme(axis.text.x = element_text(angle = 90))
ggsave(plot = p, "./dotplots/Tan2022Fig3b_Lymphocytes_SCT.pdf", width = 17, height = 7)
######################################################
#Tan2022 Fig. 3b, stromal
genes_to_plot <- c("RERGL", "MCAM", "MYH11", "STEAP4","GGT5", "RGS5")
pops_to_plot <- c("Prv-MYH11", "Prv-CCL19", "Prv-STEAP4", "VSMC")
# reorder clusters
pops_seurat <- subset(data, idents = pops_to_plot)
pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
                            levels=c("VSMC", "Prv-STEAP4", "Prv-CCL19","Prv-MYH11"))

p <- Seurat::DotPlot(
  object = pops_seurat,
  scale = T,
  features = genes_to_plot,
  idents = pops_to_plot,
) + theme(axis.text.x = element_text(angle = 90))
ggsave(plot = p, "./dotplots/Tan2022Fig3b_stromal_SCT.pdf", width = 7, height = 7)
######################################################
#Tan2022 Fig. 4b, macrophages
genes_to_plot <- c("FLOR2", "LYVE1", "MRC1", "CX3CR1", "ICAM2", "CLEC5A", "CCR2", "VEGFA", "FN1", "SPARC", "APOE", "SPP1")
pops_to_plot <- c(meta %>% filter(AnnotationMain == "myeloid") %>% pull(Symphony_Refined) %>% unique())
pops_to_plot <- c("M$\\Phi$5-activated","M$\\Phi$4-infiltrated","M$\\Phi$3-APOE","M$\\Phi$2-peritoneal","M$\\Phi$1-LYVE1")

# reorder clusters
pops_seurat <- subset(data, idents = pops_to_plot)
pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
                            levels=pops_to_plot)
p <- Seurat::DotPlot(
  object = pops_seurat,
  scale = T,
  features = genes_to_plot,
  idents = pops_to_plot,
) + theme(axis.text.x = element_text(angle = 90))
ggsave(plot = p, "./dotplots/Tan2022Fig3b_macrophages_SCT.pdf", width = 10, height = 7)
######################################################
#Tan2022 Fig. 5a, DCs
genes_to_plot <- c("CD1C", "CLEC9A", "LAMP3")
pops_to_plot <- c("mDC", "cDC1", "DC3", "cDC2", "pre-cDC2")
# reorder clusters
pops_seurat <- subset(data, idents = pops_to_plot)
pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
                            levels=pops_to_plot)
p <- Seurat::DotPlot(
  object = pops_seurat,
  scale = T,
  features = genes_to_plot,
  idents = pops_to_plot,
) + theme(axis.text.x = element_text(angle = 90))
ggsave(plot = p, "./dotplots/Tan2022Fig3b_DCs_SCT.pdf", width = 5, height = 7)

rm(data)
gc()

```

```{major cluster markers}
#Load data
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")

#Set SCT as default and remove RNA to save space
DefaultAssay(data) <- "SCT"
data@assays$RNA <- NULL
gc()

#Remove data without annotation
Idents(data) <- data$AnnotationMain
data <- subset(data, subset = AnnotationMain %in% levels(as.factor(data$AnnotationMain)))

GENES <- c("EPCAM","CDH1","PDGFRA","VWF","PECAM1","PTPRC","CD68","CD14","CD2")
p <- VlnPlot(data, features = GENES, group.by = "AnnotationMain",raster=FALSE, pt.size = 0, assay = "SCT", stack=T, flip=T)
ggsave(plot = p, "./VlnPlotMain5_SCT_notds.pdf", width = 6, height = 4.5)

GENES <- c("PDGFRA","DCN","COL1A1","COL6A1","CRISPLD2","COL6A3","LUM","COL5A1") #stomal genes, only "PDGFRA" and "CRISPLD2" are cleanly expressed in stromal only, the others are a bit mixed
p <- VlnPlot(data, features = GENES, group.by = "AnnotationMain",raster=FALSE, pt.size = 0, assay = "RNA", stack=T, flip=T)
ggsave(plot = p, "./VlnPlotMain5_stromalMG_RNA_notds.pdf", width = 6, height = 4.5)

p <- VlnPlot(data, features = GENES, group.by = "AnnotationMain",raster=FALSE, pt.size = 0, assay = "SCT", stack=T, flip=T)
ggsave(plot = p, "./VlnPlotMain5_stromalMG_SCT_notds.pdf", width = 6, height = 4.5)

GENES <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2")
p <- VlnPlot(data, features = GENES, group.by = "AnnotationMain",raster=FALSE, pt.size = 0, assay = "RNA", stack=T, flip=T)
ggsave(plot = p, "./VlnPlotMain5_RNA_notds.pdf", width = 6, height = 4.5)

table(data$AnnotationMain)

#Make DotPlots
DefaultAssay(data) <- "SCT"

GENES

######################################################
#Tan2022 main markers
genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2")
pops_to_plot <- c("stromal","epithelial","myeloid","lymphocyte","endothelial")

Seurat::DotPlot(
  object = data,
  scale = T,
  features = genes_to_plot,
  idents = pops_to_plot,
) + theme(axis.text.x = element_text(angle = 90))
ggsave(plot = last_plot(), "./dotplots/main5_main5markers_SCT.pdf", width = 10, height = 5)

#stomal markers
genes_to_plot <- c("PDGFRA","DCN","COL1A1","COL6A1","CRISPLD2","COL6A3","LUM","COL5A1") #stomal genes, only "PDGFRA" and "CRISPLD2" are cleanly expressed in stromal only, the others are a bit mixed
Seurat::DotPlot(
  object = data,
  scale = T,
  features = genes_to_plot,
  idents = pops_to_plot,
) + theme(axis.text.x = element_text(angle = 90))
ggsave(plot = last_plot(), "./dotplots/main5_stromalmarkers_SCT.pdf", width = 10, height = 5)

#all markers + stomal
genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL5A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2","DCN","COL6A1","CRISPLD2","COL6A3","LUM","COL1A1") #stomal genes, only "PDGFRA" and "CRISPLD2" are cleanly expressed in stromal only, the others are a bit mixed
Seurat::DotPlot(
  object = data,
  scale = T,
  features = genes_to_plot,
  idents = pops_to_plot,
) + theme(axis.text.x = element_text(angle = 90))
ggsave(plot = last_plot(), "./dotplots/main5_allmarkers_SCT.pdf", width = 17, height = 5)

```

```{r barplot distribution of cv splits, all cells, prolif}
meta <- read.csv("/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")

library(reshape2)
library(ggplot2)
library(forcats)

# reshape df
melted_meta <- melt(meta, id.vars = c("lane", "endo_EndometriosisGrade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
## create the bar plot for endo_EndometriosisGrade
ggplot(melted_meta, aes(x = variable, fill = endo_EndometriosisGrade)) +
  geom_bar(position = "stack", width = 0.7) +
  facet_wrap(~ fct_rev(value), ncol = 2) +
  labs(title = "Bar Plot of Samples by Endo EndometriosisGrade",
       x = "Fold",
       y = "Number of Samples") +
  scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
  theme_minimal() +
  coord_flip()
ggsave(paste0(dir_out,"Bar Plot of Samples by Endo EndometriosisGrade_allcells.pdf"), plot = last_plot(), height = 6.3, width = 7.2)

## create the bar plot for lane
ggplot(melted_meta, aes(x = variable, fill = lane)) +
  geom_bar(position = "stack", width = 0.7) +
  facet_wrap(~ fct_rev(value), ncol = 2) +
  labs(title = "Bar Plot of Samples by lane",
       x = "Fold",
       y = "Number of Samples") +
  #scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
  theme_minimal() +
  coord_flip()
ggsave(paste0(dir_out,"Bar Plot of Samples by lane_allcells.pdf"), plot = last_plot(), height = 6.3, width = 7.2)

##Upsetplot of eval samples
df <- meta %>%
  rownames_to_column(var = "name") %>%
  select(-name) %>% 
  select(fold1, fold2, fold3, fold4, fold5) %>%
  mutate_all(~ ifelse(. == 'eval', TRUE, FALSE)) %>%
  filter(rowSums(. == TRUE) > 0)
pdf(file.path(dir_out, "upset_CellCnnsig_all_evalsamples.pdf"))
upset(df,
  colnames(df)) +
  ggtitle("all cells")
dev.off()

################
#heatmap, not working yet

# Extracting the necessary columns from the 'meta' dataframe
data_matrix <- meta[, c("fold1", "fold2", "fold3", "fold4", "fold5")]
data_matrix_numeric <- apply(data_matrix, 2, function(x) as.numeric(x == "eval"))

# Adding annotation columns
annotation_data <- meta[, c("EndometriosisStatus", "lane", "batch", "endo_EndometriosisGrade")]

library(ComplexHeatmap)
library(circlize)

# Define a custom color palette
my_color_palette <- colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
png(file=paste0(dir_out,'heatmap.pdf'))
# Create the heatmap using ComplexHeatmap
heatmap <- Heatmap(
  data_matrix_numeric,
  col = my_color_palette,
  row_dend_reorder = TRUE,  # Adjust this based on your preference
  column_title = "Folds",
  name = "Fold Value",
  show_row_names = FALSE,
  show_column_names = TRUE,
  cluster_columns = FALSE
)

draw(heatmap)
dev.off()
```

```{r barplot distribution of cv splits, cf, prolif}
meta <- read.csv("/home/common/data/output/projects/ENDO/E041/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")

library(reshape2)
library(ggplot2)
library(forcats)

# reshape df
melted_meta <- melt(meta, id.vars = c("lane", "endo_EndometriosisGrade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
# create the bar plot for endo_EndometriosisGrade
ggplot(melted_meta, aes(x = variable, fill = endo_EndometriosisGrade)) +
  geom_bar(position = "stack", width = 0.7) +
  facet_wrap(~ fct_rev(value), ncol = 2) +
  labs(title = "Bar Plot of Samples by Endo EndometriosisGrade",
       x = "Fold",
       y = "Number of Samples") +
  scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
  theme_minimal() +
  coord_flip()
ggsave(paste0(dir_out,"Bar Plot of Samples by Endo EndometriosisGrade_cf.pdf"), plot = last_plot(), height = 6.3, width = 7.2)

# create the bar plot for lane
ggplot(melted_meta, aes(x = variable, fill = lane)) +
  geom_bar(position = "stack", width = 0.7) +
  facet_wrap(~ fct_rev(value), ncol = 2) +
  labs(title = "Bar Plot of Samples by lane",
       x = "Fold",
       y = "Number of Samples") +
  #scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
  theme_minimal() +
  coord_flip()
ggsave(paste0(dir_out,"Bar Plot of Samples by lane_cf.pdf"), plot = last_plot(), height = 6.3, width = 7.2)

##Upsetplot of eval samples
df <- meta %>%
  rownames_to_column(var = "name") %>%
  select(-name) %>% 
  select(fold1, fold2, fold3, fold4, fold5) %>%
  mutate_all(~ ifelse(. == 'eval', TRUE, FALSE)) %>%
  filter(rowSums(. == TRUE) > 0)
pdf(file.path(dir_out, "upset_CellCnnsig_cf_evalsamples.pdf"))
upset(df,
  colnames(df)) +
  ggtitle("all cells")
dev.off()

################
#heatmap, not working yet

# Extracting the necessary columns from the 'meta' dataframe
data_matrix <- meta[, c("fold1", "fold2", "fold3", "fold4", "fold5")]
data_matrix_numeric <- apply(data_matrix, 2, function(x) as.numeric(x == "eval"))

# Adding annotation columns
annotation_data <- meta[, c("EndometriosisStatus", "lane", "batch", "endo_EndometriosisGrade")]

library(ComplexHeatmap)
library(circlize)

# Define a custom color palette
my_color_palette <- colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
png(file=paste0(dir_out,'heatmap.pdf'))
# Create the heatmap using ComplexHeatmap
heatmap <- Heatmap(
  data_matrix_numeric,
  col = my_color_palette,
  row_dend_reorder = TRUE,  # Adjust this based on your preference
  column_title = "Folds",
  name = "Fold Value",
  show_row_names = FALSE,
  show_column_names = TRUE,
  cluster_columns = FALSE
)

draw(heatmap)
dev.off()
```

```{r DimPlot with cycling fibroblasts}
data <- readRDS("/home/common/data/output/projects/ENDO/E031/A004/proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated.RDS")
DimPlot(data, raster = FALSE)
meta <- data@meta.data
write.csv(meta, paste0(dir_out,"E031A004_proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated_meta.csv"))

# make dotplot
features <- c("NFKBIA", "SLPI", "CXCL2","CXCL3")
Idents(data) <- data$Annotation

epithelial <- subset(data, idents = "Unciliated")

DotPlot(
  epithelial,
  assay = "SCT",
  features,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  idents = NULL,
  group.by = "EndometriosisStatus",
  split.by = NULL,
  cluster.idents = FALSE,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA
)

```

```{r Expression distribution of all genes and pca signature}
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")

#CellCnn gene signature
#Summary table for cyc fibro corrgpca: 
CellCnnsig_fibro_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E041/A010/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
#Summary table for  cyc fibro pca :  
CellCnnsig_fibro_pca <- read.csv("/home/common/data/output/projects/ENDO/E041/A012/sigs/weighted_score_sig_pca_1pc.csv") 
#Summary table for  all cells corrgpca: 
CellCnnsig_all_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E042/A011/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
#Summary table for  all cells pca: 
CellCnnsig_all_pca <- read.csv("/home/common/data/output/projects/ENDO/E042/A012/sigs/weighted_score_sig_pca_1pc.csv") 
CellCnnsig <- unique(c(CellCnnsig_all_corrgpca$gene, CellCnnsig_all_pca$gene))
ScaiNETsamples <- read.csv("/home/common/data/output/projects/ENDO/E041/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")

#Make UpSet plot
# Create table with the gene sets
gene_sets <- list(
    PCA = CellCnnsig_all_pca$gene,
    CorrgPCA = CellCnnsig_all_corrgpca$gene)
all_genes <- unique(unlist(gene_sets))
df <- as.data.frame(lapply(gene_sets, function(gene_set) {
  setNames(as.logical(all_genes %in% gene_set), all_genes)
}))
# Make upset plot
pdf(file.path(dir_out, "upset_CellCnnsig_all_pca_corrgpcs.pdf"))
upset(df,
  colnames(df),
  set_sizes=(
        upset_set_size(position='right'))) +
  ggtitle("all cells")
dev.off()

#CTL and Endo samples used for the DEG
#Filter for samples with >50 cells/sample
data_metadata <- data@meta.data
df_sample_cells <- data@meta.data %>% 
  dplyr::group_by(sample, EndometriosisStatus) %>% 
  dplyr::summarize(cells = n()) %>% 
  dplyr::filter(cells >= 50) %>% 
  dplyr::filter(sample %in% ScaiNETsamples$names) 
ENDO <- df_sample_cells %>% filter(EndometriosisStatus == "Endo") %>%
  select(sample) %>%
  pull(sample)
CTL <- df_sample_cells %>% filter(EndometriosisStatus == "CTL") %>%
  select(sample) %>%
  pull(sample)
#Retrieve number of Endo/CTL samples with RNA expression
exprmatrix_data <- AggregateExpression(
  data,
  assays = "RNA",
  features = NULL,
  return.seurat = FALSE,
  group.by = c("sample"),
  slot = "counts",
  verbose = TRUE)[["RNA"]]
exprmatrix_data <- as.data.frame(exprmatrix_data)
exprmatrix_data$gene <- row.names(exprmatrix_data)
exprmatrix_data <- exprmatrix_data %>%
  mutate(Endo_expressed_n = rowSums(select(., all_of(ENDO)) > 0)) %>%
  mutate(CTL_expressed_n = rowSums(select(., all_of(CTL)) > 0)) %>%
  mutate(samples_expressed_n = Endo_expressed_n + CTL_expressed_n) %>%
  mutate(ENDOorCTL_max_expressed_n = pmax(Endo_expressed_n, CTL_expressed_n))
table(exprmatrix_data$ENDOorCTL_max_expressed_n)

exprmatrix_CellCnnsig <- exprmatrix_data %>%
  filter(row.names(exprmatrix_data) %in% CellCnnsig$gene)
table(exprmatrix_CellCnnsig$ENDOorCTL_max_expressed_n)

exprmatrix_data <- exprmatrix_data %>%
  mutate(CellCnnsig = row.names(exprmatrix_data) %in% CellCnnsig$gene)

# Create a histogram
ggplot(exprmatrix_data, aes(x = samples_expressed_n, y = after_stat(count))) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  facet_wrap(~ CellCnnsig, ncol = 2, scales = "free_y") +
  labs(x = "number of samples with gene expression", y = "number of genes expressed") +
  theme_minimal() +
  geom_text(
    aes(label = after_stat(count)),
    stat = "count",
    vjust = -0.5,
    size = 2,
    position = position_stack()
  )

################################ in work
# Assuming your data is stored in a data frame named 'CellCnnsig_all_pca'
library(ggplot2)

color_scale <- scale_fill_gradient(low = "red", high = "green")

heatmap_plot <- ggplot(CellCnnsig_all_pca, aes(x = X, y = gene, fill = rank)) +
  color_scale +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_plot <- heatmap_plot +
  geom_text(aes(label = gene), vjust = 1, hjust = 1, size = 3)

# Show the plot
print(heatmap_plot)
#####################################
```

```{r Expression distribution of all genes and pca signature for cf}
#CellCnn gene signature
#Summary table for cyc fibro corrgpca: 
CellCnnsig_fibro_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E041/A010/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
#Summary table for  cyc fibro pca :  
CellCnnsig_fibro_pca <- read.csv("/home/common/data/output/projects/ENDO/E041/A012/sigs/weighted_score_sig_pca_1pc.csv") 
CellCnnsig <- unique(c(CellCnnsig_fibro_corrgpca$gene, CellCnnsig_fibro_pca$gene))

#Make UpSet plot
# Create table with the gene sets
gene_sets <- list(
    PCA = CellCnnsig_fibro_pca$gene,
    CorrgPCA = CellCnnsig_fibro_corrgpca$gene)
all_genes <- unique(unlist(gene_sets))
df <- as.data.frame(lapply(gene_sets, function(gene_set) {
  setNames(as.logical(all_genes %in% gene_set), all_genes)
}))
# Make upset plot
pdf(file.path(dir_out, "upset_CellCnnsig_fibro_pca_corrgpcs.pdf"))
upset(df,
  colnames(df),
  set_sizes=(
        upset_set_size(position='right'))) +
  ggtitle("fibro")
dev.off()
```

```{r Plot DEGs from Tan proliferative, vulcano and DotPlot}
#RNA assay DEGs from proliferative samples
RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv") %>%
  filter(annotation_reference == "Tan") %>%
  filter(AveExpr > 18.53202)
##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes, there are no hb genes in the DEG list
mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes, no genes
rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes, 165 genes
RNAlistRDS005_df <- RNAlistRDS005_df %>%
  dplyr::filter(!gene %in% rb.genes) #Focus on protein coding genes
##Rename cluster_id to major cell types
meta <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_metadata.rds")
meta$Symphony_Refined[meta$Symphony_Refined == "M$\\Phi$1-LYVE1"] <- "M-Phi-1-LYVE1"
meta$Symphony_Refined[meta$Symphony_Refined == "glandular early-secretory"] <- "glandular-early-secretory"
meta <- meta %>% 
  select(AnnotationMain,Symphony_Refined,Symphony_Refined_A029_United) %>%
  distinct()
RNAlistRDS005_df <- RNAlistRDS005_df %>%
  mutate(AnnotationMain = cluster_id)
RNAlistRDS005_df$AnnotationMain <- case_when(
  RNAlistRDS005_df$AnnotationMain %in% meta$Symphony_Refined ~ meta$AnnotationMain[match(RNAlistRDS005_df$AnnotationMain, meta$Symphony_Refined)],
  RNAlistRDS005_df$AnnotationMain %in% meta$Symphony_Refined_A029_United ~ meta$AnnotationMain[match(RNAlistRDS005_df$AnnotationMain, meta$Symphony_Refined_A029_United)],
  TRUE ~ RNAlistRDS005_df$AnnotationMain)
RNAlistRDS005_df$AnnotationMain[RNAlistRDS005_df$AnnotationMain == "myeloid" |
                                      RNAlistRDS005_df$AnnotationMain == "lymphocyte"] <- "immune"

###Plot DEGs
#vulcano plot without gene labels
ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc), color = stage)) +
  geom_point() +
  theme_minimal() +
  #ylim(-2, 17) +
  xlim(-1, 1.2)
ggsave(paste0(dir_out,"VulcanoPlot_DEGs_stage.pdf"), 
       plot = last_plot(), 
       height = 6.3, width = 5)

#vulcano plot with gene labels
GenesToLabel <- RNAlistRDS005_df %>% 
               #distinct(gene, logFC,AveExpr,p_adj.loc, .keep_all = TRUE) %>% 
                dplyr::filter(abs(logFC) >= 0.5)
ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc), color = stage)) +
  geom_point() +
  theme_minimal() +
  #ylim(-2, 17) +
  xlim(-1, 1.2) +
  geom_text_repel(data = GenesToLabel,
                  aes(label = gene),
                  size = 2,
                  max.overlaps = 50,
                  box.padding = 0.35,
                  point.padding = 0.5,
                  segment.color = 'grey50')
#Barplot unique genes per main cell type
RNAlistRDS005_df %>% 
  select(AnnotationMain,gene,stage,cluster_id) %>%
  ggplot(aes(x = stage, fill = cluster_id)) +
  geom_bar() +
  facet_wrap(~ AnnotationMain, ncol = 4) +
  theme(axis.text.x = element_text(angle=45, vjust=.2, hjust=0.3))

#Barplot unique genes per main cell type
RNAlistRDS005_df %>% 
  select(AnnotationMain, gene, stage, cluster_id) %>%
  ggplot(aes(x = stage, fill = factor(AnnotationMain, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.3))

#Barplot unique genes per main cell type
RNAlistRDS005_df %>% 
  select(AnnotationMain, gene, stage) %>%
  distinct() %>%
  ggplot(aes(x = stage, fill = factor(AnnotationMain, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.3))
ggsave(paste0(dir_out,"BarPlot_DEGs_main4.pdf"), 
       plot = last_plot(), 
       height = 4, width = 10)

RNAlistRDS005_df %>% 
      select(AnnotationMain, gene, stage) %>%
      distinct() %>%
      count(stage, AnnotationMain)

RNAlistRDS005_df %>% 
      filter(logFC >= 0.5) %>%
      distinct(gene, stage) %>%
      count(stage)

RNAlistRDS005_df %>% 
      select(gene, stage) %>%
      distinct() %>%
      count(stage)

table <- RNAlistRDS005_df %>% 
      select(AnnotationMain, gene, stage) %>%
      distinct() %>%
      count(AnnotationMain, stage)

```

```{r DEGs vs CellCNN gene signature}
##Load dataframes
#CellCnn gene signature
#Summary table for cyc fibro corrgpca: 
CellCnnsig_fibro_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E041/A010/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
#Summary table for  cyc fibro pca :  
CellCnnsig_fibro_pca <- read.csv("/home/common/data/output/projects/ENDO/E041/A012/sigs/weighted_score_sig_pca_1pc.csv") 
#Summary table for  all cells corrgpca: 
CellCnnsig_all_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E042/A011/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
CellCnnsig_all_corrgpca$origin <- "A011corrgpca"
#Summary table for  all cells pca: 
CellCnnsig_all_pca <- read.csv("/home/common/data/output/projects/ENDO/E042/A012/sigs/weighted_score_sig_pca_1pc.csv") 
CellCnnsig_all_pca$origin <- "A012pca"

CellCnnsig <- rbind(CellCnnsig_all_corrgpca, CellCnnsig_all_pca)

#RNA assay DEGs from proliferative samples
RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")

##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
#Haemoglobin genes
hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes
CellCnnsig[CellCnnsig$gene %in% hb.genes, ]#no hb genes
RNAlistRDS005_df[RNAlistRDS005_df$gene %in% hb.genes, ]#no hb genes
#mitochondrial genes
grep("^MT-", CellCnnsig$gene, value=TRUE)#no mt genes
grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#no mt genes
#ribosomal genes
rb.genes_CellCnnsig <- grep(pattern = "^RP[SL]", CellCnnsig$gene, value=TRUE)#3 rb genes
CellCnnsig <- CellCnnsig[!(CellCnnsig$gene %in% rb.genes_CellCnnsig), ]
rb.genes_DEGs <- unique(grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE))#77 rb genes
RNAlistRDS005_df <- RNAlistRDS005_df[!(RNAlistRDS005_df$gene %in% rb.genes_DEGs), ]

##Combine the two dataframes
RNAlistRDS005_df$CellCnnsig <- factor(RNAlistRDS005_df$gene %in% CellCnnsig$gene,
                                        levels = c(FALSE, TRUE),
                                        labels = c("Non-overlapping", "overlapping"))

##Make plots
# Create a scatter plot logFC vs AveExpr
ggplot(RNAlistRDS005_df, aes(x = AveExpr, y = logFC)) +
  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "Scatter Plot",
       color = "Overlap with CellCnnsig") +
  theme_minimal()
ggplot(RNAlistRDS005_df, aes(x = AveExpr, y = logFC, color = cluster_id)) +
  geom_point(size = 1) +
  labs(title = "Scatter Plot") +
  theme_minimal()

# Create a scatter plot logFC vs AveExpr
ggplot(RNAlistRDS005_df, aes(x = p_adj.loc, y = logFC)) +
  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "Scatter Plot",
       color = "Overlap with CellCnnsig") +
  theme_minimal()
ggplot(RNAlistRDS005_df, aes(x = p_adj.loc, y = logFC, color = cluster_id)) +
  geom_point(size = 1) +
  labs(title = "Scatter Plot") +
  theme_minimal()

# Create a scatter plot
ggplot(RNAlistRDS005_df, aes(x = logFC, y = p_adj.loc)) +
  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "Scatter Plot",
       color = "Overlap with CellCnnsig E042A011A012") +
  theme_minimal()

# Create vulcano plot
ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "Scatter Plot",
       color = "Overlap with CellCnnsig E042A011A012") +
  theme_minimal()
ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc), color = cluster_id)) +
  geom_point(size = 1.2) 
ggsave(filename = paste0(dir_out, 'vulcano_clusterid_10_5.pdf'), width =6,height=3)

overlap <- RNAlistRDS005_df[(RNAlistRDS005_df$gene %in% CellCnnsig$gene),]
length(unique(overlap$gene))
#Plot 
par(mfrow = c(2, 1))
plot(density(overlap$AveExpr))
plot(density(RNAlistRDS005_df$AveExpr))

#Histogram
# Create the barplot
summary_data <- RNAlistRDS005_df %>%
  group_by(cluster_id, stage, CellCnnsig) %>%
  dplyr::summarize(Freq = n())
# Create the barplot with color by CellCnnsig

# Create the barplot with a log2 scale on the y-axis
ggplot(summary_data, aes(x = cluster_id, y = Freq, fill = CellCnnsig)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~stage, scales = "free_x", ncol = 4) +
  scale_y_break(c(100, 150), scales = 0.3)+ 
  scale_y_break(c(350, 2500), scales = 0.3)

ggplot(RNAlistRDS005_df, aes(x = cluster_id, y = p_adj.loc, color = CellCnnsig)) +
  geom_violin() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~stage, scales = "free_x", ncol = 4)
ggplot(RNAlistRDS005_df, aes(x = cluster_id, y = logFC, color = CellCnnsig)) +
  geom_violin() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~stage, scales = "free_x", ncol = 4)
ggplot(RNAlistRDS005_df, aes(x = cluster_id, y = AveExpr, color = CellCnnsig)) +
  geom_boxplot(outlier.colour="black", outlier.shape=16,
             outlier.size=0, notch=FALSE) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~stage, scales = "free_x", ncol = 4)

ggplot(summary_data, aes(x = cluster_id, y = Freq, fill = CellCnnsig)) +
  geom_bar(position = "fill", stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~stage, scales = "free_x", ncol = 4)

#Get number of overlapping DEGs per cluster
head(RNAlistRDS005_df)
overlapping <- RNAlistRDS005_df[RNAlistRDS005_df$gene %in% CellCnnsig$gene,]
table(overlapping$cluster_id)
table(RNAlistRDS005_df$cluster_id)

#Density plot for logFC
p1 <- ggplot(RNAlistRDS005_df, aes(x = logFC, fill = CellCnnsig)) +
  geom_density(alpha = 0.5) +  
  theme_minimal() + 
  scale_fill_manual(values = c("Non-overlapping" = "blue", "overlapping" = "orange"),drop = FALSE) 
p1
#Density plot for AveExpr
p2 <- ggplot(RNAlistRDS005_df, aes(x = log2(AveExpr - 17), fill = CellCnnsig)) +
  geom_density(alpha = 0.5) + 
  theme_minimal() + 
  scale_fill_manual(values = c("Non-overlapping" = "blue", "overlapping" = "orange"),drop = FALSE) +
  geom_vline(xintercept = 0.5849625)
p2
#Density plot for p_adj.loc
p3 <- ggplot(RNAlistRDS005_df, aes(x = p_adj.loc, fill = CellCnnsig)) +
  geom_density(alpha = 0.5) + 
  theme_minimal() + 
  scale_fill_manual(values = c("Non-overlapping" = "blue", "overlapping" = "orange"),drop = FALSE) 
plot_grid(p1,p2,p3, ncol=1)

```

```{r Make rank by padj.loc and logFC and evaluate in vulcano plot}
table <- RNAlistRDS005_df %>%
  mutate(logFC_rank = rank(-abs(logFC))) %>%
  mutate(p_adj.loc_rank = rank(p_adj.loc)) %>%
  mutate(padj.loc_logFC_rank = (logFC_rank + p_adj.loc_rank)) %>%
  mutate(padj.loc_logFC10_rank = (10*logFC_rank + p_adj.loc_rank)) %>%
  mutate(padj.loc_logFC100_rank = (100*logFC_rank + p_adj.loc_rank)) %>%
  arrange(AveExpr)
table_overlapping <- table[table$gene %in% CellCnnsig$gene,]

#table200 <- table %>% arrange(mean_rank) %>% slice_head(n = 300)

# Create vulcano plot
p1 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_logFC_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_logFC_rank, slice_head300") +
  theme_minimal()
p1

p2 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_logFC10_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_logFC10_rank, slice_head300") +
  theme_minimal()
p2

p3 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_logFC100_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_logFC100_rank, slice_head300") +
  theme_minimal()
p3

p4 <- ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "Scatter Plot",
       color = "CellCnnsigE042A011A012") +
  theme_minimal()
p4

p5 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
  geom_point() +
  labs(title = "RNAlistRDS005_df") +
  theme_minimal() +
  scale_colour_gradientn(colours = rev(rainbow(5)),
                       breaks = seq(18.3, 25, by = 5),
                       limits = c(18.3, 25),
                       labels = as.character(seq(18.3, 25, by = 5)))
p5
p6 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
  geom_point() +
  labs(title = "RNAlistRDS005_df") +
  theme_minimal() +
  scale_colour_gradientn(colours = rev(rainbow(5)),
                       breaks = seq(18.5, 25, by = 5),
                       limits = c(18.5, 25),
                       labels = as.character(seq(18.5, 25, by = 5)))
p6
plot_grid(p1,p2,p3,p4,p5,p6, ncol=3)

#Keep in mind to remove gene duplicates before giving ShSh the list
```

```{r Make rank by padj.loc and logFC and AveExpr and evaluate in vulcano plot}
table <- RNAlistRDS005_df %>%
  mutate(logFC_rank = rank(-abs(logFC))) %>%
  mutate(p_adj.loc_rank = rank(p_adj.loc)) %>%
  mutate(AveExpr_rank = rank(-AveExpr)) %>%
  mutate(padj.loc_logFC_AveExpr_rank = (logFC_rank + p_adj.loc_rank + AveExpr_rank)) %>%
  mutate(padj.loc_logFC10_AveExpr_rank = (10*logFC_rank + p_adj.loc_rank + AveExpr_rank)) %>%
  mutate(padj.loc_10logFC_10AveExpr_rank = (logFC_rank + p_adj.loc_rank + 10*AveExpr_rank)) %>%
  arrange(AveExpr)
table_overlapping <- table[table$gene %in% CellCnnsig$gene,]
#table200 <- table %>% arrange(mean_rank) %>% slice_head(n = 300)

# Create vulcano plot
p1 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_logFC_AveExpr_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_logFC_AveExpr_rank, slice_head300") +
  theme_minimal()
p1

p2 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_logFC10_AveExpr_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_logFC10_AveExpr_rank, slice_head300") +
  theme_minimal()
p2

p3 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_10logFC_10AveExpr_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_10logFC_10AveExpr_rank, slice_head300") +
  theme_minimal()
p3

p4 <- ggplot(table_overlapping, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
  geom_point() +
  labs(title = "table_overlapping") +
  theme_minimal() +
  scale_colour_gradientn(colours = rev(rainbow(5)),
                       breaks = seq(18.3, 25, by = 5),
                       limits = c(18.3, 25),
                       labels = as.character(seq(18.3, 25, by = 5)))
p4

p5 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
  geom_point() +
  labs(title = "RNAlistRDS005_df") +
  theme_minimal() +
  scale_colour_gradientn(colours = rev(rainbow(5)),
                       breaks = seq(18.3, 25, by = 5),
                       limits = c(18.3, 25),
                       labels = as.character(seq(18.3, 25, by = 5)))
p5
p6 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
  geom_point() +
  labs(title = "RNAlistRDS005_df") +
  theme_minimal() +
  scale_colour_gradientn(colours = rev(rainbow(5)),
                       breaks = seq(18.5, 25, by = 5),
                       limits = c(18.5, 25),
                       labels = as.character(seq(18.5, 25, by = 5)))
p6
plot_grid(p1,p2,p3,p4,p5,p6, ncol=3)

#Keep in mind to remove gene duplicates before giving ShSh the list
```

```{r Make rank by padj.loc and logFC, AveExpr > 18.5 and evaluate in vulcano plot}
table <- RNAlistRDS005_df %>%
  dplyr::filter(AveExpr > 18.5) %>%
  mutate(logFC_rank = rank(-abs(logFC))) %>%
  mutate(p_adj.loc_rank = rank(p_adj.loc)) %>%
  mutate(padj.loc_logFC_rank = (logFC_rank + p_adj.loc_rank)) %>%
  mutate(padj.loc_logFC10_rank = (10*logFC_rank + p_adj.loc_rank)) %>%
  mutate(padj.loc_logFC2_rank = (2*logFC_rank + p_adj.loc_rank)) %>%
  arrange(AveExpr)
table_overlapping <- table[table$gene %in% CellCnnsig$gene,]

#Histogram
# Create the barplot
summary_data <- table %>%
  group_by(cluster_id, stage, CellCnnsig) %>%
  dplyr::summarize(Freq = n())
# Create the barplot with color by CellCnnsig

# Create the barplot with a log2 scale on the y-axis
ggplot(summary_data, aes(x = cluster_id, y = Freq, fill = CellCnnsig)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~stage, scales = "free_x", ncol = 4) +
  scale_y_break(c(40, 50), scales = 0.3)

#table200 <- table %>% arrange(mean_rank) %>% slice_head(n = 300)

# Create vulcano plot
p1 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_logFC_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_logFC_rank, slice_head300") +
  theme_minimal()
p1

p2 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_logFC10_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_logFC10_rank, slice_head300") +
  theme_minimal()
p2

p3 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = table,
             aes(color = "Non-overlapping"), size = 1) +
  geom_point(data = table %>% arrange(padj.loc_logFC2_rank) %>% slice_head(n = 300),
             aes(color = "overlapping"), size = 2) +
  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
  labs(title = "padj.loc_logFC2_rank, slice_head300") +
  theme_minimal()
p3

p4 <- ggplot(table_overlapping, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
  geom_point() +
  labs(title = "table_overlapping") +
  theme_minimal() +
  scale_colour_gradientn(colours = rev(rainbow(5)),
                       breaks = seq(18.5, 25, by = 5),
                       limits = c(18.5, 25),
                       labels = as.character(seq(18.5, 25, by = 5)))
p4

p5 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
  geom_point() +
  labs(title = "RNAlistRDS005_df") +
  theme_minimal() +
  scale_colour_gradientn(colours = rev(rainbow(5)),
                       breaks = seq(18.3, 25, by = 5),
                       limits = c(18.3, 25),
                       labels = as.character(seq(18.3, 25, by = 5)))
p5
p6 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
  geom_point() +
  labs(title = "RNAlistRDS005_df") +
  theme_minimal() +
  scale_colour_gradientn(colours = rev(rainbow(5)),
                       breaks = seq(18.5, 25, by = 5),
                       limits = c(18.5, 25),
                       labels = as.character(seq(18.5, 25, by = 5)))
p6
plot_grid(p1,p3,p2,p4,p6, ncol=3)

#Keep in mind to remove gene duplicates before giving ShSh the list
```

```{r df of table()s }
metadata <- read.csv("/home/common/data/output/projects/ENDO/E044/A006/Metadata_allE044_filtered.csv")
df <- data.frame(unclass(table(metadata$EndometriosisStatus,metadata$CyclePhases)))
df <- df %>%
  rename(Mid.Secretory = Secretory) %>%
  select(Proliferative, Early.Secretory, Mid.Secretory, Late.Secretory)

```

```{r VlnPlots of SLPI and NFKBIA for epithelial cells}
epithelial <- readRDS("/home/common/data/output/projects/ENDO/E044/A066/epithelial.rds")
#Filter by proliferative samples used for the DEG analysis
samples <- unique(read.csv("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/celltype_freq_sample.csv")$Var2)
Idents(epithelial) <- epithelial$sample
epithelial_prolif <- subset(epithelial, idents = samples)

#VlnPlot
p1 <- VlnPlot(epithelial_prolif, features = c("SLPI"), assay = "RNA",group.by = "sample", pt.size = 0,split.by = "EndometriosisStatus") + ylim(0,100) +
stat_summary(fun.y = median, geom='point', size = 5, colour = "black", shape = 95) + ggtitle("SLPI, RNA assay")
p2 <- VlnPlot(epithelial_prolif, features = c("NFKBIA"), assay = "RNA",group.by = "sample", pt.size = 0,split.by = "EndometriosisStatus") +
stat_summary(fun.y = median, geom='point', size = 5, colour = "black", shape = 95) + ggtitle("NFKBIA, RNA assay")
p3 <- VlnPlot(epithelial_prolif, features = c("SLPI"), assay = "SCT",group.by = "sample", pt.size = 0,split.by = "EndometriosisStatus") +
stat_summary(fun.y = median, geom='point', size = 5, colour = "black", shape = 95)+ ggtitle("SLPI, SCT assay")
p4 <- VlnPlot(epithelial_prolif, features = c("NFKBIA"), assay = "SCT",group.by = "sample", pt.size = 0,split.by = "EndometriosisStatus") +
stat_summary(fun.y = median, geom='point', size = 5, colour = "black", shape = 95)+ ggtitle("NFKBIA, SCT assay")
plot_grid(p1,p3,p2,p4, ncol=2)
ggsave(plot = plot_grid(p1,p3,p2,p4, ncol=2), paste0(dir_out, "SLPI_NFKBIA_epithelial_prolif_RNASCT.png"))

##Calculate mean/median SLPI/NKKBIA expression per sample for SCT assay
expression_values <- GetAssayData(object = epithelial_prolif, assay = "SCT")[c("SLPI", "NFKBIA"), ]
expression_df <- as.data.frame(t(as.matrix(expression_values)))
expression_df$sample <- gsub("_.*", "", rownames(expression_df))
mean_values <- aggregate(. ~ sample, data = expression_df, FUN = mean)
median_values <- aggregate(. ~ sample, data = expression_df, FUN = median)
# Combine mean and median values into a single table
result_table_SCT <- data.frame(
  sample_SCT = mean_values$sample,
  SLPI_mean_SCT = mean_values$SLPI,
  SLPI_median_SCT = median_values$SLPI,
  NFKBIA_mean_SCT = mean_values$NFKBIA,
  NFKBIA_median_SCT = median_values$NFKBIA
)

##Calculate mean/median SLPI/NKKBIA expression per sample for RNA assay
expression_values <- GetAssayData(object = epithelial_prolif, assay = "RNA")[c("SLPI", "NFKBIA"), ]
expression_df <- as.data.frame(t(as.matrix(expression_values)))
expression_df$sample <- gsub("_.*", "", rownames(expression_df))
mean_values <- aggregate(. ~ sample, data = expression_df, FUN = mean)
median_values <- aggregate(. ~ sample, data = expression_df, FUN = median)
# Combine mean and median values into a single table
result_table_RNA <- data.frame(
  sample = mean_values$sample,
  SLPI_mean_RNA = mean_values$SLPI,
  SLPI_median_RNA = median_values$SLPI,
  NFKBIA_mean_RNA = mean_values$NFKBIA,
  NFKBIA_median_RNA = median_values$NFKBIA
)

#combine RNA and SCT assay and add EndometriosisStatus condition to df
result_table <- cbind(result_table_RNA,result_table_SCT)
sample_EndometriosisStatus <- epithelial_prolif@meta.data %>% select(c("sample","EndometriosisStatus")) %>% distinct()
result_table <- left_join(result_table,sample_EndometriosisStatus, by = "sample")
write.csv(result_table,"mean_median_SLPI_NFKBIA_epithelial_prolif_RNASCT.csv")

```

```{r Make vulcano plots for DEGs colored by major cell types}
##RNA assay DEGs from proliferative samples
RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")
##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes
mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes
rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes
RNAlistRDS005_df <- RNAlistRDS005_df %>%
  dplyr::filter(!gene %in% c(hb.genes, mt.genes, rb.genes)) %>%
  dplyr::filter(annotation_reference == "Tan") %>%
  dplyr::filter(AveExpr > 18.53202)
##Rename cluster_id to major cell types
meta <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_metadata.rds")
meta$Symphony_Refined[meta$Symphony_Refined == "M$\\Phi$1-LYVE1"] <- "M-Phi-1-LYVE1"
meta$Symphony_Refined[meta$Symphony_Refined == "glandular early-secretory"] <- "glandular-early-secretory"
meta <- meta %>% 
  select(AnnotationMain,Symphony_Refined,Symphony_Refined_A029_United) %>%
  distinct()

RNAlistRDS005_df <- RNAlistRDS005_df %>%
  mutate(AnnotationMain = cluster_id)
RNAlistRDS005_df$AnnotationMain <- case_when(
  RNAlistRDS005_df$AnnotationMain %in% meta$Symphony_Refined ~ meta$AnnotationMain[match(RNAlistRDS005_df$AnnotationMain, meta$Symphony_Refined)],
  RNAlistRDS005_df$AnnotationMain %in% meta$Symphony_Refined_A029_United ~ meta$AnnotationMain[match(RNAlistRDS005_df$AnnotationMain, meta$Symphony_Refined_A029_United)],
  TRUE ~ RNAlistRDS005_df$AnnotationMain)
RNAlistRDS005_df$AnnotationMain[RNAlistRDS005_df$AnnotationMain == "myeloid" |
                                      RNAlistRDS005_df$AnnotationMain == "lymphocyte"] <- "immune"

##Plot
ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc), color = AnnotationMain)) +
  geom_point(size = 1) +
  #theme_minimal() +
  facet_wrap(~stage)
ggsave(filename = paste0(dir_out, 'vulcano_DEG_stage_main4.pdf'), width =7,height=4)

ggplot(RNAlistRDS005_df, aes(x = AnnotationMain, fill = cluster_id)) +
  geom_bar(stat = "count", position = "stack", width = 0.7) +  
  # theme_minimal() +  
  facet_wrap(~stage) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(filename = paste0(dir_out, 'vulcano_DEG_stage_main5.pdf'), width =7,height=4)

```

```{r Captum scores}
#Load captum scores and combine tables
all_cells_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E042/A011/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv")
all_cells_corrgpca$dataset <- "all_cells_corrgpca"
all_cells_pca <- read.csv("/home/common/data/output/projects/ENDO/E042/A012/sigs/weighted_score_sig_pca_1pc.csv")
all_cells_pca$dataset <- "all_cells_pca"

##RNA assay DEGs from proliferative samples
RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")
##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes
mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes
rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes
RNAlistRDS005_df <- RNAlistRDS005_df %>%
  dplyr::filter(!gene %in% c(hb.genes, mt.genes, rb.genes)) %>%
  dplyr::filter(annotation_reference == "Tan") %>%
  dplyr::filter(AveExpr > 18.53202)

#Combine tables
all_cells <- rbind(all_cells_pca,all_cells_corrgpca)
all_cells <- left_join(all_cells, RNAlistRDS005_df, by = "gene") %>% 
  arrange(desc(rank))

```

```{r cycling fibroblasts}

#LTs annotation for cycling fibroblasts
data <- readRDS("/home/common/data/output/projects/ENDO/E031/A004/proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated.RDS")
meta <- data@meta.data
rm(data)
gc()
table(meta$Annotation, meta$EndometriosisStatus)

#Make cell id compatible with new data set
meta$cell_id <- paste(meta$sample, meta$cell.names, sep = '_')
rownames(meta) <- meta$cell_id

#Load entire dataset and add LTs annotation
data_ds <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
data_ds <- AddMetaData(data_ds,meta[,c("Annotation", "scds.cxds.score")])
DimPlot(data_ds_new,group.by = "Annotation")

#Highlight cycling fibroblasts
Idents(data_ds) <- data_ds$Annotation
table(data_ds$Annotation)
cyclingfibs <- WhichCells(data_ds, idents = c("Cycling Fibroblasts"))
DimPlot(data_ds, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)

```

```{r Barplots for  filter response median rank E042A027}
#Load metadata
meta_final <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")

#Set variables to loop over
DIMRED <- c("corrgpca","pca")
ANNOTATION <- c("Symphony_Refined","AnnotationMain")
MEDIAN <- c("median_rank_endo_logic003","median_rank_control_logic003","median_rank_endo_logic005","median_rank_control_logic005","median_rank_endo_logic01","median_rank_control_logic01")

for (dim in DIMRED) {
#Read filter ranking
average_filter_ranking <- readRDS(paste0("/home/common/data/output/projects/ENDO/E042/A027/response_cellcnn_manually/all_cells/",dim,"_average_filter_ranking.rds"))

#Combine tables
meta <- left_join(meta_final, average_filter_ranking %>% select(cell_ident, median_rank_endo, median_rank_control), by = "cell_ident")

#Exclude NA
metaNAfree <- meta %>%
  select(median_rank_endo, median_rank_control,Symphony_Refined,AnnotationMain) %>%
  na.omit() %>%
  filter(Symphony_Refined != "granulocyte")

endo_threshold003 <- quantile(metaNAfree$median_rank_endo, 0.03)
control_threshold003 <- quantile(metaNAfree$median_rank_control, 0.03)
endo_threshold005 <- quantile(metaNAfree$median_rank_endo, 0.05)
control_threshold005 <- quantile(metaNAfree$median_rank_control, 0.05)
endo_threshold01 <- quantile(metaNAfree$median_rank_endo, 0.1)
control_threshold01 <- quantile(metaNAfree$median_rank_control, 0.1)

# Create the new column with ranking logic threshold
metaNAfree <- metaNAfree %>%
  mutate(median_rank_endo_logic003 = median_rank_endo <= endo_threshold003) %>%
  mutate(median_rank_control_logic003 = median_rank_control <= control_threshold003) %>%
  mutate(median_rank_endo_logic005 = median_rank_endo <= endo_threshold005) %>%
  mutate(median_rank_control_logic005 = median_rank_control <= control_threshold005) %>%
  mutate(median_rank_endo_logic01 = median_rank_endo <= endo_threshold01) %>%
  mutate(median_rank_control_logic01 = median_rank_control <= control_threshold01)

# Create bar plots
dir.create(paste0(dir_out,"median_rank_plots/"))

for (ann in ANNOTATION) {
  for (med in MEDIAN) {
    # Create plot
    plot1 <- ggplot(metaNAfree, aes_string(x = ann)) +
      geom_bar(position = "identity") +
      scale_y_continuous(trans = "log10") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      ggtitle(paste0(dim, "___",ann, "___", med))
    plot2 <- ggplot(metaNAfree, aes_string(x = ann, fill = med)) +
      geom_bar(position = "fill") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
    combined_plot <- plot1 / plot2
    # Save plot
    ggsave(filename = paste0(dir_out, "/median_rank_plots/", dim, "___",ann, "___", med, ".pdf"), plot = combined_plot, width = 13, height = 10)
  }
}
}

```

```{r UMAPs for  filter response median rank E042A027}
#Load data and metadata
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
meta_final <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
average_filter_ranking <- readRDS(paste0("/home/common/data/output/projects/ENDO/E042/A027/response_cellcnn_manually/all_cells/pca_average_filter_ranking.rds"))
#Combine meta, clean up and do ranking
meta <- left_join(meta_final, average_filter_ranking %>% select(cell_ident, median_rank_endo, median_rank_control), by = "cell_ident") 
#Exclude NA
metaNAfree <- meta %>%
  select(median_rank_endo, median_rank_control,Symphony_Refined,AnnotationMain,cell_ident) %>%
  na.omit()
#Ranking
endo_threshold01 <- quantile(metaNAfree$median_rank_endo, 0.1)
control_threshold01 <- quantile(metaNAfree$median_rank_control, 0.1)
metaNAfree <- metaNAfree %>%
  mutate(median_rank_endo_logic01 = median_rank_endo <= endo_threshold01) %>%
  mutate(median_rank_control_logic01 = median_rank_control <= control_threshold01) %>%
  select(Symphony_Refined, median_rank_endo_logic01, median_rank_control_logic01,cell_ident)
#Make summary table
summary_table <- metaNAfree %>%
  select(Symphony_Refined, median_rank_endo_logic01, median_rank_control_logic01) %>%
  group_by(Symphony_Refined) %>%
  summarize(
    percent_endo_TRUE = mean(median_rank_endo_logic01) * 100,
    percent_control_TRUE = mean(median_rank_control_logic01) * 100
  )
write.csv(summary_table, paste0(dir_out,"summary_pca_average_filter_ranking_10perc.csv"))

#Highlight 10% cells
endo_true_cells <- subset(metaNAfree, median_rank_endo_logic01 == TRUE)$cell_ident
control_true_cells <- subset(metaNAfree, median_rank_control_logic01 == TRUE)$cell_ident
p1 <- DimPlot(data, reduction = "umap", cells.highlight = endo_true_cells, cols.highlight = c("darkred"), cols= "grey", raster = FALSE) + ggtitle("endo_true_cells")
p2 <- DimPlot(data, reduction = "umap", cells.highlight = control_true_cells, cols.highlight = c("darkred"), cols= "grey", raster = FALSE) + ggtitle("control_true_cells")

```

```{r make DEG table for paper}
#RNA assay DEGs from proliferative samples
RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv") %>%
  filter(annotation_reference == "Tan") %>%
  filter(AveExpr > 18.53202)
##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes, there are no hb genes in the DEG list
RNAlistRDS005_df[RNAlistRDS005_df$gene %in% hb.genes, ]#no hb genes
#mitochondrial genes
mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes, no genes
rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes, 77 genes
RNAlistRDS005_df <- RNAlistRDS005_df %>%
  dplyr::filter(!gene %in% rb.genes)

write.csv(RNAlistRDS005_df, paste0(dir_out,"RNAlistRDS005_AveExpr18.53202_noRBgenes.csv"))

length(unique(RNAlistRDS005_df$gene)) #427
```

```{r}

metadata <- read.csv('/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv')
metadata$new_column <- NULL
metadata$batch <- NULL

# Remove 'names' column as it will be used as row labels
row_names <- metadata$names
metadata <- metadata[, -1]

# Convert non-numeric columns to factors
metadata[, c("EndometriosisStatus", "lane",  "endo_EndometriosisGrade",  "fold1", "fold2", "fold3", "fold4", "fold5")] <- lapply(metadata[, c("EndometriosisStatus", "lane", "endo_EndometriosisGrade", "fold1", "fold2", "fold3", "fold4", "fold5")], as.factor)

# Create heatmap using ComplexHeatmap
pdf(paste0(dir_out, "metadata_prolif_eval_fold_all_heatmap.pdf"), width = 7, height = 9)
Heatmap(metadata,
        name = "Metadata",
        row_names_side = "left",
        row_names_gp = gpar(fontsize = 8),
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        show_column_names = TRUE,
        show_row_names = TRUE,
        column_title = "Columns",
        row_title = "Row Names",
        heatmap_legend_param = list(title = "Legend")
)
dev.off()

######################
metadata <- read.csv('/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv')
metadata$new_column <- NULL
metadata$batch <- NULL

# Remove 'names' column as it will be used as row labels
row_names <- metadata$names
metadata <- metadata[, -1]

# Convert non-numeric columns to factors
metadata[, c("EndometriosisStatus", "lane","endo_EndometriosisGrade",  "fold1", "fold2", "fold3", "fold4", "fold5")] <- lapply(metadata[, c("EndometriosisStatus", "lane", "endo_EndometriosisGrade", "fold1", "fold2", "fold3", "fold4", "fold5")], as.factor)

# Convert remaining columns to a matrix
mat <- data.matrix(metadata)

# Create heatmap
pdf(paste0(dir_out, "metadata_prolif_eval_fold_all_heatmap2.pdf"), width = 7, height = 9)
heatmap.2(mat,
          Rowv=FALSE, Colv=FALSE,
          dendrogram="none",
          trace="none",
          col = heat.colors(256),
          margins=c(8,10),
          labRow=row_names,
          scale="column",
          key=TRUE,
          keysize=1.5,
          key.title="Color Key",
          key.xlab="Value",
          key.ylab=NULL,
          density.info="none",
          main="Heatmap of Metadata")
dev.off()

```

### 2025-01-24 05b_Functional_enrichment_analysis/DEG_gProfiler.Rmd
Code update

---
title: "DEG_RNA_summary"
author: "LD"
date: "2025-01-19"
output: html_document
---

```{r setup}
library(gprofiler2)
library(plyr)
library(dplyr)
library(ggplot2)
library(tibble)
library(cowplot)

dir_out <- "/home/common/data/output/projects/ENDO/E044/A031/"
setwd(dir_out)
```

Determine inflection point based on unfiltered DEGs
```{r}
##Combine muscat output into one dataframe
PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
RNAlistRDS <- lapply(PATHS, readRDS)
names(RNAlistRDS) <- print(PATHS)
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
RNAlistRDS <- lapply(RNAlistRDS,plyr::rbind.fill)
RNAlistRDSdf <- as.data.frame(do.call(rbind, RNAlistRDS))

##exclude ribosomal genes (comment: does not change the inflection point)
rb.genes_DEGs <- unique(grep(pattern = "^RP[SL]", RNAlistRDSdf$gene, value=TRUE)) #around 150 unique rb genes
RNAlistRDSdf <- RNAlistRDSdf[!(RNAlistRDSdf$gene %in% rb.genes_DEGs), ]

##Determine inflection point
#Determine Inflection Point, No log2 scale, bw = 0.02
plot(density(RNAlistRDSdf$AveExpr))
# Create the density plot and store it in an object
density_plot <- density(RNAlistRDSdf$AveExpr, bw = 0.02) #good values 0.025 or 0.02, 0.01 is too small
plot(density_plot)
# Calculate the second derivative of the density curve
second_derivative <- diff(diff(density_plot$y)) / diff(density_plot$x)^2
# Find where the second derivative crosses zero (inflection point)
inflection_points <- density_plot$x[which(diff(sign(second_derivative)) != 0)]
# Print or plot the inflection point(s)
cat("Inflection Point(s):", inflection_points, "\n")
# You can also plot the density curve with the inflection point(s) marked
plot(density_plot)
abline(v = inflection_points, col = "red", lty = 2)
abline(v = 18.53202, col = "green", lty = 2)
text(x=18.45, y=1.5, srt=90, '18.53202')

#Best result: no log2 scale, bw = 0.02 with resulting inflection point 18.53202

length(RNAlistRDSdf$gene)
RNAlistRDSdf_filt <- RNAlistRDSdf %>% dplyr::filter(AveExpr > 18.53202) 
length(RNAlistRDSdf_filt$gene)
```

summarize DEGs into df and plot
```{r RNA assay}
#RNA assay
PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
print(PATHS)
RNAlistRDS <- lapply(PATHS, readRDS)
names(RNAlistRDS) <- print(PATHS)
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
RNAlistRDS <- lapply(RNAlistRDS,plyr::rbind.fill)
RNAlistRDSdf <- as.data.frame(do.call(rbind, RNAlistRDS))
RNAlistRDS005 <- lapply(RNAlistRDS, function(x) dplyr::filter(x, p_adj.loc < 0.05))
RNAlistRDS005 <- as.data.frame(do.call(rbind, RNAlistRDS005))
#Check for cluster_ids that have or don't have DEGs with p_adj.loc < 0.05
intersect(levels(factor(RNAlistRDSdf$cluster_id)),levels(factor(RNAlistRDS005$cluster_id)))
setdiff(levels(factor(RNAlistRDSdf$cluster_id)),levels(factor(RNAlistRDS005$cluster_id)))
#Add columns stage, clustering, annotation_reference
library(stringr)
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(stage = case_when(
    str_detect(rownames(RNAlistRDS005), "_severe") ~ "severe",
    str_detect(rownames(RNAlistRDS005), "_mild") ~ "mild",
    TRUE ~ "all"
  ))
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(clustering = case_when(
    str_detect(rownames(RNAlistRDS005), "major") ~ "major",
    str_detect(rownames(RNAlistRDS005), "united") ~ "united",
    TRUE ~ "RefinedCluster"
  ))
RNAlistRDS005 <- RNAlistRDS005 %>%
  dplyr::mutate(annotation_reference = case_when(
    str_detect(rownames(RNAlistRDS005), "Tan") ~ "Tan",
    str_detect(rownames(RNAlistRDS005), "Fonseca") ~ "Fonseca"
  ))

length(unique(RNAlistRDS005$gene)) #12115

write.csv(RNAlistRDS005,"/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")
#RNAlistRDS005 <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv", row.names = 1)

###Delete below ###########

RNAlistRDS005 %>% dplyr::filter(logFC > 0.5) %>% nrow()
RNAlistRDS005 %>% dplyr::filter(logFC < -0.5) %>% nrow()

RNAlistRDS005 <- RNAlistRDS005 %>% dplyr::filter(annotation_reference == "Tan")
###plotting
plot(density(RNAlistRDS005$AveExpr))
# bar plot of DEGs by condition
df <- RNAlistRDS005 %>%
  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
  ungroup()
ggplot(df, aes(x=stage, y = DEGs, fill=cluster_id)) +
  geom_bar(stat='identity') +
  scale_y_continuous(expand = c(0,0)) +
  RotatedAxis() +
  xlab('Sample ID') +
  facet_wrap(~clustering,scales = "free_x", ncol = 4)
ggsave(paste0(dir_out, "Barplot_cluster_id_stage_clustering.pdf"),
  plot = last_plot(),
  width = 15, height = 6)
sum(df$DEGs)
# bar plot of DEGs by condition with AveExpr > 18.5
#There are extremely many DEGs in the clusters eF, epithelial and C7_Activated_Fibro, too many to be reliable. Median AveExpr seems to be lower in those clusters and gProfiler gave good results with a threshold of AveExpr>18.5. Therefore, I set the cut-off at AveExpr>18.5
df18_3 <- RNAlistRDS005 %>%
  dplyr::filter(AveExpr > 18.3) %>%
  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
  ungroup()
df18_5 <- RNAlistRDS005 %>%
  dplyr::filter(AveExpr > 18.5) %>%
  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
  ungroup()
ggplot(df18_5, aes(x=stage, y = DEGs, fill=cluster_id)) +
  geom_bar(stat='identity') +
  scale_y_continuous(expand = c(0,0)) +
  RotatedAxis() +
  xlab('Sample ID') +
  facet_wrap(~clustering,scales = "free_x", ncol = 4) +
  ggtitle("AveExpr > 18.5")
ggsave(paste0(dir_out, "Barplot_cluster_id_stage_clustering_AveExpr185.pdf"),
  plot = last_plot(),
  width = 15, height = 6)
ggplot(df18_5, aes(x=stage, y = DEGs, fill=cluster_id)) +
  geom_bar(stat='identity') +
  scale_y_continuous(expand = c(0,0)) +
  RotatedAxis() +
  xlab('Sample ID')+
  ggtitle("AveExpr > 18.5") 
ggsave(paste0(dir_out, "Barplot_cluster_id_stage_AveExpr185.pdf"),
  plot = last_plot(),
  width = 6, height = 4)
sum(df18_5$DEGs)

# Count the number of genes in each combination of clustering and stage
RNAcounts <- table(RNAlistRDS005$clustering, RNAlistRDS005$stage)
pheatmap(RNAcounts)
# Create the grouped bar plot
ggplot(as.data.frame(RNAcounts), aes(x = Var1, y = Freq, fill = Var2)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Clustering", y = "Count", fill = "Stage") +
  ggtitle("Count of Genes by Clustering and Stage") +
  theme_bw() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Create vulcano plot
ggplot(RNAlistRDS005, aes(x = logFC, y = -log10(p_adj.loc), color = cluster_id)) +
  geom_point(data = RNAlistRDS005) +
  theme_minimal() +
  xlim(-0.8,1.3)
ggsave(paste0(dir_out, "Vulcano_logFC_padjloc_clusterid.pdf"),
  plot = last_plot(),
  width = 6, height = 4)

#scatter plot logFC, AveExpr
ggplot(RNAlistRDS005, aes(x = AveExpr, y = logFC, color = cluster_id)) +
  geom_point(data = RNAlistRDS005) +
  theme_minimal()
ggsave(paste0(dir_out, "Scatter_logFC_AveExpr_clusterid.pdf"),
  plot = last_plot(),
  width = 6, height = 4)
plot(density(log2(RNAlistRDS005$AveExpr-16.97529)))
```

gProfiler based on RNA/SCT assay and per annotation with gene expression >18.53202
```{r}
##function to run gProfiler and save output
gProfiler_FUN <- function(x,ANALYSIS,GRADE,ORDERED_QUERY,ASSAY,dir_out) {
#gProfiler
custom_bg <- unique(x$gene)
query <- x %>%
  dplyr::filter(AveExpr > 18.53202) %>%
  dplyr::filter(p_adj.loc < 0.05) %>%
  dplyr::arrange(p_adj.loc) #arrange by lowest p_adj.loc
if(nrow(query) > 0){
query <- distinct(query, gene, .keep_all= TRUE) #remove duplicates of gene with higher p_adj.loc
query <- query$gene
gostres <<- gost(query = query, 
                organism = "hsapiens", 
                ordered_query = ORDERED_QUERY, #can use TRUE, p_adj.loc has been arranged 
                multi_query = FALSE, 
                significant = TRUE, #only statistically significant results should be returned
                exclude_iea = FALSE, 
                measure_underrepresentation = FALSE, 
                evcodes = TRUE,
                user_threshold = 0.05, 
                correction_method = "g_SCS", #g_SCS is default
                domain_scope = "annotated", #if custom_bg is set, I believe this shouldn't matter too much
                custom_bg = custom_bg, #custom background, use!
                numeric_ns = "", 
                sources = NULL, #by default all sources are analyzed
                as_short_link = FALSE)
if(length(gostres) > 0){
results <- gostres$result
results$cluster_id <- levels(as.factor(x$cluster_id))
results$analysis <- print(ANALYSIS)
results$assay <- print(ASSAY)
results$grade <- print(GRADE) 
results$ordered_query <- print(ORDERED_QUERY)
results$parents <- NULL
write.csv(results,paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
print(paste0("Saved gotres as gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
plot <- gostplot(gostres, capped = TRUE, interactive = FALSE) + ggtitle(paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY)))
ggsave(plot = plot, filename = paste0(dir_out,"gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".png"), width = 7, height = 5)
print("Saved plot")
}
}
}
##function to run gProfiler_FUN on different grades, analysis and assay
RUN_gProfier <- function(GRADE, ANALYSIS,ASSAY,SPECIFICATION){
dir_out <- paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ANALYSIS,"/",ASSAY,"assay_",SPECIFICATION,"/")
dir.create(dir_out, recursive = TRUE)
setwd(dir_out)
#Read paths to DEG output .rds files
if (ASSAY == "RNA") { 
    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
    } else if (ASSAY == "SCT") {
    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "mm_dream_sct_counts_all.rds", recursive = TRUE, full.names = TRUE)
    }
#Select paths based on all/mild/severe
if (GRADE == "all") { 
    PATHS <- PATHS[!grepl("_severe", PATHS)]
    PATHS <- PATHS[!grepl("_mild", PATHS)]
    print(PATHS)
    } else if (GRADE == "mild") {
    PATHS <- PATHS[grepl("_mild", PATHS)]
    print(PATHS)
    } else if  (GRADE == "severe") {
    PATHS <- PATHS[grepl("_severe", PATHS)]
    print(PATHS)
    } 
RNAlistRDS <- lapply(PATHS, readRDS)
RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
RNAlistRDS <- RNAlistRDS[!duplicated(RNAlistRDS)]
lapply(RNAlistRDS, gProfiler_FUN, ANALYSIS = ANALYSIS,GRADE = GRADE,ORDERED_QUERY = FALSE, ASSAY = ASSAY, dir_out = dir_out)
lapply(RNAlistRDS, gProfiler_FUN, ANALYSIS = ANALYSIS,GRADE = GRADE,ORDERED_QUERY = TRUE, ASSAY = ASSAY, dir_out = dir_out)
}

RUN_gProfier(GRADE = "all", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
RUN_gProfier(GRADE = "severe", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
RUN_gProfier(GRADE = "mild", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
RUN_gProfier(GRADE = "all", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
RUN_gProfier(GRADE = "severe", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
RUN_gProfier(GRADE = "mild", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
```

```{r Combine output tables from gProfiler}
#functions
summarize_gProfiler_PerAnnotation <- function(x){
ASSAY = "A030"
##Combine all results
gProfilerlist <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ASSAY,"/",x), pattern = "TRUE.csv$|FALSE.csv$", full.names = TRUE)
gProfilerlist <- lapply(gProfilerlist, read.csv) #read .rds
#Combine all gProfiler results into one df, save
gProfiler <- plyr::rbind.fill(gProfilerlist) #bind deg lists into one df
gProfiler$specification <- print(x)
#save csv
write.csv(gProfiler, paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ASSAY,"/gProfiler_",ASSAY,"_",x,"_summary.csv"))
table(gProfiler$grade)
#table(gProfiler$cluster_id,gProfiler$grade)
}

ASSAY = "A030"
DIRSLIST <- list.dirs(path = paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ASSAY),full.names = FALSE)
DIRSLIST_Merged <- DIRSLIST[grepl("Merged", DIRSLIST)]
DIRSLIST_PerAnnotation <- DIRSLIST[grepl("PerAnnotation", DIRSLIST)]
lapply(DIRSLIST_PerAnnotation,summarize_gProfiler_PerAnnotation)
```

```{r setup}
library(ggplot2)
library(dplyr)
library(purrr)
library(Seurat)
library(tidyr)
library(ggrepel)

dir_out <- "/home/common/data/output/projects/ENDO/E044/A031/InterestingDEGs/AveExpr18_53202/"
dir.create(dir_out, recursive = TRUE)
```

```{r}
###Read in gProfiler_A030_RNAassay_PerAnnotation_AveExpr18_53202_summary and filter for Tan annotation
gProfiler_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044A030//gProfiler_A030_RNAassay_PerAnnotation_AveExpr18_53202_summary.csv")
gProfiler_df <- gProfiler_df %>% dplyr::filter(source != "HPA") #HPA is pretty useless for our analysis
Tan_clusters <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv") %>%
  filter(annotation_reference == "Tan") %>%
  distinct(cluster_id) %>%
  pull()
gProfiler_df <- gProfiler_df %>% 
  filter(cluster_id %in% Tan_clusters)

###interesting term_names
#Immune System and inflammation
ImmuneTermsGREP <- c("TGF", "TNF", "NOD", "IL-", "immun", "cytokine", "chemokine", "NF-kappa")
ImmuneTerms <- gProfiler_df %>%
  filter(grepl(paste(ImmuneTermsGREP, collapse = "|"), term_name, ignore.case = TRUE)) %>% 
  pull(term_name) %>% 
  unique()
#angiogenesis
angiogenesisTerms  <- c(unique(gProfiler_df$term_name[grepl("VEGF", gProfiler_df$term_name, ignore.case = TRUE)]),
                       unique(gProfiler_df$term_name[grepl("angiog", gProfiler_df$term_name, ignore.case = TRUE)]))
#extracellullar terms
extracellularTerms <- unique(gProfiler_df$term_name[grepl("extracellular", gProfiler_df$term_name, ignore.case = TRUE)])
#cell adhesion
adhesionTermsGREP <- c("adhesion", "junction")
adhesionTerms <- gProfiler_df %>%
  filter(grepl(paste(adhesionTermsGREP, collapse = "|"), term_name, ignore.case = TRUE)) %>% 
  pull(term_name) %>% 
  unique()
#growth factors
insulinTerms <- unique(gProfiler_df$term_name[grepl("insulin", gProfiler_df$term_name, ignore.case = TRUE)])
#Other interesting terms
otherTerms <- c('CTNNB1-DDX5-HDGF complex') #Hepatoma-Derived Growth Factor and DDX5 Promote Carcinogenesis and Progression of Endometrial Cancer by Activating β-Catenin
# Define the list of interesting terms
interesting_terms <- list(
  ImmuneTerms = ImmuneTerms,
  AngiogenesisTerms = angiogenesisTerms,
  ExtracellularTerms = extracellularTerms,
  AdhesionTerms = adhesionTerms,
  InsulinTerms = insulinTerms
  #,OtherTerms = otherTerms
)
# Loop through the list and assign categories
gProfiler_df$DEG_category <- NA
for (category_name in names(interesting_terms)) {
  terms <- interesting_terms[[category_name]]
  for (term in terms) {
    gProfiler_df$DEG_category[gProfiler_df$term_name == term] <- category_name
  }
}
table(gProfiler_df$DEG_category)

gProfiler_df %>% 
  #dplyr::select(p_value, term_size, query_size, intersection_size, term_id, term_name) %>% 
  dplyr::filter(term_size > 25 & term_size < 750) %>% #larger terms are likely to be excessively broad and prone to false positives
  distinct(cluster_id,analysis,assay,grade,term_name, .keep_all = TRUE) %>%
  #filter(grade == "all")  %>%
  filter(!grepl(pattern = "HPA|HP|CORUM|MIRNA", x = term_id)) %>%
  dplyr::arrange(p_value) %>% 
  #dplyr::slice_head(n = 20) %>% 
  ggplot2::ggplot() + 
  ggplot2::aes(y = reorder(term_name, -log(p_value)), x = -log(p_value), color = grade, size = intersection_size/term_size * 100) + #, shape = ordered_query
  ggplot2::geom_point() + 
  #ggplot2::scale_color_continuous(name = "-Log(adj. p-value)") + 
  ggplot2::scale_size_continuous("Intersection size / Term size * 100 (%)") + 
  ggplot2::theme_classic() + 
  ggplot2::xlab("-Log(adj. p-value)") + 
  ggplot2::ylab("enriched terms") + 
  ggplot2::ggtitle("all cells")
ggsave(paste0(dir_out, "GOterms_all_severe.pdf"),
  plot = last_plot(),
  width = 14, height = 14)

#save Table
gProfiler_df_save <- gProfiler_df %>% 
  #dplyr::select(p_value, term_size, query_size, intersection_size, term_id, term_name) %>% 
  #dplyr::filter(term_size > 25 & term_size < 750) %>% #larger terms are likely to be excessively broad and prone to false positives
  distinct(cluster_id,analysis,assay,grade,term_name, .keep_all = TRUE) %>%
  #filter(grade == "all")  %>%
  #filter(!grepl(pattern = "HPA|HP|CORUM|MIRNA", x = term_id)) %>%
  dplyr::arrange(p_value)
table(gProfiler_df_save$term_id)
write.csv(gProfiler_df_save, paste0(dir_out, "GOterms_all_severe_minimalfiltering.csv"))

######
##GO plot with cell type label
gProfiler_df %>% 
  #dplyr::select(p_value, term_size, query_size, intersection_size, term_id, term_name) %>% 
  dplyr::filter(term_size > 25 & term_size < 750) %>% #larger terms are likely to be excessively broad and prone to false positives
  distinct(cluster_id,analysis,assay,grade,term_name, .keep_all = TRUE) %>%
  #filter(grade == "all")  %>%
  filter(!grepl(pattern = "HPA|HP|CORUM|MIRNA", x = term_id)) %>%
  dplyr::arrange(p_value) %>% 
  #dplyr::slice_head(n = 20) %>% 
  ggplot2::ggplot() + 
  ggplot2::aes(y = reorder(term_name, -log(p_value)), x = -log(p_value), color = cluster_id, size = intersection_size/term_size * 100) + #, shape = ordered_query
  ggplot2::geom_point() + 
  #ggplot2::scale_color_continuous(name = "-Log(adj. p-value)") + 
  ggplot2::scale_size_continuous("Intersection size / Term size * 100 (%)") + 
  ggplot2::theme_classic() + 
  ggplot2::xlab("-Log(adj. p-value)") + 
  ggplot2::ylab("enriched terms") + 
  ggplot2::ggtitle("all cells")
ggsave(paste0(dir_out, "GOterms_all_severe_colorCellID.pdf"),
  plot = last_plot(),
  width = 14, height = 14)

##############################################################################################################################
##Merge the DEG table with the Terms

#RNA assay DEGs from proliferative samples
RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv") %>%
  filter(annotation_reference == "Tan") %>%
  filter(AveExpr > 18.53202)
##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes, there are no hb genes in the DEG list
mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes, no genes
rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes, 165 genes
RNAlistRDS005_df <- RNAlistRDS005_df %>%
  dplyr::filter(!gene %in% rb.genes) #Focus on protein coding genes

#Mutate so each row contains only a single gene in the "gene" column
gProfiler_df <- gProfiler_df %>%
  mutate(gene = strsplit(intersection, ",")) %>%
  unnest(gene) %>%
  dplyr::rename(stage = grade) %>%
  dplyr::filter(assay == "RNA")

# left_join RNAlistRDS005_df with Terms by c("cluster_id", "gene", "stage"), no multiple match warning
result_df <- left_join(RNAlistRDS005_df, gProfiler_df, by = c("cluster_id", "gene","stage"))

###Plot DEGs
# Create vulcano plot
ggplot(result_df, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
  geom_point(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
             aes(color = DEG_category), size = 2) +
  theme_minimal() +
  facet_wrap(~stage)
ggsave(filename = paste0(dir_out, 'vulcano_DEG_category_InterestingTerms.pdf'), width =8,height=6)

# Create vulcano plot with gene labels
#logFC vs AveExpr with gene labels
InterestingTerms <- result_df[,c(1:12,34)] %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms") %>% 
  distinct(gene, logFC,AveExpr,p_adj.loc, .keep_all = TRUE) %>% 
  #dplyr::filter(logFC > 0.5 | logFC < -0.5) %>% 
  dplyr::filter(stage == "all", annotation_reference == "Tan", AveExpr > 18.5, logFC > 0.2)
plot <- ggplot(result_df, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
  geom_point(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
             aes(color = DEG_category), size = 2) +
  theme_minimal()+
  ylim(0,18) +
  xlim(-.9,1.2) +
  geom_text_repel(data = InterestingTerms,
                  aes(label = gene),
                  size = 2,
                  max.overlaps = 50,
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50')
plot
ggsave(plot, filename = paste0(dir_out, 'vulcano_DEG_category_InterestingTerms_labelled.pdf'), width =16,height=12)

# Create vulcano plot with gene labels
#logFC vs AveExpr with gene labels
InterestingTerms <- result_df[,c(1:12,34)] %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms") %>% 
  distinct(gene, logFC,AveExpr,p_adj.loc, .keep_all = TRUE) %>% 
  #dplyr::filter(logFC > 0.5 | logFC < -0.5) %>% 
  dplyr::filter(stage == "all", annotation_reference == "Tan", abs(logFC) > 0.5)
result_df_all <- filter(result_df,stage == "all")
ggplot(result_df_all, aes(x = logFC, y = -log10(p_adj.loc))) +
  geom_point(aes(color = "Non-overlapping"), size = 1, color = "grey") +
  geom_point(data = result_df_all %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
             aes(x = logFC, y = -log10(p_adj.loc), color = DEG_category), size = 2) +
  theme_minimal() +
  ylim(-2, 17) +
  xlim(-0.9, 1.2) +
  geom_text_repel(data = result_df_all[2:12] %>% dplyr::filter(abs(logFC) > 0.5) %>% distinct(),
                  aes(label = gene),
                  size = 2,
                  max.overlaps = 50,
                  box.padding = 0.35,
                  point.padding = 0.5,
                  segment.color = 'grey50') +
  facet_wrap(~stage)
ggsave(filename = paste0(dir_out, 'vulcano_DEG_category_InterestingTerms_labelledlogFC05.pdf'), width =8,height=6)

results_df_logFC05 <- result_df[2:12] %>% dplyr::filter(abs(logFC) > 0.5) %>% distinct()
write.csv(results_df_logFC05, paste0(dir_out,'TermslogFC05.csv'))

#logFC vs AveExpr
ggplot(result_df, aes(x = log2(AveExpr-17.00183), y = logFC)) +
  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
  geom_point(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
             aes(color = DEG_category), size = 2) +
  theme_minimal() +
  geom_text(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"), aes(label = gene))
ggsave(filename = paste0(dir_out, 'AveExprLOG2_logFC_InterestingTerms.pdf'), width =8,height=6)

#logFC vs AveExpr with gene labels
InterestingTerms <- result_df[,c(1:12,34)] %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms") %>% 
  distinct(gene, logFC,AveExpr,p_adj.loc, .keep_all = TRUE) %>% 
  #dplyr::filter(logFC > 0.5 | logFC < -0.5) %>% 
  dplyr::filter(stage == "all", annotation_reference == "Tan", AveExpr > 18.5)

ggplot(result_df, aes(x = log2(AveExpr-17.00183), y = logFC)) +
  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
  geom_point(data = InterestingTerms , aes(color = DEG_category), size = 2) +
  theme_minimal() +
  geom_text_repel(data = InterestingTerms,
                  aes(label = gene),
                  size = 2,
                  max.overlaps = 35,
                  box.padding   = 0.35, 
                  point.padding = 0.5,
                  segment.color = 'grey50')
ggsave(filename = paste0(dir_out, 'AveExprLOG2_logFC_InterestingTerms_labelled.pdf'), width =14,height=8)

#logFC vs AveExpr
ggplot(result_df, aes(x = AveExpr, y = logFC)) +
  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
  geom_point(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
             aes(color = DEG_category), size = 2) +
  theme_minimal()
 ggsave(filename = paste0(dir_out, 'AveExpr_logFC_InterestingTerms.pdf'), width =8,height=6)

```

```{r Make DotPlots}
#Mutate so each row contains only a single gene in the "gene" column
gProfiler_df <- gProfiler_df %>%
  dplyr::filter(!is.na(DEG_category))
table(gProfiler_df$term_name,gProfiler_df$cluster_id)

table <- left_join(gProfiler_df, RNAlistRDS005_df[,c("gene","logFC","AveExpr","p_adj.loc","cluster_id","stage")], by = c("gene","cluster_id","stage"))

#x = -log(p_adj.loc)
table %>%
  arrange(DEG_category) %>%
  #dplyr::filter(term_name == "extracellular organelle") %>%
  ggplot2::ggplot() + 
  ggplot2::aes(y = reorder(gene, -log(logFC)), x = -log(p_adj.loc), color = stage, size = logFC) + 
  ggplot2::geom_point() + 
  ggplot2::scale_size_continuous("logFC") + 
  ggplot2::scale_x_continuous(limits = c(1, 6)) +
  ggplot2::theme_classic() +
  ggplot2::facet_wrap(~term_name, scales = "free_y", ncol = 13) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 8))
 ggsave(filename = paste0(dir_out, 'DotPlot_DEGcategories_DEGs_logFC_pval.pdf'), width =25,height=12)

#x = p_adj.loc
table %>%
  arrange(DEG_category) %>%
  #dplyr::filter(term_name == "extracellular organelle") %>%
  ggplot2::ggplot() + 
  ggplot2::aes(y = reorder(gene, -log(logFC)), x = p_adj.loc, color = stage, size = logFC) + 
  ggplot2::geom_point() + 
  ggplot2::scale_size_continuous("logFC") + 
  ggplot2::theme_classic() +
  ggplot2::facet_wrap(~term_name, scales = "free_y", ncol = 13) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 8))

#x = logFC
table %>%
  arrange(DEG_category) %>%
  #dplyr::filter(term_name == "extracellular organelle") %>%
  ggplot2::ggplot() + 
  ggplot2::aes(y = reorder(gene, -log(logFC)), x = logFC, color = stage, size = -log(p_adj.loc)) + 
  ggplot2::geom_point() + 
  #ggplot2::scale_color_continuous(name = "-Log(p_adj.loc)") + 
  #ggplot2::scale_size_continuous("logFC") + 
  ggplot2::theme_classic() +
  ggplot2::facet_wrap(~term_name, scales = "free_y", ncol = 13) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 8))
  #ggplot2::xlab("-Log(adj. p-value)") + 
  #ggplot2::ylab("GO term") + 
  #ggplot2::ggtitle("GO enrichment - all cells")

levels(factor(gProfiler_df$term_name))
table(gProfiler_df$term_name,gProfiler_df$term_size,grade)
n <- gProfiler_df %>% select(term_name, term_size,stage,cluster_id) %>% distinct() %>% arrange(term_size) %>% arrange(stage)

write.csv(table, paste0(dir_out,"table.csv"))
write.csv(RNAlistRDS005_df, paste0(dir_out,"RNAlistRDS005_df_Tan_AveExpr18.53202.csv"))

```

### 2025-01-24 03_PCA_analysis/PCAtools.R

@@ -1,86 +1,147 @@
 
 ###############################################
-### Load Libraries and define dir_out
+### Load Libraries and set path links
 ###############################################
 
+#Load libraries
 library(PCAtools)
 library(Seurat)
-library(dplyr)
-library(tibble)
-library(stringr)
-library(ggplot2)
-library(cowplot)
-library(ggbreak) 
-library(patchwork)
-library(ComplexUpset)
-library(ggrepel)
 
-dir_out <- ""
+
+##Set paths
+dir_ENDO <- "/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/"
+dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")
 dir.create(dir_out, recursive = TRUE)
 
 
+###############################################
+### Load Data
+###############################################
+##Load Data
+EndoAtlas <- readRDS(paste0(dir_ENDO, "_Data/EndoAtlas.rds"))
+
 
 ###############################################
-### Prepare Data and run PCA
+### Entire EndoAtlas: Data preparation and Conduct PCA
 ###############################################
 
-##Aggregate Expression per sample 
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
-bulk <- AggregateExpression(data, group.by = "sample", assay = "integrated", return.seurat = TRUE)
-
+##Data preparation
+#Aggregate Expression per sample 
+bulk <-  AggregateExpression(EndoAtlas, group.by = "sample", assay = "SCT", return.seurat = TRUE)
 # Extract counts matrix
-mat <- bulk[["integrated"]]@data
+mat <- bulk[["SCT"]]$data
+
+
+##Modify Metadata to match mat
+EndoAtlas_meta <- EndoAtlas@meta.data  %>%
+  distinct()
+# Reorder EndoAtlas_meta to match the order of columns in mat
+EndoAtlas_meta <- EndoAtlas_meta[match(colnames(mat), EndoAtlas_meta$sample), ]
+# Set the row names of EndoAtlas_meta to be the reordered sample column
+rownames(EndoAtlas_meta) <- EndoAtlas_meta$sample
+## check that sample names match exactly between pdata and expression data 
+all(colnames(mat) == rownames(EndoAtlas_meta))
+## [1] TRUE
 
[…]

### 2025-01-27 Figures/PaperFiguresextended.Rmd
Added comments

@@
 
-dir_out <- "/home/common/data/output/projects/ENDO/E044/A033/"
-dir.create(dir_out, recursive = TRUE)
-setwd(dir_out)
-```
-
-In script
-Replaced ProlifPeriSecFinal with MenstrualCyclePhase_main
-
-```{r save downsampled data}
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
-
-# Downsample the number of cells per identity class
-Idents(data) <- data$sample
-data_ds <- subset(x = data, downsample = 200)
-saveRDS(data_ds, "/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
-
-#Some plotting
-DimPlot(data, group.by = "AnnotationMain", raster = FALSE)
-DimPlot(data, group.by = "Symphony_Refined_A029", raster = FALSE)
-DimPlot(data, group.by = "Symphony_Refined_A029_United", raster = FALSE)
-
-#Save object subsetted for ScaiNet samples
-ScaiNETsamples <- read.csv("/home/common/data/output/projects/ENDO/E041/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")$names
-Idents(data) <- data$sample
-subset <- subset(data, idents = ScaiNETsamples)
-rm(data)
-gc()
-meta <- subset@meta.data
-saveRDS(subset, "/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_ScaiNETsamples.rds")
-rm(subset)
-gc()
-
-meta$AnnotationMain
-
-```
-
-Remove? TODO
-```{r order of samples by menstrual cycle phase and pseudotime}
-meta_final <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
-
-MenstrualCyclePhase_sampleOrder <- meta_final %>%
-  arrange(MenstrualCyclePhase,epithelial.pseudotime) %>%
-  distinct(sample,epithelial.pseudotime, MenstrualCyclePhase,CycleDay) %>%
-  pull(sample)
-saveRDS(MenstrualCyclePhase_sampleOrder, paste0(dir_out,"sampleorder_MenstrualCyclePhase_greenCirclemedian.RDS"))
-```
+```
+
+ln -s /home/common/data/output/projects/ENDO/E044/A033/EndoAtlas_GEO_meta_complete_downsampled.rds EndoAtlas.rds
+EndoAtlas_meta <- readRDS(paste0(dir_data,"EndoAtlas_meta.rds"))
+EndoAtlas <- readRDS(paste0(dir_data,"EndoAtlas.rds"))
+EndoAtlas <- AddMetaData(EndoAtlas, EndoAtlas_meta)
+saveRDS(EndoAtlas,paste0(dir_data,"EndoAtlas.rds") )
 
@@
 #Set directory links and create directories
-dir_data <- "L:/GitHubDesktop/endometriosis_endometrium_scRNA_atlas/_Data/"
-dir_out <- "L:/GitHubDesktop/endometriosis_endometrium_scRNA_atlas/Figures_out/"
+dir_ENDO <- "/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/"
+
+dir_data <- paste0(dir_ENDO,"_Data/")
+dir_out <- paste0(dir_ENDO,"Figures_out/")
 if (!dir.exists(dir_out)) {
-  dir.create(dir_out)
+  dir.create(dir_out, recursive = TRUE)
 }
@@
 #Load data
-#A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_metadata_ptime_snn_Final.rds")
-meta <- readRDS(paste0(dir_data, "EndoAtlas_GEO_meta_complete.rds"))
+EndoAtlas <- readRDS(paste0(dir_data, "EndoAtlas.rds"))
+EndoAtlas_meta <- EndoAtlas@meta.data
+
+#Preserve cell names
+EndoAtlas_meta$cell_names <- rownames(EndoAtlas_meta)
 
@@
 # Create column MenstrualCyclePhase_main based on MenstrualCyclePhase by merging the secretory sub-phases from 
-meta$MenstrualCyclePhase_main <- as.character(meta$MenstrualCyclePhase)
-meta$MenstrualCyclePhase_main[meta$MenstrualCyclePhase_main %in% c("secretory_early", "secretory_mid", "secretory_late")] <- "secretory"
+EndoAtlas_meta$MenstrualCyclePhase_main <- as.character(EndoAtlas_meta$MenstrualCyclePhase)
+EndoAtlas_meta$MenstrualCyclePhase_main[EndoAtlas_meta$MenstrualCyclePhase_main %in% c("secretory_early", "secretory_mid", "secretory_late")] <- "secretory"
 #Factor samples by MenstrualCyclePhase_main for plotting
-meta$MenstrualCyclePhase_main <- factor(meta$MenstrualCyclePhase_main, levels = c("proliferative","periovulatory","secretory"))
+EndoAtlas_meta$MenstrualCyclePhase_main <- factor(EndoAtlas_meta$MenstrualCyclePhase_main, levels = c("proliferative","periovulatory","secretory"))
 
@@
 # Perform a left join of the `meta` dataframe with `SupplementaryTable1`
-meta <- dplyr::left_join(
-  meta,
+EndoAtlas_meta <- dplyr::left_join(
+  EndoAtlas_meta,
   SupplementaryTable1 %>% 
@@
       sequencing.batch,
+      X10x.capturing.batch,
       Minor.exclusion.criteria.logical,
@@
 
-#Double-check, TODO
-table(meta$sample, meta$Minor.exclusion.criteria.logical)
-prolifsamples <- meta%>%
-  dplyr::filter(MenstrualCyclePhase_main == "proliferative") %>%
-  select(sample) %>%
-  distinct() %>%
-  arrange(sample) %>%
-  pull(sample)
-DEGsamples <- unique(read.csv("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/celltype_freq_sample.csv")$Var2)
-DEGsamples <- DEGsamples[order(DEGsamples)]
-setdiff(prolifsamples,DEGsamples) #MinorExclusion samples overlap with samples excluded in DEG analysis, sample87 is late secretory
-
 ##Determine order of samples by MenstrualCyclePhase and epithelial.pseudotime
-MenstrualCyclePhase_sampleOrder <- meta%>%
+MenstrualCyclePhase_sampleOrder <- EndoAtlas_meta%>%
   select(sample, MenstrualCyclePhase,epithelial.pseudotime) %>%
@@
 #Factor samples by MenstrualCyclePhase and epithelial.pseudotime for plotting
-meta$sample <- factor(meta$sample, levels = c(MenstrualCyclePhase_sampleOrder$sample))
+EndoAtlas_meta$sample <- factor(EndoAtlas_meta$sample, levels = c(MenstrualCyclePhase_sampleOrder$sample))
+
+#Add cell names as row names
+rownames(EndoAtlas_meta) <- EndoAtlas_meta$cell_names
+
+#Save EndoAtlas_meta as .rds
+saveRDS(EndoAtlas_meta, paste0(dir_data, "EndoAtlas_meta.rds"))
+
+#EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))
 
@@
 
-Supplementary Data Figure 4 a
+Save epithelial object and upload to GEO.
+```{r}
+#Load monocl object
+data_monocl <- readRDS("/home/common/data/output/projects/ENDO/E044/A062/trajectory_day5-6-7_greenCircle/trajectory_timepoint0.rds")
+listData <- (data_monocl@colData@listData)
+listData <- listData["cell_ident"]
+
+EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))
+colnames(EndoAtlas_meta)
+
+########################
+# Extract cell.ident from listData
+cell_ids <- listData[["cell_ident"]]
+
+# Ensure cell_ids is not empty
+if (length(cell_ids) == 0) {
+  stop("cell.ident in listData is empty. Ensure it is populated before proceeding.")
+}
+
+# Filter and order EndoAtlas_meta by cell_ids
+filtered_meta <- EndoAtlas_meta[EndoAtlas_meta$cell_names %in% cell_ids, ]
+ordered_meta <- filtered_meta[match(cell_ids, filtered_meta$cell_names), ]
+
+# Verify alignment
+if (!all(ordered_meta$cell_names == cell_ids)) {
+  stop("Ordering of filtered EndoAtlas_meta does not match cell_ids.")
+}
+
+# Add columns from ordered_meta to listData
+columns_to_add <- c("sample", "EndometriosisStatus", "EndometriosisGrade", "CycleDay", 
+                    "ProgesteroneSerum", "MenstrualCyclePhase", "AnnotationMain", 
+                    "AnnotationRefined", "cell_names", "MenstrualCyclePhase_main")
+
+for (col in columns_to_add) {
+  if (col %in% colnames(ordered_meta)) {
+    listData[[col]] <- ordered_meta[[col]]
+  } else {
+    warning(paste("Column", col, "not found in EndoAtlas_meta. Skipping."))
+  }
+}
+
+# Update the main object
+data_monocl@colData@listData <- listData
+
+data_monocl@principal_graph_aux@listData[["UMAP"]][["pseudotime"]]
+
+saveRDS(data_monocl, paste0(dir_data, "epithelial_cells_monocle3.rds"))
+```
+
+###########################################
+Main Figures
+###########################################
+
+Figure 1c
+UMAP of the entire endometrium single-cell atlas from women with and without endometriosis, colored by cell types (n=63).
+```{r}
+# Define order of cell types
+order <- EndoAtlas_meta %>%
+  select(AnnotationMain, AnnotationRefined) %>%
+  filter(!is.na(AnnotationMain)) %>%
+  distinct() %>%
+  arrange(AnnotationRefined) %>%
+  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
+  arrange(AnnotationMain)
+EndoAtlas$AnnotationRefined <- factor(x = EndoAtlas$AnnotationRefined, levels = order$AnnotationRefined)
+# DimPlot
+Fig1c <- DimPlot(EndoAtlas, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
+Fig1c
+# Save DimPlot
+ggsave(filename = paste0(dir_out, 'Fig1c.pdf'),
+       plot = Fig1c,
+       width =20, height=9)
+
+```
+
+Figure 1d
+Main cell type frequencies per patient (n=60, box plot, left) and entire dataset (bar plot, middle), with mesenchymal cells being the most abundant cell type.
+The overall cell number distribution between non-ENDO, mild-ENDO and severe-ENDO is balanced (bar plot, right).
 ```{r plotting of cell types per sample}
-#########################################################################
-# bar plot of the number of AnnotationRefined cells in each sample
-df <- meta%>%
+## Fig1d_1 Cells per patient
+# Boxplot
+Fig1d_1 <- EndoAtlas_meta %>%
+  group_by(AnnotationMain) %>%
+  summarise(row_count = n()) %>% 
+  ggplot(aes(x = AnnotationMain, y = row_count)) +
+  geom_col() +
+  coord_flip()
+Fig1d_1
+# Save Boxplot
+ggsave(paste0(dir_out, "Fig1d_1.pdf"),
+  plot = last_plot(),
+  width = 15, height = 6)
+
+## Fig1d_2 Cells per main cell type
+# Barplot
+Fig1d_2 <- EndoAtlas_meta %>%
+  group_by(sample, AnnotationMain) %>%
+  summarise(row_count = n()) %>%
+  ggplot(aes(x = AnnotationMain, y = row_count)) +
+  geom_boxplot() +
+  scale_y_log10() +
+  coord_flip() +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
+Fig1d_2
+# Save Barplot
+ggsave(paste0(dir_out, "Fig1d_2.pdf"),
+         plot = last_plot(),
+         width = 15, height = 6)
+
+##Fig1d_3 cell number distribution between non-ENDO, mild-ENDO and severe-ENDO
+# Barplot
+Fig1d_3 <- EndoAtlas_meta%>%
+  filter(!is.na(AnnotationMain)) %>%
+  group_by(AnnotationMain, EndometriosisGrade) %>%
+  summarise(row_count = n()) %>%
+  ggplot(aes(x = AnnotationMain, y = row_count, fill = EndometriosisGrade)) +
+  geom_bar(stat = 'identity', position = 'fill') +
+  scale_y_continuous(expand = c(0, 0)) +
+  coord_flip()
+Fig1d_3
+# Save Barplot
+ggsave(paste0(dir_out, "Fig1d_3.pdf"),
+  plot = last_plot(),
+  width = 6, height = 4)
+
+```
+
+Figure 1e
+Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
+
+CODE running, need to extend the PCA analysis folder
+dir_out <- "/home/common/data/output/projects/ENDO/E044/A072/allcells_EndoAtlas_SCTassay/"
+```{r}
+## Figure 1e
+# Load data
+p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
+# Plot and save eigencorplot
+pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
+eigencorplot(p,
+    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
+    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
+    corMultipleTestCorrection = "BH"
+)
+dev.off()
+```
+
+Figure 2a
+Scores plot of the first two principal components from PCA of sample-wise whole transcriptome.
+```{r}
+## Figure 2a
+# Load data
+p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
+# biplot
+biplot(p,
+    colby = 'MenstrualCyclePhase',
+    # ellipse config
+      ellipse = TRUE,
+      ellipseType = 't',
+      ellipseLevel = 0.90,
+      ellipseFill = TRUE,
+      ellipseAlpha = 1/4,
+      ellipseLineSize = 1.0,
+    labSize = 0,drawConnectors = FALSE,
+    legendPosition = 'top', legendLabSize = 16, legendIconSize = 8.0)
+```
+
+Figure 2b
+UMAPs of the entire dataset (left) and the re-integrated epithelial cells (right), colored by menstrual cycle phase, show a clear visual separation between proliferative/periovulatory, early-, mid-, and late-secretory samples, particularly in mesenchymal (endometrial fibroblasts, C7 fibroblasts, and mural cells), epithelial (glandular), and endothelial cells.
+```{r}
+## Figure 2b entire atlas
+#DimPlot
+Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
+Fig2b_1
+#Save DimPlot
+ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
+       plot = Fig2b_1,
+       width =20, height=9)
+
+## Figure 2b epithelial cells
+#DimPlot
+Fig2b_2 <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
+Fig2b_2
+#Save DimPlot
+ggsave(filename = paste0(dir_out, 'Fig2b_2.pdf'),
+       plot = Fig2b_2,
+       width =20, height=9)
+
+```
+
+Figure 2c
+UMAP of the re-integrated epithelial cells, colored by pseudo time prediction from monocle3.
+```{r}
+# Figure 2c epithelial cells
+epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))
+
+# DimPlot of epithelial cells with pseudotime
+Fig2c <- plot_cells(epithelial_cells_monocle3, color_cells_by = "pseudotime",
+                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
+                     graph_label_size=3, cell_size = 1)
+Fig2c
+# Save DimPlot
+ggsave(filename = paste0(dir_out, 'Fig2c.pdf'),
+       plot = Fig2c,
+       width =20, height=9)
+```
+
+Figure 2d
+Heatmap of selected marker gene expression with high log fold changes over the key menstrual cycle phases (proliferative, periovulatory, early-, mid-, and late-secretory).
+```{r}
+#Remove samples 65 and 105, since these samples are just in between menstrual cycle phases and cannot be clearly assigned
+Idents(EndoAtlas) <- EndoAtlas$sample
+EndoAtlas_filtereds65s105 <- subset(EndoAtlas, subset = sample %in% setdiff(unique(EndoAtlas$sample), c("sample65", "sample105")))
+Idents(EndoAtlas_filtereds65s105) <- EndoAtlas_filtereds65s105$AnnotationMain
+
+#### Plot heatmaps for main cell type specific markers for the different menstrual cycle phases. Compare also to Wang et al 2020 marker genes (Wang et al. 2020, DOI: 10.1038/s41591-020-1040-z)
+
+###########################
+### main epithelial marker genes for menstrual cycle phases
+
+# Subset celltype of interest
+EndoAtlas_filtereds65s105 <- subset(EndoAtlas_filtereds65s105, idents = epithelial)
+#main epithelial marker genes for menstrual cycle phases
+epithelial_main <- c("SFRP4", "LAMC2", "MMP11", "MMP7", "SPRY1", "TFPI2","STEAP4","ADAMTS8","CAPN6","SLC37A2","SCGB1D2", "MT1F", "MT1X","RIMKLB", "S100P", "DEFB1","G0S2","NNMT","GPX3", "PAEP")
+
+# Do MultiBarHeatmap
+DoMultiBarHeatmap(object = data, features = epithelial_main, group.by="MenstrualCyclePhase", additional.group.by="epithelial.pseudotime")
+# Save MultiBarHeatmap
+ggsave(
+    filename = paste0(dir_out, epithelial, "Fig2d_epithelial_DoMultiBarHeatmap.pdf"),
+    plot = last_plot(),
+    width = 15,
+    height = 8)
+
+##Plot our marker genes and Wang et al. 2020 marker genes (Wang et al. 2020, DOI: 10.1038/s41591-020-1040-z)
+#Wang et al. 2020 menstrual phase marker genes
+Wang_epithelial_mens_markers <- c("PLAU", "MMP7", "THBS1", "CADM1", "NPAS3", "ATP1A1", "ANK3", "ALPL", "TRAK1", "SCGB1D2", "MT1F", "MT1X", "MT1E", "MT1G", "CXCL14", "MAOA", "DPP4", "NUPR1", "GPX3", "PAEP")
+#Plot first our genes, then Wang genes
+heatmap <- DoHeatmap(
+  object = data,
+  features = c(setdiff(epithelial_main ,Wang_epithelial_mens_markers), Wang_epithelial_mens_markers),
+  cells = NULL,
+  group.by = "MenstrualCyclePhase",
+  group.bar = TRUE)
+#heatmap
+ggsave(
+    filename = paste0(dir_out, epithelial, "DoMultiBarHeatmap_epithelial_OursAndWangMarkers.pdf"),
+    plot = last_plot(),
+    width = 15,
+    height = 8)
+
+###########################
+### main stromal marker genes for menstrual cycle phases
+
+# Subset celltype of interest
+EndoAtlas_filtereds65s105 <- subset(EndoAtlas_filtereds65s105, idents = stromal)
+#main stromal marker genes for menstrual cycle phases
+stromal_main <- c("SFRP1", "H1-1", "H3C2", "WNT5A","CILP","BRINP1", "CFD","PTGDS","SLIT3","RPRM", "TIMP3", "IGFBP1","LEFTY2","LRRC15","LUM")
+# Do MultiBarHeatmap
+DoMultiBarHeatmap(object = data, features = stromal_main, group.by="MenstrualCyclePhase", additional.group.by="epithelial.pseudotime")
+# Save MultiBarHeatmap
+ggsave(
+    filename = paste0(dir_out, stromal, "Fig2d_stromal_DoMultiBarHeatmap.pdf"),
+    plot = last_plot(),
+    width = 15,
+    height = 10)
+
+##Plot our marker genes and Wang et al. 2020 marker genes (Wang et al. 2020, DOI: 10.1038/s41591-020-1040-z)
+#Wang et al. 2020 menstrual phase marker genes
+Wang_stromal_mens_markers <- c('STC1', 'NFATC2', 'BMP2' ,'PMAIP1' ,'MMP11' ,'SFRP1', 'WNT5A' ,'ZFYVE21' ,'CILP' ,'SLF2' ,'MATN2', 'S100A4' ,'DKK1' ,'CRYAB', 'FOXO1', 'IL15', 'FGF7', 'LMCD1')
+#Plot first our genes, then Wang genes
+heatmap <- DoHeatmap(
+  object = data,
+  features = c(setdiff(stromal_main ,Wang_stromal_mens_markers), Wang_stromal_mens_markers),
+  cells = NULL,
+  group.by = "MenstrualCyclePhase",
+  group.bar = TRUE)
+#heatmap
+ggsave(
+    filename = paste0(dir_out, stromal, "DoMultiBarHeatmap_stromal_OursAndWangMarkers.pdf"),
+    plot = last_plot(),
+    width = 15,
+    height = 15)
+
+###########################
+### main endothelial, myeloid and lymphocyte marker genes for menstrual cycle phases
+endothelial_main <- c("RGS16","DUOX1","CFD","MYC","TIMP3","LEFTY2","IL1B")
+myeloid_main <- c("S100A10","TREM2","STXBP2","CCL2","GNLY","CXCL3","CXCL2")
+lymphocyte_main <- c("IL7R","DUSP4","MMP26","MT1G","CXCL3","CXCL2","G0S2")
+
+#Make heatmaps in loop
+CELLTYPES <- c("endothelial","lymphocyte","myeloid") 
+for (CELLTYPE in CELLTYPES) {
+  # Subset celltype of interest
+  EndoAtlas_filtereds65s105 <- subset(EndoAtlas_filtereds65s105, idents = CELLTYPE)
+  # Do MultiBarHeatmap
+  DoMultiBarHeatmap(object = EndoAtlas_filtereds65s105, 
+                  features = if (CELLTYPE == "endothelial") endothelial_main
+                              else if (CELLTYPE == "myeloid") myeloid_main
+                              else if (CELLTYPE == "lymphocyte") lymphocyte_main, 
+                  group.by="MenstrualCyclePhase", 
+                  additional.group.by="epithelial.pseudotime")
+  # Save MultiBarHeatmap
+  ggsave(
+    filename = paste0(dir_out, "Fig2d_", CELLTYPE, "_DoMultiBarHeatmap.pdf"),
+    plot = last_plot(),
+    width = 15,
+    height = 3)
+
+}
+
+###########################
+## Additional strong Menstrual cycle phase markers
+stromal_extended <- c("H1-1","H1-2","H1-4","H1-5","H2AC13","H2Ac14","H3C2","H4C3","MKI67","TNC", #prolif/periov
+                      "CALB2", "PENK","POSTN",#periovulatory
+                      "CAPN6", "CYP26A1","HGD","HLA-DMB","MMP26","UPK1B","TFPI2","ENPP3","PKHD1L1","CILP","CHODL","GDF15","SCGB1D2","SCGB1D4","SCGB2A1","CERNA2","CD9",#secretory_early
+                      "MT1E","MT1F","MT1G","MT1H","MT1M", "LINC01320",#secretory_early/_mid
+                      "ACKR1", #secretory_mid
+                      "BRINP1","C1QC","CCN3","CFD","GPX3","MPTL2","MUC16","MYC","PAEP","PDK4","PTGDS","RPRM","SLIT3","SLPI","SPP1","TGM2","TIMP3","TM1E","TM4SF1",
+                      "IGFBP1","IGFBP6","KRT7","LEFTY2","LMCD1","LRRC15","LUM","NID2","PLAGL1","PLAT","PRLR","RBP4","BAMBI","CAB39L","COMP", #secretory_late
+                      "APOD","CLDN4","CLU","CP","CXCL13","CXCL14","DEFB1","DEPP1","DKK1") #secretory_early/_mid/_late
+epithelial_extended <- unique(c("SFRP4", "LAMC2", "MMP11", "TMEM107", "C12orf75", "UBE2C", "CENPW", "H2AC11", "H3C3", "H2AC13", "CSF3", "CENPE","SPRY1", "TFPI2", "STEAP4", "CKB", "LRRC26", "SNX30", "ADAMTS8", "CAPN6", "FXYD4", "XDH","CCN3", "SLC37A2","MMP26", "SCGB1D2", "SCGB1D4", "CYP26A1", "ENPP3", "PTGDS",  "PLEKHH2", "ZBTB16", "STC1", "IGFBP5", "PDGFRA", "REV3L", "HSPA1A",   "MT1F", "MT1G", "PAEP", "RIMKLB", "S100P", "DEFB1", "G0S2", "GDF15", "C12orf75", "DEFB1", "SPP1", "DKK1", "NNMT", "RBP4"))
+endothelial_extended <- c("POSTN","RGS16","TFPI2", "SCGB1D4",  "CYP26A1","MT1G", "CFD", "FABP5","APLN","TIMP3","MYC","MUC16", "NAPSB", "LEFTY2","IL1B", "COMP")
+myeloid_extended <- c("TREM2",  "PHYHIPL", "PAEP", "RIMKLB", "LEFTY2", "IL2RB", "APOD", "NNMT",  "STXBP2","GNLY", "S100A10", "OLR1", "MUC16", "CXCL13",  "INSIG1", "GZMA",  "PLAUR",  "MMP26",  "RGCC", "IGSF6",  "C15orf48", "UBD", "CCL19",  "FBP1", "CCL2", "FCN1", "CXCL3", "CORO1A","MT1G","INHBA", "CXCL5", "MMP10","SERPINB2","MMP1", "CCL18","SLC7A5","CCL20","THBS1","AREG", "APOC1","SERPINA1")
+lymphocyte_extended <- c("CAPN6","MT1H",  "ENPP3",  "TFPI2", "CLDN3", "DUSP4", "PDK4", "MMP26", "CXCL13", "GIMAP4",  "RIMKLB", "MT1G", "DEFB1", "CFD","IL7R", "LDLRAD4", "LEFTY2", "DOCK10", "CYP26A1", "G0S2","MTND1P23")
+
+```
+
+Figure 2e
+Boxplots of sample-wise mesenchymal and epithelial cell frequency as a percentage of total cells, split by menstrual cycle phase (stringent patient exclusion criteria, 32 ENDO, and 22 non-ENDO samples).
+```{r plotting of cell types per sample}
+##Boxplots of main cell types cell frequency as a percentage of total cells, split by menstrual cycle phase and endometriosis status
+
+#Prepare data for boxplot
+df_Fig2e <- EndoAtlas_meta%>%
+  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
+  group_by(sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main, EndometriosisStatus) %>%
+  summarise(row_count = n()) %>%
+  ungroup() %>%
+  filter(!is.na(AnnotationMain)) %>%
+  group_by(sample) %>%
+  mutate(AnnotationMain_perc = row_count / sum(row_count) * 100) %>%
+  ungroup()
+
+#BoxPlots percentage, EndometriosisStatus 
+Fig2e <- ggplot(df_Fig2e, aes(x = MenstrualCyclePhase_main, y = AnnotationMain_perc, fill = EndometriosisStatus)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_point(aes(colour = EndometriosisStatus),
+             colour="black",pch=21, 
+             position = position_jitterdodge()) +
+  RotatedAxis() +
+  xlab('Sample ID') +
+  facet_wrap(~AnnotationMain,scales = "free", ncol = 6)
+ggsave(paste0(dir_out, "Fig2e_boxplot.pdf"),
+  plot = Fig2e,
+  width = 12, height = 4.4)
+
+###Significance testing CTL vs Endo
+# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
+df_Fig2e_ttest_EndometriosisStatus <- df_Fig2e %>%
+  group_by(AnnotationMain, MenstrualCyclePhase_main) %>%
+  summarise(p.value = t.test(AnnotationMain_perc ~ EndometriosisStatus)$p.value) %>%
+  ungroup()
+#epithelial, proliferative just not significant (p.value = 0.0898)
+
+###Significance testing prolif vs periov vs sec, CTL + Endo
+## Perform t-test Prolif vs Peri for each cell type
+df_Fig2e_ttest_ProlifPeri <- df_Fig2e %>%
+  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
+  group_by(AnnotationMain) %>%
+  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
+  ungroup()
+#none significant
+## Perform t-test Prolif vs Sec for each cell type
+df_Fig2e_ttest_ProlifSec <- df_Fig2e %>%
+  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
+  group_by(AnnotationMain) %>%
+  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
+  ungroup()
+#epi (0.00977) and str (0.0255) significant
+
+## Perform t-test periovulatory vs sec for each cell type
+df_Fig2e_ttest_PeriSec <- df_Fig2e %>%
+  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
+  group_by(AnnotationMain) %>%
+  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
+  ungroup()
+#epi (0.00794) and str (0.0137) significant
+```
+
+Make better comments! xxx
+Figure 2f
+Boxplot of sample-wise secretory mural cell frequency as a percentage of total cells.
+```{r plotting of cell types per sample}
+#Early and Mid Secretory Prv_VSMCs 
+df_Fig2f <- EndoAtlas_meta%>%
+  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
+  group_by(sample, AnnotationRefined) %>%
+  summarise(AnnotationRefined_counts = n()) %>%
+  ungroup() %>%
+  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
+  filter(!is.na(AnnotationRefined)) %>%
+  group_by(sample) %>%
+  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
+  ungroup() %>% #60 samples * 63 cell types = 3780 rows, correct
+  filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
+df_Fig2f <- left_join(df_Fig2f, dplyr::distinct(EndoAtlas_meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main","MenstrualCyclePhase")]), by = "sample")
+
+secEarylMid_Prv_VSMC_MenstrualCyclePhase_main <- df_Fig2f %>%
+  filter(AnnotationRefined %in% c("Prv_VSMC secretory", "Prv_VSMC secretory_late")) %>% #same result with and without "Prv_VSMC secretory_late"
+  dplyr::filter(MenstrualCyclePhase %in% c("proliferative","periovulatory","secretory_mid","secretory_early")) %>%
+  group_by(sample, EndometriosisStatus, MenstrualCyclePhase_main) %>%
+  summarize(
+    total_counts_MenstrualCyclePhase_main = sum(AnnotationRefined_counts, na.rm = TRUE),
+    total_perc_MenstrualCyclePhase_main = sum(AnnotationRefined_perc, na.rm = TRUE)
+  ) %>%
+  ungroup()
+
+ggplot(secEarylMid_Prv_VSMC_MenstrualCyclePhase_main, aes(x = MenstrualCyclePhase_main, y = total_perc_MenstrualCyclePhase_main, fill = EndometriosisStatus)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_point(aes(colour = EndometriosisStatus),
+             colour="black",pch=21, 
+             position = position_jitterdodge(jitter.width = 0.2))
+ggsave(paste0(dir_out, "Fig2f_boxplot.pdf"),
+  plot = last_plot(),
+  width = 17, height = 12)
+
+#Statistical testing
+secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
+  dplyr::filter(MenstrualCyclePhase_main %in% c("secretory")) %>% 
+  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
+  ungroup()
+#significant, p.value 0.0261
+
+secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval_prolif <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
+  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative")) %>% 
+  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
+  ungroup()
+#not significant, p.value NA
+
+secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval_peri <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
+  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory")) %>% 
+  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
+  ungroup()
+#not significant, p.value 0.7639853
+
+#Multiple testing correction
+p.adjust(c(0.0261), method = "BH", n = length(3)) #Value does not change
+p.adjust(c(0.7639853), method = "BH", n = length(3)) #Value does not change
+```
+
+Figure 3a and 3b
+3a Barplot showing number of unique differentially expressed genes (DEGs, adjusted p-value < 0.05, average expression > 18.53202) between non-ENDO (n = 11) and all-ENDO (n = 12), mild-ENDO (n = 4) and severe-ENDO (n = 8) samples within immune, endothelial, epithelial, and mesenchymal cell types, analyzed by Muscat.
+3b Selected terms enriched in the cluster- and stage-wise functional enrichment analysis of DEGs, with term sizes between 25 and 750.
+```{r}
+
+DEGs
+E044A031
+
+```
+
+xxx
+Figure 3c-h
+(c) Selected signaling pathways with overall enriched activity in ENDO (n = 12) in comparison to non-ENDO (n = 11).
+(d) Circle plot of TNF signaling pathway between all cell types in ENDO (left, n = 12) and non-ENDO (right, n = 11).
+(e-h) Bubble plots of selected communications with significant interaction probabilities (unadjusted p-value < 0.01) within the ICAM (e), CXCL (f), IGFBP (g) and VEGF (h) signaling pathways for ENDO (n = 12) and non-ENDO (n = 11).
+Further p-value adjustment is not recommended by the CellChat author.
+```{r}
+
+CellChat
+E044A076
+#cellchat <-readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
+dir_out <- "/home/common/data/output/projects/ENDO/E044/A076/"
+
+#Figure 3c and Supplementary Figure 5b
+gk1 <- rankNet(cellchat, mode = "comparison", stacked = T, do.stat = TRUE)
+gk1 
+ggsave(plot = gk1,
+       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12)
+
+#Figure 3d (or was it done separately? xxx)
+setwd(paste0(dir_out,"output_comparison_Refined"))
+###circle plot
+pathways.show <- c("TNF") 
+object.list <- list(Endo = cellchat.Endo, Cntrl = cellchat.Cntrl)
+weight.max <- getMaxWeight(object.list, slot.name = c("netP"), attribute = pathways.show) 
+par(mfrow = c(1,2), xpd=TRUE)
+for (i in 1:length(object.list)) {
+  netVisual_aggregate(object.list[[i]], signaling = pathways.show, layout = "circle", edge.weight.max = weight.max[1], edge.width.max = 10, signaling.name = paste(pathways.show, names(object.list)[i]))
+}
+
+# Save plot
+pdf(file = "EndoCtl_TNF_circle_plot_edge10.pdf")
+par(mfrow = c(1, length(object.list)), xpd = TRUE)
+for (i in 1:length(object.list)) {
+  netVisual_aggregate(object.list[[i]], signaling = pathways.show, layout = "circle", edge.weight.max = weight.max[1], edge.width.max = 10, signaling.name = paste(pathways.show, names(object.list)[i]))
+}
+dev.off()
+
+# Save plot
+pdf(file = "EndoCtl_TNF_circle_plot_edge15.pdf")
+par(mfrow = c(1, length(object.list)), xpd = TRUE)
+for (i in 1:length(object.list)) {
+  netVisual_aggregate(object.list[[i]], signaling = pathways.show, layout = "circle", edge.weight.max = weight.max[1], edge.width.max = 10, signaling.name = paste(pathways.show, names(object.list)[i]))
+}
+dev.off()
+####################
+
+#Load CellChat data
+cellchat <- readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
+
+#Figure 3e ICAM
+netVisual_bubble(cellchat, comparison = c(1, 2),signaling = c("ICAM"), remove.isolate = FALSE) 
+ggsave(plot = last_plot(),
+       filename = paste0(dir_out, 'output_comparison_Refined/netVisual_bubble_ICAM.pdf'), width =25,height=5)
+
+#Figure 3f CXCL
+#Load data
+netVisual_bubble(cellchat, comparison = c(1, 2),signaling = c("CXCL"), remove.isolate = FALSE) 
+ggsave(plot = last_plot(),
+       filename = paste0(dir_out, 'output_comparison_Refined/netVisual_bubble_CXCL.pdf'), width =25,height=7)
+
+#Figure 3g VEGF
+netVisual_bubble(cellchat, comparison = c(1, 2),signaling = c("VEGF"), remove.isolate = FALSE) # sources.use = 4, targets.use = c(5:11),
+ggsave(plot = last_plot(),
+       filename = paste0(dir_out, 'output_comparison_Refined/netVisual_bubble_VEGF.pdf'), width =15,height=5)
+
+#Figure 3h IGFBP
+netVisual_bubble(cellchat, comparison = c(1, 2),signaling = c("IGFBP"), remove.isolate = FALSE,targets.use = c(5:11)) 
+ggsave(plot = last_plot(),
+       filename = paste0(dir_out, 'output_comparison_Refined/netVisual_bubble_IGFBP.pdf'), width =15,height=5)
+
+```
+
+###########################################
+Supplementary Data Figures
+###########################################
+
+Supplementary Data Figure 1a and 1b
+Violin plots illustrating counts and features from RNA assay (a) and SCT assay (b), show no discernable differences between ENDO and non-ENDO.
+```{r VlnPlots of nCounts and nFeatures}
+#VlnPlots for "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"
+VlnPlot(EndoAtlas,
+  features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"),
+  group.by='EndometriosisStatus', ncol = 1, pt.size=0)
+# Save VlnPlots
+ggsave(paste0(dir_out, "VlnPlot_GroupByEndometriosisStatus_nFeature_nCounts_RNA_SCT_percMT.pdf"),
+  plot = last_plot(),
+  width = 4, height = 13)
+```
+
+Supplementary Data Figure 1c and d
+UMAPs of the entire endometrium single-cell atlas, colored by menstrual cycle day (c) and progesterone in serum (d), show a clear visual separation of cells by menstrual cycle day and serum progesterone concentration.
+```{r}
+## SFig1c
+# Featureplot
+SFig1c <- FeaturePlot(EndoAtlas, features = "CycleDay", raster = FALSE)
+SFig1c
+# Save Featureplot
+ggsave(filename = paste0(dir_out, 'SFig1c.pdf'),
+       plot = SFig1c,
+       width =20, height=9)
+
+## SFig1d
+# Featureplot
+SFig1d <- FeaturePlot(EndoAtlas, features = "ProgesteroneSerum", raster = FALSE)
+SFig1d
+# Save Featureplot
+ggsave(filename = paste0(dir_out, 'SFig1d.pdf'),
+       plot = SFig1d,
+       width =20, height=9)
+
+```
+
+xxx double-check code
+Supplementary Data Figure 1e
+Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) of the proliferative samples with strict exclusion criteria with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
+```{r}
+dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")
+
+p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_DEGsamples.rds"))
+
+#Figure 1e
+pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
+eigencorplot(p,
+    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
+    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
+    corMultipleTestCorrection = "BH"
+)
+dev.off()
+```
+
+Supplementary Data Figure 1f
+Eigencor plots showing the effect of different co-variants of the atlas dataset on principal components (PCs) of the main 5 cell types (epithelial, mesenchymal, endothelial, myeloid, lymphoid) with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
+"/home/common/data/output/projects/ENDO/E044/A082/SCTassay/"
+```{r}
+
+#Define subsets of main annotation
+subsets <- c("myeloid", "endothelial", "epithelial", "lymphocyte","stromal")
+
+#Plot eigencorplots for the main annotations
+plotting_PCA <- function(SUBSET) {
+    p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_",SUBSET,".rds"))
+
+    pdf(paste0(dir_out,"eigencorplot_final_BH_",SUBSET,".pdf"), width = 10, height = 5)
+    eigencorplot(p,
+    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
+    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
+    corMultipleTestCorrection = "BH"
+    )
+    dev.off()
+}
+
+lapply(subsets, plotting_PCA)
+```
+
+Supplementary Data Figure 2
+Expression of the marker genes for the different cell types corresponds to the cell type-specific expression shown in Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5).
+```{refined cluster markers}
+
+# Set SCT as default and remove RNA to save space
+DefaultAssay(EndoAtlas) <- "SCT"
+EndoAtlas@assays$RNA <- NULL
+gc()
+# Remove data without annotation (no NAs!)
+Idents(EndoAtlas) <- EndoAtlas$AnnotationMain
+EndoAtlas <- subset(EndoAtlas, subset = AnnotationMain %in% levels(as.factor(AnnotationMain$Symphony_Refined)))
+gc()
+
+### AnnotationMain markers
+genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2") #correct xxx
+pops_to_plot <- c("stromal","epithelial","myeloid","lymphocyte","endothelial")
+# DotPlot
+Seurat::DotPlot(
+  object = EndoAtlas,
+  scale = T,
+  features = genes_to_plot,
+  idents = pops_to_plot,
+) + theme(axis.text.x = element_text(angle = 90))
+# Save DotPlot
+ggsave(plot = last_plot(), "./SFig2_AnnotationMain.pdf", width = 10, height = 5)
+
+###Set identity to AnnotationRefined
+Idents(EndoAtlas) <- EndoAtlas$AnnotationRefined
+
+### DC markers (from Tan2022 Fig. 5a)
+genes_to_plot <- c("CD1C", "CLEC9A", "LAMP3")
+pops_to_plot <- c("mDC", "cDC1", "DC3", "cDC2", "pre-cDC2")
+# reorder clusters
+pops_seurat <- subset(EndoAtlas, idents = pops_to_plot)
+pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
+                            levels=pops_to_plot)
+# DotPlot
+p <- Seurat::DotPlot(
+  object = pops_seurat,
+  scale = T,
+  features = genes_to_plot,
+  idents = pops_to_plot,
+) + theme(axis.text.x = element_text(angle = 90))
+# Save DotPlot
+ggsave(plot = p, "./dotplots/Tan2022Fig3b_DCs_SCT.pdf", width = 5, height = 7)
+
+### Lymphocytes markers (from Tan2022 Fig. 4b)
+genes_to_plot <- c("FCGR3A","KLRD1","ITGA1","LILRB1","ITGAX","KLRC1","KLRC2","EPAS1","KIR2DL4","IGFBP2","CD160","GZMH","KLRF1","FGFBP2","TBX21","CD8A","CD8B","CRTAM","CCR5","NCR3","RORC","IL7R","TCF7","CD4","CCR7","SELL","LEF1","FOXP3","CTLA4","IL2RA","CXCR6","ZNF683","ITGAE","XCL1","SLC4A10","DPP4","GZMK","IGKC", "MZB1", "DERL3", "CD79A", "CD79B", "MS4A1", "BANK1")
+pops_to_plot <- c(EndoAtlas@meta.data %>% filter(AnnotationMain == "lymphocyte") %>% pull(AnnotationRefined) %>% unique())
+
+# reorder clusters
+pops_seurat <- subset(EndoAtlas, idents = pops_to_plot)
+pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
+                           levels=c("B cell", "plasma", "CD8 T$_{RM}$","CD4 T$_{RM}$","T$_{Reg}$", "T$_N$/T$_{CM}$" , "ILC" , "T$_{EM}$","CTL", "pNK", "NK3","NK2","NK1"))
+# DotPlot
+p <- Seurat::DotPlot(
+  object = pops_seurat,
+  scale = T,
+  features = genes_to_plot,
+  idents = pops_to_plot,
+) + theme(axis.text.x = element_text(angle = 90))
+# Save DotPlot
+ggsave(plot = p, "./dotplots/Tan2022Fig3b_Lymphocytes_SCT.pdf", width = 17, height = 7)
+
+### epithelial markers (from Tan2022 Sup Fig. 7c)
+genes_to_plot <- c("TPPP3", "PIFO", "FOXJ1", "TP63", "KRT5", "SOX9", "LGR5", "MUC5B", "TFF3", "RUNX3", "SAA1", "SIX1", "PROM1", "PROM2", "ANPEP", "KIAA1324", "SPDEF", "SMAD9", "CD36", "HSD17B2", "SCGB2A2", "GDA", "FGFR2", "WNT7A", "FGF9", "PTGS1", "MSLN", "UPK3B", "CALB2", "PGR", "ESR1", "MMP7", "TIMP1", "IDO1", "PSAT1", "ENPP3", "GNG11", "CREB3L1", "IHH", "TOP2A", "MT1F", "MT1E", "MT1X", "SERPINA5", "SPP1", "PAEP", "CXCL14", "DPP4")
+pops_to_plot <- c("mesothelial", "ciliated", "mid-secretory", "lumenal 2", "lumenal 1", "lumenal", "glandular early-secretory", "glandular", "TP63+/KRT5+", "MUC5B+")
+# reorder clusters
+pops_seurat <- subset(EndoAtlas, idents = pops_to_plot)
+pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
+                            levels=pops_to_plot)
+#DotPlot
+p <- Seurat::DotPlot(
+  object = pops_seurat,
+  scale = T,
+  features = genes_to_plot,
+  idents = pops_to_plot,
+) + theme(axis.text.x = element_text(angle = 90))
+# Save DotPlot
+ggsave(plot = p, "./dotplots/Tan2022Fig3b_epithelial_SCT.pdf", width = 17, height = 7)
+
+#myeloid markers ( from Tan2022 Fig. 4b)
+genes_to_plot <- c("FLOR2", "LYVE1", "MRC1", "CX3CR1", "ICAM2", "CLEC5A", "CCR2", "VEGFA", "FN1", "SPARC", "APOE", "SPP1")
+pops_to_plot <- c(EndoAtlas@meta.data %>% filter(AnnotationMain == "myeloid") %>% pull(AnnotationRefined) %>% unique())
+pops_to_plot <- c("M$\\Phi$5-activated","M$\\Phi$4-infiltrated","M$\\Phi$3-APOE","M$\\Phi$2-peritoneal","M$\\Phi$1-LYVE1")
+# reorder clusters
+pops_seurat <- subset(EndoAtlas, idents = pops_to_plot)
+pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
+                            levels=pops_to_plot)
+# DotPlot
+p <- Seurat::DotPlot(
+  object = pops_seurat,
+  scale = T,
+  features = genes_to_plot,
+  idents = pops_to_plot,
+) + theme(axis.text.x = element_text(angle = 90))
+# Save DotPlot
+ggsave(plot = p, "./dotplots/Tan2022Fig3b_macrophages_SCT.pdf", width = 10, height = 7)
+
+#Prv and VSMC markers (from Tan2022 Fig. 3b)
+genes_to_plot <- c("RERGL", "MCAM", "MYH11", "STEAP4","GGT5", "RGS5")
+pops_to_plot <- c("Prv-MYH11", "Prv-CCL19", "Prv-STEAP4", "VSMC")
+# reorder clusters
+pops_seurat <- subset(EndoAtlas, idents = pops_to_plot)
+pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
+                            levels=c("VSMC", "Prv-STEAP4", "Prv-CCL19","Prv-MYH11"))
+# DotPlot
+p <- Seurat::DotPlot(
+  object = pops_seurat,
+  scale = T,
+  features = genes_to_plot,
+  idents = pops_to_plot,
+) + theme(axis.text.x = element_text(angle = 90))
+# Save DotPlot
+ggsave(plot = p, "./dotplots/Tan2022Fig3b_stromal_SCT.pdf", width = 7, height = 7)
+
+```
+
+Supplementary Figure 3a
+Scores plot of the first two principal components PC1 and PC2 from sample-wise whole transcriptome PCA of the SCT assay, colored by endometriosis status.
+```{r}
+# Load data
+p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
+# biplot
+SFig3a <- biplot(p,
+    colby = 'EndometriosisStatus',
+    labSize = 0,drawConnectors = FALSE, 
+    hline = 0, vline = 0,
+    legendPosition = 'right')
+ggsave(plot = SFig3a, "./SFig3a.pdf", width = 7, height = 7)
+```
+
+Supplementary Figure 3b and e
+UMAP of re-integrated epithelial cells colored by menstrual cycle day and menstrual cycle phase Wang
+```{r}
+epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))
+
+#Menstrual Cycle Day
+SFig3b <- plot_cells(epithelial_cells_monocle3, color_cells_by = "CycleDay",
+                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
+                     graph_label_size=3, cell_size = 1)
+SFig3b
+ggsave(filename = paste0(dir_out, 'SFig3b.pdf'),
+       plot = SFig3b,
+       width =20, height=9)
+
+#Menstual Cycle Phase Wang
+SFig3e <- plot_cells(epithelial_cells_monocle3 , color_cells_by = "WangMenstrualCyclePhase",
+                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
+                     graph_label_size=3, cell_size = 1)
+SFig3e
+ggsave(filename = paste0(dir_out, 'SFig3e.pdf'),
+       plot = SFig3e,
+       width =20, height=9)
+
+```
+
+xxxsee above for object name in GEO
+Supplementary Data Figure 3c
+Look in IBU folder for object with trajectory line or A063/A066
+```{r}
+#Figure S3c epithelial cells
+FeaturePlot
+Fig3c <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCycleDay", raster = FALSE)
+Fig3c
+ggsave(filename = paste0(dir_out, 'Fig3c.pdf'),
+       plot = Fig3c,
+       width =20, height=9)
+```
+
+Supplementary Data Figure 3f
+CODE OK
+```{r}
+#Figure 2b entire atlas
+Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "Wang2020Phase", raster = FALSE)
+Fig2b_1
+ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
+       plot = Fig2b_1,
+       width =20, height=9)
+
+```
+
+Supplementary Data Figure 3 g
+CODE OK
+```{r}
+####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
+#cd endometriosis_endometrium_scRNA_atlas/_Data
+#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad
+
+####### Convert .h5ad files to .h5seurat files
+library(SeuratDisk)
+Convert(paste0(dir_data,"endo-2022_epithelial.h5ad"), dest = paste0(dir_data,"endo-2022_epithelial.h5seurat"), assay = "RNA", overwrite = TRUE)
+Tan_epithelial <- LoadH5Seurat(paste0(dir_data,"endo-2022_epithelial.h5seurat"), assays = "RNA")
+
+#Create scale.data slot for the RNA assay required for heatmapping
+Tan_epithelial_features <- Tan_epithelial@assays[["RNA"]]@data@Dimnames[[1]]
+Tan_epithelial <- ScaleData(object = Tan_epithelial, features = Tan_epithelial_features)
+
+#Subset cells from the endometrium (exclude cells from endometriosis lesions)
+Idents(Tan_epithelial) <- Tan_epithelial$sample_type_rename
+Tan_epithelial <- subset(Tan_epithelial, idents = c("Ctrl", "EuE"))
+
+# factor levels in better order (E10 and E11 don't have endometrium cells)
+Tan_epithelial$PID <- factor(Tan_epithelial$PID, levels = c("C01", "C02", "C03", "E01", "E02", "E03", "E04", "E05", "E06", "E07", "E08", "E09"))
+
+# Generate heatmap of the endometrium cells from Tan et al. 2022 with the same epithelial menstrual marker genes as in our manuscript Figure 2d, for comparison
+feature_list_figure2 <- c("SFRP4", "LAMC2", "MMP11", "MMP7", "SPRY1", "TFPI2","STEAP4","ADAMTS8","CAPN6","SLC37A2","SCGB1D2", "MT1F", "MT1X","RIMKLB", "S100P", "DEFB1","G0S2","NNMT","GPX3", "PAEP")
+
+heatmap <- DoHeatmap(
+  object = Tan_epithelial,
+  features = intersect(feature_list_figure2, Tan_epithelial_features),
+  cells = NULL,
+  group.by = "PID",
+  assay = "RNA",
+  group.bar = TRUE) + 
+  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
+heatmap
+ggsave(
+    filename = paste0(dir_out, "Tan_epithelial_Fig2EpiMarkers.pdf"),
+    plot = last_plot(),
+    width = 12.3,
+    height = 5)
+
+#As an alternative control heatmap of menstrual marker genes from Wang et al. 2020 (DOI: 10.1038/s41591-020-1040-z)
+feature_list_Wang <- c("PLAU", "MMP7", "THBS1", "CADM1", "NPAS3", "ATP1A1", "ANK3", "ALPL", "TRAK1", "SCGB1D2", "MT1F", "MT1X", "MT1E", "MT1G", "CXCL14", "MAOA", "DPP4", "NUPR1", "GPX3", "PAEP")
+
+heatmap <- DoHeatmap(
+  object = Tan_epithelial,
+  features = intersect(feature_list_Wang, Tan_epithelial_features),
+  cells = NULL,
+  group.by = "PID",
+  assay = "RNA",
+  group.bar = TRUE) + 
+  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
+heatmap
+ggsave(
+    filename = paste0(dir_out, "Tan_epithelial_WangEpiMarkers.pdf"),
+    plot = last_plot(),
+    width = 15,
+    height = 15)
+
+#The menstrual marker gene expression varies largely between the endometrium samples in the Tan et al.
+2022 dataset, strongly indicating not homogeneous menstrual cycle phase between the endometrium samples.
+#The expression distribution of menstrual marker genes from our manuscript figure 2d and Wang et al. 2020 look comparable.
+
+```
+
+Supplementary Data Figure 3 h
+CODE OK
+```{r}
+####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
+#cd endometriosis_endometrium_scRNA_atlas/_Data
+#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad
+
+####### Convert .h5ad files to .h5seurat files
+library(SeuratDisk)
+Convert(paste0(dir_data,"endo-2022_stromal.h5ad"), dest = paste0(dir_data,"endo-2022_stromal.h5seurat"), assay = "RNA", overwrite = TRUE)
+Tan_stromal <- LoadH5Seurat(paste0(dir_data,"endo-2022_stromal.h5seurat"), assays = "RNA")
+
+#Create scale.data slot for the RNA assay required for heatmapping
+Tan_stromal_features <- Tan_stromal@assays[["RNA"]]@data@Dimnames[[1]]
+Tan_stromal <- ScaleData(object = Tan_stromal, features = Tan_stromal_features)
+
+#Subset cells from the endometrium (exclude cells from endometriosis lesions)
+Idents(Tan_stromal) <- Tan_stromal$sample_type_rename
+Tan_stromal <- subset(Tan_stromal, idents = c("Ctrl", "EuE"))
+
+# factor levels in better order (E10 and E11 don't have endometrium cells)
+Tan_stromal$PID <- factor(Tan_stromal$PID, levels = c("C01", "C02", "C03", "E01", "E02", "E03", "E04", "E05", "E06", "E07", "E08", "E09"))
+
+# Generate heatmap of the endometrium cells from Tan et al. 2022 with the same epithelial menstrual marker genes as in our manuscript Figure 2d, for comparison
+feature_list_figure2_stromal <- c("SFRP1", "H1-1", "H3C2", "WNT5A","CILP","BRINP1", "CFD","PTGDS","SLIT3","RPRM", "TIMP3", "IGFBP1","LEFTY2","LRRC15","LUM")
+
+heatmap <- DoHeatmap(
+  object = Tan_stromal,
+  features = intersect(feature_list_figure2_stromal, Tan_stromal_features),
+  cells = NULL,
+  group.by = "PID",
+  assay = "RNA",
+  group.bar = TRUE) + 
+  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
+heatmap
+ggsave(
+    filename = paste0(dir_out, "Tan_stromal_Fig2EpiMarkers.pdf"),
+    plot = last_plot(),
+    width = 12.3,
+    height = 5)
+
+#As an alternative control heatmap of menstrual marker genes from Wang et al. 2020 (DOI: 10.1038/s41591-020-1040-z)
+feature_list_Wang_stromal <- c('STC1', 'NFATC2', 'BMP2' ,'PMAIP1' ,'MMP11' ,'SFRP1', 'WNT5A' ,'ZFYVE21' ,'CILP' ,'SLF2' ,'MATN2', 'S100A4' ,'DKK1' ,'CRYAB', 'FOXO1', 'IL15', 'FGF7', 'LMCD1')
+
+heatmap <- DoHeatmap(
+  object = Tan_stromal,
+  features = intersect(feature_list_Wang_stromal, Tan_stromal_features),
+  cells = NULL,
+  group.by = "PID",
+  assay = "RNA",
+  group.bar = TRUE) + 
+  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
+heatmap
+ggsave(
+    filename = paste0(dir_out, "Tan_stromal_WangEpiMarkers.pdf"),
+    plot = last_plot(),
+    width = 15,
+    height = 15)
+
+#The menstrual marker gene expression varies largely between the endometrium samples in the Tan et al.
+2022 dataset, strongly indicating not homogeneous menstrual cycle phases between the endometrium samples.
+#The expression distribution of menstrual marker genes from our manuscript figure 2d and Wang et al. 2020 look comparable.
+
+```
+
+Supplementary Data Figure 4a
+CODE OK
+```{r plotting of cell types per sample}
+##Barplots of cell type frequency split by the main 5 cell types
+
+#Set colour palette
+P63 = createPalette(63, c("#ff0000", "#00ff00", "#0000ff")) 
+swatch(P63)
+names(P63) <- NULL
+
+#Make df with number of AnnotationRefined cells in each sample
+df <- EndoAtlas_meta%>%
   filter(!is.na(AnnotationMain)) %>%
@@
 
-#Bar plot
-#####use this for paper supplement
-#Bar plot split by main5 
-#MenstrualCyclePhase_sampleOrder <- readRDS(paste0(dir_out,"sampleorder_MenstrualCyclePhase_greenCirclemedian.RDS"))
-#df$sample <- factor(df$sample, levels = MenstrualCyclePhase_sampleOrder)
-P63 = createPalette(63, c("#ff0000", "#00ff00", "#0000ff")) 
-swatch(P63)
-names(P63) <- NULL
-#Make barplots with different seeds
-seed <- 1 #find in dir.create(paste0(dir_out, "/seedopt/"))
-set.seed(seed)  
+#Make bar plot
 p <- ggplot(df, aes(x = sample, y = row_count, fill = AnnotationRefined)) +
@@
   facet_grid(AnnotationMain ~ ., scales = "free_y")
-ggsave(paste0(dir_out, "/seedopt/Barplot_sample_AnnotationRefined_main5_", seed, ".pdf"),
+p
+#Save barplot
+ggsave(paste0(dir_out, "SFig4a_barplot.pdf"),
        plot = p,
@@
 
-##Consider leaving out
-a1 <- ggplot(df %>% select(sample, EndometriosisStatus) %>% distinct (), aes(x = sample, y = EndometriosisStatus, fill = EndometriosisStatus)) +
-  geom_bar(stat = 'identity', position = 'fill') +
-  xlab('Sample ID') +
-  theme(
-    axis.text.x = element_text(angle = 90, hjust = 1),
-    panel.grid.minor = element_blank(),
-    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
-  ) 
-a2 <- ggplot(df %>% select(sample, MenstrualCyclePhase) %>% distinct (), aes(x = sample, y = MenstrualCyclePhase, fill = MenstrualCyclePhase)) +
-  geom_bar(stat = 'identity', position = 'fill') +
-  xlab('Sample ID') +
-  theme(
-    axis.text.x = element_text(angle = 90, hjust = 1),
-    panel.grid.minor = element_blank(),
-    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
-  )
-a1 + a2 + plot_layout(nrow = 2)
-ggsave(paste0(dir_out, "Barplot_sample_AnnotationRefined_main5_annotation.pdf"),
-  plot = last_plot(),
-  width = 15, height = 6)
-
-```
-
-Figure 2 e
-```{r plotting of cell types per sample}
-##########################################################################
-#BoxPlots sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main
-df2 <- meta%>%
-  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
-  group_by(sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main, EndometriosisStatus) %>% #no sample, AnnotationMain combinations are dropped
-  summarise(row_count = n()) %>%
-  ungroup() %>%
-  filter(!is.na(AnnotationMain)) %>%
-  group_by(sample) %>%
-  mutate(AnnotationMain_perc = row_count / sum(row_count) * 100) %>%
-  ungroup()
-
-#BoxPlots percentage, EndometriosisStatus #for publication
-ggplot(df2, aes(x = MenstrualCyclePhase_main, y = AnnotationMain_perc, fill = EndometriosisStatus)) +
-  geom_boxplot(outlier.shape = NA) +
-  geom_point(aes(colour = EndometriosisStatus),
-             colour="black",pch=21, 
-             position = position_jitterdodge()) +
-  RotatedAxis() +
-  xlab('Sample ID') +
-  facet_wrap(~AnnotationMain,scales = "free", ncol = 6)
-ggsave(paste0(dir_out, "Boxplot_MenstrualCyclePhase_main_main5_sample_EndometriosisStatus_perc_publication.pdf"),
-  plot = last_plot(),
-  width = 12, height = 4.4)
-
-###Significance testing CTL vs Endo
-# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
-df2_ttest_EndometriosisStatus <- df2 %>%
-  group_by(AnnotationMain, MenstrualCyclePhase_main) %>%
-  summarise(p.value = t.test(AnnotationMain_perc ~ EndometriosisStatus)$p.value) %>%
-  ungroup()
-#epithelial, proliferative just not significant (p.value = 0.0898)
-
-###Significance testing prolif vs periov vs sec, CTL + Endo
-## Perform t-test Prolif vs Peri for each cell type
-df2_ttest_ProlifPeri <- df2 %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
-  group_by(AnnotationMain) %>%
-  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
-  ungroup()
-#none significant
-## Perform t-test Prolif vs Sec for each cell type
-df2_ttest_ProlifSec <- df2 %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
-  group_by(AnnotationMain) %>%
-  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
-  ungroup()
-#epi (0.00977) and str (0.0255) significant
-
-## Perform t-test periovulatory vs sec for each cell type
-df2_ttest_PeriSec <- df2 %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
-  group_by(AnnotationMain) %>%
-  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
-  ungroup()
-#epi (0.00794) and str (0.0137) significant
-```
-
-Supplementary Data Figure 4b, delete?
-```{r plotting of cell types per sample}
-##########################################################################################
-##Plot differences for DEG analysis samples
-#Extract sample names of samples used for DEG analysis
-#DEGsamples <- unique(read.csv("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/celltype_freq_sample.csv")$Var2)
-
-DEGsamples <- SupplementaryTable1 %>%
-  filter(DEG.analysis..Figure.3a.b. == TRUE) %>%
-  pull(Patient.sample.ID)
-
-#Make df 
-#Introduce 0 for the ones with 0 count!
-df3 <- meta%>%
-  group_by(sample, AnnotationRefined) %>%
-  summarise(AnnotationRefined_counts = n()) %>%
-  ungroup() %>%
-  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
-  filter(!is.na(AnnotationRefined)) %>%
-  group_by(sample) %>%
-  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
-  ungroup()  %>% #60 samples * 63 cell types = 3780 rows, correct
-  filter(sample %in% DEGsamples) #23 samples * 63 cell types = 1449 rows, correct
-df3 <- left_join(df3, dplyr::distinct(meta[,c("sample","EndometriosisStatus")]), by = "sample")
-
-# Perform t-test for each cell type
-df3_ttest <- df3 %>%
-  group_by(AnnotationRefined) %>%
-  summarise(p.value = t.test(AnnotationRefined_perc ~ EndometriosisStatus)$p.value) %>%
-  ungroup()
-#No significant difference between the EndometriosisStatus in any cell type
-
-###Plot
-#BoxPlots percentage, EndometriosisStatus, percentage
-ggplot(df3, aes(x = AnnotationRefined, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
-  geom_boxplot(outlier.shape = NA) +
-  geom_point(aes(colour = EndometriosisStatus),
-             colour="black",pch=21, 
-             position = position_jitterdodge(jitter.width = 0.2)) +
-  facet_wrap(~AnnotationRefined, scales = "free", ncol = 7)
-ggsave(paste0(dir_out, "Boxplot_DEGsamples_SymphonyRefinedFinal_EndometriosisStatus_perc.pdf"),
-  plot = last_plot(),
-  width = 17, height = 12)
 ```
@@
 Supplementary Data Figure 4b
+CODE OK
 ```{r plotting of cell types per sample}
@@
 #Introduce 0 for the ones with 0 count!
-df4 <- meta%>%
+SFig4b <- EndoAtlas_meta%>%
   filter(Minor.exclusion.criteria.logical == "FALSE") %>%
@@
   filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
-df4 <- left_join(df4, dplyr::distinct(meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main")]), by = "sample")
-write.csv(df4, paste0(dir_out,"df4_AnnotationRefined_frequency.csv"))
-
-# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
-df4_ttest_EndometriosisStatus <- df4 %>%
-  group_by(AnnotationRefined, MenstrualCyclePhase_main) %>%
-  summarise(p.value = t.test(AnnotationRefined_perc ~ EndometriosisStatus)$p.value) %>%
-  ungroup()
-#Prv_VSMC secretory, secretory is the only significant one (pval 0.009075467)!!
-df4_ttest_EndometriosisStatus$p.value_adjBH <- p.adjust(df4_ttest_EndometriosisStatus$p.value, method = "BH", n = length(df4_ttest_EndometriosisStatus$p.value))
-#After correction for multiple testing Prv_VSMC secretory is not significant anymore
-
-df4_ttest_EndometriosisStatus_secretory <- df4_ttest_EndometriosisStatus %>%
-                      filter(MenstrualCyclePhase_main == "secretory")
-df4_ttest_EndometriosisStatus_secretory$p.value_adjBH <- p.adjust(df4_ttest_EndometriosisStatus_secretory$p.value, method = "BH", n = length(df4_ttest_EndometriosisStatus_secretory$p.value))
-#Even when subsetting secretory phase, after correction for multiple testing Prv_VSMC secretory is not significant anymore
-
-## Perform t-test Prolif vs Peri for each cell type
-df4_ttest_ProlifPeri <- df4 %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
-  group_by(AnnotationRefined) %>%
-  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
-  ungroup()
-#Few significant
-## Perform t-test Prolif vs Sec for each cell type
-df4_ttest_ProlifSec <- df4 %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
-  group_by(AnnotationRefined) %>%
-  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
-  ungroup()
-write.csv(df4_ttest_ProlifSec, paste0(dir_out,"df4_ttest_ProlifSec.csv"))
-df4_ttest_ProlifSec <- read.csv(paste0(dir_out,"df4_ttest_ProlifSec.csv"))
-df4_ttest_ProlifSec$p.value_adjBH <- p.adjust(df4_ttest_ProlifSec$p.value, method = "BH", n = length(df4_ttest_ProlifSec$p.value))
-write.csv(df4_ttest_ProlifSec, paste0(dir_out,"df4_ttest_ProlifSec_adjpval.csv"))
-
-#About half are significant, show in paper supplement
-## Perform t-test periovulatory vs sec for each cell type
-df4_ttest_PeriSec <- df4 %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
-  group_by(AnnotationRefined) %>%
-  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
-  ungroup()
-#Less than half are significant, don't show in paper supplement
+SFig4b <- left_join(SFig4b, dplyr::distinct(EndoAtlas_meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main")]), by = "sample")
+write.csv(SFig4b, paste0(dir_out,"SFig4b_AnnotationRefined_frequency.csv"))
 
 #BoxPlots percentage, EndometriosisStatus, percentage ###for paper Extended Data Fig. S4
-ggplot(df4, aes(x = MenstrualCyclePhase_main, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
+ggplot(SFig4b, aes(x = MenstrualCyclePhase_main, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
   geom_boxplot(outlier.shape = NA) +
@@
 
-ggsave(paste0(dir_out, "Boxplot_DEGsamples_SymphonyRefinedFinal_EndometriosisStatus_perc.pdf"),
+ggsave(paste0(dir_out, "SFig4b_EndometriosisStatus_perc.pdf"),
   plot = last_plot(),
   width = 20, height = 12)
-```
-
-Figure 2f
-```{r plotting of cell types per sample}
-#Early and Mid Secretory Prv_VSMCs 
-df4 <- meta%>%
-  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
-  group_by(sample, AnnotationRefined) %>%
-  summarise(AnnotationRefined_counts = n()) %>%
-  ungroup() %>%
-  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
-  filter(!is.na(AnnotationRefined)) %>%
-  group_by(sample) %>%
-  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
-  ungroup() %>% #60 samples * 63 cell types = 3780 rows, correct
-  filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
-df4 <- left_join(df4, dplyr::distinct(meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main","MenstrualCyclePhase")]), by = "sample")
-
-secEarylMid_Prv_VSMC_MenstrualCyclePhase_main <- df4 %>%
-  filter(AnnotationRefined %in% c("Prv_VSMC secretory", "Prv_VSMC secretory_late")) %>% #same result with and without "Prv_VSMC secretory_late"
-  dplyr::filter(MenstrualCyclePhase %in% c("proliferative","periovulatory","secretory_mid","secretory_early")) %>%
-  group_by(sample, EndometriosisStatus, MenstrualCyclePhase_main) %>%
-  summarize(
-    total_counts_MenstrualCyclePhase_main = sum(AnnotationRefined_counts, na.rm = TRUE),
-    total_perc_MenstrualCyclePhase_main = sum(AnnotationRefined_perc, na.rm = TRUE)
-  ) %>%
+
+## Perform t-test Prolif vs Sec for each cell type
+SFig4b_ttest_ProlifSec <- SFig4b %>%
+  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
+  group_by(AnnotationRefined) %>%
+  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
   ungroup()
-
-ggplot(secEarylMid_Prv_VSMC_MenstrualCyclePhase_main, aes(x = MenstrualCyclePhase_main, y = total_perc_MenstrualCyclePhase_main, fill = EndometriosisStatus)) +
-  geom_boxplot(outlier.shape = NA) +
-  geom_point(aes(colour = EndometriosisStatus),
-             colour="black",pch=21, 
-             position = position_jitterdodge(jitter.width = 0.2))
-ggsave(paste0(dir_out, "Boxplot_sec_Prv_VSMC_NOsecretory_late_total_perc_MenstrualCyclePhase_main_EndometriosisStatus.pdf"),
-  plot = last_plot(),
-  width = 17, height = 12)
-
-secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("secretory")) %>% 
-  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
+SFig4b_ttest_ProlifSec$p.value_adjBH <- p.adjust(SFig4b_ttest_ProlifSec$p.value, method = "BH", n = length(SFig4b_ttest_ProlifSec$p.value))
+write.csv(SFig4b_ttest_ProlifSec, paste0(dir_out,"SFig4b_ttest_ProlifSec_adjpval.csv"))
+
+# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
+SFig4b_ttest_EndometriosisStatus <- SFig4b %>%
+  group_by(AnnotationRefined, MenstrualCyclePhase_main) %>%
+  summarise(p.value = t.test(AnnotationRefined_perc ~ EndometriosisStatus)$p.value) %>%
   ungroup()
-#significant, p.value 0.0261 -> paper
-
-secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval_prolif <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative")) %>% 
-  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
-  ungroup()
-#significant, p.value NA
-
-secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval_peri <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
-  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory")) %>% 
-  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
-  ungroup()
-#significant, p.value 0.7639853
-
-p.adjust(c(0.0261), method = "BH", n = length(3)) #Value does not change
-p.adjust(c(0.7639853), method = "BH", n = length(3)) #Value does not change
-
-#Calculate fold change ENDO vs nonENDO secretory
-# Assuming secEarylMid_Prv_VSMC_MenstrualCyclePhase_main is your dataframe
-
-library(dplyr)
-
-# Filter the dataframe for 'secretory' in MenstrualCyclePhase_main and then group by EndometriosisStatus
-mean_total_perc <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
-  filter(MenstrualCyclePhase_main == "secretory") %>%
-  group_by(EndometriosisStatus) %>%
-  summarize(mean_total_perc = mean(total_perc_MenstrualCyclePhase_main))
-```
-
-Figure 1d
-```{r plotting of cell types per sample}
-p5 <- ggplot(df5, aes(x = AnnotationMain, y = row_count)) +
-  geom_col() +
-  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
-  coord_flip()
-p5
-ggsave(paste0(dir_out, "Barplot_EndometriosisGrade_main5.pdf"),
-  plot = last_plot(),
-  width = 15, height = 6)
-
-#main5 boxplot 
-df6 <- meta%>%
-  group_by(sample, AnnotationMain, EndometriosisGrade) %>%
-  summarise(row_count = n()) %>%
-  ungroup()
-p6 <- ggplot(df6, aes(x = AnnotationMain, y = row_count)) +
-  geom_boxplot() +
-  #geom_jitter(position = position_jitter(width = 0.2, height = 0), size = 0.5) +
-  scale_y_log10() +
-  coord_flip() +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1))
-p6
-ggsave(paste0(dir_out, "Boxplot_EndometriosisGrade_main5_colEndometriosisGrade.pdf"),
-         plot = last_plot(),
-         width = 15, height = 6)
-
-# bar plot of the number of AnnotationRefined cells in each sample
-df7 <- meta%>%
-  filter(!is.na(AnnotationMain)) %>%
-  group_by(AnnotationMain, EndometriosisGrade) %>%
-  summarise(row_count = n()) %>%
-  ungroup() 
-ggplot(df7, aes(x = AnnotationMain, y = row_count, fill = EndometriosisGrade)) +
-  geom_bar(stat = 'identity', position = 'fill') +
-  scale_y_continuous(expand = c(0, 0)) +
-  coord_flip()
-ggsave(paste0(dir_out, "Barplot_AnnotationMain_EndometriosisGrade.pdf"),
-  plot = last_plot(),
-  width = 6, height = 4)
-
-```
-
-UMAPs
-
-epithelial UMAP
-2b,2c,3b,3e,
-
-atlas UMAP
-S3f, S7a
-
-Figure 1c
-```{r}
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
-#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
-A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
-table(A003A033$AnnotationRefined)
-data <- AddMetaData(data,A003A033)
-E031A004meta <- read.csv("/home/common/data/output/projects/ENDO/E044/A033/E031A004_proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated_meta.csv")
-E031A004meta$cell_id <- paste(E031A004meta$sample, E031A004meta$cell.names, sep = "_")
-rownames(E031A004meta) <- E031A004meta$cell_id 
-data <- AddMetaData(data, E031A004meta[, "Annotation", drop = FALSE])
-
-order <- A003A033 %>%
-  select(AnnotationMain, AnnotationRefined) %>%
-  filter(!is.na(AnnotationMain)) %>%
-  distinct() %>%
-  arrange(AnnotationRefined) %>%
-  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
-  arrange(AnnotationMain)
-data$AnnotationRefined <- factor(x = data$AnnotationRefined, levels = order$AnnotationRefined)
-p24 <- DimPlot(data, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
-p24
-
-```
-
-Figure 2b
-```{r}
-#Figure 2b entire atlas
-Fig2b1 <- DimPlot(data, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
-Fig2b1
-ggsave(filename = paste0(dir_out, 'Fig2b1.pdf'),
-       plot = Fig2b1,
-       width =20, height=9)
-
-#Figure 2b epithelial cells
-Fig2b2 <- DimPlot(data, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
-Fig2b2
-ggsave(filename = paste0(dir_out, 'Fig2b2.pdf'),
-       plot = Fig2b2,
-       width =20, height=9)
-
-```
-
-Supplementary Data Figure 1c
-```{r}
-SFig1c <- FeaturePlot(data, features = "CycleDay", raster = FALSE)
-SFig1c
-ggsave(filename = paste0(dir_out, 'SFig1c.pdf'),
-       plot = SFig1c,
-       width =20, height=9)
-```
-
-Supplementary Data Figure 1d
-```{r}
-SFig1d <- FeaturePlot(data, features = "ProgesteroneSerum", raster = FALSE)
-SFig1d
-ggsave(filename = paste0(dir_out, 'SFig1d.pdf'),
-       plot = SFig1d,
-       width =20, height=9)
-
-```
-
-Supplementary Data Figure 3e
-/home/common/data/output/projects/ENDO/E044/A066/phase_sc_Fig3_Wang_epi_pseudot/evaluation/ #S3c/d
-```{r plott umaps of epithelial cells}
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A062/E044A050_seed621_filteredWO_ciliatedMesothelialMUC5B.rds")
-A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
-table(A003A033$AnnotationRefined)
-data <- AddMetaData(data,A003A033)
-data <- AddMetaData(data, meta_ptime)
-data$epithelial.pseudotime <- as.integer(data$epithelial.pseudotime)
-data$Fig3_Wang_epi_pseudot[is.na(data$Fig3_Wang_epi_pseudot)] <- 1 #checked samples
-
-#General Plotting
-p1 <- FeaturePlot(data, features = "ProgesteroneSerum", split.by = "EndometriosisStatus")
-p2 <- DimPlot(data, group.by = "EndometriosisStatus", split.by = "CyclePhase")
-p3 <- DimPlot(data, split.by = "Clusters", group.by = "EndometriosisStatus")
-#plot_grid(p1,p2,p3, ncol=1)
-ggsave(filename = paste0(dir_out, 'EpiUMAP_general.pdf'),
-       plot = plot_grid(p1,p2,p3, ncol=1),
-       width =8 ,height=10)
-
-#Plotting of mens cycle info
-p1 <- DimPlot(data, group.by = "Fig3_Wang_epi_pseudot")
-p2 <- DimPlot(data, group.by = "MenstrualCyclePhase")
-
-#FeaturePlot(seurat_obj, features = "trajectory_day567_epithelial.pseudotime")
-#plot_grid(p1,p2,p3,p4,p5,p6, ncol=3)
-
-ggsave(filename = paste0(dir_out, 'EpiUMAP_PhasePrediction_w11.pdf'),
-       plot = plot_grid(p1,p2, ncol=2),
-       width =11 ,height=5)
-
-```
-
-```{r plotting umaps of entire dataset}
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
-#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
-A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
-table(A003A033$AnnotationRefined)
-data <- AddMetaData(data,A003A033)
-E031A004meta <- read.csv("/home/common/data/output/projects/ENDO/E044/A033/E031A004_proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated_meta.csv")
-E031A004meta$cell_id <- paste(E031A004meta$sample, E031A004meta$cell.names, sep = "_")
-rownames(E031A004meta) <- E031A004meta$cell_id 
-data <- AddMetaData(data, E031A004meta[, "Annotation", drop = FALSE])
-
-#Plot clusters, batch,sample EndometriosisStatus, cycle phase
-p1 <- DimPlot(data, reduction = "umap", group.by = "EndometriosisStatus", raster = FALSE)
-p2 <- DimPlot(data, reduction = "umap", group.by = "batch", raster = FALSE)
-p3 <- DimPlot(data, reduction = "umap", group.by = "sample", raster = FALSE) +
-  theme(legend.text = element_text(size=4))+
-  theme(legend.key.size = unit(0.2, 'cm'))+  guides(color = guide_legend(ncol=1) )
-p4 <- DimPlot(data, reduction = "umap", group.by = "celline", raster = FALSE)
-p5 <- DimPlot(data, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
-p6 <- DimPlot(data, reduction = "umap", group.by = "AnnotationMain", raster = FALSE)
-p7 <- DimPlot(data, reduction = "umap", group.by = "Symphony_Refined_A029", raster = FALSE)
-p8 <- DimPlot(data, reduction = "umap", group.by = "CyclePhase", raster = FALSE)
-p9 <- FeaturePlot(data, features = "ProgesteroneSerum", raster = FALSE)
-ggsave(filename = paste0(dir_out, 'UMAPs_p1p9.pdf'),
-       plot = plot_grid(p1,p2,p3,p4,p6,p7,p5,p8,p9, ncol=3),
-       width =28 ,height=24)
-ggsave(filename = paste0(dir_out, 'UMAPs_p1p9.png'),
-       plot = plot_grid(p1,p2,p3,p4,p6,p7,p5,p8,p9, ncol=3),
-       width =28 ,height=24)
-
-p10 <- DimPlot(data, reduction = "umap", group.by = "AnnotationMain", raster = FALSE)
-p11 <- FeaturePlot(data, reduction = "umap", features = "Symphony_Global_prob", raster = FALSE)
-p12 <- DimPlot(data, reduction = "umap", group.by = "Symphony_Refined", raster = FALSE)
-p13 <- FeaturePlot(data, reduction = "umap", features = "Symphony_Refined_prob", raster = FALSE)
-p14 <- DimPlot(data, reduction = "umap", group.by = "AnnotationRefined", raster = FALSE)
-ggsave(filename = paste0(dir_out, 'UMAPs_p10p14.pdf'),
-       plot = plot_grid(p10,p11,p12,p13,p14, ncol=3),
-       width =28 ,height=16)
-
-p20 <- DimPlot(data, reduction = "umap", group.by = "batch", raster = FALSE)
-p21 <- FeaturePlot(data, reduction = "umap", features = "ProgesteroneSerum", raster = FALSE)
-p22 <- FeaturePlot(data, reduction = "umap", features = "CycleDay", raster = FALSE)
-p23 <- DimPlot(data, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
-
-#Order data by main5
-###In progress
-order <- A003A033 %>%
-  select(AnnotationMain, AnnotationRefined) %>%
-  filter(!is.na(AnnotationMain)) %>%
-  distinct() %>%
-  arrange(AnnotationRefined) %>%
-  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
-  arrange(AnnotationMain)
-data$AnnotationRefined <- factor(x = data$AnnotationRefined, levels = order$AnnotationRefined)
-p24 <- DimPlot(data, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
-p24
-data$Fig3_Wang_epi_pseudot[is.na(data$Fig3_Wang_epi_pseudot)] <- 1 #checked samples
-p25 <- DimPlot(data, reduction = "umap", group.by = "Fig3_Wang_epi_pseudot",na.value = "",  raster = FALSE)
-p25
-#Highlight cycling fibroblasts
-Idents(data) <- data$Annotation
-table(data$Annotation)
-cyclingfibs <- WhichCells(data, idents = c("Cycling Fibroblasts"))
-p26 <- DimPlot(data, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
-p26
-
-plot_grid(p20,p21,p22,p23, ncol=2)
-ggsave(filename = paste0(dir_out, 'UMAPs_p20p23.pdf'),
-       plot = plot_grid(p20,p21,p22,p23, ncol=2),
-       width =24, height=18)
-ggsave(filename = paste0(dir_out, 'UMAPs_p20p23.png'),
-       plot = plot_grid(p20,p21,p22,p23, ncol=2),
-       width =24, height=18)
-ggsave(filename = paste0(dir_out, 'UMAPs_p24.pdf'),
-       plot = p24,
-       width =20, height=9)
-ggsave(filename = paste0(dir_out, 'UMAPs_p24.png'),
-       plot = p24,
-       width =20, height=9)
-ggsave(filename = paste0(dir_out, 'UMAPs_p25.pdf'),
-       plot = p25,
-       width =13, height=9)
-ggsave(filename = paste0(dir_out, 'UMAPs_p25.png'),
-       plot = p25,
-       width =13, height=9)
-ggsave(filename = paste0(dir_out, 'UMAPs_p26.pdf'),
-       plot = p26,
-       width =13, height=9)
-ggsave(filename = paste0(dir_out, 'UMAPs_p26.png'),
-       plot = p26,
-       width =13, height=9)
-
-#Plot objects with split.by sample
-p15 <- DimPlot(data, split.by = "sample", raster = FALSE)
-ggsave(plot = p15, filename = paste0(dir_out,'A003_samples.pdf'),width =45 ,height=5)
-
-#Plot objects with split.by EndometriosisStatus
-p16 <- DimPlot(data, split.by = "EndometriosisStatus", raster = FALSE)
-ggsave(plot = p16, filename = paste0(dir_out,'A003_EndometriosisStatus.pdf'),width =15 ,height=5)
-
-rm(data)
-gc()
-```
-
-Supplementary Data Figure 3 g
-```{r}
-####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
-#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad
-
-####### Convert .h5ad files to .h5seurat files
-library(SeuratDisk)
-Convert("endo-2022_epithelial.h5ad", dest = "h5seurat", assay = "RNA", overwrite = TRUE)
-
-Tan_epithelial <- LoadH5Seurat(paste0("endo-2022_epithelial.h5seurat"), assays = "RNA")
-
-Tan_epithelial <- ScaleData(object = Tan_epithelial, features = data@assays[["RNA"]]@data@Dimnames[[1]])
-
-#Subset cells from the endometrium (exclude cells from endometriosis lesions)
-Idents(Tan_epithelial) <- Tan_epithelial$sample_type_rename
-Tan_epithelial <- subset(Tan_epithelial, idents = c("Ctrl", "EuE"))
-
-# factor levels in better order (E10 and E11 don't have endometrium cells)
-Tan_epithelial$sample <- factor(Tan_epithelial$sample, levels = c("C01", "C02", "C03", "E01", "E02", "E03", "E04", "E05", "E06", "E07", "E08", "E09"))
-
-# Generate heatmap of the endometrium cells from Tan et al. 2022 with the same epithelial menstrual marker genes as in our manuscript Figure 2d, for comparison
-feature_list_figure2 <- c("SFRP4", "LAMC2", "MMP11", "MMP7", "SPRY1", "TFPI2","STEAP4","ADAMTS8","CAPN6","SLC37A2","SCGB1D2", "MT1F", "MT1X","RIMKLB", "S100P", "DEFB1","G0S2","NNMT","GPX3", "PAEP")
-
-heatmap <- DoHeatmap(
-  object = Tan_epithelial,
-  features = intersect(feature_list_figure2,names),
-  cells = NULL,
-  group.by = "sample",
-  assay = "RNA",
-  group.bar = TRUE) + 
-  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
-heatmap
-ggsave(
-    filename = paste0(dir_out, "Tan_epithelial_Fig2EpiMarkers.pdf"),
-    plot = last_plot(),
-    width = 12.3,
-    height = 5)
-
-#As control also plot menstrual marker genes from Wang et al. 2020 (DOI: 10.1038/s41591-020-1040-z)
-feature_list_Wang <- c("PLAU", "MMP7", "THBS1", "CADM1", "NPAS3", "ATP1A1", "ANK3", "ALPL", "TRAK1", "SCGB1D2", "MT1F", "MT1X", "MT1E", "MT1G", "CXCL14", "MAOA", "DPP4", "NUPR1", "GPX3", "PAEP")
-
-heatmap <- DoHeatmap(
-  object = Tan_epithelial,
-  features = intersect(feature_list_Wang, names),
-  cells = NULL,
-  group.by = "sample",
-  assay = "RNA",
-  group.bar = TRUE) + 
-  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
-heatmap
-ggsave(
-    filename = paste0(dir_out, "Tan_epithelial_WangEpiMarkers.pdf"),
-    plot = last_plot(),
-    width = 15,
-    height = 15)
-
-#The menstrual marker gene expression distribution of each sample looks comparable to our manuscript figure 2d, confirming that the different samples in Tan et al.
-2022 express menstrual marker genes from different menstrual cycle phases
-
-```
-
-Supplementary Data Figure 3 h
-```{r}
-####### Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) through the command line
-#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_stromal.h5ad
-
-####### Convert .h5ad files to .h5seurat files
-library(SeuratDisk)
-Convert("endo-2022_stromal.h5ad", dest = "h5seurat", assay = "RNA", overwrite = TRUE)
-Tan_stromal <- LoadH5Seurat(paste0("endo-2022_stromal.h5seurat"), assays = "RNA")
-
-# 
-FEATURES <- data@assays[["RNA"]]@data@Dimnames[[1]]
-Tan_stromal <- ScaleData(object = Tan_stromal, features = FEATURES)
-
-#Subset cells from the endometrium (exclude cells from endometriosis lesions)
-Idents(Tan_stromal) <- Tan_stromal$sample_type_rename
-Tan_stromal <- subset(Tan_stromal, idents = c("Ctrl", "EuE"))
-
-# factor levels in better order (E10 and E11 don't have endometrium cells)
-Tan_stromal$sample <- factor(Tan_stromal$sample, levels = c("C01", "C02", "C03", "E01", "E02", "E03", "E04", "E05", "E06", "E07", "E08", "E09"))
-
-# Generate heatmap of the endometrium cells from Tan et al. 2022 with the same stromal menstrual marker genes as in our manuscript Figure 2d, for comparison
-feature_list_figure2_stromal <- c("SFRP1", "H1-1", "H3C2", "WNT5A","CILP","BRINP1", "CFD","PTGDS","SLIT3","RPRM", "TIMP3", "IGFBP1","LEFTY2","LRRC15","LUM")
-
-#Generate heatmap
-
-heatmap <- DoHeatmap(
-  object = Tan_stromal,
-  features = intersect(feature_list_figure2_stromal,FEATURES),
-  cells = NULL,
-  group.by = "sample",
-  assay = "RNA",
-  group.bar = TRUE) + 
-  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
-heatmap
-ggsave(
-    filename = paste0(dir_out, "Tan_stromal_Fig2stromalMarkers.pdf"),
-    plot = last_plot(),
-    width = 12.3,
-    height = 5)
-
-#Wang genes
-feature_list_Wang_stromal <- c('STC1', 'NFATC2', 'BMP2' ,'PMAIP1' ,'MMP11' ,'SFRP1', 'WNT5A' ,'ZFYVE21' ,'CILP' ,'SLF2' ,'MATN2', 'S100A4' ,'DKK1' ,'CRYAB', 'FOXO1', 'IL15', 'FGF7', 'LMCD1')
-
-heatmap <- DoHeatmap(
-  object = Tan_stromal,
-  features = intersect(feature_list_Wang_stromal, FEATURES),
-  cells = NULL,
-  group.by = "sample",
-  assay = "RNA",
-  group.bar = TRUE) + 
-  scale_fill_gradientn(limits = c(-2, 2), colors = c("#FF00FF", "#000000", "#FFFF00"))
-heatmap
-ggsave(
-    filename = paste0(dir_out, "Tan_stromal_WangstromalMarkers.pdf"),
-    plot = last_plot(),
-    width = 15,
-    height = 15)
-
-#The menstrual marker gene expression distribution of each sample looks comparable to our manuscript figure 2d, confirming that the different samples in Tan et al.
-2022 express menstrual marker genes from different menstrual cycle phases
-
-```
-
-Supplementary Data Figure 1a and 1b
-```{r VlnPlots of nCounts and nFeatures}
-##VlnPlots for "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT","percent.mt"
-VlnPlot(data,
-  features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT","percent.mt"),
-  group.by='EndometriosisStatus', ncol = 1, pt.size=0)
-ggsave(paste0(dir_out, "VlnPlot_GroupByEndometriosisStatus_nFeature_nCounts_RNA_SCT_percMT.pdf"),
-  plot = last_plot(),
-  width = 4, height = 13)
-
-##VlnPlots grouped by sample of "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT","percent.mt" 
-# default level sort order is alphabetical, therefore refactor before plotting
-data[["EndometriosisStatus"]] <- factor(unlist(data[["EndometriosisStatus"]])) 
-#VlnPlot
-VlnPlot(data,
-  features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT","percent.mt"),
-  group.by='sample', ncol = 1, pt.size=0)
-ggsave(paste0(dir_out, "VlnPlot_GroupBySample_nFeature_nCounts_RNA_SCT_percMT.pdf"),
-  plot = last_plot(),
-  width = 25, height = 15)
-#Plot VlnPlots with boxplots and save them
-VlnPlot(data, features = "nCount_RNA", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
-VlnPlot(data, features = "nFeature_RNA", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
-VlnPlot(data, features = "nCount_SCT", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
-VlnPlot(data, features = "nFeature_SCT", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
-VlnPlot(data, features = "percent.mt", pt.size = 0) + geom_boxplot(width=0.1,fill="white",outlier.size=0)
-
-```
-
-PCA 
-A072 1e, 2a, S1e, S3a
-A082 S1f
-
-```{r different counts and means/medians}
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_metadata.rds")
-
-#Percentage of main 5
-table <- as.data.frame(table(data$AnnotationMain))
-table$Percentage <- round((table$Freq / sum(table$Freq)) * 100, 1)
-
-#Number of cells for CTL and Endo
-table(data$EndometriosisStatus)
-
-#Number of patients for CTL and Endo
-data %>% select(sample,EndometriosisStatus) %>%
-  distinct() %>%
-  group_by(EndometriosisStatus) %>%
-  summarize(number_of_rows = n())
-
-#mean number of cells per sample
-mean(table(data$sample))
-
-#median count/feature
-median_summary <- data %>%
-  select(nCount_RNA, nCount_SCT, nFeature_RNA, nFeature_SCT) %>%
-  summarize(
-    median_nCount_RNA = median(nCount_RNA),
-    median_nCount_SCT = median(nCount_SCT),
-    median_nFeature_RNA = median(nFeature_RNA),
-    median_nFeature_SCT = median(nFeature_SCT)
-  )
-
-```
-
-Supplementary Data Figure 2
-```{refined cluster markers}
-#Load data
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
-#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
-meta <- data@meta.data
-
-#Set SCT as default and remove RNA to save space
-DefaultAssay(data) <- "SCT"
-data@assays$RNA <- NULL
-gc()
-
-#Remove data without annotation (no NAs!)
-Idents(data) <- data$Symphony_Refined
-data <- subset(data, subset = Symphony_Refined %in% levels(as.factor(data$Symphony_Refined)))
-gc()
-
-#setdiff(pops_to_plot,unique(data$Symphony_Refined))
-######################################################
-#Tan2022 Sup Fig. 7c, epithelial
-genes_to_plot <- c("TPPP3", "PIFO", "FOXJ1", "TP63", "KRT5", "SOX9", "LGR5", "MUC5B", "TFF3", "RUNX3", "SAA1", "SIX1", "PROM1", "PROM2", "ANPEP", "KIAA1324", "SPDEF", "SMAD9", "CD36", "HSD17B2", "SCGB2A2", "GDA", "FGFR2", "WNT7A", "FGF9", "PTGS1", "MSLN", "UPK3B", "CALB2", "PGR", "ESR1", "MMP7", "TIMP1", "IDO1", "PSAT1", "ENPP3", "GNG11", "CREB3L1", "IHH", "TOP2A", "MT1F", "MT1E", "MT1X", "SERPINA5", "SPP1", "PAEP", "CXCL14", "DPP4")
-pops_to_plot <- c("mesothelial", "ciliated", "mid-secretory", "lumenal 2", "lumenal 1", "lumenal", "glandular early-secretory", "glandular", "TP63+/KRT5+", "MUC5B+")
-
-# reorder clusters
-pops_seurat <- subset(data, idents = pops_to_plot)
-pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
-                            levels=pops_to_plot)
-p <- Seurat::DotPlot(
-  object = pops_seurat,
-  scale = T,
-  features = genes_to_plot,
-  idents = pops_to_plot,
-) + theme(axis.text.x = element_text(angle = 90))
-ggsave(plot = p, "./dotplots/Tan2022Fig3b_epithelial_SCT.pdf", width = 17, height = 7)
-######################################################
-#DotPlot of features from Tan2022 Fig. 4b, Lymphocytes
-genes_to_plot <- c("FCGR3A","KLRD1","ITGA1","LILRB1","ITGAX","KLRC1","KLRC2","EPAS1","KIR2DL4","IGFBP2","CD160","GZMH","KLRF1","FGFBP2","TBX21","CD8A","CD8B","CRTAM","CCR5","NCR3","RORC","IL7R","TCF7","CD4","CCR7","SELL","LEF1","FOXP3","CTLA4","IL2RA","CXCR6","ZNF683","ITGAE","XCL1","SLC4A10","DPP4","GZMK","IGKC", "MZB1", "DERL3", "CD79A", "CD79B", "MS4A1", "BANK1")
-pops_to_plot <- c(meta %>% filter(AnnotationMain == "lymphocyte") %>% pull(Symphony_Refined) %>% unique())
-
-# reorder clusters
-pops_seurat <- subset(data, idents = pops_to_plot)
-pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
-                           levels=c("B cell", "plasma", "CD8 T$_{RM}$","CD4 T$_{RM}$","T$_{Reg}$", "T$_N$/T$_{CM}$" , "ILC" , "T$_{EM}$","CTL", "pNK", "NK3","NK2","NK1"))
-
-p <- Seurat::DotPlot(
-  object = pops_seurat,
-  scale = T,
-  features = genes_to_plot,
-  idents = pops_to_plot,
-) + theme(axis.text.x = element_text(angle = 90))
-ggsave(plot = p, "./dotplots/Tan2022Fig3b_Lymphocytes_SCT.pdf", width = 17, height = 7)
-######################################################
-#Tan2022 Fig. 3b, stromal
-genes_to_plot <- c("RERGL", "MCAM", "MYH11", "STEAP4","GGT5", "RGS5")
-pops_to_plot <- c("Prv-MYH11", "Prv-CCL19", "Prv-STEAP4", "VSMC")
-# reorder clusters
-pops_seurat <- subset(data, idents = pops_to_plot)
-pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
-                            levels=c("VSMC", "Prv-STEAP4", "Prv-CCL19","Prv-MYH11"))
-
-p <- Seurat::DotPlot(
-  object = pops_seurat,
-  scale = T,
-  features = genes_to_plot,
-  idents = pops_to_plot,
-) + theme(axis.text.x = element_text(angle = 90))
-ggsave(plot = p, "./dotplots/Tan2022Fig3b_stromal_SCT.pdf", width = 7, height = 7)
-######################################################
-#Tan2022 Fig. 4b, macrophages
-genes_to_plot <- c("FLOR2", "LYVE1", "MRC1", "CX3CR1", "ICAM2", "CLEC5A", "CCR2", "VEGFA", "FN1", "SPARC", "APOE", "SPP1")
-pops_to_plot <- c(meta %>% filter(AnnotationMain == "myeloid") %>% pull(Symphony_Refined) %>% unique())
-pops_to_plot <- c("M$\\Phi$5-activated","M$\\Phi$4-infiltrated","M$\\Phi$3-APOE","M$\\Phi$2-peritoneal","M$\\Phi$1-LYVE1")
-
-# reorder clusters
-pops_seurat <- subset(data, idents = pops_to_plot)
-pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
-                            levels=pops_to_plot)
-p <- Seurat::DotPlot(
-  object = pops_seurat,
-  scale = T,
-  features = genes_to_plot,
-  idents = pops_to_plot,
-) + theme(axis.text.x = element_text(angle = 90))
-ggsave(plot = p, "./dotplots/Tan2022Fig3b_macrophages_SCT.pdf", width = 10, height = 7)
-######################################################
-#Tan2022 Fig. 5a, DCs
-genes_to_plot <- c("CD1C", "CLEC9A", "LAMP3")
-pops_to_plot <- c("mDC", "cDC1", "DC3", "cDC2", "pre-cDC2")
-# reorder clusters
-pops_seurat <- subset(data, idents = pops_to_plot)
-pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
-                            levels=pops_to_plot)
-p <- Seurat::DotPlot(
-  object = pops_seurat,
-  scale = T,
-  features = genes_to_plot,
-  idents = pops_to_plot,
-) + theme(axis.text.x = element_text(angle = 90))
-ggsave(plot = p, "./dotplots/Tan2022Fig3b_DCs_SCT.pdf", width = 5, height = 7)
-
-rm(data)
-gc()
-
-```
-
-```{major cluster markers}
-#Load data
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
-#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
-
-#Set SCT as default and remove RNA to save space
-DefaultAssay(data) <- "SCT"
-data@assays$RNA <- NULL
-gc()
-
-#Remove data without annotation
-Idents(data) <- data$AnnotationMain
-data <- subset(data, subset = AnnotationMain %in% levels(as.factor(data$AnnotationMain)))
-
-GENES <- c("EPCAM","CDH1","PDGFRA","VWF","PECAM1","PTPRC","CD68","CD14","CD2")
-p <- VlnPlot(data, features = GENES, group.by = "AnnotationMain",raster=FALSE, pt.size = 0, assay = "SCT", stack=T, flip=T)
-ggsave(plot = p, "./VlnPlotMain5_SCT_notds.pdf", width = 6, height = 4.5)
-
-GENES <- c("PDGFRA","DCN","COL1A1","COL6A1","CRISPLD2","COL6A3","LUM","COL5A1") #stomal genes, only "PDGFRA" and "CRISPLD2" are cleanly expressed in stromal only, the others are a bit mixed
-p <- VlnPlot(data, features = GENES, group.by = "AnnotationMain",raster=FALSE, pt.size = 0, assay = "RNA", stack=T, flip=T)
-ggsave(plot = p, "./VlnPlotMain5_stromalMG_RNA_notds.pdf", width = 6, height = 4.5)
-
-p <- VlnPlot(data, features = GENES, group.by = "AnnotationMain",raster=FALSE, pt.size = 0, assay = "SCT", stack=T, flip=T)
-ggsave(plot = p, "./VlnPlotMain5_stromalMG_SCT_notds.pdf", width = 6, height = 4.5)
-
-GENES <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2")
-p <- VlnPlot(data, features = GENES, group.by = "AnnotationMain",raster=FALSE, pt.size = 0, assay = "RNA", stack=T, flip=T)
-ggsave(plot = p, "./VlnPlotMain5_RNA_notds.pdf", width = 6, height = 4.5)
-
-table(data$AnnotationMain)
-
-#Make DotPlots
-DefaultAssay(data) <- "SCT"
-
-GENES
-
-######################################################
-#Tan2022 main markers
-genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2")
-pops_to_plot <- c("stromal","epithelial","myeloid","lymphocyte","endothelial")
-
-Seurat::DotPlot(
-  object = data,
-  scale = T,
-  features = genes_to_plot,
-  idents = pops_to_plot,
-) + theme(axis.text.x = element_text(angle = 90))
-ggsave(plot = last_plot(), "./dotplots/main5_main5markers_SCT.pdf", width = 10, height = 5)
-
-#stomal markers
-genes_to_plot <- c("PDGFRA","DCN","COL1A1","COL6A1","CRISPLD2","COL6A3","LUM","COL5A1") #stomal genes, only "PDGFRA" and "CRISPLD2" are cleanly expressed in stromal only, the others are a bit mixed
-Seurat::DotPlot(
-  object = data,
-  scale = T,
-  features = genes_to_plot,
-  idents = pops_to_plot,
-) + theme(axis.text.x = element_text(angle = 90))
-ggsave(plot = last_plot(), "./dotplots/main5_stromalmarkers_SCT.pdf", width = 10, height = 5)
-
-#all markers + stomal
-genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL5A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2","DCN","COL6A1","CRISPLD2","COL6A3","LUM","COL1A1") #stomal genes, only "PDGFRA" and "CRISPLD2" are cleanly expressed in stromal only, the others are a bit mixed
-Seurat::DotPlot(
-  object = data,
-  scale = T,
-  features = genes_to_plot,
-  idents = pops_to_plot,
-) + theme(axis.text.x = element_text(angle = 90))
-ggsave(plot = last_plot(), "./dotplots/main5_allmarkers_SCT.pdf", width = 17, height = 5)
-
-```
-
+#Prv_VSMC secretory pval 0.009075467
+SFig4b_ttest_EndometriosisStatus$p.value_adjBH <- p.adjust(SFig4b_ttest_EndometriosisStatus$p.value, method = "BH", n = length(SFig4b_ttest_EndometriosisStatus$p.value))
+write.csv(SFig4b_ttest_EndometriosisStatus, paste0(dir_out,"SFig4b_ttest_EndometriosisStatus_adjpval.csv"))
+#After correction for multiple testing Prv_VSMC secretory is not significant anymore
+
+```
+
+Supplementary Data Figure 5a
+```{r}
+DEGs
+
+```
+
+Supplementary Data Figure 5b-d
+```{r}
+
+#Figure 3c and Supplementary Figure 5b
+gk1 <- rankNet(cellchat, mode = "comparison", stacked = T, do.stat = TRUE)
+gk1 
+ggsave(plot = gk1,
+       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12)
+
+#Load CellChat data
+cellchat <- readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
+
+#Supplementary Data Figure 5d TNF
+netVisual_bubble(cellchat, comparison = c(1, 2),signaling = c("TNF"), remove.isolate = FALSE) 
+ggsave(plot = last_plot(),
+       filename = paste0(dir_out, 'output_comparison_Refined/netVisual_bubble_TNF.pdf'), width =25,height=5)
+
+```
+
+##################################################################################################################
+Scailyte part
+#Supplementary Data Figure 6a, b
 ```{r barplot distribution of cv splits, all cells, prolif}
@@
 
+#Supplementary Data Figure 6a
 # reshape df
-melted_meta <- melt(meta, id.vars = c("lane", "endo_EndometriosisGrade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
+melted_meta <- melt(meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
 ## create the bar plot for endo_EndometriosisGrade
-ggplot(melted_meta, aes(x = variable, fill = endo_EndometriosisGrade)) +
+ggplot(melted_meta, aes(x = variable, fill = endo_grade)) +
   geom_bar(position = "stack", width = 0.7) +
@@
 
-## create the bar plot for lane
+#Supplementary Data Figure 6b
 ggplot(melted_meta, aes(x = variable, fill = lane)) +
@@
 
+#Supplementary Data Figure 6e
 ##Upsetplot of eval samples
@@
 
-################
+################?/???????
 #heatmap, not working yet
@@
 draw(heatmap)
-dev.off()
-```
-
-```{r barplot distribution of cv splits, cf, prolif}
-meta <- read.csv("/home/common/data/output/projects/ENDO/E041/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")
-
-library(reshape2)
-library(ggplot2)
-library(forcats)
-
-# reshape df
-melted_meta <- melt(meta, id.vars = c("lane", "endo_EndometriosisGrade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
-# create the bar plot for endo_EndometriosisGrade
-ggplot(melted_meta, aes(x = variable, fill = endo_EndometriosisGrade)) +
-  geom_bar(position = "stack", width = 0.7) +
-  facet_wrap(~ fct_rev(value), ncol = 2) +
-  labs(title = "Bar Plot of Samples by Endo EndometriosisGrade",
-       x = "Fold",
-       y = "Number of Samples") +
-  scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
-  theme_minimal() +
-  coord_flip()
-ggsave(paste0(dir_out,"Bar Plot of Samples by Endo EndometriosisGrade_cf.pdf"), plot = last_plot(), height = 6.3, width = 7.2)
-
-# create the bar plot for lane
-ggplot(melted_meta, aes(x = variable, fill = lane)) +
-  geom_bar(position = "stack", width = 0.7) +
-  facet_wrap(~ fct_rev(value), ncol = 2) +
-  labs(title = "Bar Plot of Samples by lane",
-       x = "Fold",
-       y = "Number of Samples") +
-  #scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
-  theme_minimal() +
-  coord_flip()
-ggsave(paste0(dir_out,"Bar Plot of Samples by lane_cf.pdf"), plot = last_plot(), height = 6.3, width = 7.2)
-
-##Upsetplot of eval samples
-df <- meta %>%
-  rownames_to_column(var = "name") %>%
-  select(-name) %>% 
-  select(fold1, fold2, fold3, fold4, fold5) %>%
-  mutate_all(~ ifelse(. == 'eval', TRUE, FALSE)) %>%
-  filter(rowSums(. == TRUE) > 0)
-pdf(file.path(dir_out, "upset_CellCnnsig_cf_evalsamples.pdf"))
-upset(df,
-  colnames(df)) +
-  ggtitle("all cells")
-dev.off()
-
-################
-#heatmap, not working yet
-
-# Extracting the necessary columns from the 'meta' dataframe
-data_matrix <- meta[, c("fold1", "fold2", "fold3", "fold4", "fold5")]
-data_matrix_numeric <- apply(data_matrix, 2, function(x) as.numeric(x == "eval"))
-
-# Adding annotation columns
-annotation_data <- meta[, c("EndometriosisStatus", "lane", "batch", "endo_EndometriosisGrade")]
-
-library(ComplexHeatmap)
-library(circlize)
-
-# Define a custom color palette
-my_color_palette <- colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
-png(file=paste0(dir_out,'heatmap.pdf'))
-# Create the heatmap using ComplexHeatmap
-heatmap <- Heatmap(
-  data_matrix_numeric,
-  col = my_color_palette,
-  row_dend_reorder = TRUE,  # Adjust this based on your preference
-  column_title = "Folds",
-  name = "Fold Value",
-  show_row_names = FALSE,
-  show_column_names = TRUE,
-  cluster_columns = FALSE
-)
-
-draw(heatmap)
-dev.off()
-```
-
-```{r DimPlot with cycling fibroblasts}
-data <- readRDS("/home/common/data/output/projects/ENDO/E031/A004/proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated.RDS")
-DimPlot(data, raster = FALSE)
-meta <- data@meta.data
-write.csv(meta, paste0(dir_out,"E031A004_proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated_meta.csv"))
-
-# make dotplot
-features <- c("NFKBIA", "SLPI", "CXCL2","CXCL3")
-Idents(data) <- data$Annotation
-
-epithelial <- subset(data, idents = "Unciliated")
-
-DotPlot(
-  epithelial,
-  assay = "SCT",
-  features,
-  cols = c("lightgrey", "blue"),
-  col.min = -2.5,
-  col.max = 2.5,
-  dot.min = 0,
-  dot.scale = 6,
-  idents = NULL,
-  group.by = "EndometriosisStatus",
-  split.by = NULL,
-  cluster.idents = FALSE,
-  scale = TRUE,
-  scale.by = "radius",
-  scale.min = NA,
-  scale.max = NA
-)
-
-```
-
-```{r Expression distribution of all genes and pca signature}
-data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
-#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
-
-#CellCnn gene signature
-#Summary table for cyc fibro corrgpca: 
-CellCnnsig_fibro_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E041/A010/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
-#Summary table for  cyc fibro pca :  
-CellCnnsig_fibro_pca <- read.csv("/home/common/data/output/projects/ENDO/E041/A012/sigs/weighted_score_sig_pca_1pc.csv") 
-#Summary table for  all cells corrgpca: 
-CellCnnsig_all_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E042/A011/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
-#Summary table for  all cells pca: 
-CellCnnsig_all_pca <- read.csv("/home/common/data/output/projects/ENDO/E042/A012/sigs/weighted_score_sig_pca_1pc.csv") 
-CellCnnsig <- unique(c(CellCnnsig_all_corrgpca$gene, CellCnnsig_all_pca$gene))
-ScaiNETsamples <- read.csv("/home/common/data/output/projects/ENDO/E041/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")
-
-#Make UpSet plot
-# Create table with the gene sets
-gene_sets <- list(
-    PCA = CellCnnsig_all_pca$gene,
-    CorrgPCA = CellCnnsig_all_corrgpca$gene)
-all_genes <- unique(unlist(gene_sets))
-df <- as.data.frame(lapply(gene_sets, function(gene_set) {
-  setNames(as.logical(all_genes %in% gene_set), all_genes)
-}))
-# Make upset plot
-pdf(file.path(dir_out, "upset_CellCnnsig_all_pca_corrgpcs.pdf"))
-upset(df,
-  colnames(df),
-  set_sizes=(
-        upset_set_size(position='right'))) +
-  ggtitle("all cells")
-dev.off()
-
-#CTL and Endo samples used for the DEG
-#Filter for samples with >50 cells/sample
-data_metadata <- data@meta.data
-df_sample_cells <- data@meta.data %>% 
-  dplyr::group_by(sample, EndometriosisStatus) %>% 
-  dplyr::summarize(cells = n()) %>% 
-  dplyr::filter(cells >= 50) %>% 
-  dplyr::filter(sample %in% ScaiNETsamples$names) 
-ENDO <- df_sample_cells %>% filter(EndometriosisStatus == "Endo") %>%
-  select(sample) %>%
-  pull(sample)
-CTL <- df_sample_cells %>% filter(EndometriosisStatus == "CTL") %>%
-  select(sample) %>%
-  pull(sample)
-#Retrieve number of Endo/CTL samples with RNA expression
-exprmatrix_data <- AggregateExpression(
-  data,
-  assays = "RNA",
-  features = NULL,
-  return.seurat = FALSE,
-  group.by = c("sample"),
-  slot = "counts",
-  verbose = TRUE)[["RNA"]]
-exprmatrix_data <- as.data.frame(exprmatrix_data)
-exprmatrix_data$gene <- row.names(exprmatrix_data)
-exprmatrix_data <- exprmatrix_data %>%
-  mutate(Endo_expressed_n = rowSums(select(., all_of(ENDO)) > 0)) %>%
-  mutate(CTL_expressed_n = rowSums(select(., all_of(CTL)) > 0)) %>%
-  mutate(samples_expressed_n = Endo_expressed_n + CTL_expressed_n) %>%
-  mutate(ENDOorCTL_max_expressed_n = pmax(Endo_expressed_n, CTL_expressed_n))
-table(exprmatrix_data$ENDOorCTL_max_expressed_n)
-
-exprmatrix_CellCnnsig <- exprmatrix_data %>%
-  filter(row.names(exprmatrix_data) %in% CellCnnsig$gene)
-table(exprmatrix_CellCnnsig$ENDOorCTL_max_expressed_n)
-
-exprmatrix_data <- exprmatrix_data %>%
-  mutate(CellCnnsig = row.names(exprmatrix_data) %in% CellCnnsig$gene)
-
-# Create a histogram
-ggplot(exprmatrix_data, aes(x = samples_expressed_n, y = after_stat(count))) +
-  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
-  facet_wrap(~ CellCnnsig, ncol = 2, scales = "free_y") +
-  labs(x = "number of samples with gene expression", y = "number of genes expressed") +
-  theme_minimal() +
-  geom_text(
-    aes(label = after_stat(count)),
-    stat = "count",
-    vjust = -0.5,
-    size = 2,
-    position = position_stack()
-  )
-
-################################ in work
-# Assuming your data is stored in a data frame named 'CellCnnsig_all_pca'
-library(ggplot2)
-
-color_scale <- scale_fill_gradient(low = "red", high = "green")
-
-heatmap_plot <- ggplot(CellCnnsig_all_pca, aes(x = X, y = gene, fill = rank)) +
-  color_scale +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-heatmap_plot <- heatmap_plot +
-  geom_text(aes(label = gene), vjust = 1, hjust = 1, size = 3)
-
-# Show the plot
-print(heatmap_plot)
-#####################################
-```
-
-```{r Expression distribution of all genes and pca signature for cf}
-#CellCnn gene signature
-#Summary table for cyc fibro corrgpca: 
-CellCnnsig_fibro_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E041/A010/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
-#Summary table for  cyc fibro pca :  
-CellCnnsig_fibro_pca <- read.csv("/home/common/data/output/projects/ENDO/E041/A012/sigs/weighted_score_sig_pca_1pc.csv") 
-CellCnnsig <- unique(c(CellCnnsig_fibro_corrgpca$gene, CellCnnsig_fibro_pca$gene))
-
-#Make UpSet plot
-# Create table with the gene sets
-gene_sets <- list(
-    PCA = CellCnnsig_fibro_pca$gene,
-    CorrgPCA = CellCnnsig_fibro_corrgpca$gene)
-all_genes <- unique(unlist(gene_sets))
-df <- as.data.frame(lapply(gene_sets, function(gene_set) {
-  setNames(as.logical(all_genes %in% gene_set), all_genes)
-}))
-# Make upset plot
-pdf(file.path(dir_out, "upset_CellCnnsig_fibro_pca_corrgpcs.pdf"))
-upset(df,
-  colnames(df),
-  set_sizes=(
-        upset_set_size(position='right'))) +
-  ggtitle("fibro")
 dev.off()
@@
 meta <- meta %>% 
-  select(AnnotationMain,Symphony_Refined,Symphony_Refined_A029_United) %>%
+  select(Symphony_Global_main5,Symphony_Refined,Symphony_Refined_A029_United) %>%
   distinct()
 RNAlistRDS005_df <- RNAlistRDS005_df %>%
-  mutate(AnnotationMain = cluster_id)
-RNAlistRDS005_df$AnnotationMain <- case_when(
-  RNAlistRDS005_df$AnnotationMain %in% meta$Symphony_Refined ~ meta$AnnotationMain[match(RNAlistRDS005_df$AnnotationMain, meta$Symphony_Refined)],
-  RNAlistRDS005_df$AnnotationMain %in% meta$Symphony_Refined_A029_United ~ meta$AnnotationMain[match(RNAlistRDS005_df$AnnotationMain, meta$Symphony_Refined_A029_United)],
-  TRUE ~ RNAlistRDS005_df$AnnotationMain)
-RNAlistRDS005_df$AnnotationMain[RNAlistRDS005_df$AnnotationMain == "myeloid" |
-                                      RNAlistRDS005_df$AnnotationMain == "lymphocyte"] <- "immune"
-
-###Plot DEGs
-#vulcano plot without gene labels
-ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc), color = stage)) +
-  geom_point() +
-  theme_minimal() +
-  #ylim(-2, 17) +
-  xlim(-1, 1.2)
-ggsave(paste0(dir_out,"VulcanoPlot_DEGs_stage.pdf"), 
-       plot = last_plot(), 
-       height = 6.3, width = 5)
-
-#vulcano plot with gene labels
-GenesToLabel <- RNAlistRDS005_df %>% 
-               #distinct(gene, logFC,AveExpr,p_adj.loc, .keep_all = TRUE) %>% 
-                dplyr::filter(abs(logFC) >= 0.5)
-ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc), color = stage)) +
-  geom_point() +
-  theme_minimal() +
-  #ylim(-2, 17) +
-  xlim(-1, 1.2) +
-  geom_text_repel(data = GenesToLabel,
-                  aes(label = gene),
-                  size = 2,
-                  max.overlaps = 50,
-                  box.padding = 0.35,
-                  point.padding = 0.5,
-                  segment.color = 'grey50')
+  mutate(Symphony_Global_main5 = cluster_id)
+RNAlistRDS005_df$Symphony_Global_main5 <- case_when(
+  RNAlistRDS005_df$Symphony_Global_main5 %in% meta$Symphony_Refined ~ meta$Symphony_Global_main5[match(RNAlistRDS005_df$Symphony_Global_main5, meta$Symphony_Refined)],
+  RNAlistRDS005_df$Symphony_Global_main5 %in% meta$Symphony_Refined_A029_United ~ meta$Symphony_Global_main5[match(RNAlistRDS005_df$Symphony_Global_main5, meta$Symphony_Refined_A029_United)],
+  TRUE ~ RNAlistRDS005_df$Symphony_Global_main5)
+RNAlistRDS005_df$Symphony_Global_main5[RNAlistRDS005_df$Symphony_Global_main5 == "myeloid" |
+                                      RNAlistRDS005_df$Symphony_Global_main5 == "lymphocyte"] <- "immune"
+
 #Barplot unique genes per main cell type
 RNAlistRDS005_df %>% 
-  select(AnnotationMain,gene,stage,cluster_id) %>%
-  ggplot(aes(x = stage, fill = cluster_id)) +
-  geom_bar() +
-  facet_wrap(~ AnnotationMain, ncol = 4) +
-  theme(axis.text.x = element_text(angle=45, vjust=.2, hjust=0.3))
-
-#Barplot unique genes per main cell type
-RNAlistRDS005_df %>% 
-  select(AnnotationMain, gene, stage, cluster_id) %>%
-  ggplot(aes(x = stage, fill = factor(AnnotationMain, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
+  select(Symphony_Global_main5, gene, stage, cluster_id) %>%
+  ggplot(aes(x = stage, fill = factor(Symphony_Global_main5, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
   geom_bar() +
@@
 
-#Barplot unique genes per main cell type
+#Barplot unique genes per main cell type, filter for duplicates. Doublecheck numbers
 RNAlistRDS005_df %>% 
-  select(AnnotationMain, gene, stage) %>%
+  select(Symphony_Global_main5, gene, stage) %>%
   distinct() %>%
-  ggplot(aes(x = stage, fill = factor(AnnotationMain, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
+  ggplot(aes(x = stage, fill = factor(Symphony_Global_main5, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
   geom_bar() +
@@
 RNAlistRDS005_df %>% 
-      select(AnnotationMain, gene, stage) %>%
+      select(Symphony_Global_main5, gene, stage) %>%
       distinct() %>%
-      count(stage, AnnotationMain)
+      count(stage, Symphony_Global_main5)
 
@@
 table <- RNAlistRDS005_df %>% 
-      select(AnnotationMain, gene, stage) %>%
+      select(Symphony_Global_main5, gene, stage) %>%
       distinct() %>%
-      count(AnnotationMain, stage)
-
-```
-
-```{r DEGs vs CellCNN gene signature}
-##Load dataframes
-#CellCnn gene signature
-#Summary table for cyc fibro corrgpca: 
-CellCnnsig_fibro_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E041/A010/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
-#Summary table for  cyc fibro pca :  
-CellCnnsig_fibro_pca <- read.csv("/home/common/data/output/projects/ENDO/E041/A012/sigs/weighted_score_sig_pca_1pc.csv") 
-#Summary table for  all cells corrgpca: 
-CellCnnsig_all_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E042/A011/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv") 
-CellCnnsig_all_corrgpca$origin <- "A011corrgpca"
-#Summary table for  all cells pca: 
-CellCnnsig_all_pca <- read.csv("/home/common/data/output/projects/ENDO/E042/A012/sigs/weighted_score_sig_pca_1pc.csv") 
-CellCnnsig_all_pca$origin <- "A012pca"
-
-CellCnnsig <- rbind(CellCnnsig_all_corrgpca, CellCnnsig_all_pca)
-
-#RNA assay DEGs from proliferative samples
-RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")
-
-##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
-#Haemoglobin genes
-hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes
-CellCnnsig[CellCnnsig$gene %in% hb.genes, ]#no hb genes
-RNAlistRDS005_df[RNAlistRDS005_df$gene %in% hb.genes, ]#no hb genes
-#mitochondrial genes
-grep("^MT-", CellCnnsig$gene, value=TRUE)#no mt genes
-grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#no mt genes
-#ribosomal genes
-rb.genes_CellCnnsig <- grep(pattern = "^RP[SL]", CellCnnsig$gene, value=TRUE)#3 rb genes
-CellCnnsig <- CellCnnsig[!(CellCnnsig$gene %in% rb.genes_CellCnnsig), ]
-rb.genes_DEGs <- unique(grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE))#77 rb genes
-RNAlistRDS005_df <- RNAlistRDS005_df[!(RNAlistRDS005_df$gene %in% rb.genes_DEGs), ]
-
-##Combine the two dataframes
-RNAlistRDS005_df$CellCnnsig <- factor(RNAlistRDS005_df$gene %in% CellCnnsig$gene,
-                                        levels = c(FALSE, TRUE),
-                                        labels = c("Non-overlapping", "overlapping"))
-
-##Make plots
-# Create a scatter plot logFC vs AveExpr
-ggplot(RNAlistRDS005_df, aes(x = AveExpr, y = logFC)) +
-  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "Scatter Plot",
-       color = "Overlap with CellCnnsig") +
-  theme_minimal()
-ggplot(RNAlistRDS005_df, aes(x = AveExpr, y = logFC, color = cluster_id)) +
-  geom_point(size = 1) +
-  labs(title = "Scatter Plot") +
-  theme_minimal()
-
-# Create a scatter plot logFC vs AveExpr
-ggplot(RNAlistRDS005_df, aes(x = p_adj.loc, y = logFC)) +
-  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "Scatter Plot",
-       color = "Overlap with CellCnnsig") +
-  theme_minimal()
-ggplot(RNAlistRDS005_df, aes(x = p_adj.loc, y = logFC, color = cluster_id)) +
-  geom_point(size = 1) +
-  labs(title = "Scatter Plot") +
-  theme_minimal()
-
-# Create a scatter plot
-ggplot(RNAlistRDS005_df, aes(x = logFC, y = p_adj.loc)) +
-  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "Scatter Plot",
-       color = "Overlap with CellCnnsig E042A011A012") +
-  theme_minimal()
-
-# Create vulcano plot
-ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "Scatter Plot",
-       color = "Overlap with CellCnnsig E042A011A012") +
-  theme_minimal()
-ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc), color = cluster_id)) +
-  geom_point(size = 1.2) 
-ggsave(filename = paste0(dir_out, 'vulcano_clusterid_10_5.pdf'), width =6,height=3)
-
-overlap <- RNAlistRDS005_df[(RNAlistRDS005_df$gene %in% CellCnnsig$gene),]
-length(unique(overlap$gene))
-#Plot 
-par(mfrow = c(2, 1))
-plot(density(overlap$AveExpr))
-plot(density(RNAlistRDS005_df$AveExpr))
-
-#Histogram
-# Create the barplot
-summary_data <- RNAlistRDS005_df %>%
-  group_by(cluster_id, stage, CellCnnsig) %>%
-  dplyr::summarize(Freq = n())
-# Create the barplot with color by CellCnnsig
-
-# Create the barplot with a log2 scale on the y-axis
-ggplot(summary_data, aes(x = cluster_id, y = Freq, fill = CellCnnsig)) +
-  geom_bar(position = "stack", stat = "identity") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_wrap(~stage, scales = "free_x", ncol = 4) +
-  scale_y_break(c(100, 150), scales = 0.3)+ 
-  scale_y_break(c(350, 2500), scales = 0.3)
-
-ggplot(RNAlistRDS005_df, aes(x = cluster_id, y = p_adj.loc, color = CellCnnsig)) +
-  geom_violin() +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_wrap(~stage, scales = "free_x", ncol = 4)
-ggplot(RNAlistRDS005_df, aes(x = cluster_id, y = logFC, color = CellCnnsig)) +
-  geom_violin() +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_wrap(~stage, scales = "free_x", ncol = 4)
-ggplot(RNAlistRDS005_df, aes(x = cluster_id, y = AveExpr, color = CellCnnsig)) +
-  geom_boxplot(outlier.colour="black", outlier.shape=16,
-             outlier.size=0, notch=FALSE) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_wrap(~stage, scales = "free_x", ncol = 4)
-
-ggplot(summary_data, aes(x = cluster_id, y = Freq, fill = CellCnnsig)) +
-  geom_bar(position = "fill", stat = "identity") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_wrap(~stage, scales = "free_x", ncol = 4)
-
-#Get number of overlapping DEGs per cluster
-head(RNAlistRDS005_df)
-overlapping <- RNAlistRDS005_df[RNAlistRDS005_df$gene %in% CellCnnsig$gene,]
-table(overlapping$cluster_id)
-table(RNAlistRDS005_df$cluster_id)
-
-#Density plot for logFC
-p1 <- ggplot(RNAlistRDS005_df, aes(x = logFC, fill = CellCnnsig)) +
-  geom_density(alpha = 0.5) +  
-  theme_minimal() + 
-  scale_fill_manual(values = c("Non-overlapping" = "blue", "overlapping" = "orange"),drop = FALSE) 
-p1
-#Density plot for AveExpr
-p2 <- ggplot(RNAlistRDS005_df, aes(x = log2(AveExpr - 17), fill = CellCnnsig)) +
-  geom_density(alpha = 0.5) + 
-  theme_minimal() + 
-  scale_fill_manual(values = c("Non-overlapping" = "blue", "overlapping" = "orange"),drop = FALSE) +
-  geom_vline(xintercept = 0.5849625)
-p2
-#Density plot for p_adj.loc
-p3 <- ggplot(RNAlistRDS005_df, aes(x = p_adj.loc, fill = CellCnnsig)) +
-  geom_density(alpha = 0.5) + 
-  theme_minimal() + 
-  scale_fill_manual(values = c("Non-overlapping" = "blue", "overlapping" = "orange"),drop = FALSE) 
-plot_grid(p1,p2,p3, ncol=1)
-
-```
-
-```{r Make rank by padj.loc and logFC and evaluate in vulcano plot}
-table <- RNAlistRDS005_df %>%
-  mutate(logFC_rank = rank(-abs(logFC))) %>%
-  mutate(p_adj.loc_rank = rank(p_adj.loc)) %>%
-  mutate(padj.loc_logFC_rank = (logFC_rank + p_adj.loc_rank)) %>%
-  mutate(padj.loc_logFC10_rank = (10*logFC_rank + p_adj.loc_rank)) %>%
-  mutate(padj.loc_logFC100_rank = (100*logFC_rank + p_adj.loc_rank)) %>%
-  arrange(AveExpr)
-table_overlapping <- table[table$gene %in% CellCnnsig$gene,]
-
-#table200 <- table %>% arrange(mean_rank) %>% slice_head(n = 300)
-
-# Create vulcano plot
-p1 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_logFC_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_logFC_rank, slice_head300") +
-  theme_minimal()
-p1
-
-p2 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_logFC10_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_logFC10_rank, slice_head300") +
-  theme_minimal()
-p2
-
-p3 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_logFC100_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_logFC100_rank, slice_head300") +
-  theme_minimal()
-p3
-
-p4 <- ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = RNAlistRDS005_df, aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = RNAlistRDS005_df %>% dplyr::filter(CellCnnsig == "overlapping"),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "Scatter Plot",
-       color = "CellCnnsigE042A011A012") +
-  theme_minimal()
-p4
-
-p5 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
-  geom_point() +
-  labs(title = "RNAlistRDS005_df") +
-  theme_minimal() +
-  scale_colour_gradientn(colours = rev(rainbow(5)),
-                       breaks = seq(18.3, 25, by = 5),
-                       limits = c(18.3, 25),
-                       labels = as.character(seq(18.3, 25, by = 5)))
-p5
-p6 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
-  geom_point() +
-  labs(title = "RNAlistRDS005_df") +
-  theme_minimal() +
-  scale_colour_gradientn(colours = rev(rainbow(5)),
-                       breaks = seq(18.5, 25, by = 5),
-                       limits = c(18.5, 25),
-                       labels = as.character(seq(18.5, 25, by = 5)))
-p6
-plot_grid(p1,p2,p3,p4,p5,p6, ncol=3)
-
-#Keep in mind to remove gene duplicates before giving ShSh the list
-```
-
-```{r Make rank by padj.loc and logFC and AveExpr and evaluate in vulcano plot}
-table <- RNAlistRDS005_df %>%
-  mutate(logFC_rank = rank(-abs(logFC))) %>%
-  mutate(p_adj.loc_rank = rank(p_adj.loc)) %>%
-  mutate(AveExpr_rank = rank(-AveExpr)) %>%
-  mutate(padj.loc_logFC_AveExpr_rank = (logFC_rank + p_adj.loc_rank + AveExpr_rank)) %>%
-  mutate(padj.loc_logFC10_AveExpr_rank = (10*logFC_rank + p_adj.loc_rank + AveExpr_rank)) %>%
-  mutate(padj.loc_10logFC_10AveExpr_rank = (logFC_rank + p_adj.loc_rank + 10*AveExpr_rank)) %>%
-  arrange(AveExpr)
-table_overlapping <- table[table$gene %in% CellCnnsig$gene,]
-#table200 <- table %>% arrange(mean_rank) %>% slice_head(n = 300)
-
-# Create vulcano plot
-p1 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_logFC_AveExpr_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_logFC_AveExpr_rank, slice_head300") +
-  theme_minimal()
-p1
-
-p2 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_logFC10_AveExpr_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_logFC10_AveExpr_rank, slice_head300") +
-  theme_minimal()
-p2
-
-p3 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_10logFC_10AveExpr_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_10logFC_10AveExpr_rank, slice_head300") +
-  theme_minimal()
-p3
-
-p4 <- ggplot(table_overlapping, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
-  geom_point() +
-  labs(title = "table_overlapping") +
-  theme_minimal() +
-  scale_colour_gradientn(colours = rev(rainbow(5)),
-                       breaks = seq(18.3, 25, by = 5),
-                       limits = c(18.3, 25),
-                       labels = as.character(seq(18.3, 25, by = 5)))
-p4
-
-p5 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
-  geom_point() +
-  labs(title = "RNAlistRDS005_df") +
-  theme_minimal() +
-  scale_colour_gradientn(colours = rev(rainbow(5)),
-                       breaks = seq(18.3, 25, by = 5),
-                       limits = c(18.3, 25),
-                       labels = as.character(seq(18.3, 25, by = 5)))
-p5
-p6 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
-  geom_point() +
-  labs(title = "RNAlistRDS005_df") +
-  theme_minimal() +
-  scale_colour_gradientn(colours = rev(rainbow(5)),
-                       breaks = seq(18.5, 25, by = 5),
-                       limits = c(18.5, 25),
-                       labels = as.character(seq(18.5, 25, by = 5)))
-p6
-plot_grid(p1,p2,p3,p4,p5,p6, ncol=3)
-
-#Keep in mind to remove gene duplicates before giving ShSh the list
-```
-
-```{r Make rank by padj.loc and logFC, AveExpr > 18.5 and evaluate in vulcano plot}
-table <- RNAlistRDS005_df %>%
-  dplyr::filter(AveExpr > 18.5) %>%
-  mutate(logFC_rank = rank(-abs(logFC))) %>%
-  mutate(p_adj.loc_rank = rank(p_adj.loc)) %>%
-  mutate(padj.loc_logFC_rank = (logFC_rank + p_adj.loc_rank)) %>%
-  mutate(padj.loc_logFC10_rank = (10*logFC_rank + p_adj.loc_rank)) %>%
-  mutate(padj.loc_logFC2_rank = (2*logFC_rank + p_adj.loc_rank)) %>%
-  arrange(AveExpr)
-table_overlapping <- table[table$gene %in% CellCnnsig$gene,]
-
-#Histogram
-# Create the barplot
-summary_data <- table %>%
-  group_by(cluster_id, stage, CellCnnsig) %>%
-  dplyr::summarize(Freq = n())
-# Create the barplot with color by CellCnnsig
-
-# Create the barplot with a log2 scale on the y-axis
-ggplot(summary_data, aes(x = cluster_id, y = Freq, fill = CellCnnsig)) +
-  geom_bar(position = "stack", stat = "identity") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  facet_wrap(~stage, scales = "free_x", ncol = 4) +
-  scale_y_break(c(40, 50), scales = 0.3)
-
-#table200 <- table %>% arrange(mean_rank) %>% slice_head(n = 300)
-
-# Create vulcano plot
-p1 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_logFC_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_logFC_rank, slice_head300") +
-  theme_minimal()
-p1
-
-p2 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_logFC10_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_logFC10_rank, slice_head300") +
-  theme_minimal()
-p2
-
-p3 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = table,
-             aes(color = "Non-overlapping"), size = 1) +
-  geom_point(data = table %>% arrange(padj.loc_logFC2_rank) %>% slice_head(n = 300),
-             aes(color = "overlapping"), size = 2) +
-  scale_color_manual(values = c("Non-overlapping" = "black", "overlapping" = "purple")) +
-  labs(title = "padj.loc_logFC2_rank, slice_head300") +
-  theme_minimal()
-p3
-
-p4 <- ggplot(table_overlapping, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
-  geom_point() +
-  labs(title = "table_overlapping") +
-  theme_minimal() +
-  scale_colour_gradientn(colours = rev(rainbow(5)),
-                       breaks = seq(18.5, 25, by = 5),
-                       limits = c(18.5, 25),
-                       labels = as.character(seq(18.5, 25, by = 5)))
-p4
-
-p5 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
-  geom_point() +
-  labs(title = "RNAlistRDS005_df") +
-  theme_minimal() +
-  scale_colour_gradientn(colours = rev(rainbow(5)),
-                       breaks = seq(18.3, 25, by = 5),
-                       limits = c(18.3, 25),
-                       labels = as.character(seq(18.3, 25, by = 5)))
-p5
-p6 <- ggplot(table, aes(x = logFC, y = -log10(p_adj.loc), color = AveExpr)) +
-  geom_point() +
-  labs(title = "RNAlistRDS005_df") +
-  theme_minimal() +
-  scale_colour_gradientn(colours = rev(rainbow(5)),
-                       breaks = seq(18.5, 25, by = 5),
-                       limits = c(18.5, 25),
-                       labels = as.character(seq(18.5, 25, by = 5)))
-p6
-plot_grid(p1,p3,p2,p4,p6, ncol=3)
-
-#Keep in mind to remove gene duplicates before giving ShSh the list
-```
-
-```{r df of table()s }
-metadata <- read.csv("/home/common/data/output/projects/ENDO/E044/A006/Metadata_allE044_filtered.csv")
-df <- data.frame(unclass(table(metadata$EndometriosisStatus,metadata$CyclePhases)))
-df <- df %>%
-  rename(Mid.Secretory = Secretory) %>%
-  select(Proliferative, Early.Secretory, Mid.Secretory, Late.Secretory)
-
-```
-
-```{r VlnPlots of SLPI and NFKBIA for epithelial cells}
-epithelial <- readRDS("/home/common/data/output/projects/ENDO/E044/A066/epithelial.rds")
-#Filter by proliferative samples used for the DEG analysis
-samples <- unique(read.csv("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/celltype_freq_sample.csv")$Var2)
-Idents(epithelial) <- epithelial$sample
-epithelial_prolif <- subset(epithelial, idents = samples)
-
-#VlnPlot
-p1 <- VlnPlot(epithelial_prolif, features = c("SLPI"), assay = "RNA",group.by = "sample", pt.size = 0,split.by = "EndometriosisStatus") + ylim(0,100) +
-stat_summary(fun.y = median, geom='point', size = 5, colour = "black", shape = 95) + ggtitle("SLPI, RNA assay")
-p2 <- VlnPlot(epithelial_prolif, features = c("NFKBIA"), assay = "RNA",group.by = "sample", pt.size = 0,split.by = "EndometriosisStatus") +
-stat_summary(fun.y = median, geom='point', size = 5, colour = "black", shape = 95) + ggtitle("NFKBIA, RNA assay")
-p3 <- VlnPlot(epithelial_prolif, features = c("SLPI"), assay = "SCT",group.by = "sample", pt.size = 0,split.by = "EndometriosisStatus") +
-stat_summary(fun.y = median, geom='point', size = 5, colour = "black", shape = 95)+ ggtitle("SLPI, SCT assay")
-p4 <- VlnPlot(epithelial_prolif, features = c("NFKBIA"), assay = "SCT",group.by = "sample", pt.size = 0,split.by = "EndometriosisStatus") +
-stat_summary(fun.y = median, geom='point', size = 5, colour = "black", shape = 95)+ ggtitle("NFKBIA, SCT assay")
-plot_grid(p1,p3,p2,p4, ncol=2)
-ggsave(plot = plot_grid(p1,p3,p2,p4, ncol=2), paste0(dir_out, "SLPI_NFKBIA_epithelial_prolif_RNASCT.png"))
-
-##Calculate mean/median SLPI/NKKBIA expression per sample for SCT assay
-expression_values <- GetAssayData(object = epithelial_prolif, assay = "SCT")[c("SLPI", "NFKBIA"), ]
-expression_df <- as.data.frame(t(as.matrix(expression_values)))
-expression_df$sample <- gsub("_.*", "", rownames(expression_df))
-mean_values <- aggregate(. ~ sample, data = expression_df, FUN = mean)
-median_values <- aggregate(. ~ sample, data = expression_df, FUN = median)
-# Combine mean and median values into a single table
-result_table_SCT <- data.frame(
-  sample_SCT = mean_values$sample,
-  SLPI_mean_SCT = mean_values$SLPI,
-  SLPI_median_SCT = median_values$SLPI,
-  NFKBIA_mean_SCT = mean_values$NFKBIA,
-  NFKBIA_median_SCT = median_values$NFKBIA
-)
-
-##Calculate mean/median SLPI/NKKBIA expression per sample for RNA assay
-expression_values <- GetAssayData(object = epithelial_prolif, assay = "RNA")[c("SLPI", "NFKBIA"), ]
-expression_df <- as.data.frame(t(as.matrix(expression_values)))
-expression_df$sample <- gsub("_.*", "", rownames(expression_df))
-mean_values <- aggregate(. ~ sample, data = expression_df, FUN = mean)
-median_values <- aggregate(. ~ sample, data = expression_df, FUN = median)
-# Combine mean and median values into a single table
-result_table_RNA <- data.frame(
-  sample = mean_values$sample,
-  SLPI_mean_RNA = mean_values$SLPI,
-  SLPI_median_RNA = median_values$SLPI,
-  NFKBIA_mean_RNA = mean_values$NFKBIA,
-  NFKBIA_median_RNA = median_values$NFKBIA
-)
-
-#combine RNA and SCT assay and add EndometriosisStatus condition to df
-result_table <- cbind(result_table_RNA,result_table_SCT)
-sample_EndometriosisStatus <- epithelial_prolif@meta.data %>% select(c("sample","EndometriosisStatus")) %>% distinct()
-result_table <- left_join(result_table,sample_EndometriosisStatus, by = "sample")
-write.csv(result_table,"mean_median_SLPI_NFKBIA_epithelial_prolif_RNASCT.csv")
-
-```
-
-```{r Make vulcano plots for DEGs colored by major cell types}
-##RNA assay DEGs from proliferative samples
-RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")
-##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
-hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes
-mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes
-rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes
-RNAlistRDS005_df <- RNAlistRDS005_df %>%
-  dplyr::filter(!gene %in% c(hb.genes, mt.genes, rb.genes)) %>%
-  dplyr::filter(annotation_reference == "Tan") %>%
-  dplyr::filter(AveExpr > 18.53202)
-##Rename cluster_id to major cell types
-meta <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_metadata.rds")
-meta$Symphony_Refined[meta$Symphony_Refined == "M$\\Phi$1-LYVE1"] <- "M-Phi-1-LYVE1"
-meta$Symphony_Refined[meta$Symphony_Refined == "glandular early-secretory"] <- "glandular-early-secretory"
-meta <- meta %>% 
-  select(AnnotationMain,Symphony_Refined,Symphony_Refined_A029_United) %>%
-  distinct()
-
-RNAlistRDS005_df <- RNAlistRDS005_df %>%
-  mutate(AnnotationMain = cluster_id)
-RNAlistRDS005_df$AnnotationMain <- case_when(
-  RNAlistRDS005_df$AnnotationMain %in% meta$Symphony_Refined ~ meta$AnnotationMain[match(RNAlistRDS005_df$AnnotationMain, meta$Symphony_Refined)],
-  RNAlistRDS005_df$AnnotationMain %in% meta$Symphony_Refined_A029_United ~ meta$AnnotationMain[match(RNAlistRDS005_df$AnnotationMain, meta$Symphony_Refined_A029_United)],
-  TRUE ~ RNAlistRDS005_df$AnnotationMain)
-RNAlistRDS005_df$AnnotationMain[RNAlistRDS005_df$AnnotationMain == "myeloid" |
-                                      RNAlistRDS005_df$AnnotationMain == "lymphocyte"] <- "immune"
-
-##Plot
-ggplot(RNAlistRDS005_df, aes(x = logFC, y = -log10(p_adj.loc), color = AnnotationMain)) +
-  geom_point(size = 1) +
-  #theme_minimal() +
-  facet_wrap(~stage)
-ggsave(filename = paste0(dir_out, 'vulcano_DEG_stage_main4.pdf'), width =7,height=4)
-
-ggplot(RNAlistRDS005_df, aes(x = AnnotationMain, fill = cluster_id)) +
-  geom_bar(stat = "count", position = "stack", width = 0.7) +  
-  # theme_minimal() +  
-  facet_wrap(~stage) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
-
-ggsave(filename = paste0(dir_out, 'vulcano_DEG_stage_main5.pdf'), width =7,height=4)
-
-```
-
-```{r Captum scores}
-#Load captum scores and combine tables
-all_cells_corrgpca <- read.csv("/home/common/data/output/projects/ENDO/E042/A011/sigs/weighted_score_sig_corrgpca_eval_20pc_pcs.csv")
-all_cells_corrgpca$dataset <- "all_cells_corrgpca"
-all_cells_pca <- read.csv("/home/common/data/output/projects/ENDO/E042/A012/sigs/weighted_score_sig_pca_1pc.csv")
-all_cells_pca$dataset <- "all_cells_pca"
-
-##RNA assay DEGs from proliferative samples
-RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")
-##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
-hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes
-mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes
-rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes
-RNAlistRDS005_df <- RNAlistRDS005_df %>%
-  dplyr::filter(!gene %in% c(hb.genes, mt.genes, rb.genes)) %>%
-  dplyr::filter(annotation_reference == "Tan") %>%
-  dplyr::filter(AveExpr > 18.53202)
-
-#Combine tables
-all_cells <- rbind(all_cells_pca,all_cells_corrgpca)
-all_cells <- left_join(all_cells, RNAlistRDS005_df, by = "gene") %>% 
-  arrange(desc(rank))
-
-```
-
+      count(Symphony_Global_main5, stage)
+
+```
+
+Supplementary Data Fig. 7a
 ```{r cycling fibroblasts}
-
 #LTs annotation for cycling fibroblasts
@@
 
+Figure 4e
 ```{r Barplots for  filter response median rank E042A027}
@@
 
-```{r make DEG table for paper}
-#RNA assay DEGs from proliferative samples
-RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv") %>%
-  filter(annotation_reference == "Tan") %>%
-  filter(AveExpr > 18.53202)
-##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
-hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes, there are no hb genes in the DEG list
-RNAlistRDS005_df[RNAlistRDS005_df$gene %in% hb.genes, ]#no hb genes
-#mitochondrial genes
-mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes, no genes
-rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes, 77 genes
-RNAlistRDS005_df <- RNAlistRDS005_df %>%
-  dplyr::filter(!gene %in% rb.genes)
-
-write.csv(RNAlistRDS005_df, paste0(dir_out,"RNAlistRDS005_AveExpr18.53202_noRBgenes.csv"))
-
-length(unique(RNAlistRDS005_df$gene)) #427
-```
-
+Supplementary Data Figure 6c
 ```{r}
@@
 ```
+
+make DEG table for paper
+```{r}
+#RNA assay DEGs from proliferative samples
+RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv") %>%
+  filter(annotation_reference == "Tan") %>%
+  filter(AveExpr > 18.53202)
+##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
+hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes, there are no hb genes in the DEG list
+RNAlistRDS005_df[RNAlistRDS005_df$gene %in% hb.genes, ]#no hb genes
+#mitochondrial genes
+mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes, no genes
+rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes, 77 genes
+RNAlistRDS005_df <- RNAlistRDS005_df %>%
+  dplyr::filter(!gene %in% rb.genes)
+
+write.csv(RNAlistRDS005_df, paste0(dir_out,"RNAlistRDS005_AveExpr18.53202_noRBgenes.csv"))
+
+length(unique(RNAlistRDS005_df$gene)) #427
+```

### 2025-01-28 Figures/PaperFiguresextended.Rmd
Added comments

@@
 ```{r}
+# Set dir_out
 dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")
-
+# Load CellChat object
 p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_DEGsamples.rds"))
 
-#Figure 1e
+# eigencorplot for Figure 1e
 pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
@@
 plotting_PCA <- function(SUBSET) {
+    # Load data    
     p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_",SUBSET,".rds"))
-
+    # plot eigencorplot and save  
     pdf(paste0(dir_out,"eigencorplot_final_BH_",SUBSET,".pdf"), width = 10, height = 5)
@@
     legendPosition = 'right')
+# save biplot
 ggsave(plot = SFig3a, "./SFig3a.pdf", width = 7, height = 7)
@@
 Supplementary Figure 3b and e
-UMAP of re-integrated epithelial cells colored by menstrual cycle day and menstrual cycle phase Wang
-```{r}
+UMAP of re-integrated epithelial cells colored by menstrual cycle day and menstrual cycle phase Wang.
+UMAP of the entire dataset colored by the Symphony transferred Wang phases, show a clear visual separation between phase 1/2, phase 3, and phase 4.
+```{r}
+# Load data
 epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))
 
-#Menstrual Cycle Day
+#Plot UMAP with Menstrual Cycle Day
 SFig3b <- plot_cells(epithelial_cells_monocle3, color_cells_by = "CycleDay",
@@
 SFig3b
+# save UMAP
 ggsave(filename = paste0(dir_out, 'SFig3b.pdf'),
@@
 
-#Menstual Cycle Phase Wang
+#Plot UMAP with Menstual Cycle Phase Wang
 SFig3e <- plot_cells(epithelial_cells_monocle3 , color_cells_by = "WangMenstrualCyclePhase",
@@
 SFig3e
+# save UMAP
 ggsave(filename = paste0(dir_out, 'SFig3e.pdf'),
@@
 
-xxxsee above for object name in GEO
 Supplementary Data Figure 3c
-Look in IBU folder for object with trajectory line or A063/A066
-```{r}
-#Figure S3c epithelial cells
-FeaturePlot
-Fig3c <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCycleDay", raster = FALSE)
-Fig3c
-ggsave(filename = paste0(dir_out, 'Fig3c.pdf'),
-       plot = Fig3c,
-       width =20, height=9)
+Scatter plots showing the correlation of menstrual cycle day and serum progesterone with our menstrual cycle staging (c) and median Wang phases (d), transferred with Symphony (d).
+```{r}
+xxx missing
+
 ```
@@
 Supplementary Data Figure 3f
-CODE OK
-```{r}
-#Figure 2b entire atlas
-Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "Wang2020Phase", raster = FALSE)
+UMAP of the entire dataset colored by the Symphony transferred Wang phases, show a clear visual separation between phase 1/2, phase 3, and phase 4. 
+```{r}
+## Figure 2b UMAP entire atlas
+Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "WangMenstrualCyclePhase", raster = FALSE)
 Fig2b_1
+# Save UMAP
 ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
@@
 Supplementary Data Figure 3 g
-CODE OK
+Heatmap showing the expression of menstrual cycle phase markers (as in Fig. 2d) of the eutopic endometrial epithelial cells (g) from the Tan et al. 2022 single-cell atlas.
 ```{r}
@@
 Supplementary Data Figure 3 h
-CODE OK
+Heatmaps showing the expression of menstrual cycle phase markers (as in Fig. 2d) of the eutopic endometrial mesenchymal cells (h) from the Tan et al. 2022 single-cell atlas.
 ```{r}
@@
 Supplementary Data Figure 4a
-CODE OK
+Bar plot of cell type frequencies per sample, sorted by menstrual cycle phase and pseudo time and divided by the main 5 cell types. 
 ```{r plotting of cell types per sample}
@@
 #Make df with number of AnnotationRefined cells in each sample
-df <- EndoAtlas_meta%>%
+SFig4a_df <- EndoAtlas_meta%>%
   filter(!is.na(AnnotationMain)) %>%
@@
 #Make bar plot
-p <- ggplot(df, aes(x = sample, y = row_count, fill = AnnotationRefined)) +
+SFig4a <- ggplot(SFig4a_df, aes(x = sample, y = row_count, fill = AnnotationRefined)) +
   geom_bar(stat = 'identity', position = 'fill') +
@@
   facet_grid(AnnotationMain ~ ., scales = "free_y")
-p
+SFig4a
 #Save barplot
 ggsave(paste0(dir_out, "SFig4a_barplot.pdf"),
-       plot = p,
+       plot = SFig4a,
        width = 15, height = 6)
@@
 Supplementary Data Figure 4b
-CODE OK
+Box plots of refined cell type frequencies between the main menstrual cycle phases. 
 ```{r plotting of cell types per sample}
@@
 
-#BoxPlots percentage, EndometriosisStatus, percentage ###for paper Extended Data Fig. S4
+# BoxPlots percentage, EndometriosisStatus, percentage ###for paper Extended Data Fig. S4
 ggplot(SFig4b, aes(x = MenstrualCyclePhase_main, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
@@
   expand_limits(y = 0)
-
+# Save BoxPlots
 ggsave(paste0(dir_out, "SFig4b_EndometriosisStatus_perc.pdf"),
@@
 Supplementary Data Figure 5a
+Genes underlying selected pathways enriched in the cluster- and stage-wise functional enrichment analysis.
 ```{r}
@@
 Supplementary Data Figure 5b-d
-```{r}
+(b) Signaling pathways with relative information flow from CellChat analysis, ranked by overall information flow within the inferred networks between ENDO and non-ENDO.
+(c) Circle plot of TGFb signaling pathway between all cell types in ENDO (top) and non-ENDO (bottom, no significant interactions).
+(d) Bubble plot of selected communication probabilities with significant interaction probabilities (unadjusted p-value < 0.01)  within the TNF signaling pathway.
+```{r}
+#Load CellChat data
+cellchat <- readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
 
@@
        filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12)
-
-#Load CellChat data
-cellchat <- readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
 

### 2025-02-04 05b_Functional_enrichment_analysis/DEG_gProfiler.Rmd
Cleaned up code

@@
 setwd(dir_out)
+```
+
+For Shaoline, also looking at BH readjustment
+```{r}
+##Combine muscat output into one dataframe
+PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
+PATHS <- c("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/data/res_mm_dream.rds",
+           "/home/common/data/output/projects/ENDO/E044/A030/myeloid_lymphocyte_epithelial_endothelial_Tan/data/res_mm_dream.rds",
+           "/home/common/data/output/projects/ENDO/E044/A030/stromal_Tan/data/res_mm_dream.rds",
+           "/home/common/data/output/projects/ENDO/E044/A030/united_celltypes_Tan/data/res_mm_dream.rds")
+
+# Load data from files
+print(PATHS)
+RNAlistRDS <- lapply(PATHS, readRDS)
+
+# Assign names based on PATHS
+names(RNAlistRDS) <- PATHS
+
+#combine sub-lists
+RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE)
+
+# Filter without BH adjustment 
+##BH adjustment before AveExpr filtering is identical to p_adj.loc, BH adjustment after AveExpr filtering leads to very few significant genes
+RNAlistRDS005 <- lapply(RNAlistRDS, function(x) {
+  dplyr::filter(x, AveExpr > 18.53202 & p_adj.loc < 0.05)
+})
+RNAlistRDS005 <- as.data.frame(do.call(rbind, RNAlistRDS005))
+
+##exclude ribosomal genes (comment: does not change the inflection point)
+rb.genes_DEGs <- unique(grep(pattern = "^RP[SL]", RNAlistRDS005$gene, value=TRUE)) #around 150 unique rb genes
+RNAlistRDS005 <- RNAlistRDS005[!(RNAlistRDS005$gene %in% rb.genes_DEGs), ]
+
+# Filter with BH adjustment after AveExpr filtering leads to very few significant genes
+RNAlistRDS005BH <- lapply(RNAlistRDS, function(x) {
+  x <- dplyr::filter(x, AveExpr > 18.53202)
+  x$p_adj.BH <- p.adjust(x$p_val, method = "BH")
+  dplyr::filter(x, p_adj.BH < 0.05)
+})
+RNAlistRDS005BH <- as.data.frame(do.call(rbind, RNAlistRDS005BH))
+
 ```
@@
 
-###Delete below ###########
-
-RNAlistRDS005 %>% dplyr::filter(logFC > 0.5) %>% nrow()
-RNAlistRDS005 %>% dplyr::filter(logFC < -0.5) %>% nrow()
-
-RNAlistRDS005 <- RNAlistRDS005 %>% dplyr::filter(annotation_reference == "Tan")
-###plotting
-plot(density(RNAlistRDS005$AveExpr))
-# bar plot of DEGs by condition
-df <- RNAlistRDS005 %>%
-  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
-  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
-  ungroup()
-ggplot(df, aes(x=stage, y = DEGs, fill=cluster_id)) +
-  geom_bar(stat='identity') +
-  scale_y_continuous(expand = c(0,0)) +
-  RotatedAxis() +
-  xlab('Sample ID') +
-  facet_wrap(~clustering,scales = "free_x", ncol = 4)
-ggsave(paste0(dir_out, "Barplot_cluster_id_stage_clustering.pdf"),
-  plot = last_plot(),
-  width = 15, height = 6)
-sum(df$DEGs)
-# bar plot of DEGs by condition with AveExpr > 18.5
-#There are extremely many DEGs in the clusters eF, epithelial and C7_Activated_Fibro, too many to be reliable.
-Median AveExpr seems to be lower in those clusters and gProfiler gave good results with a threshold of AveExpr>18.5.
-Therefore, I set the cut-off at AveExpr>18.5
-df18_3 <- RNAlistRDS005 %>%
-  dplyr::filter(AveExpr > 18.3) %>%
-  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
-  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
-  ungroup()
-df18_5 <- RNAlistRDS005 %>%
-  dplyr::filter(AveExpr > 18.5) %>%
-  dplyr::group_by(cluster_id, stage, clustering,annotation_reference) %>%
-  dplyr::summarise(DEGs = n(), median_logFC = median(logFC), median_AveExpr = median(AveExpr), median_padjloc = median(p_adj.loc)) %>%
-  ungroup()
-ggplot(df18_5, aes(x=stage, y = DEGs, fill=cluster_id)) +
-  geom_bar(stat='identity') +
-  scale_y_continuous(expand = c(0,0)) +
-  RotatedAxis() +
-  xlab('Sample ID') +
-  facet_wrap(~clustering,scales = "free_x", ncol = 4) +
-  ggtitle("AveExpr > 18.5")
-ggsave(paste0(dir_out, "Barplot_cluster_id_stage_clustering_AveExpr185.pdf"),
-  plot = last_plot(),
-  width = 15, height = 6)
-ggplot(df18_5, aes(x=stage, y = DEGs, fill=cluster_id)) +
-  geom_bar(stat='identity') +
-  scale_y_continuous(expand = c(0,0)) +
-  RotatedAxis() +
-  xlab('Sample ID')+
-  ggtitle("AveExpr > 18.5") 
-ggsave(paste0(dir_out, "Barplot_cluster_id_stage_AveExpr185.pdf"),
-  plot = last_plot(),
-  width = 6, height = 4)
-sum(df18_5$DEGs)
-
-# Count the number of genes in each combination of clustering and stage
-RNAcounts <- table(RNAlistRDS005$clustering, RNAlistRDS005$stage)
-pheatmap(RNAcounts)
-# Create the grouped bar plot
-ggplot(as.data.frame(RNAcounts), aes(x = Var1, y = Freq, fill = Var2)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(x = "Clustering", y = "Count", fill = "Stage") +
-  ggtitle("Count of Genes by Clustering and Stage") +
-  theme_bw() +
-  theme(plot.title = element_text(size = 14, face = "bold"),
-        axis.text.x = element_text(angle = 45, hjust = 1))
-
-# Create vulcano plot
-ggplot(RNAlistRDS005, aes(x = logFC, y = -log10(p_adj.loc), color = cluster_id)) +
-  geom_point(data = RNAlistRDS005) +
-  theme_minimal() +
-  xlim(-0.8,1.3)
-ggsave(paste0(dir_out, "Vulcano_logFC_padjloc_clusterid.pdf"),
-  plot = last_plot(),
-  width = 6, height = 4)
-
-#scatter plot logFC, AveExpr
-ggplot(RNAlistRDS005, aes(x = AveExpr, y = logFC, color = cluster_id)) +
-  geom_point(data = RNAlistRDS005) +
-  theme_minimal()
-ggsave(paste0(dir_out, "Scatter_logFC_AveExpr_clusterid.pdf"),
-  plot = last_plot(),
-  width = 6, height = 4)
-plot(density(log2(RNAlistRDS005$AveExpr-16.97529)))
 ```
@@
 gProfiler based on RNA/SCT assay and per annotation with gene expression >18.53202
+```{r}
+##function to run gProfiler and save output
+gProfiler_FUN <- function(x,ANALYSIS,GRADE,ORDERED_QUERY,ASSAY,dir_out) {
+#gProfiler
+custom_bg <- unique(x$gene)
+query <- x %>%
+  dplyr::filter(AveExpr > 18.53202) %>%
+  dplyr::filter(p_adj.loc < 0.05) %>%
+  dplyr::arrange(p_adj.loc) #arrange by lowest p_adj.loc
+if(nrow(query) > 0){
+query <- distinct(query, gene, .keep_all= TRUE) #remove duplicates of gene with higher p_adj.loc
+query <- query$gene
+gostres <<- gost(query = query, 
+                organism = "hsapiens", 
+                ordered_query = ORDERED_QUERY, #can use TRUE, p_adj.loc has been arranged 
+                multi_query = FALSE, 
+                significant = TRUE, #only statistically significant results should be returned
+                exclude_iea = FALSE, 
+                measure_underrepresentation = FALSE, 
+                evcodes = TRUE,
+                user_threshold = 0.05, 
+                correction_method = "g_SCS", #g_SCS is default
+                domain_scope = "annotated", #if custom_bg is set, I believe this shouldn't matter too much
+                custom_bg = custom_bg, #custom background, use!
+                numeric_ns = "", 
+                sources = NULL, #by default all sources are analyzed
+                as_short_link = FALSE)
+if(length(gostres) > 0){
+results <- gostres$result
+results$cluster_id <- levels(as.factor(x$cluster_id))
+results$analysis <- print(ANALYSIS)
+results$assay <- print(ASSAY)
+results$grade <- print(GRADE) 
+results$ordered_query <- print(ORDERED_QUERY)
+results$parents <- NULL
+write.csv(results,paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
+print(paste0("Saved gotres as gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
+plot <- gostplot(gostres, capped = TRUE, interactive = FALSE) + ggtitle(paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY)))
+ggsave(plot = plot, filename = paste0(dir_out,"gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".png"), width = 7, height = 5)
+print("Saved plot")
+}
+}
+}
+##function to run gProfiler_FUN on different grades, analysis and assay
+RUN_gProfier <- function(GRADE, ANALYSIS,ASSAY,SPECIFICATION){
+dir_out <- paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ANALYSIS,"/",ASSAY,"assay_",SPECIFICATION,"/")
+dir.create(dir_out, recursive = TRUE)
+setwd(dir_out)
+#Read paths to DEG output .rds files
+if (ASSAY == "RNA") { 
+    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
+    } else if (ASSAY == "SCT") {
+    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "mm_dream_sct_counts_all.rds", recursive = TRUE, full.names = TRUE)
+    }
+#Select paths based on all/mild/severe
+if (GRADE == "all") { 
+    PATHS <- PATHS[!grepl("_severe", PATHS)]
+    PATHS <- PATHS[!grepl("_mild", PATHS)]
+    print(PATHS)
+    } else if (GRADE == "mild") {
+    PATHS <- PATHS[grepl("_mild", PATHS)]
+    print(PATHS)
+    } else if  (GRADE == "severe") {
+    PATHS <- PATHS[grepl("_severe", PATHS)]
+    print(PATHS)
+    } 
+RNAlistRDS <- lapply(PATHS, readRDS)
+RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
+RNAlistRDS <- RNAlistRDS[!duplicated(RNAlistRDS)]
+lapply(RNAlistRDS, gProfiler_FUN, ANALYSIS = ANALYSIS,GRADE = GRADE,ORDERED_QUERY = FALSE, ASSAY = ASSAY, dir_out = dir_out)
+lapply(RNAlistRDS, gProfiler_FUN, ANALYSIS = ANALYSIS,GRADE = GRADE,ORDERED_QUERY = TRUE, ASSAY = ASSAY, dir_out = dir_out)
+}
+
+RUN_gProfier(GRADE = "all", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
+RUN_gProfier(GRADE = "severe", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
+RUN_gProfier(GRADE = "mild", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
+RUN_gProfier(GRADE = "all", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
+RUN_gProfier(GRADE = "severe", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
+RUN_gProfier(GRADE = "mild", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
+```
+
+##Ongoing xxxx
+gProfiler based on RNA assay and per annotation with gene expression >18.53202 and BH pval adjusted
 ```{r}

### 2025-02-04 Figures/PaperFiguresextended.Rmd
Added code for DEG and CellChat analysis

--- 2025-01-27 Figures/PaperFiguresextended.Rmd
@@
 Figure 1c
-UMAP of the entire endometrium single-cell atlas from women with and without endometriosis, colored by cell types (n=63).
-```{r}
-# Define order of cell types
+CODE OK
+```{r}
+
 order <- EndoAtlas_meta %>%
@@
 EndoAtlas$AnnotationRefined <- factor(x = EndoAtlas$AnnotationRefined, levels = order$AnnotationRefined)
-# DimPlot
 Fig1c <- DimPlot(EndoAtlas, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
 Fig1c
-# Save DimPlot
 ggsave(filename = paste0(dir_out, 'Fig1c.pdf'),
@@
 Figure 1d
-Main cell type frequencies per patient (n=60, box plot, left) and entire dataset (bar plot, middle), with mesenchymal cells being the most abundant cell type.
-The overall cell number distribution between non-ENDO, mild-ENDO and severe-ENDO is balanced (bar plot, right).
+CODE OK
 ```{r plotting of cell types per sample}
-## Fig1d_1 Cells per patient
-# Boxplot
+#Fig1d_1
 Fig1d_1 <- EndoAtlas_meta %>%
@@
 Fig1d_1
-# Save Boxplot
 ggsave(paste0(dir_out, "Fig1d_1.pdf"),
@@
 
-## Fig1d_2 Cells per main cell type
-# Barplot
+#Fig1d_2
 Fig1d_2 <- EndoAtlas_meta %>%
@@
 Fig1d_2
-# Save Barplot
 ggsave(paste0(dir_out, "Fig1d_2.pdf"),
@@
 
-##Fig1d_3 cell number distribution between non-ENDO, mild-ENDO and severe-ENDO
-# Barplot
+#Fig1d_3 
 Fig1d_3 <- EndoAtlas_meta%>%
@@
 Fig1d_3
-# Save Barplot
 ggsave(paste0(dir_out, "Fig1d_3.pdf"),
@@
 
-Figure 1e
-Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
-
+Figure 1e, 2a, S3a
 CODE running, need to extend the PCA analysis folder
 dir_out <- "/home/common/data/output/projects/ENDO/E044/A072/allcells_EndoAtlas_SCTassay/"
-```{r}
-## Figure 1e
-# Load data
+
+```{r}
 p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
-# Plot and save eigencorplot
+#Figure 1e
 pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
@@
 dev.off()
-```
-
-Figure 2a
-Scores plot of the first two principal components from PCA of sample-wise whole transcriptome.
-```{r}
-## Figure 2a
-# Load data
-p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
-# biplot
+
+#Figure 2a
 biplot(p,
@@
     legendPosition = 'top', legendLabSize = 16, legendIconSize = 8.0)
+
+#Supplementary Data Figure 3a
+biplot(p,
+    colby = 'EndometriosisStatus',
+    labSize = 0,drawConnectors = FALSE, 
+    hline = 0, vline = 0,
+    legendPosition = 'right')
 ```
@@
 Figure 2b
-UMAPs of the entire dataset (left) and the re-integrated epithelial cells (right), colored by menstrual cycle phase, show a clear visual separation between proliferative/periovulatory, early-, mid-, and late-secretory samples, particularly in mesenchymal (endometrial fibroblasts, C7 fibroblasts, and mural cells), epithelial (glandular), and endothelial cells.
-```{r}
-## Figure 2b entire atlas
-#DimPlot
+CODE OK
+```{r}
+#Figure 2b entire atlas
 Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
 Fig2b_1
-#Save DimPlot
 ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
@@
 
-## Figure 2b epithelial cells
-#DimPlot
+#Figure 2b epithelial cells
 Fig2b_2 <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
 Fig2b_2
-#Save DimPlot
 ggsave(filename = paste0(dir_out, 'Fig2b_2.pdf'),
@@
 Figure 2c
-UMAP of the re-integrated epithelial cells, colored by pseudo time prediction from monocle3.
-```{r}
-# Figure 2c epithelial cells
+Look in IBU folder for object with trajectory line or A063/A066
+```{r}
+#Figure 2c epithelial cells
 epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))
 
-# DimPlot of epithelial cells with pseudotime
+#Pseudotime
 Fig2c <- plot_cells(epithelial_cells_monocle3, color_cells_by = "pseudotime",
@@
 Fig2c
-# Save DimPlot
 ggsave(filename = paste0(dir_out, 'Fig2c.pdf'),
@@
 Figure 2d
-Heatmap of selected marker gene expression with high log fold changes over the key menstrual cycle phases (proliferative, periovulatory, early-, mid-, and late-secretory).
+CODE OK
 ```{r}
@@
 
-#### Plot heatmaps for main cell type specific markers for the different menstrual cycle phases. Compare also to Wang et al 2020 marker genes (Wang et al. 2020, DOI: 10.1038/s41591-020-1040-z)
+#### Plot heatmaps for main cell type specific markers for the different menstrual cycle phases. Compare also Wang et al 2020 marker genes (Wang et al. 2020, DOI: 10.1038/s41591-020-1040-z)
 
@@
 Figure 2e
-Boxplots of sample-wise mesenchymal and epithelial cell frequency as a percentage of total cells, split by menstrual cycle phase (stringent patient exclusion criteria, 32 ENDO, and 22 non-ENDO samples).
+CODE OK
 ```{r plotting of cell types per sample}
@@
 #Prepare data for boxplot
-df_Fig2e <- EndoAtlas_meta%>%
+df2 <- EndoAtlas_meta%>%
   filter(Minor.exclusion.criteria.logical == "FALSE") %>%
@@
 #BoxPlots percentage, EndometriosisStatus 
-Fig2e <- ggplot(df_Fig2e, aes(x = MenstrualCyclePhase_main, y = AnnotationMain_perc, fill = EndometriosisStatus)) +
+ggplot(df2, aes(x = MenstrualCyclePhase_main, y = AnnotationMain_perc, fill = EndometriosisStatus)) +
   geom_boxplot(outlier.shape = NA) +
@@
 ggsave(paste0(dir_out, "Fig2e_boxplot.pdf"),
-  plot = Fig2e,
+  plot = last_plot(),
   width = 12, height = 4.4)
@@
 # Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
-df_Fig2e_ttest_EndometriosisStatus <- df_Fig2e %>%
+df2_ttest_EndometriosisStatus <- df2 %>%
   group_by(AnnotationMain, MenstrualCyclePhase_main) %>%
@@
 ## Perform t-test Prolif vs Peri for each cell type
-df_Fig2e_ttest_ProlifPeri <- df_Fig2e %>%
+df2_ttest_ProlifPeri <- df2 %>%
   dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
@@
 ## Perform t-test Prolif vs Sec for each cell type
-df_Fig2e_ttest_ProlifSec <- df_Fig2e %>%
+df2_ttest_ProlifSec <- df2 %>%
   dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
@@
 ## Perform t-test periovulatory vs sec for each cell type
-df_Fig2e_ttest_PeriSec <- df_Fig2e %>%
+df2_ttest_PeriSec <- df2 %>%
   dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
@@
 
-Make better comments! xxx
 Figure 2f
-Boxplot of sample-wise secretory mural cell frequency as a percentage of total cells.
+CODE OK
 ```{r plotting of cell types per sample}
 #Early and Mid Secretory Prv_VSMCs 
-df_Fig2f <- EndoAtlas_meta%>%
+Fig2f <- EndoAtlas_meta%>%
   filter(Minor.exclusion.criteria.logical == "FALSE") %>%
@@
   filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
-df_Fig2f <- left_join(df_Fig2f, dplyr::distinct(EndoAtlas_meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main","MenstrualCyclePhase")]), by = "sample")
-
-secEarylMid_Prv_VSMC_MenstrualCyclePhase_main <- df_Fig2f %>%
+Fig2f <- left_join(Fig2f, dplyr::distinct(EndoAtlas_meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main","MenstrualCyclePhase")]), by = "sample")
+
+secEarylMid_Prv_VSMC_MenstrualCyclePhase_main <- Fig2f %>%
   filter(AnnotationRefined %in% c("Prv_VSMC secretory", "Prv_VSMC secretory_late")) %>% #same result with and without "Prv_VSMC secretory_late"
@@
 p.adjust(c(0.7639853), method = "BH", n = length(3)) #Value does not change
+
 ```
@@
 Figure 3a and 3b
-3a Barplot showing number of unique differentially expressed genes (DEGs, adjusted p-value < 0.05, average expression > 18.53202) between non-ENDO (n = 11) and all-ENDO (n = 12), mild-ENDO (n = 4) and severe-ENDO (n = 8) samples within immune, endothelial, epithelial, and mesenchymal cell types, analyzed by Muscat.
-3b Selected terms enriched in the cluster- and stage-wise functional enrichment analysis of DEGs, with term sizes between 25 and 750.
 ```{r}
@@
 
-xxx
 Figure 3c-h
-(c) Selected signaling pathways with overall enriched activity in ENDO (n = 12) in comparison to non-ENDO (n = 11).
-(d) Circle plot of TNF signaling pathway between all cell types in ENDO (left, n = 12) and non-ENDO (right, n = 11).
-(e-h) Bubble plots of selected communications with significant interaction probabilities (unadjusted p-value < 0.01) within the ICAM (e), CXCL (f), IGFBP (g) and VEGF (h) signaling pathways for ENDO (n = 12) and non-ENDO (n = 11).
-Further p-value adjustment is not recommended by the CellChat author.
 ```{r}
@@
 Supplementary Data Figure 1a and 1b
-Violin plots illustrating counts and features from RNA assay (a) and SCT assay (b), show no discernable differences between ENDO and non-ENDO.
+CODE OK
 ```{r VlnPlots of nCounts and nFeatures}
-#VlnPlots for "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"
+##VlnPlots for "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"
 VlnPlot(EndoAtlas,
@@
   group.by='EndometriosisStatus', ncol = 1, pt.size=0)
-# Save VlnPlots
 ggsave(paste0(dir_out, "VlnPlot_GroupByEndometriosisStatus_nFeature_nCounts_RNA_SCT_percMT.pdf"),
@@
   width = 4, height = 13)
-```
-
-Supplementary Data Figure 1c and d
-UMAPs of the entire endometrium single-cell atlas, colored by menstrual cycle day (c) and progesterone in serum (d), show a clear visual separation of cells by menstrual cycle day and serum progesterone concentration.
-```{r}
-## SFig1c
-# Featureplot
+
+```
+
+```{r different counts and means/medians}
+#Percentage of main 5
+table <- as.data.frame(table(EndoAtlas_meta$AnnotationMain))
+table$Percentage <- round((table$Freq / sum(table$Freq)) * 100, 1)
+
+#Number of cells for CTL and Endo
+table(EndoAtlas_meta$EndometriosisStatus)
+
+#Number of patients for CTL and Endo
+EndoAtlas_meta %>% select(sample,EndometriosisStatus) %>%
+  distinct() %>%
+  group_by(EndometriosisStatus) %>%
+  summarize(number_of_rows = n())
+
+#mean number of cells per sample
+mean(table(EndoAtlas_meta$sample))
+
+#median count/feature
+median_summary <- EndoAtlas_meta %>%
+  select(nCount_RNA, nCount_SCT, nFeature_RNA, nFeature_SCT) %>%
+  summarize(
+    median_nCount_RNA = median(nCount_RNA),
+    median_nCount_SCT = median(nCount_SCT),
+    median_nFeature_RNA = median(nFeature_RNA),
+    median_nFeature_SCT = median(nFeature_SCT)
+  )
+```
+
+Supplementary Data Figure 1c
+CODE OK
+```{r}
 SFig1c <- FeaturePlot(EndoAtlas, features = "CycleDay", raster = FALSE)
 SFig1c
-# Save Featureplot
 ggsave(filename = paste0(dir_out, 'SFig1c.pdf'),
@@
        width =20, height=9)
-
-## SFig1d
-# Featureplot
+```
+
+Supplementary Data Figure 1d
+CODE OK
+```{r}
 SFig1d <- FeaturePlot(EndoAtlas, features = "ProgesteroneSerum", raster = FALSE)
 SFig1d
-# Save Featureplot
 ggsave(filename = paste0(dir_out, 'SFig1d.pdf'),
@@
 
-xxx double-check code
 Supplementary Data Figure 1e
-Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) of the proliferative samples with strict exclusion criteria with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
+
 ```{r}
@@
 Supplementary Data Figure 1f
-Eigencor plots showing the effect of different co-variants of the atlas dataset on principal components (PCs) of the main 5 cell types (epithelial, mesenchymal, endothelial, myeloid, lymphoid) with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
+CODE OK
 "/home/common/data/output/projects/ENDO/E044/A082/SCTassay/"
@@
 Supplementary Data Figure 2
-Expression of the marker genes for the different cell types corresponds to the cell type-specific expression shown in Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5).
+CODE OK
 ```{refined cluster markers}
@@
 
-Supplementary Figure 3a
-Scores plot of the first two principal components PC1 and PC2 from sample-wise whole transcriptome PCA of the SCT assay, colored by endometriosis status.
-```{r}
-# Load data
-p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
-# biplot
-SFig3a <- biplot(p,
-    colby = 'EndometriosisStatus',
-    labSize = 0,drawConnectors = FALSE, 
-    hline = 0, vline = 0,
-    legendPosition = 'right')
-ggsave(plot = SFig3a, "./SFig3a.pdf", width = 7, height = 7)
-```
-
 Supplementary Figure 3b and e
-UMAP of re-integrated epithelial cells colored by menstrual cycle day and menstrual cycle phase Wang
+CODE running
 ```{r}
@@
 
-xxxsee above for object name in GEO
 Supplementary Data Figure 3c

### 2025-02-04 05b_Functional_enrichment_analysis/DEG_gProfiler.Rmd

@@
 setwd(dir_out)
-```
-
-For Shaoline, also looking at BH readjustment
-```{r}
-##Combine muscat output into one dataframe
-PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
-PATHS <- c("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/data/res_mm_dream.rds",
-           "/home/common/data/output/projects/ENDO/E044/A030/myeloid_lymphocyte_epithelial_endothelial_Tan/data/res_mm_dream.rds",
-           "/home/common/data/output/projects/ENDO/E044/A030/stromal_Tan/data/res_mm_dream.rds",
-           "/home/common/data/output/projects/ENDO/E044/A030/united_celltypes_Tan/data/res_mm_dream.rds")
-
-# Load data from files
-print(PATHS)
-RNAlistRDS <- lapply(PATHS, readRDS)
-
-# Assign names based on PATHS
-names(RNAlistRDS) <- PATHS
-
-#combine sub-lists
-RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE)
-
-# Filter without BH adjustment 
-##BH adjustment before AveExpr filtering is identical to p_adj.loc, BH adjustment after AveExpr filtering leads to very few significant genes
-RNAlistRDS005 <- lapply(RNAlistRDS, function(x) {
-  dplyr::filter(x, AveExpr > 18.53202 & p_adj.loc < 0.05)
-})
-RNAlistRDS005 <- as.data.frame(do.call(rbind, RNAlistRDS005))
-
-##exclude ribosomal genes (comment: does not change the inflection point)
-rb.genes_DEGs <- unique(grep(pattern = "^RP[SL]", RNAlistRDS005$gene, value=TRUE)) #around 150 unique rb genes
-RNAlistRDS005 <- RNAlistRDS005[!(RNAlistRDS005$gene %in% rb.genes_DEGs), ]
-
-# Filter with BH adjustment after AveExpr filtering leads to very few significant genes
-RNAlistRDS005BH <- lapply(RNAlistRDS, function(x) {
-  x <- dplyr::filter(x, AveExpr > 18.53202)
-  x$p_adj.BH <- p.adjust(x$p_val, method = "BH")
-  dplyr::filter(x, p_adj.BH < 0.05)
-})
-RNAlistRDS005BH <- as.data.frame(do.call(rbind, RNAlistRDS005BH))
-
 ```

### 2025-02-17 01_Atlas_generation/Integration/02_run_seuratv3.R
Added snakemake integration pipeline

@@ -1,81 +1,0 @@
-####################################
-## 00. Load library and environment
-####################################
-
-# load packages
-suppressPackageStartupMessages({
-  library(dplyr)
-  library(Seurat)
-  library(future)
-  library(RColorBrewer)
-  library(remotes)
-  library(ggplot2)
-  library(SingleCellExperiment)
-})
-
-# load split seurat object
-seurat_obj_split_input <- readRDS("./seurat_obj_split.rds")
-print("input loaded")
-
-# specify output directory
-dir_out <- snakemake@config$integration_output_dir
-
-####################################
-## 01. Run integration
-####################################
-
-# RPCA integration on SCT transformed counts
-features_integration <- SelectIntegrationFeatures(object.list = seurat_object_list,
-                                                  nfeatures = 3000)
-seurat_object_list <- Seurat::PrepSCTIntegration(object.list = seurat_object_list,
-                                                 assay = "SCT",
-                                                 anchor.features = features_integration,
-                                                 verbose = TRUE
-)
-seurat_object_list <- lapply(seurat_object_list,
-                             FUN = Seurat::RunPCA,
-                             anchor.features = features_integration
-)
-
-
-reference_dataset <- which(names(seurat_object_list) == "batch17")
-
-
-integration_anchors <- FindIntegrationAnchors(object.list = seurat_object_list,
-                                              normalization.method = "SCT",
-                                              anchor.features = features_integration,
-                                              dims = seq_len(20),
-                                              reduction = "rpca",
-                                              k.anchor = 20,
-                                              reference = reference_dataset,
-                                              verbose = TRUE
-)
-
-seurat_object_list <- IntegrateData(anchorset = integration_anchors,
-                                    normalization.method = "SCT",
-                                    dims = seq_len(snakemake@config$n_dimred_integration),
-                                    verbose = TRUE
-)
-
-DefaultAssay(integrated_data) <- "integrated"
-
-seed <- 122334
-
-# Perform linear dimensional reduction by PCA on HVGs
-integrated_data <- RunPCA(object = integrated_data,
-                     features = VariableFeatures(integrated_data),
-                     seed.use = seed,
-                     npcs = 50)
-# Run UMAP
-integrated_data <- RunUMAP(integrated_data,  dims = 1:30, seed.use = seed)
-
-#Save seurat
[…]

### 2025-02-17 01_Atlas_generation/Integration/scripts/01b_run_normalization_split.R

@@ -2,6 +2,14 @@
 ## 00. Load library and environment
 ####################################
 
+# log
+log <- file(snakemake@log$log_file, open="wt")
+sink(log)
+sink(log, type="message")
+
+# image
+save.image(file = paste0(snakemake@log$log_object) )
+
 # load packages
 suppressPackageStartupMessages({
   library(dplyr)
@@ -12,15 +20,17 @@ suppressPackageStartupMessages({
 
 
 # load seurat object
-seurat_obj <- readRDS(./seurat.rds)
+seurat_obj <- readRDS(snakemake@input[[1]])
 Idents(seurat_obj) <- seurat_obj@meta.data$sample
+
 
 ####################################
 ## 01. Normalization split per batch
 ####################################
 
-# SCTransform and regress out percent.mt and sample
-regr.vars <- c("percent.mt","sample")
+regr.vars <- strsplit(snakemake@params[[1]], split='_')[[1]]
+
+# split by batch
 seurat_obj <- SCTransform(seurat_obj,
                           vars.to.regress = regr.vars,
                           conserve.memory=FALSE,
@@ -28,10 +38,8 @@ seurat_obj <- SCTransform(seurat_obj,
                           vst.flavor = "v2",
                           return.only.var.genes=TRUE)
 
-# split by 10x capture batch
-seurat_obj_split <- SplitObject(seurat_obj, split.by = "batch")
+seurat_obj_split <- SplitObject(seurat_obj, split.by = snakemake@params[[2]])
 
-# save split seurat object
-saveRDS(seurat_obj_split, "./seurat_obj_split.rds")
+saveRDS(seurat_obj_split, snakemake@output[[1]])
 sessionInfo()
 date()

### 2025-02-17 01_Atlas_generation/Integration/scripts/02_run_seuratv3.R

@@ -1,0 +1,99 @@
+####################################
+## 00. Load library and environment
+####################################
+
+# log
+log <- file(snakemake@log$log_file, open="wt")
+sink(log)
+sink(log, type="message")
+
+# image
+save.image(file = paste0(snakemake@log$log_object) )
+
+# load packages
+suppressPackageStartupMessages({
+  library(dplyr)
+  library(Seurat)
+  library(future)
+  library(RColorBrewer)
+  library(remotes)
+  library(ggplot2)
+  library(SingleCellExperiment)
+})
+
+# dir_out
+dir_out <- snakemake@config$integration_output_dir
+
+
+# load seurat object
+seurat_obj_split_input <- readRDS(snakemake@input[[1]])
+print("input loaded")
+
+
+
+####################################
+## 01. Run integration
+####################################
+
+rpca_integration = function(seurat_object_list, snakemake, features_integration, verbose = TRUE){
+
+  # RPCA integration on SCT transformed counts
+  features_integration <- SelectIntegrationFeatures(object.list = seurat_obj_split_input,
+                                                    nfeatures = 3000)
+  seurat_object_list <- Seurat::PrepSCTIntegration(object.list = seurat_object_list,
+                                                   assay = "SCT",
+                                                   anchor.features = features_integration,
+                                                   verbose = verbose
+  )
+  seurat_object_list <- lapply(seurat_object_list,
+                               FUN = Seurat::RunPCA,
+                               anchor.features = features_integration
+  )
+
+
+  reference_dataset <- which(names(seurat_object_list) == snakemake@config$reference_dataset)
+
+
+  integration_anchors <- FindIntegrationAnchors(object.list = seurat_object_list,
+                                                normalization.method = "SCT",
+                                                anchor.features = features_integration,
+                                                dims = seq_len(snakemake@config$n_dimred_integration),
+                                                reduction = "rpca",
+                                                k.anchor = 20,
+                                                reference = reference_dataset,
+                                                verbose = verbose
+  )
+  print(paste0('Length of features used for integration',
+               length(integration_anchors@anchor.features)))
+  print(paste0('Number of cells used for integration ',
+               dim(integration_anchors@anchors)[1]))
+  seurat_object_list <- IntegrateData(anchorset = integration_anchors,
+                                      normalization.method = "SCT",
[…]

### 2025-02-17 01_Atlas_generation/Integration_epithelial_only/scripts/01b_run_normalization_split.R
Added code for integration of epithelial only cells

@@ -1,0 +1,57 @@
+####################################
+## 00. Load library and environment
+####################################
+
+# log
+log <- file(snakemake@log$log_file, open="wt")
+sink(log)
+sink(log, type="message")
+
+# image
+save.image(file = paste0(snakemake@log$log_object) )
+
+# load packages
+suppressPackageStartupMessages({
+  library(dplyr)
+  library(Seurat)
+  library(future)
+  library(RColorBrewer)
+})
+
+
+# load seurat object
+seurat_obj <- readRDS(snakemake@input[[1]])
+
+# Subset epithelial cells
+Idents(seurat_obj) <- seurat_obj@meta.data$AnnotationMain
+seurat_obj <- subset(seurat_obj, idents = "epithelial")
+
+#Remove samples with less than 100 cells, otherwise the subsequent integration will fail
+samples_cells <- as.data.frame(table(seurat_obj@meta.data[["sample"]])) 
+print(samples_cells)
+samples_integration <- samples_cells %>% filter(Freq > 99)
+samples_integration <- samples_integration$Var1
+print("samples used for integration")
+print(samples_integration)
+seurat_obj <- subset(seurat_obj, idents = samples_integration)
+
+####################################
+## 01. Normalization split per batch
+####################################
+
+regr.vars <- strsplit(snakemake@params[[1]], split='_')[[1]]
+
+# split by batch
+Idents(seurat_obj) <- seurat_obj@meta.data$sample
+seurat_obj <- SCTransform(seurat_obj,
+                          vars.to.regress = regr.vars,
+                          conserve.memory=FALSE,
+                          seed.use = 123,
+                          vst.flavor = "v2",
+                          return.only.var.genes=TRUE)
+
+seurat_obj_split <- SplitObject(seurat_obj, split.by = snakemake@params[[2]])
+
+saveRDS(seurat_obj_split, snakemake@output[[1]])
+sessionInfo()
+date()

### 2025-02-17 01_Atlas_generation/Integration_epithelial_only/scripts/02_run_seuratv3.R

@@ -1,0 +1,122 @@
+####################################
+## 00. Load library and environment
+####################################
+
+# log
+log <- file(snakemake@log$log_file, open="wt")
+sink(log)
+sink(log, type="message")
+
+# image
+save.image(file = paste0(snakemake@log$log_object) )
+
+# load packages
+suppressPackageStartupMessages({
+  library(dplyr)
+  library(Seurat)
+  library(future)
+  library(RColorBrewer)
+  library(remotes)
+  library(ggplot2)
+  library(SingleCellExperiment)
+})
+
+# load seurat object
+seurat_obj_split_input <- readRDS(snakemake@input[[1]])
+print("input loaded")
+#####filter hvgs
+features_integration <- SelectIntegrationFeatures(object.list = seurat_obj_split_input,
+                                                    nfeatures = 3000)
+print("integration features loaded")
+
+##Gene lists
+#Haemoglobin genes
+hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes 
+#mitochondrial genes
+mt.genes <- grep("^MT-", features_integration, value=TRUE)#mitochondrial genes
+#ribosomal genes
+rb.genes <- grep(pattern = "^RP[SL]", features_integration, value=TRUE)
+##Core stress response genes#https://github.com/kieranrcampbell/scrnaseq-digestion-paper/blob/master/data/deliverables/coregene_df-FALSE-v3.csv
+stress.genes <- read.csv("/home/common/data/output/projects/ENDO/E039/A012/Tan22databases/coregene_stress-response.csv")$gene_symbol
+##Cell cycle genes #Giotti, B. et al. Assembly of a parts list of the human mitotic cell cycle machinery. J. Mol. Cell. Biol. 11, 703718 (2019).
+CellCycleGenes <- read.csv("/home/common/data/output/projects/ENDO/E039/A012/Tan22databases/cell_cycle.csv")$Genes
+
+print("gene lists loaded")
+
+#filter hvgs
+suffix='hvgs_filtered'
+features_integration <- Reduce(setdiff, list(features_integration, hb.genes,mt.genes,rb.genes))
+
+print("hvgs loaded")
+
+
+# dir_out
+dir_out <- snakemake@config$integration_output_dir
+
+
+####################################
+## 01. Run integration
+####################################
+
+rpca_integration = function(seurat_object_list, snakemake, features_integration, verbose = TRUE){
+
+  # RPCA integration on SCT transformed counts
[…]

### 2025-02-19 01_Atlas_generation/Integration_epithelial_only/scripts/02_run_seuratv3.R
updated seed

@@ -101,7 +101,7 @@ integrated_data <- rpca_integration(seurat_obj_split_input, snakemake, features_
 DefaultAssay(integrated_data) <- "integrated"
 
 
-seed <- 122334
+seed <- 621
 
 # Perform linear dimensional reduction by PCA on HVGs
 integrated_data <- RunPCA(object = integrated_data,

### 2025-02-19 02_Cell_annotation/01_symphony.R
updated paths

@@ -1,0 +1,78 @@
+########################################################
+##### 00. Load packages and utils
+#######################################################
+
+# load packages
+suppressPackageStartupMessages({
+  source('/endometriosis_endometrium_scRNA_atlas/02_Cell_annotation/01_libs.R')# imports
+}) 
+
+#source utils
+source('/endometriosis_endometrium_scRNA_atlas/02_Cell_annotation/01_utils.R') # color definitions, symphony function and plotting functions
+
+
+
+########################################################
+##### 01. Run Symphony on entire EndoAtlas
+#######################################################
+
+#00 global annotation
+symphony(
+  CELLTYPEref = "Tan_global",
+  CELLTYPEquery = "ENDO_global"
+)
+
+
+########################################################
+##### 02. Save EndoAtlas split into the 5 main cell types
+#######################################################
+
+#Load integrated EndoAtlas
+data <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/ENDO_global.rds")
+Idents(data) <- data$AnnotationMain
+
+# Define the vector of annotations
+ANNOTATIONMAIN <- c("stromal", "epithelial", "endothelial", "lymphocyte", "myeloid")
+# Loop over each annotation type
+for (annotation in ANNOTATIONMAIN) {
+  # Subset data based on the current annotation
+  AnnotationMain_subset <- subset(data, idents = annotation)
+  # Save the subsetted data to an RDS file
+  saveRDS(AnnotationMain_subset, paste0("/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/", annotation, ".rds"))
+}
+
+
+########################################################
+##### 03. Run Symphony on 5 main cell types separately for refined annotation
+#######################################################
+
+
+#01 lymphocyte
+symphony(
+  CELLTYPEref = "Tan_lymphocyte",
+  CELLTYPEquery = "ENDO_lymphocyte"
+)
+
+#02 myeloid
+symphony(
+  CELLTYPEref = "Tan_myeloid",
+  CELLTYPEquery = "ENDO_myeloid"
+)
+
+#03 epithelial
+symphony(
+  CELLTYPEref = "Tan_epithelial",
+  CELLTYPEquery = "ENDO_epithelial"
+)
+
+#04 endothelial
+symphony(
+  CELLTYPEref = "Tan_endothelial",
[…]

### 2025-02-19 02_Cell_annotation/01_utils.R

@@ -37,7 +37,7 @@ symphony <- function(CELLTYPEref,
   ######################################################## 
   
   #create directory
-  dir_out <- paste0("/home/common/data/output/projects/ENDO/E044/A047/prolif_sec_",CELLTYPEref,"ref_",CELLTYPEquery,"query","/")
+  dir_out <- paste0("/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/",CELLTYPEref,"ref_",CELLTYPEquery,"query","/")
   dir.create(dir_out, recursive = TRUE)
   setwd(dir_out)
   
@@ -46,7 +46,7 @@ symphony <- function(CELLTYPEref,
   ########################################################
   
   #Load ref data set
-  cells_ref <- readRDS(paste0("/home/common/data/output/projects/ENDO/E039/A001/",CELLTYPEref,".rds"))
+  cells_ref <- readRDS(paste0("/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/",CELLTYPEref,".rds"))
   cells_ref@meta.data$dataset <- "ref"
   cells_ref <- NormalizeData(cells_ref, normalization.method = "LogNormalize", scale.factor = 10000, assay = "RNA")
   ref_exp_full <- cells_ref@assays$RNA@data
@@ -56,7 +56,7 @@ symphony <- function(CELLTYPEref,
   ref_metadata <- setnames(ref_metadata, old = c("subtypes", "phase"), new = c("cell_type", "batch"))
   
   #Load query data set
-  cells_query <- readRDS(paste0(dir_out,"/prolif_sec_",CELLTYPEquery,".rds"))
+  cells_query <- readRDS(paste0(dir_out,"/EndoAtlas_",CELLTYPEquery,".rds"))
   DefaultAssay(cells_query) <- "RNA"
   cells_query <- DietSeurat(cells_query, assay="RNA", dimreducs = "umap")
   cells_query@meta.data$dataset <- "query"

### 2025-02-19 02_Cell_annotation/02_MenstrualCycleSpecificCelltypes.R

@@ -13,7 +13,7 @@ library(ComplexUpset)
 library(ggrepel)
 library(PCAtools)
 library(scCustomize)
-dir_out <- "ENDO/E044/A074/"
+dir_out <- "/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/"
 dir.create(dir_out, recursive = TRUE)
 setwd(dir_out)
 
@@ -21,7 +21,7 @@ setwd(dir_out)
 #### 01. Load data and plot
 #######################
 ###Load data
-data <- readRDS("./ENDOatlas.rds")
+data <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/01_Atlas_generation/Integration/ENDOatlas.rds")
 
 ###Plotting for overview
 DimPlot_scCustom(data, reduction = "umap", group.by = "MenstrualCyclePhase")
@@ -175,7 +175,7 @@ table(meta_final$Symphony_Refined_Final) #sanity check
 #### 04. Menstrual cycle specific epithelial cells
 #######################
 ###Look into difference between glandular (proliferative), (glandular) mid-secretory, glandular early-secretory
-data <- readRDS("./epithelial_reintegrated.rds")
+data <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/01_Atlas_generation/Integration_epithelial_only/epithelial_reintegrated.rds")
 data <- AddMetaData(data, meta_final)
 DimPlot(data, split.by = "MenstrualCyclePhase", group.by = "Symphony_Refined_Final")
 #there are quite some glandular early-secretory already in proliferative/periovulatory phase and a lot in secretory_early. lumenal 2 are mainly in lumenal 2. mid-secretory are in secretory_mid/late, this matches

### 2025-02-19 02_Cell_annotation/symphony.R

@@ -1,51 +1,0 @@
-########################################################
-##### 00. Load packages and utils
-#######################################################
-
-# load packages
-suppressPackageStartupMessages({
-  source('./libs.R')# imports
-}) 
-
-#source utils
-source('./utils.R') # color definitions, symphony function and plotting functions
-
-########################################################
-##### 01. Run Symphony
-#######################################################
-
-#00 global
-symphony(
-  CELLTYPEref = "Tan_global",
-  CELLTYPEquery = "ENDO_global"
-)
-
-#01 lymphocyte
-symphony(
-  CELLTYPEref = "Tan_lymphocyte",
-  CELLTYPEquery = "ENDO_lymphocyte"
-)
-
-#02 myeloid
-symphony(
-  CELLTYPEref = "Tan_myeloid",
-  CELLTYPEquery = "ENDO_myeloid"
-)
-
-#03 epithelial
-symphony(
-  CELLTYPEref = "Tan_epithelial",
-  CELLTYPEquery = "ENDO_epithelial"
-)
-
-#04 endothelial
-symphony(
-  CELLTYPEref = "Tan_endothelial",
-  CELLTYPEquery = "ENDO_endothelial"
-)
-
-#05 stromal fibroblasts
-symphony(
-  CELLTYPEref = "Tan_stromal",
-  CELLTYPEquery = "ENDO_stromal"
-)

### 2025-02-19 03_PCA_analysis/PCAtools.R

@@ -9,7 +9,7 @@ library(Seurat)
 
 
 ##Set paths
-dir_ENDO <- "/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/"
+dir_ENDO <- "/endometriosis_endometrium_scRNA_atlas/"
 dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")
 dir.create(dir_out, recursive = TRUE)
 

### 2025-02-19 03_PCA_analysis/PCAtools_original_vignette.R

@@ -1,578 +1,0 @@
-## ----VROOM_CONNECTION_SIZE, eval = TRUE, echo = FALSE-------------------------
-
-Sys.setenv(VROOM_CONNECTION_SIZE='512000')
-
-
-## ----getPackage, eval = FALSE-------------------------------------------------
-#  
-#    if (!requireNamespace('BiocManager', quietly = TRUE))
-#      install.packages('BiocManager')
-#  
-#    BiocManager::install('PCAtools')
-#  
-
-## ----getPackageDevel, eval = FALSE--------------------------------------------
-#  
-#    if (!requireNamespace('remotes', quietly = TRUE))
-#      install.packages('remotes')
-#  
-#    remotes::install_github('kevinblighe/PCAtools')
-#  
-
-## ----Load, message = FALSE----------------------------------------------------
-
-library(PCAtools)
-
-
-## ----message = FALSE----------------------------------------------------------
-
-library(airway)
-library(magrittr)
-
-data('airway')
-airway$dex %<>% relevel('untrt')
-
-
-## ----message = FALSE----------------------------------------------------------
-
-ens <- rownames(airway)
-
-library(org.Hs.eg.db)
-symbols <- mapIds(org.Hs.eg.db, keys = ens,
-                  column = c('SYMBOL'), keytype = 'ENSEMBL')
-symbols <- symbols[!is.na(symbols)]
-symbols <- symbols[match(rownames(airway), names(symbols))]
-rownames(airway) <- symbols
-keep <- !is.na(rownames(airway))
-airway <- airway[keep,]
-
-
-## ----message = FALSE, warning = FALSE-----------------------------------------
-
-library('DESeq2')
-
-dds <- DESeqDataSet(airway, design = ~ cell + dex)
-dds <- DESeq(dds)
-vst <- assay(vst(dds))
-
-
-## -----------------------------------------------------------------------------
-
-p <- pca(vst, metadata = colData(airway), removeVar = 0.1)
-
-
-## ----ex1, warning = FALSE, fig.height = 7, fig.width = 6, fig.cap = 'Figure 1: A scree plot'----
-
-screeplot(p, axisLabSize = 18, titleLabSize = 22)
-
-
-## ----ex2a, eval = FALSE-------------------------------------------------------
-#  
-#    biplot(p)
-#  
-
[…]

### 2025-02-19 04_Time_trajectory_analysis/trajectoryAnalysis_monocle3.r

@@ -14,16 +14,14 @@ library(dplyr)
 library(ggplot2)
 
 ##Set paths
-workPath <- "/ENDO"
-analysisFolder <- paste0(workPath, "/ENDO/monocle3")
-out <- "epithelial"
-rawData <- paste0(workPath, "/raw_data/epithelial.rds")
+analysisFolder <- "endometriosis_endometrium_scRNA_atlas/_Data/04_Time_trajectory_analysis/"
+rawData <- "/endometriosis_endometrium_scRNA_atlas/_Data/01_Atlas_generation/Integration_epithelial_only/EndoAtlas_epithelial.rds"
 metadataFile <- ""
 subsetClusters <- c()  
 #-------------
 
 #Load data
-dir.create(paste0(analysisFolder, "/", out))
+dir.create(analysisFolder)
 
 data <- readRDS(rawData)
 

### 2025-02-19 Figures/PaperFiguresextended.Rmd
Added code for S6

@@
 
+#Supplementary Data Figure 6a, b
+#Need to change path
+```{r barplot distribution of cv splits, all cells, prolif}
+meta <- read.csv("/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")
+
+library(reshape2)
+library(ggplot2)
+library(forcats)
+
+#Supplementary Data Figure 6a
+# reshape df
+melted_meta <- melt(meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
+## create the bar plot for endo_EndometriosisGrade
+SFig6a <- ggplot(melted_meta, aes(x = variable, fill = endo_grade)) +
+  geom_bar(position = "stack", width = 0.7) +
+  facet_wrap(~ fct_rev(value), ncol = 2) +
+  labs(title = "Bar Plot of Samples by Endo EndometriosisGrade",
+       x = "Fold",
+       y = "Number of Samples") +
+  scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
+  theme_minimal() +
+  coord_flip()
+SFig6a
+ggsave(paste0(dir_out,"SFig6a_barplot.pdf"), plot = SFig6a, height = 6.3, width = 7.2)
+
+#Supplementary Data Figure 6b
+SFig6b <- ggplot(melted_meta, aes(x = variable, fill = lane)) +
+  geom_bar(position = "stack", width = 0.7) +
+  facet_wrap(~ fct_rev(value), ncol = 2) +
+  labs(title = "Bar Plot of Samples by lane",
+       x = "Fold",
+       y = "Number of Samples") +
+  #scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
+  theme_minimal() +
+  coord_flip()
+SFig6b
+ggsave(paste0(dir_out,"SFig6b_barplot.pdf"), plot = SFig6b, height = 6.3, width = 7.2)
+
+```
+
+Supplementary Data Figure 6c
+#Need to change path
+```{r}
+
+metadata <- read.csv('/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv')
+metadata$new_column <- NULL
+metadata$batch <- NULL
+
+# Remove 'names' column as it will be used as row labels
+row_names <- metadata$names
+metadata <- metadata[, -1]
+
+# Convert non-numeric columns to factors
+metadata[, c("EndometriosisStatus", "lane",  "endo_EndometriosisGrade",  "fold1", "fold2", "fold3", "fold4", "fold5")] <- lapply(metadata[, c("EndometriosisStatus", "lane", "endo_EndometriosisGrade", "fold1", "fold2", "fold3", "fold4", "fold5")], as.factor)
+
+# Create heatmap using ComplexHeatmap
+pdf(paste0(dir_out, "metadata_prolif_eval_fold_all_heatmap.pdf"), width = 7, height = 9)
+Heatmap(metadata,
+        name = "Metadata",
+        row_names_side = "left",
+        row_names_gp = gpar(fontsize = 8),
+        cluster_rows = FALSE,
+        cluster_columns = FALSE,
+        show_column_names = TRUE,
+        show_row_names = TRUE,
+        column_title = "Columns",
+        row_title = "Row Names",
+        heatmap_legend_param = list(title = "Legend")
+)
+dev.off()
+
+######################
+metadata <- read.csv('/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv')
+metadata$new_column <- NULL
+metadata$batch <- NULL
+
+# Remove 'names' column as it will be used as row labels
+row_names <- metadata$names
+metadata <- metadata[, -1]
+
+# Convert non-numeric columns to factors
+metadata[, c("EndometriosisStatus", "lane","endo_EndometriosisGrade",  "fold1", "fold2", "fold3", "fold4", "fold5")] <- lapply(metadata[, c("EndometriosisStatus", "lane", "endo_EndometriosisGrade", "fold1", "fold2", "fold3", "fold4", "fold5")], as.factor)
+
+# Convert remaining columns to a matrix
+mat <- data.matrix(metadata)
+
+# Create heatmap
+pdf(paste0(dir_out, "metadata_prolif_eval_fold_all_heatmap2.pdf"), width = 7, height = 9)
+heatmap.2(mat,
+          Rowv=FALSE, Colv=FALSE,
+          dendrogram="none",
+          trace="none",
+          col = heat.colors(256),
+          margins=c(8,10),
+          labRow=row_names,
+          scale="column",
+          key=TRUE,
+          keysize=1.5,
+          key.title="Color Key",
+          key.xlab="Value",
+          key.ylab=NULL,
+          density.info="none",
+          main="Heatmap of Metadata")
+dev.off()
+
+```
+
 ##################################################################################################################
-Scailyte part
+Scailyte part, extra code to delete
+
 #Supplementary Data Figure 6a, b
@@
 
-Supplementary Data Figure 6c
-```{r}
-
-metadata <- read.csv('/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv')
-metadata$new_column <- NULL
-metadata$batch <- NULL
-
-# Remove 'names' column as it will be used as row labels
-row_names <- metadata$names
-metadata <- metadata[, -1]
-
-# Convert non-numeric columns to factors
-metadata[, c("EndometriosisStatus", "lane",  "endo_EndometriosisGrade",  "fold1", "fold2", "fold3", "fold4", "fold5")] <- lapply(metadata[, c("EndometriosisStatus", "lane", "endo_EndometriosisGrade", "fold1", "fold2", "fold3", "fold4", "fold5")], as.factor)
-
-# Create heatmap using ComplexHeatmap
-pdf(paste0(dir_out, "metadata_prolif_eval_fold_all_heatmap.pdf"), width = 7, height = 9)
-Heatmap(metadata,
-        name = "Metadata",
-        row_names_side = "left",
-        row_names_gp = gpar(fontsize = 8),
-        cluster_rows = FALSE,
-        cluster_columns = FALSE,
-        show_column_names = TRUE,
-        show_row_names = TRUE,
-        column_title = "Columns",
-        row_title = "Row Names",
-        heatmap_legend_param = list(title = "Legend")
-)
-dev.off()
-
-######################
-metadata <- read.csv('/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv')
-metadata$new_column <- NULL
-metadata$batch <- NULL
-
-# Remove 'names' column as it will be used as row labels
-row_names <- metadata$names
-metadata <- metadata[, -1]
-
-# Convert non-numeric columns to factors
-metadata[, c("EndometriosisStatus", "lane","endo_EndometriosisGrade",  "fold1", "fold2", "fold3", "fold4", "fold5")] <- lapply(metadata[, c("EndometriosisStatus", "lane", "endo_EndometriosisGrade", "fold1", "fold2", "fold3", "fold4", "fold5")], as.factor)
-
-# Convert remaining columns to a matrix
-mat <- data.matrix(metadata)
-
-# Create heatmap
-pdf(paste0(dir_out, "metadata_prolif_eval_fold_all_heatmap2.pdf"), width = 7, height = 9)
-heatmap.2(mat,
-          Rowv=FALSE, Colv=FALSE,
-          dendrogram="none",
-          trace="none",
-          col = heat.colors(256),
-          margins=c(8,10),
-          labRow=row_names,
-          scale="column",
-          key=TRUE,
-          keysize=1.5,
-          key.title="Color Key",
-          key.xlab="Value",
-          key.ylab=NULL,
-          density.info="none",
-          main="Heatmap of Metadata")
-dev.off()
-
-```
-
 make DEG table for paper

### 2025-02-19 Figures/PaperFiguresextended.Rmd
updated heatmap S6c

@@
 )
-dev.off()
-
-######################
-metadata <- read.csv('/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv')
-metadata$new_column <- NULL
-metadata$batch <- NULL
-
-# Remove 'names' column as it will be used as row labels
-row_names <- metadata$names
-metadata <- metadata[, -1]
-
-# Convert non-numeric columns to factors
-metadata[, c("EndometriosisStatus", "lane","endo_EndometriosisGrade",  "fold1", "fold2", "fold3", "fold4", "fold5")] <- lapply(metadata[, c("EndometriosisStatus", "lane", "endo_EndometriosisGrade", "fold1", "fold2", "fold3", "fold4", "fold5")], as.factor)
-
-# Convert remaining columns to a matrix
-mat <- data.matrix(metadata)
-
-# Create heatmap
-pdf(paste0(dir_out, "metadata_prolif_eval_fold_all_heatmap2.pdf"), width = 7, height = 9)
-heatmap.2(mat,
-          Rowv=FALSE, Colv=FALSE,
-          dendrogram="none",
-          trace="none",
-          col = heat.colors(256),
-          margins=c(8,10),
-          labRow=row_names,
-          scale="column",
-          key=TRUE,
-          keysize=1.5,
-          key.title="Color Key",
-          key.xlab="Value",
-          key.ylab=NULL,
-          density.info="none",
-          main="Heatmap of Metadata")
 dev.off()

### 2025-02-19 Figures/PaperFiguresextended.Rmd
cleaned code

@@
 
-################?/???????
-#heatmap, not working yet
-
-# Extracting the necessary columns from the 'meta' dataframe
-data_matrix <- meta[, c("fold1", "fold2", "fold3", "fold4", "fold5")]
-data_matrix_numeric <- apply(data_matrix, 2, function(x) as.numeric(x == "eval"))
-
-# Adding annotation columns
-annotation_data <- meta[, c("EndometriosisStatus", "lane", "batch", "endo_EndometriosisGrade")]
-
-library(ComplexHeatmap)
-library(circlize)
-
-# Define a custom color palette
-my_color_palette <- colorRamp2(c(0, 0.5, 1), c("blue", "white", "red"))
-png(file=paste0(dir_out,'heatmap.pdf'))
-# Create the heatmap using ComplexHeatmap
-heatmap <- Heatmap(
-  data_matrix_numeric,
-  col = my_color_palette,
-  row_dend_reorder = TRUE,  # Adjust this based on your preference
-  column_title = "Folds",
-  name = "Fold Value",
-  show_row_names = FALSE,
-  show_column_names = TRUE,
-  cluster_columns = FALSE
-)
-
-draw(heatmap)
-dev.off()
 ```

### 2025-02-20 Figures/01_Generation_Figure4.R
add figure code and tables for Scailyte Figures

@@ -1,0 +1,92 @@
+# Aim : Code used for the generation of figure 4 
+# Author: Shaoline Sheppard
+
+## ---------------------------------- ##
+## Setup
+## ---------------------------------- ##
+
+# load librairies
+library(ggplot2)
+library(rlang)
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(cowplot)
+library(gridExtra)
+library(patchwork)
+library(tibble)
+library(tinter)
+
+# input path 
+input_path <- '../Tables' 
+
+# source common functions 
+source('./common_functions.R')
+
+## ---------------------------------- ##
+##  Figure 4.b
+## ---------------------------------- ##
+
+#1. Get plot for validation 
+#load performance summary table validation 
+validation_metrics <- read.csv(file.path(input_path,'Performance_all_cells_validation.csv'))
+validation_metrics$dimred <- factor(validation_metrics$dimred,levels = c('pca','corrgpca','hvg','diffgen'))
+validation_metrics_list <- split(validation_metrics, validation_metrics$dimred)
+list_plot_validation_auc <- lapply(validation_metrics_list,plot_boxplot,y_variable='auc')
+
+
+#2. Get plot for evaluation
+#load performance summary table evaluation  
+evaluation_metrics <- read.csv(file.path(input_path,'Performance_all_cells_evaluation.csv'))
+evaluation_metrics$dimred <- factor(evaluation_metrics$dimred,levels = c('pca','corrgpca','hvg','diffgen'))
+evaluation_metrics_list <- split(evaluation_metrics, evaluation_metrics$dimred)
+list_plot_evaluation_auc <- lapply(evaluation_metrics_list,plot_scatterplot,y_variable='auc')
+
+#3. Combine the 2 plots
+dimreds <- names(list_plot_validation_auc)
+for (i in 1:length(dimreds)){
+  list_plot_validation_auc[[i]]$title <- dimreds[i]
+}
+list_plot_validation_auc <- lapply(list_plot_validation_auc,function(x){x + ggtitle(x$title)})
+plot_grid(plotlist = c(list_plot_validation_auc,
+                       list_plot_evaluation_auc),nrow = 2)
+validation_metrics %>%
+  group_by(dimred) %>%
+  summarize(median(auc))
+## ---------------------------------- ##
+##  Figure 4.c
+## ---------------------------------- ##
+
[…]

### 2025-02-20 Figures/02_Generation_Supplementary_Figure6.R

@@ -1,0 +1,79 @@
+## ---------------------------------- ##
+##  Setup
+## ---------------------------------- ##
+
+# load librairies
+library(ggplot2)
+library(rlang)
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(cowplot)
+library(gridExtra)
+library(patchwork)
+library(tibble)
+library(tinter)
+library(UpSetR)
+library(forcats)
+
+# input path 
+input_path<- '../Tables' 
+  
+# source common functions 
+source('./common_functions.R')
+
+## ---------------------------------- ##
+##  Figure 6.a
+## ---------------------------------- ##
+
+## to do lea 
+
+## ---------------------------------- ##
+##  Figure 6.b
+## ---------------------------------- ##
+
+## to do lea 
+
+## ---------------------------------- ##
+##  Figure 6.c
+## ---------------------------------- ##
+
+## to do lea 
+
+## ---------------------------------- ##
+##  Figure 6.d
+## ---------------------------------- ##
+
+output_proba_best_learners_corrgpca <- read.csv(file.path(input_path,'output_probabilities_sample_corrgpca.csv'))
+output_proba_best_learners_corrgpca <- output_proba_best_learners_corrgpca %>%
+  mutate(grade = recode(grade, "CTL" = "non-ENDO", 
+                        "severe" = "severe-ENDO", 
+                        "mild" = "mild-ENDO"))
+plot_violin(data = output_proba_best_learners_corrgpca,
+            y_variable='median_output_probability')
+## ---------------------------------- ##
+##  Figure 6.e
+## ---------------------------------- ##
+
+pca_signature <- read.csv(file.path(input_path,'SupplementaryTable6_PCAsignature.csv')) %>%
+  dplyr::pull(gene)
+corrgpca_signature <- read.csv(file.path(input_path,'SupplementaryTable6_CorrgPCA_signature.csv')) %>%
+  dplyr::pull(gene)
+
+list_signatures <- list(CorrgPCA = corrgpca_signature,
+                        PCA = pca_signature
+                        )
+upset(fromList(list_signatures), order.by = "freq",text.scale = 2)
+
+## ---------------------------------- ##
+##  Figure 6.f
+## ---------------------------------- ##
+
+## Predictive score for each of the genes of the 11 gene signature 
[…]

### 2025-02-20 Figures/04_generate_summary_learners_tables.R

@@ -1,0 +1,77 @@
+## ---------------------------------- ##
+##  Setup
+## ---------------------------------- ##
+
+# load librairies
+library(ggplot2)
+library(rlang)
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(cowplot)
+library(gridExtra)
+library(patchwork)
+library(tibble)
+library(tinter)
+library(Seurat)
+
+# input path 
+input_path <- '../Tables'
+
+## ---------------------------------- ##
+##  Generate summary tables 
+## ---------------------------------- ##
+
+read_in_tables <- function(file_path,learner_name='learner_uid'){ 
+  data <- read.csv(file_path)
+  data %>%
+    mutate(cvsplit = gsub(".*(CVsplit\\d+).*", "\\1", !!sym(learner_name)),
+           learner_id = gsub(".*(learner\\d+).*", "\\1", !!sym(learner_name))) %>%
+    select(!c(!!sym(learner_name)))
+}
+
+
+performance_11_genes <- read_in_tables(file.path(input_path,'Performance_11_gene_signature.csv')) %>%
+  dplyr::mutate(dimred = '11_genes_signature',
+         cell_selection = 'all_cells',
+         Model_training_phase = 'validation')
+
+performance_allcells_signature_confirmation <- read_in_tables(file.path(input_path,'Performance_all_cells_signature_confirmation.csv')) %>%
+  dplyr::mutate(dimred = paste0('signature_confirmation_',dimred),
+         Model_training_phase = 'validation',
+         cell_selection = 'all_cells')
+
+performance_cf_signature_confirmation <- read_in_tables(file.path(input_path,'Performance_cycling_fibroblasts_signature_confirmation.csv')) %>%
+  dplyr::mutate(cell_selection = 'cycling_fibroblasts',
+         Model_training_phase = 'validation',
+         dimred = paste0('signature_confirmation_',dimred))
+
+performance_all_cells_validation <- read.csv(file.path(input_path,'Performance_all_cells_validation.csv')) %>%
+  dplyr::mutate(cell_selection = 'all_cells',
+         Model_training_phase = 'validation') %>%
+  dplyr::rename('learner_id'='learner')
+
+performance_all_cells_evaluation <- read_in_tables(file.path(input_path,'Performance_all_cells_evaluation.csv'),learner_name='learner') %>%
+  dplyr::mutate(cell_selection = 'all_cells',
+         Model_training_phase = 'evaluation')
+
[…]

### 2025-02-20 Figures/04_generate_summary_learners_tables.R
remove script not meant to be added

@@ -1,77 +1,0 @@
-## ---------------------------------- ##
-##  Setup
-## ---------------------------------- ##
-
-# load librairies
-library(ggplot2)
-library(rlang)
-library(stringr)
-library(dplyr)
-library(tidyr)
-library(cowplot)
-library(gridExtra)
-library(patchwork)
-library(tibble)
-library(tinter)
-library(Seurat)
-
-# input path 
-input_path <- '../Tables'
-
-## ---------------------------------- ##
-##  Generate summary tables 
-## ---------------------------------- ##
-
-read_in_tables <- function(file_path,learner_name='learner_uid'){ 
-  data <- read.csv(file_path)
-  data %>%
-    mutate(cvsplit = gsub(".*(CVsplit\\d+).*", "\\1", !!sym(learner_name)),
-           learner_id = gsub(".*(learner\\d+).*", "\\1", !!sym(learner_name))) %>%
-    select(!c(!!sym(learner_name)))
-}
-
-
-performance_11_genes <- read_in_tables(file.path(input_path,'Performance_11_gene_signature.csv')) %>%
-  dplyr::mutate(dimred = '11_genes_signature',
-         cell_selection = 'all_cells',
-         Model_training_phase = 'validation')
-
-performance_allcells_signature_confirmation <- read_in_tables(file.path(input_path,'Performance_all_cells_signature_confirmation.csv')) %>%
-  dplyr::mutate(dimred = paste0('signature_confirmation_',dimred),
-         Model_training_phase = 'validation',
-         cell_selection = 'all_cells')
-
-performance_cf_signature_confirmation <- read_in_tables(file.path(input_path,'Performance_cycling_fibroblasts_signature_confirmation.csv')) %>%
-  dplyr::mutate(cell_selection = 'cycling_fibroblasts',
-         Model_training_phase = 'validation',
-         dimred = paste0('signature_confirmation_',dimred))
-
-performance_all_cells_validation <- read.csv(file.path(input_path,'Performance_all_cells_validation.csv')) %>%
-  dplyr::mutate(cell_selection = 'all_cells',
-         Model_training_phase = 'validation') %>%
-  dplyr::rename('learner_id'='learner')
-
-performance_all_cells_evaluation <- read_in_tables(file.path(input_path,'Performance_all_cells_evaluation.csv'),learner_name='learner') %>%
-  dplyr::mutate(cell_selection = 'all_cells',
-         Model_training_phase = 'evaluation')
-
[…]

### 2025-02-22 Figures/01_Generation_Figure4.R
add code for figure 4e

@@ -80,8 +80,28 @@ go_enrichment_plot(go_results_all_cells)
 ##  Figure 4.e
 ## ---------------------------------- ##
 
-## predictive cells enrichment 
+df <- read.csv(file = file.path(input_path,'cell_ranking_figure4e.csv'))
 
+# define the order in which the x axis variable are displayed on the plot
+df$Symphony_Refined <- factor(df$Symphony_Refined,
+                                                 levels = c('overall','VSMC',
+                                                            'Prv-CCL19','Prv-MYH11','Prv-STEAP4',
+                                                            'ciliated',
+                                                            "M$\\Phi$1-LYVE1","M$\\Phi$3-APOE","M$\\Phi$5-activated"))
+
+# Create a stacked bar plot for the filtered data, showing percentage of TRUE and FALSE for 'median_rank_control_logic01'
+Fig4e <- ggplot(df, aes(x = Symphony_Refined, 
+                                           y = percentage, 
+                                           fill = as.factor(median_rank_control_logic01))) +
+  geom_bar(stat = "identity", position = "fill") +  # Stack the bars to show proportions
+  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Format y-axis as percentages
+  scale_fill_manual(values=c("TRUE"='navyblue',"FALSE"='grey'),
+                    labels = c('TRUE'='top 90%','FALSE'='lower 10%')) +
+  geom_hline(yintercept = 0.1,linewidth = 1.5)+
+  theme_minimal() +  # Use minimal theme
+  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  # Rotate x-axis labels for readability
+  labs(x = "", y = "", fill = "Median cell rank")  # Label the plot
+Fig4e
 
 ## ---------------------------------- ##
 ##  Figure 4.f

### 2025-02-23 05a_DEG_analysis_muscat/job_env.sh
replaced code with muscat pipeline

@@ -1,0 +1,2 @@
+NAMESPACE=endo
+PIPELINE=endo

### 2025-02-23 05a_DEG_analysis_muscat/run_snakemake.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+snakemake --profile /endometriosis_endometrium_scRNA_atlas/05a_DEG_analysis_muscat/config/profile 

### 2025-02-23 05a_DEG_analysis_muscat/scripts/DGE_muscat_prolif_stringent.R

@@ -1,0 +1,72 @@
+####################################
+## 00. Load library and environment
+####################################
+
+
+suppressPackageStartupMessages({
+library(ggplot2)
+library(dplyr)
+library(limma)
+library(muscat)
+library(purrr)
+library(Seurat)
+library(scater)
+library(SingleCellExperiment)
+library(Matrix)
+})
+
+####################################
+## 01. Prepare Data
+####################################
+
+#load seurat object 
+data <- readRDS(snakemake@input[[1]])
+print("loaded data")
+
+#Filter for proliferative samples with stringent exclusion criteria
+Idents(data) <- data$DEG.analysis.Figure.3a.b
+data <- subset(data, idents = "TRUE")
+
+
+
+# List of annotations to loop through
+ANNOTATIONS <- c("AnnotationMain", "AnnotationRefined", "AnnotationUnited")
+
+# Loop through each annotation
+for (Annotation in ANNOTATIONS) {
+
+	#make SingleCellExperiment from RNA assay
+	sce <- as.SingleCellExperiment(data, assay = "RNA")
+
+	# remove undetected genes
+	sce <- sce[rowSums(counts(sce) > 0) > 0, ]
+
+	# remove lowly expressed genes (less than 10 counts)
+	sce <- sce[rowSums(counts(sce) > 1) >= 10, ]
+
+	##Data preparation
+	sce$id <- paste0(sce$treatment, sce$sample)
+
+	(sce <- prepSCE(sce,
+					kid = Annotation, # subpopulation assignments
+					gid = "condition",  # group IDs (ENDO/non-ENDO)
+					sid = "sample",   # sample IDs 
+					drop = FALSE))  
+
+	nk <- length(kids <- levels(sce$cluster_id))
+	ns <- length(sids <- levels(sce$sample_id))
+	names(kids) <- kids; names(sids) <- sids
+
+	####################################
+	## 02. Cell-level analysis: Mixed models
+	####################################
+
+	print("start DE testing:")
+		   mm_dream <- mmDS(sce, method = "dream", coef = "group_idEndo",
+					 n_cells = 50, n_samples = 3, covs = "batch")
+
+	saveRDS(mm_dream, paste0("./mm_dream",Annotation,".csv"))
+}
+
+sessionInfo()
+date()

### 2025-02-23 05a_DEG_analysis_muscat/DGE_muscat.R
Added secretory DEG

@@ -1,0 +1,37 @@
+########################################################
+##### 00. Load packages and utils
+#######################################################
+
+# load packages
+suppressPackageStartupMessages({
+  source('/endometriosis_endometrium_scRNA_atlas/02_Cell_annotation/01_libs.R')# imports
+}) 
+
+#source utils
+source('/endometriosis_endometrium_scRNA_atlas/02_Cell_annotation/01_utils.R') # color definitions, symphony function and plotting functions
+
+
+
+########################################################
+##### 01. Run DEG analysis with muscat 
+#######################################################
+
+##Proliferative Phase
+#Proliferative phase samples with strict exclusion criteria, AnnotationMain
+DEGanalysis_muscat(MenstrualCyclePhase = "Proliferative", Annotation = "AnnotationMain")
+  
+#Proliferative phase samples with strict exclusion criteria, AnnotationRefined
+DEGanalysis_muscat(MenstrualCyclePhase = "Proliferative", Annotation = "AnnotationRefined")
+
+#Proliferative phase samples with strict exclusion criteria, AnnotationUnited
+DEGanalysis_muscat(MenstrualCyclePhase = "Proliferative", Annotation = "AnnotationUnited")
+
+##Secretory Phase
+#Secretory phase samples with strict exclusion criteria, AnnotationMain
+DEGanalysis_muscat(MenstrualCyclePhase = "Secretory", Annotation = "AnnotationMain")
+
+#Secretory phase samples with strict exclusion criteria, AnnotationRefined
+DEGanalysis_muscat(MenstrualCyclePhase = "Secretory", Annotation = "AnnotationRefined")
+
+#Secretory phase samples with strict exclusion criteria, AnnotationUnited
+DEGanalysis_muscat(MenstrualCyclePhase = "Secretory", Annotation = "AnnotationUnited")

### 2025-02-23 05a_DEG_analysis_muscat/libs.R

@@ -1,0 +1,11 @@
+# Analysis
+library(ggplot2)
+library(dplyr)
+library(limma)
+library(muscat)
+library(purrr)
+library(Seurat)
+library(scater)
+library(SingleCellExperiment)
+library(Matrix)
+

### 2025-02-23 05a_DEG_analysis_muscat/utils.R

@@ -1,0 +1,105 @@
+####################################
+## 00. Set path
+####################################
+
+##Set paths
+dir_out <- paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/")
+dir.create(dir_out, recursive = TRUE, showWarnings = FALSE)
+
+
+DEGanalysis_muscat <- function(MenstrualCyclePhase, Annotation) {
+
+####################################
+## 01. Prepare Data
+####################################
+
+##load seurat object 
+data <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas.rds")
+print("loaded data")
+
+##Subset samples by menstrual cycle phase
+Idents(data) <- data$sample
+
+# Proliferative phase
+if (MenstrualCyclePhase == "Proliferative") {
+  proliferativeDEG_samples <- data$DEG.analysis.Figure.3a.b
+  # Alternatively, manually specify the samples
+  # proliferativeDEG_samples <- c("sample76", "sample83", "sample84", "sample91", "sample94", 
+  #                               "sample106", "sample111", "sample125", "sample126", "sample130", 
+  #                               "sample134", "sample135", "sample70", "sample71", "sample73", 
+  #                               "sample82", "sample95", "sample96", "sample110", "sample112", 
+  #                               "sample114", "sample116", "sample123")
+  
+  data <- subset(data, idents = proliferativeDEG_samples)
+}
+
+# Secretory phase
+if (MenstrualCyclePhase == "Secretory") {
+  #Subset of balanced secretory phase samples in terms of EndometriosisStatus, CycleDay and MenstrualCyclePhase_Refined
+  secretoryDEG_samples <- c("sample133", "sample113", "sample117", "sample88", 
+                            "sample128", "sample132", "sample127", "sample75")
+  data <- subset(data, idents = secretoryDEG_samples)
+}
+
+##Rename required Annotation to 'Annotation'
+if (MenstrualCyclePhase == "AnnotationMain") {
+  data$Annotation <- data$AnnotationMain
+}
+if (MenstrualCyclePhase == "AnnotationRefined") {
+  data$Annotation <- data$AnnotationRefined
+}
+if (MenstrualCyclePhase == "AnnotationUnited") {
[…]

### 2025-02-23 Figures/PaperFiguresextended.Rmd
Complemented Figure script

@@
 
-ln -s /home/common/data/output/projects/ENDO/E044/A033/EndoAtlas_GEO_meta_complete_downsampled.rds EndoAtlas.rds
-EndoAtlas_meta <- readRDS(paste0(dir_data,"EndoAtlas_meta.rds"))
-EndoAtlas <- readRDS(paste0(dir_data,"EndoAtlas.rds"))
-EndoAtlas <- AddMetaData(EndoAtlas, EndoAtlas_meta)
-saveRDS(EndoAtlas,paste0(dir_data,"EndoAtlas.rds") )
-
-Add additional data to the Seurat metadata and reformat
-```{r plotting of cell types per sample}
-#Set directory links and create directories
-dir_ENDO <- "/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/"
-
-dir_data <- paste0(dir_ENDO,"_Data/")
-dir_out <- paste0(dir_ENDO,"Figures_out/")
-if (!dir.exists(dir_out)) {
-  dir.create(dir_out, recursive = TRUE)
-}
-
-#Load data
-EndoAtlas <- readRDS(paste0(dir_data, "EndoAtlas.rds"))
-EndoAtlas_meta <- EndoAtlas@meta.data
-
-#Preserve cell names
-EndoAtlas_meta$cell_names <- rownames(EndoAtlas_meta)
-
-##Modify menstrual cycle information
-# Create column MenstrualCyclePhase_main based on MenstrualCyclePhase by merging the secretory sub-phases from 
-EndoAtlas_meta$MenstrualCyclePhase_main <- as.character(EndoAtlas_meta$MenstrualCyclePhase)
-EndoAtlas_meta$MenstrualCyclePhase_main[EndoAtlas_meta$MenstrualCyclePhase_main %in% c("secretory_early", "secretory_mid", "secretory_late")] <- "secretory"
-#Factor samples by MenstrualCyclePhase_main for plotting
-EndoAtlas_meta$MenstrualCyclePhase_main <- factor(EndoAtlas_meta$MenstrualCyclePhase_main, levels = c("proliferative","periovulatory","secretory"))
-
-##Add additional metadata from manuscript Supplementary Table 1
-# Read the tab-delimited text file
-SupplementaryTable1 <- read.delim(paste0(dir_data, "SupplementaryTable1.txt"), header = TRUE, stringsAsFactors = FALSE)
-# Perform a left join of the `meta` dataframe with `SupplementaryTable1`
-EndoAtlas_meta <- dplyr::left_join(
-  EndoAtlas_meta,
-  SupplementaryTable1 %>% 
-    dplyr::select(
-      Patient.sample.ID,
-      epithelial.pseudotime,
-      sequencing.batch,
-      X10x.capturing.batch,
-      Minor.exclusion.criteria.logical,
-      Menstrual.cycle.phase.marker.analysis,
-      Cell.type.frequency.analysis..Figure.2e.f.,
-      DEG.analysis..Figure.3a.b.,
-      Ligand.receptor.analysis..Figure.3c.h.
-    ),
-  by = c("sample" = "Patient.sample.ID")
-)
-
-##Determine order of samples by MenstrualCyclePhase and epithelial.pseudotime
-MenstrualCyclePhase_sampleOrder <- EndoAtlas_meta%>%
-  select(sample, MenstrualCyclePhase,epithelial.pseudotime) %>%
-  distinct() %>%
-  arrange(MenstrualCyclePhase, epithelial.pseudotime) %>%
-  mutate(MenstrualCyclePhase_sampleOrder = row_number()) %>%
-  ungroup()
-#Factor samples by MenstrualCyclePhase and epithelial.pseudotime for plotting
-EndoAtlas_meta$sample <- factor(EndoAtlas_meta$sample, levels = c(MenstrualCyclePhase_sampleOrder$sample))
-
-#Add cell names as row names
-rownames(EndoAtlas_meta) <- EndoAtlas_meta$cell_names
-
-#Save EndoAtlas_meta as .rds
-saveRDS(EndoAtlas_meta, paste0(dir_data, "EndoAtlas_meta.rds"))
-
-#EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))
-
-#Save color pals
-qual_col_pals <- brewer.pal.info[brewer.pal.info$category == "qual", ]
-col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
-```
-
-Save epithelial object and upload to GEO.
-```{r}
-#Load monocl object
-data_monocl <- readRDS("/home/common/data/output/projects/ENDO/E044/A062/trajectory_day5-6-7_greenCircle/trajectory_timepoint0.rds")
-listData <- (data_monocl@colData@listData)
-listData <- listData["cell_ident"]
-
-EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))
-colnames(EndoAtlas_meta)
-
-########################
-# Extract cell.ident from listData
-cell_ids <- listData[["cell_ident"]]
-
-# Ensure cell_ids is not empty
-if (length(cell_ids) == 0) {
-  stop("cell.ident in listData is empty. Ensure it is populated before proceeding.")
-}
-
-# Filter and order EndoAtlas_meta by cell_ids
-filtered_meta <- EndoAtlas_meta[EndoAtlas_meta$cell_names %in% cell_ids, ]
-ordered_meta <- filtered_meta[match(cell_ids, filtered_meta$cell_names), ]
-
-# Verify alignment
-if (!all(ordered_meta$cell_names == cell_ids)) {
-  stop("Ordering of filtered EndoAtlas_meta does not match cell_ids.")
-}
-
-# Add columns from ordered_meta to listData
-columns_to_add <- c("sample", "EndometriosisStatus", "EndometriosisGrade", "CycleDay", 
-                    "ProgesteroneSerum", "MenstrualCyclePhase", "AnnotationMain", 
-                    "AnnotationRefined", "cell_names", "MenstrualCyclePhase_main")
-
-for (col in columns_to_add) {
-  if (col %in% colnames(ordered_meta)) {
-    listData[[col]] <- ordered_meta[[col]]
-  } else {
-    warning(paste("Column", col, "not found in EndoAtlas_meta. Skipping."))
-  }
-}
-
-# Update the main object
-data_monocl@colData@listData <- listData
-
-data_monocl@principal_graph_aux@listData[["UMAP"]][["pseudotime"]]
-
-saveRDS(data_monocl, paste0(dir_data, "epithelial_cells_monocle3.rds"))
-```
-
 ###########################################
@@
 DimPlot(data_ds, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
-
-```
-
-Figure 4e
-```{r Barplots for  filter response median rank E042A027}
-#Load metadata
-meta_final <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
-
-#Set variables to loop over
-DIMRED <- c("corrgpca","pca")
-ANNOTATION <- c("Symphony_Refined","AnnotationMain")
-MEDIAN <- c("median_rank_endo_logic003","median_rank_control_logic003","median_rank_endo_logic005","median_rank_control_logic005","median_rank_endo_logic01","median_rank_control_logic01")
-
-for (dim in DIMRED) {
-#Read filter ranking
-average_filter_ranking <- readRDS(paste0("/home/common/data/output/projects/ENDO/E042/A027/response_cellcnn_manually/all_cells/",dim,"_average_filter_ranking.rds"))
-
-#Combine tables
-meta <- left_join(meta_final, average_filter_ranking %>% select(cell_ident, median_rank_endo, median_rank_control), by = "cell_ident")
-
-#Exclude NA
-metaNAfree <- meta %>%
-  select(median_rank_endo, median_rank_control,Symphony_Refined,AnnotationMain) %>%
-  na.omit() %>%
-  filter(Symphony_Refined != "granulocyte")
-
-endo_threshold003 <- quantile(metaNAfree$median_rank_endo, 0.03)
-control_threshold003 <- quantile(metaNAfree$median_rank_control, 0.03)
-endo_threshold005 <- quantile(metaNAfree$median_rank_endo, 0.05)
-control_threshold005 <- quantile(metaNAfree$median_rank_control, 0.05)
-endo_threshold01 <- quantile(metaNAfree$median_rank_endo, 0.1)
-control_threshold01 <- quantile(metaNAfree$median_rank_control, 0.1)
-
-# Create the new column with ranking logic threshold
-metaNAfree <- metaNAfree %>%
-  mutate(median_rank_endo_logic003 = median_rank_endo <= endo_threshold003) %>%
-  mutate(median_rank_control_logic003 = median_rank_control <= control_threshold003) %>%
-  mutate(median_rank_endo_logic005 = median_rank_endo <= endo_threshold005) %>%
-  mutate(median_rank_control_logic005 = median_rank_control <= control_threshold005) %>%
-  mutate(median_rank_endo_logic01 = median_rank_endo <= endo_threshold01) %>%
-  mutate(median_rank_control_logic01 = median_rank_control <= control_threshold01)
-
-# Create bar plots
-dir.create(paste0(dir_out,"median_rank_plots/"))
-
-for (ann in ANNOTATION) {
-  for (med in MEDIAN) {
-    # Create plot
-    plot1 <- ggplot(metaNAfree, aes_string(x = ann)) +
-      geom_bar(position = "identity") +
-      scale_y_continuous(trans = "log10") +
-      theme_minimal() +
-      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
-      ggtitle(paste0(dim, "___",ann, "___", med))
-    plot2 <- ggplot(metaNAfree, aes_string(x = ann, fill = med)) +
-      geom_bar(position = "fill") +
-      theme_minimal() +
-      theme(axis.text.x = element_text(angle = 90, hjust = 1))
-    combined_plot <- plot1 / plot2
-    # Save plot
-    ggsave(filename = paste0(dir_out, "/median_rank_plots/", dim, "___",ann, "___", med, ".pdf"), plot = combined_plot, width = 13, height = 10)
-  }
-}
-}
 

### 2025-02-23 Figures/Preprocessing_metadata.Rmd
Complemented Figure script

```{r setup}
library(Seurat)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(cowplot)
library(ggbreak) 
library(tidyr)
library(patchwork)
library(ComplexUpset)
library(ggrepel)
library(Polychrome)
library(ComplexHeatmap)
library(gplots)
library(RColorBrewer)
library(randomcoloR)

```

ln -s /home/common/data/output/projects/ENDO/E044/A033/EndoAtlas_GEO_meta_complete_downsampled.rds EndoAtlas.rds
EndoAtlas_meta <- readRDS(paste0(dir_data,"EndoAtlas_meta.rds"))
EndoAtlas <- readRDS(paste0(dir_data,"EndoAtlas.rds"))
EndoAtlas <- AddMetaData(EndoAtlas, EndoAtlas_meta)
saveRDS(EndoAtlas,paste0(dir_data,"EndoAtlas.rds") )

Add additional data to the Seurat metadata and reformat
```{r plotting of cell types per sample}
#Set directory links and create directories
dir_ENDO <- "/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/"

dir_data <- paste0(dir_ENDO,"_Data/")
dir_out <- paste0(dir_ENDO,"Figures_out/")
if (!dir.exists(dir_out)) {
  dir.create(dir_out, recursive = TRUE)
}

#Load data
EndoAtlas <- readRDS(paste0(dir_data, "EndoAtlas.rds"))
EndoAtlas_meta <- EndoAtlas@meta.data

#Preserve cell names
EndoAtlas_meta$cell_names <- rownames(EndoAtlas_meta)

##Modify menstrual cycle information
# Create column MenstrualCyclePhase_main based on MenstrualCyclePhase by merging the secretory sub-phases from 
EndoAtlas_meta$MenstrualCyclePhase_main <- as.character(EndoAtlas_meta$MenstrualCyclePhase)
EndoAtlas_meta$MenstrualCyclePhase_main[EndoAtlas_meta$MenstrualCyclePhase_main %in% c("secretory_early", "secretory_mid", "secretory_late")] <- "secretory"
#Factor samples by MenstrualCyclePhase_main for plotting
EndoAtlas_meta$MenstrualCyclePhase_main <- factor(EndoAtlas_meta$MenstrualCyclePhase_main, levels = c("proliferative","periovulatory","secretory"))

##Add additional metadata from manuscript Supplementary Table 1
# Read the tab-delimited text file
SupplementaryTable1 <- read.delim(paste0(dir_data, "SupplementaryTable1.txt"), header = TRUE, stringsAsFactors = FALSE)
# Perform a left join of the `meta` dataframe with `SupplementaryTable1`
EndoAtlas_meta <- dplyr::left_join(
  EndoAtlas_meta,
  SupplementaryTable1 %>% 
    dplyr::select(
      Patient.sample.ID,
      epithelial.pseudotime,
      sequencing.batch,
      X10x.capturing.batch,
      Minor.exclusion.criteria.logical,
      Menstrual.cycle.phase.marker.analysis,
      Cell.type.frequency.analysis..Figure.2e.f.,
      DEG.analysis..Figure.3a.b.,
      Ligand.receptor.analysis..Figure.3c.h.
    ),
  by = c("sample" = "Patient.sample.ID")
)

##Determine order of samples by MenstrualCyclePhase and epithelial.pseudotime
MenstrualCyclePhase_sampleOrder <- EndoAtlas_meta%>%
  select(sample, MenstrualCyclePhase,epithelial.pseudotime) %>%
  distinct() %>%
  arrange(MenstrualCyclePhase, epithelial.pseudotime) %>%
  mutate(MenstrualCyclePhase_sampleOrder = row_number()) %>%
  ungroup()
#Factor samples by MenstrualCyclePhase and epithelial.pseudotime for plotting
EndoAtlas_meta$sample <- factor(EndoAtlas_meta$sample, levels = c(MenstrualCyclePhase_sampleOrder$sample))

#Add cell names as row names
rownames(EndoAtlas_meta) <- EndoAtlas_meta$cell_names

#Save EndoAtlas_meta as .rds
saveRDS(EndoAtlas_meta, paste0(dir_data, "EndoAtlas_meta.rds"))

#EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))

#Save color pals
qual_col_pals <- brewer.pal.info[brewer.pal.info$category == "qual", ]
col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
```

Save epithelial object and upload to GEO.
```{r}
#Load monocl object
data_monocl <- readRDS("/home/common/data/output/projects/ENDO/E044/A062/trajectory_day5-6-7_greenCircle/trajectory_timepoint0.rds")
listData <- (data_monocl@colData@listData)
listData <- listData["cell_ident"]

EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))
colnames(EndoAtlas_meta)

########################
# Extract cell.ident from listData
cell_ids <- listData[["cell_ident"]]

# Ensure cell_ids is not empty
if (length(cell_ids) == 0) {
  stop("cell.ident in listData is empty. Ensure it is populated before proceeding.")
}

# Filter and order EndoAtlas_meta by cell_ids
filtered_meta <- EndoAtlas_meta[EndoAtlas_meta$cell_names %in% cell_ids, ]
ordered_meta <- filtered_meta[match(cell_ids, filtered_meta$cell_names), ]

# Verify alignment
if (!all(ordered_meta$cell_names == cell_ids)) {
  stop("Ordering of filtered EndoAtlas_meta does not match cell_ids.")
}

# Add columns from ordered_meta to listData
columns_to_add <- c("sample", "EndometriosisStatus", "EndometriosisGrade", "CycleDay", 
                    "ProgesteroneSerum", "MenstrualCyclePhase", "AnnotationMain", 
                    "AnnotationRefined", "cell_names", "MenstrualCyclePhase_main")

for (col in columns_to_add) {
  if (col %in% colnames(ordered_meta)) {
    listData[[col]] <- ordered_meta[[col]]
  } else {
    warning(paste("Column", col, "not found in EndoAtlas_meta. Skipping."))
  }
}

# Update the main object
data_monocl@colData@listData <- listData

data_monocl@principal_graph_aux@listData[["UMAP"]][["pseudotime"]]

saveRDS(data_monocl, paste0(dir_data, "epithelial_cells_monocle3.rds"))
```

### 2025-02-23 05a_DEG_analysis_muscat/DGE_muscat.R
corrected paths

@@ -4,11 +4,11 @@
 
 # load packages
 suppressPackageStartupMessages({
-  source('/endometriosis_endometrium_scRNA_atlas/02_Cell_annotation/01_libs.R')# imports
+  source('/endometriosis_endometrium_scRNA_atlas/05a_DEG_analysis_muscat/01_libs.R')# imports
 }) 
 
 #source utils
-source('/endometriosis_endometrium_scRNA_atlas/02_Cell_annotation/01_utils.R') # color definitions, symphony function and plotting functions
+source('/endometriosis_endometrium_scRNA_atlas/05a_DEG_analysis_muscat/01_utils.R') # color definitions, symphony function and plotting functions
 
 
 

### 2025-02-23 05a_DEG_analysis_muscat/DGE_muscat.R
deleted unnecessary comment

@@ -4,11 +4,11 @@
 
 # load packages
 suppressPackageStartupMessages({
-  source('/endometriosis_endometrium_scRNA_atlas/05a_DEG_analysis_muscat/01_libs.R')# imports
+  source('/endometriosis_endometrium_scRNA_atlas/05a_DEG_analysis_muscat/01_libs.R')
 }) 
 
 #source utils
-source('/endometriosis_endometrium_scRNA_atlas/05a_DEG_analysis_muscat/01_utils.R') # color definitions, symphony function and plotting functions
+source('/endometriosis_endometrium_scRNA_atlas/05a_DEG_analysis_muscat/01_utils.R') 
 
 
 

### 2025-02-23 05a_DEG_analysis_muscat/DGE_muscat.R
formatting of code

@@ -26,6 +26,8 @@ DEGanalysis_muscat(MenstrualCyclePhase = "Proliferative", Annotation = "Annotati
 #Proliferative phase samples with strict exclusion criteria, AnnotationUnited
 DEGanalysis_muscat(MenstrualCyclePhase = "Proliferative", Annotation = "AnnotationUnited")
 
+
+
 ##Secretory Phase
 #Secretory phase samples with strict exclusion criteria, AnnotationMain
 DEGanalysis_muscat(MenstrualCyclePhase = "Secretory", Annotation = "AnnotationMain")

### 2025-02-24 01_Atlas_generation/AlevinQC/AlevinQC.R
chore: rm deprecaed scripts

@@ -1,86 +1,0 @@
-## ----v1, include = FALSE------------------------------------------------------
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>",
-  eval = TRUE,
-  crop = NULL
-)
-
-## ----install, eval=FALSE------------------------------------------------------
-#  if (!requireNamespace("BiocManager", quietly = TRUE))
-#      install.packages("BiocManager")
-#  BiocManager::install("alevinQC")
-
-## ----v2, message=FALSE--------------------------------------------------------
-library(alevinQC)
-
-## ----v3a, eval=FALSE, results='asis'------------------------------------------
-#  baseDir
-#    |--alevin
-#    |    |--featureDump.txt
-#    |    |--filtered_cb_frequency.txt
-#    |    |--MappedUmi.txt
-#    |    |--quants_mat_cols.txt
-#    |    |--quants_mat_rows.txt
-#    |    |--quants_mat.gz
-#    |    |--raw_cb_frequency.txt
-#    |    |--whitelist.txt
-#    |--aux_info
-#    |    |--meta_info.json
-#    |--cmd_info.json
-
-## ----v3b, eval=FALSE, results='asis'------------------------------------------
-#  baseDir
-#    |--alevin
-#    |    |--featureDump.txt
-#    |    |--raw_cb_frequency.txt
-#    |    |--whitelist.txt  (depending on how alevin was run)
-#    |--aux_info
-#    |    |--meta_info.json
-#    |    |--alevin_meta_info.json
-#    |--cmd_info.json
-
-## ----v4-----------------------------------------------------------------------
-baseDir <- system.file("extdata/alevin_example_v0.14", package = "alevinQC")
-checkAlevinInputFiles(baseDir = baseDir)
-
-## ----v5, message=FALSE, warning=FALSE-----------------------------------------
-outputDir <- tempdir()
-alevinQCReport(baseDir = baseDir, sampleId = "testSample", 
-               outputFile = "alevinReport.html", 
-               outputFormat = "html_document",
-               outputDir = outputDir, forceOverwrite = TRUE)
-
-## ----v7, message=FALSE--------------------------------------------------------
-app <- alevinQCShiny(baseDir = baseDir, sampleId = "testSample")
-
-## ----v8, eval=FALSE-----------------------------------------------------------
-#  shiny::runApp(app)
-
-## -----------------------------------------------------------------------------
-if (interactive()) {
-  out <- shiny::runApp(app)
-}
-
[…]

### 2025-02-24 01_Atlas_generation/AlevinQC/scrnaseqqc/pipelines/create_seurat_object/scripts/gene_lookup.R
feat: add qc pipeline

@@ -1,0 +1,63 @@
+options(error = traceback)
+# Create a lookup table containg the gene names in the ensemble and HUGO format
+# log
+log <- file(snakemake@log[[1]], open="wt")
+sink(log)
+sink(log, type="message")
+# mage
+save.image(file = paste0(snakemake@log[[2]]) )
+
+# Load libraries
+suppressPackageStartupMessages({
+  library(dplyr)
+  library(Seurat)
+  library(Matrix)
+  library(biomaRt)
+  library(testthat)
+})
+
+##################
+# load paths
+##################
+print(snakemake@input)
+print(snakemake@output)
+
+###########################
+# helper functions
+###########################
+source(paste(snakemake@scriptdir, "helperfunctions", "general.R", sep = "/"))
+
+######################################################################
+# load data
+######################################################################
+
+path <- paste0(snakemake@input[[1]])
+data <- readRDS(path)
+
+ifelse(snakemake@config$counting == TRUE, 
+       count_matrix <- data$counts,
+       count_matrix <-  data)
+#########################################################################################################
+# create gene lookuptable using the gene annotation
+#########################################################################################################
+gene_look_table <- entrez_lookup(matrix = count_matrix,
+                                 config = snakemake@config)
+
+# what are the ensembl_gene_id returning NAs in entrez
+ensembles_genes_na <-  gene_look_table[is.na(gene_look_table$entrezgene),]
+table(ensembles_genes_na$gene_biotype)
+
+# sanity check: Gene look table is not a empty data frame, columns entrezgene_id  andexternal_gene_name are numeric and character"
+stopifnot({
+  expect_is(gene_look_table, "data.frame")
+  expect_gt(dim(gene_look_table)[1], 1)
+  expect_is(gene_look_table$entrezgene_id, "integer")
+  expect_is(gene_look_table$external_gene_name, "character")
+})
+
+# save results
+saveRDS(gene_look_table, file = snakemake@output[[1]])
+
+sessionInfo()
+date()
+

### 2025-02-24 01_Atlas_generation/AlevinQC/scrnaseqqc/pipelines/create_seurat_object/scripts/helperfunctions/general.R

@@ -1,0 +1,181 @@
+# Helper Functions for core pipe
+
+###########################################################
+# Function for Normalisation, HVG selection and scaling
+###########################################################
+# I: data, snakemake config table
+# O: normalised data as seurat object
+normalisation_scaling <- function(data, config ) {
+  switch((snakemake@config$regr.vars == "NA") + 1, 
+         regr.vars <- strsplit(snakemake@config$regr.vars, ",")[[1]], 
+         regr.vars <- NULL )
+  # load the vars to regress for
+  if (paste0(config$norm.method) == c("standart")) {
+    n.features <- config$nfeatures
+    # data normalisation
+    ###########################################################
+    # data normalization by the gene expression measurements for each cell by the total expression (lib size normalisation), log transformation
+    data <- NormalizeData(object = data, normalization.method = "LogNormalize", 
+                          scale.factor = config$scale.factor)
+    # Variable feature selection
+    ###########################################################
+    # Detection of variable genes across the single cells by genewise mean and variance, fitting a model to the relationship of log(variance) and log(mean) using local polynomial regression (loess).The feature values are standartized by the observed mean and expected variance (given by the fitted line).# Feature variance is then calculated on the standardized values.
+    data <- FindVariableFeatures(object = data, selection.method = "vst",
+                                 nfeatures = n.features )
+    # Scale Data
+    ###########################################################
+    # Scaling the data and remove sources of unwanted variation
+    data <- ScaleData(object =   data,
+                      features = VariableFeatures(object = data),
+                      vars.to.regress = regr.vars,
+                      do.scale = TRUE,
+                      do.center = TRUE)
+    DefaultAssay(data) <- "RNA"
+    
+  } else if ( paste0(config$norm.method) == "sctransform") { 
+    # data normalization by SCtansform using residuals, log transformation
+    data <- SCTransform(object = data,  vars.to.regress = regr.vars,
+                        conserve.memory=FALSE,
[…]

### 2025-02-24 01_Atlas_generation/AlevinQC/scrnaseqqc/pipelines/create_seurat_object/scripts/seurat_import.R

@@ -1,0 +1,69 @@
+options(error = traceback)
+# Samplewise importing the count matrices into a Seurat object 
+# log
+log <- file(snakemake@log[[1]], open="wt")
+sink(log)
+sink(log, type="message")
+# image
+save.image(file = paste0(snakemake@log[[2]]) )
+
+# Load libraries
+suppressPackageStartupMessages({
+  library(dplyr)
+  library(Seurat)
+  library(Matrix)
+  library(biomaRt)
+  library(testthat)
+})
+
+##################
+# load paths
+##################
+print(snakemake@input)
+print(snakemake@output)
+
+###########################
+# load helper functions
+###########################
+source(paste(snakemake@scriptdir, "helperfunctions", "general.R", sep = "/"))
+
+######################################################################
+# import alevin quants and create Seurat object
+# parameters min.cells and min genes are filtering thresholds
+# project is defined as the read sample name given from snakemake 
+######################################################################
+
+path_data <- paste0(snakemake@input[[1]])
+path_lookup <- paste0(snakemake@input[[2]])
+
+data <- readRDS(path_data)
+gene_lookup_table <- readRDS(path_lookup)
+
+ifelse(snakemake@config$counting == TRUE, 
+       count_matrix <- data$counts,
+       count_matrix <- data)
+#########################################################################################################
+# rename the gene names to external gene names and create lookup table for gene names
+#########################################################################################################
+count_matrix  <- rename_features(matrix=count_matrix, 
+                                 lookup=gene_lookup_table, 
+                                 config = snakemake@config )
+
+# create a seurat object from count matrix
+data <- CreateSeuratObject(counts = count_matrix, 
+                           min.cells = snakemake@config$min.cells, 
+                           min.features = snakemake@config$min.genes, 
+                           project = paste0(snakemake@wildcards$sample))
+
+# sanity check: Object is a Seurat object,  not empty, contains features and cells in the raw data slot
+stopifnot({
+  expect_is(data, "Seurat")
+  expect_gt(dim(data@assays$RNA)[1], 1)
+  expect_gt(dim(data@assays$RNA)[2], 1)
+})
+# save results
+saveRDS(data, file = snakemake@output[[1]])
+
+sessionInfo()
+date()
+

### 2025-02-25 05b_Functional_enrichment_analysis/DEG_gProfiler.Rmd
cleaned up code

@@
 library(cowplot)
+library(ggrepel)
 
@@
 
-length(RNAlistRDSdf$gene)
-RNAlistRDSdf_filt <- RNAlistRDSdf %>% dplyr::filter(AveExpr > 18.53202) 
-length(RNAlistRDSdf_filt$gene)
 ```
@@
 
-gProfiler based on RNA/SCT assay and per annotation with gene expression >18.53202
+gProfiler based on RNA assay and per annotation with gene expression >18.53202
 ```{r}
@@
 #Read paths to DEG output .rds files
-if (ASSAY == "RNA") { 
 PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
-    } else if (ASSAY == "SCT") {
-    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "mm_dream_sct_counts_all.rds", recursive = TRUE, full.names = TRUE)
-    }
+
 #Select paths based on all/mild/severe
@@
 RUN_gProfier(GRADE = "mild", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-RUN_gProfier(GRADE = "all", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-RUN_gProfier(GRADE = "severe", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-RUN_gProfier(GRADE = "mild", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-```
-
-##Ongoing xxxx
-gProfiler based on RNA assay and per annotation with gene expression >18.53202 and BH pval adjusted
-```{r}
-##function to run gProfiler and save output
-gProfiler_FUN <- function(x,ANALYSIS,GRADE,ORDERED_QUERY,ASSAY,dir_out) {
-#gProfiler
-custom_bg <- unique(x$gene)
-query <- x %>%
-  dplyr::filter(AveExpr > 18.53202) %>%
-  dplyr::filter(p_adj.loc < 0.05) %>%
-  dplyr::arrange(p_adj.loc) #arrange by lowest p_adj.loc
-if(nrow(query) > 0){
-query <- distinct(query, gene, .keep_all= TRUE) #remove duplicates of gene with higher p_adj.loc
-query <- query$gene
-gostres <<- gost(query = query, 
-                organism = "hsapiens", 
-                ordered_query = ORDERED_QUERY, #can use TRUE, p_adj.loc has been arranged 
-                multi_query = FALSE, 
-                significant = TRUE, #only statistically significant results should be returned
-                exclude_iea = FALSE, 
-                measure_underrepresentation = FALSE, 
-                evcodes = TRUE,
-                user_threshold = 0.05, 
-                correction_method = "g_SCS", #g_SCS is default
-                domain_scope = "annotated", #if custom_bg is set, I believe this shouldn't matter too much
-                custom_bg = custom_bg, #custom background, use!
-                numeric_ns = "", 
-                sources = NULL, #by default all sources are analyzed
-                as_short_link = FALSE)
-if(length(gostres) > 0){
-results <- gostres$result
-results$cluster_id <- levels(as.factor(x$cluster_id))
-results$analysis <- print(ANALYSIS)
-results$assay <- print(ASSAY)
-results$grade <- print(GRADE) 
-results$ordered_query <- print(ORDERED_QUERY)
-results$parents <- NULL
-write.csv(results,paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
-print(paste0("Saved gotres as gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".csv"))
-plot <- gostplot(gostres, capped = TRUE, interactive = FALSE) + ggtitle(paste0("gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY)))
-ggsave(plot = plot, filename = paste0(dir_out,"gProfiler_",ANALYSIS,"_",GRADE,"_",levels(as.factor(x$cluster_id)),"_OrderedQuery",as.character(ORDERED_QUERY),".png"), width = 7, height = 5)
-print("Saved plot")
-}
-}
-}
-##function to run gProfiler_FUN on different grades, analysis and assay
-RUN_gProfier <- function(GRADE, ANALYSIS,ASSAY,SPECIFICATION){
-dir_out <- paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ANALYSIS,"/",ASSAY,"assay_",SPECIFICATION,"/")
-dir.create(dir_out, recursive = TRUE)
-setwd(dir_out)
-#Read paths to DEG output .rds files
-if (ASSAY == "RNA") { 
-    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
-    } else if (ASSAY == "SCT") {
-    PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/",ANALYSIS), pattern = "mm_dream_sct_counts_all.rds", recursive = TRUE, full.names = TRUE)
-    }
-#Select paths based on all/mild/severe
-if (GRADE == "all") { 
-    PATHS <- PATHS[!grepl("_severe", PATHS)]
-    PATHS <- PATHS[!grepl("_mild", PATHS)]
-    print(PATHS)
-    } else if (GRADE == "mild") {
-    PATHS <- PATHS[grepl("_mild", PATHS)]
-    print(PATHS)
-    } else if  (GRADE == "severe") {
-    PATHS <- PATHS[grepl("_severe", PATHS)]
-    print(PATHS)
-    } 
-RNAlistRDS <- lapply(PATHS, readRDS)
-RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
-RNAlistRDS <- RNAlistRDS[!duplicated(RNAlistRDS)]
-lapply(RNAlistRDS, gProfiler_FUN, ANALYSIS = ANALYSIS,GRADE = GRADE,ORDERED_QUERY = FALSE, ASSAY = ASSAY, dir_out = dir_out)
-lapply(RNAlistRDS, gProfiler_FUN, ANALYSIS = ANALYSIS,GRADE = GRADE,ORDERED_QUERY = TRUE, ASSAY = ASSAY, dir_out = dir_out)
-}
-
-RUN_gProfier(GRADE = "all", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-RUN_gProfier(GRADE = "severe", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-RUN_gProfier(GRADE = "mild", ANALYSIS = "A030", ASSAY = "RNA",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-RUN_gProfier(GRADE = "all", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-RUN_gProfier(GRADE = "severe", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-RUN_gProfier(GRADE = "mild", ANALYSIS = "A030", ASSAY = "SCT",SPECIFICATION = "PerAnnotation_AveExpr18_53202")
-```
-
-```{r Combine output tables from gProfiler}
-#functions
-summarize_gProfiler_PerAnnotation <- function(x){
-ASSAY = "A030"
-##Combine all results
-gProfilerlist <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ASSAY,"/",x), pattern = "TRUE.csv$|FALSE.csv$", full.names = TRUE)
-gProfilerlist <- lapply(gProfilerlist, read.csv) #read .rds
-#Combine all gProfiler results into one df, save
-gProfiler <- plyr::rbind.fill(gProfilerlist) #bind deg lists into one df
-gProfiler$specification <- print(x)
-#save csv
-write.csv(gProfiler, paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ASSAY,"/gProfiler_",ASSAY,"_",x,"_summary.csv"))
-table(gProfiler$grade)
-#table(gProfiler$cluster_id,gProfiler$grade)
-}
-
-ASSAY = "A030"
-DIRSLIST <- list.dirs(path = paste0("/home/common/data/output/projects/ENDO/E044/A031/gProfiler/E044",ASSAY),full.names = FALSE)
-DIRSLIST_Merged <- DIRSLIST[grepl("Merged", DIRSLIST)]
-DIRSLIST_PerAnnotation <- DIRSLIST[grepl("PerAnnotation", DIRSLIST)]
-lapply(DIRSLIST_PerAnnotation,summarize_gProfiler_PerAnnotation)
 ```
@@
   filter(cluster_id %in% Tan_clusters)
-
-###interesting term_names
-#Immune System and inflammation
-ImmuneTermsGREP <- c("TGF", "TNF", "NOD", "IL-", "immun", "cytokine", "chemokine", "NF-kappa")
-ImmuneTerms <- gProfiler_df %>%
-  filter(grepl(paste(ImmuneTermsGREP, collapse = "|"), term_name, ignore.case = TRUE)) %>% 
-  pull(term_name) %>% 
-  unique()
-#angiogenesis
-angiogenesisTerms  <- c(unique(gProfiler_df$term_name[grepl("VEGF", gProfiler_df$term_name, ignore.case = TRUE)]),
-                       unique(gProfiler_df$term_name[grepl("angiog", gProfiler_df$term_name, ignore.case = TRUE)]))
-#extracellullar terms
-extracellularTerms <- unique(gProfiler_df$term_name[grepl("extracellular", gProfiler_df$term_name, ignore.case = TRUE)])
-#cell adhesion
-adhesionTermsGREP <- c("adhesion", "junction")
-adhesionTerms <- gProfiler_df %>%
-  filter(grepl(paste(adhesionTermsGREP, collapse = "|"), term_name, ignore.case = TRUE)) %>% 
-  pull(term_name) %>% 
-  unique()
-#growth factors
-insulinTerms <- unique(gProfiler_df$term_name[grepl("insulin", gProfiler_df$term_name, ignore.case = TRUE)])
-#Other interesting terms
-otherTerms <- c('CTNNB1-DDX5-HDGF complex') #Hepatoma-Derived Growth Factor and DDX5 Promote Carcinogenesis and Progression of Endometrial Cancer by Activating β-Catenin
-# Define the list of interesting terms
-interesting_terms <- list(
-  ImmuneTerms = ImmuneTerms,
-  AngiogenesisTerms = angiogenesisTerms,
-  ExtracellularTerms = extracellularTerms,
-  AdhesionTerms = adhesionTerms,
-  InsulinTerms = insulinTerms
-  #,OtherTerms = otherTerms
-)
-# Loop through the list and assign categories
-gProfiler_df$DEG_category <- NA
-for (category_name in names(interesting_terms)) {
-  terms <- interesting_terms[[category_name]]
-  for (term in terms) {
-    gProfiler_df$DEG_category[gProfiler_df$term_name == term] <- category_name
-  }
-}
-table(gProfiler_df$DEG_category)
 
@@
   width = 14, height = 14)
-
-##############################################################################################################################
-##Merge the DEG table with the Terms
-
-#RNA assay DEGs from proliferative samples
-RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv") %>%
-  filter(annotation_reference == "Tan") %>%
-  filter(AveExpr > 18.53202)
-##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
-hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes, there are no hb genes in the DEG list
-mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes, no genes
-rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes, 165 genes
-RNAlistRDS005_df <- RNAlistRDS005_df %>%
-  dplyr::filter(!gene %in% rb.genes) #Focus on protein coding genes
-
-#Mutate so each row contains only a single gene in the "gene" column
-gProfiler_df <- gProfiler_df %>%
-  mutate(gene = strsplit(intersection, ",")) %>%
-  unnest(gene) %>%
-  dplyr::rename(stage = grade) %>%
-  dplyr::filter(assay == "RNA")
-
-# left_join RNAlistRDS005_df with Terms by c("cluster_id", "gene", "stage"), no multiple match warning
-result_df <- left_join(RNAlistRDS005_df, gProfiler_df, by = c("cluster_id", "gene","stage"))
-
-###Plot DEGs
-# Create vulcano plot
-ggplot(result_df, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
-  geom_point(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
-             aes(color = DEG_category), size = 2) +
-  theme_minimal() +
-  facet_wrap(~stage)
-ggsave(filename = paste0(dir_out, 'vulcano_DEG_category_InterestingTerms.pdf'), width =8,height=6)
-
-# Create vulcano plot with gene labels
-#logFC vs AveExpr with gene labels
-InterestingTerms <- result_df[,c(1:12,34)] %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms") %>% 
-  distinct(gene, logFC,AveExpr,p_adj.loc, .keep_all = TRUE) %>% 
-  #dplyr::filter(logFC > 0.5 | logFC < -0.5) %>% 
-  dplyr::filter(stage == "all", annotation_reference == "Tan", AveExpr > 18.5, logFC > 0.2)
-plot <- ggplot(result_df, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
-  geom_point(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
-             aes(color = DEG_category), size = 2) +
-  theme_minimal()+
-  ylim(0,18) +
-  xlim(-.9,1.2) +
-  geom_text_repel(data = InterestingTerms,
-                  aes(label = gene),
-                  size = 2,
-                  max.overlaps = 50,
-                  box.padding   = 0.35, 
-                  point.padding = 0.5,
-                  segment.color = 'grey50')
-plot
-ggsave(plot, filename = paste0(dir_out, 'vulcano_DEG_category_InterestingTerms_labelled.pdf'), width =16,height=12)
-
-# Create vulcano plot with gene labels
-#logFC vs AveExpr with gene labels
-InterestingTerms <- result_df[,c(1:12,34)] %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms") %>% 
-  distinct(gene, logFC,AveExpr,p_adj.loc, .keep_all = TRUE) %>% 
-  #dplyr::filter(logFC > 0.5 | logFC < -0.5) %>% 
-  dplyr::filter(stage == "all", annotation_reference == "Tan", abs(logFC) > 0.5)
-result_df_all <- filter(result_df,stage == "all")
-ggplot(result_df_all, aes(x = logFC, y = -log10(p_adj.loc))) +
-  geom_point(aes(color = "Non-overlapping"), size = 1, color = "grey") +
-  geom_point(data = result_df_all %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
-             aes(x = logFC, y = -log10(p_adj.loc), color = DEG_category), size = 2) +
-  theme_minimal() +
-  ylim(-2, 17) +
-  xlim(-0.9, 1.2) +
-  geom_text_repel(data = result_df_all[2:12] %>% dplyr::filter(abs(logFC) > 0.5) %>% distinct(),
-                  aes(label = gene),
-                  size = 2,
-                  max.overlaps = 50,
-                  box.padding = 0.35,
-                  point.padding = 0.5,
-                  segment.color = 'grey50') +
-  facet_wrap(~stage)
-ggsave(filename = paste0(dir_out, 'vulcano_DEG_category_InterestingTerms_labelledlogFC05.pdf'), width =8,height=6)
-
-results_df_logFC05 <- result_df[2:12] %>% dplyr::filter(abs(logFC) > 0.5) %>% distinct()
-write.csv(results_df_logFC05, paste0(dir_out,'TermslogFC05.csv'))
-
-#logFC vs AveExpr
-ggplot(result_df, aes(x = log2(AveExpr-17.00183), y = logFC)) +
-  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
-  geom_point(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
-             aes(color = DEG_category), size = 2) +
-  theme_minimal() +
-  geom_text(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"), aes(label = gene))
-ggsave(filename = paste0(dir_out, 'AveExprLOG2_logFC_InterestingTerms.pdf'), width =8,height=6)
-
-#logFC vs AveExpr with gene labels
-InterestingTerms <- result_df[,c(1:12,34)] %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms") %>% 
-  distinct(gene, logFC,AveExpr,p_adj.loc, .keep_all = TRUE) %>% 
-  #dplyr::filter(logFC > 0.5 | logFC < -0.5) %>% 
-  dplyr::filter(stage == "all", annotation_reference == "Tan", AveExpr > 18.5)
-
-ggplot(result_df, aes(x = log2(AveExpr-17.00183), y = logFC)) +
-  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
-  geom_point(data = InterestingTerms , aes(color = DEG_category), size = 2) +
-  theme_minimal() +
-  geom_text_repel(data = InterestingTerms,
-                  aes(label = gene),
-                  size = 2,
-                  max.overlaps = 35,
-                  box.padding   = 0.35, 
-                  point.padding = 0.5,
-                  segment.color = 'grey50')
-ggsave(filename = paste0(dir_out, 'AveExprLOG2_logFC_InterestingTerms_labelled.pdf'), width =14,height=8)
-
-#logFC vs AveExpr
-ggplot(result_df, aes(x = AveExpr, y = logFC)) +
-  geom_point(data = result_df, aes(color = "Non-overlapping"), size = 1, color = "grey") +
-  geom_point(data = result_df %>% dplyr::filter(!is.na(DEG_category), DEG_category != "EndometriumTerms"),
-             aes(color = DEG_category), size = 2) +
-  theme_minimal()
- ggsave(filename = paste0(dir_out, 'AveExpr_logFC_InterestingTerms.pdf'), width =8,height=6)
 

### 2025-02-25 Figures/PaperFiguresextended.Rmd
cleaned up code

@@
 
-Figure 1e, 2a, S3a
+Figure 1e, 2a
 CODE running, need to extend the PCA analysis folder
@@
     legendPosition = 'top', legendLabSize = 16, legendIconSize = 8.0)
-
-#Supplementary Data Figure 3a
-biplot(p,
-    colby = 'EndometriosisStatus',
-    labSize = 0,drawConnectors = FALSE, 
-    hline = 0, vline = 0,
-    legendPosition = 'right')
 ```
@@
 
+```
+
+Supplementary Figure 3a
+CODE running, need to extend the PCA analysis folder
+dir_out <- "/home/common/data/output/projects/ENDO/E044/A072/allcells_EndoAtlas_SCTassay/"
+
+```{r}
+
+p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
+
+#Supplementary Data Figure 3a
+biplot(p,
+    colby = 'EndometriosisStatus',
+    labSize = 0,drawConnectors = FALSE, 
+    hline = 0, vline = 0,
+    legendPosition = 'right')
 ```

### 2025-02-27 01_Atlas_generation/Integration/01b_run_normalization_split.R
replaced integration muscat pipelines with extracted script

@@ -2,14 +2,6 @@
 ## 00. Load library and environment
 ####################################
 
-# log
-log <- file(snakemake@log$log_file, open="wt")
-sink(log)
-sink(log, type="message")
-
-# image
-save.image(file = paste0(snakemake@log$log_object) )
-
 # load packages
 suppressPackageStartupMessages({
   library(dplyr)
@@ -20,17 +12,15 @@ suppressPackageStartupMessages({
 
 
 # load seurat object
-seurat_obj <- readRDS(snakemake@input[[1]])
+seurat_obj <- readRDS(./seurat.rds)
 Idents(seurat_obj) <- seurat_obj@meta.data$sample
-
 
 ####################################
 ## 01. Normalization split per batch
 ####################################
 
-regr.vars <- strsplit(snakemake@params[[1]], split='_')[[1]]
-
-# split by batch
+# SCTransform and regress out percent.mt and sample
+regr.vars <- c("percent.mt","sample")
 seurat_obj <- SCTransform(seurat_obj,
                           vars.to.regress = regr.vars,
                           conserve.memory=FALSE,
@@ -38,8 +28,10 @@ seurat_obj <- SCTransform(seurat_obj,
                           vst.flavor = "v2",
                           return.only.var.genes=TRUE)
 
-seurat_obj_split <- SplitObject(seurat_obj, split.by = snakemake@params[[2]])
+# split by 10x capture batch
+seurat_obj_split <- SplitObject(seurat_obj, split.by = "batch")
 
-saveRDS(seurat_obj_split, snakemake@output[[1]])
+# save split seurat object
+saveRDS(seurat_obj_split, "./seurat_obj_split.rds")
 sessionInfo()
 date()

### 2025-02-27 01_Atlas_generation/Integration/02_run_seuratv3.R

@@ -1,0 +1,81 @@
+####################################
+## 00. Load library and environment
+####################################
+
+# load packages
+suppressPackageStartupMessages({
+  library(dplyr)
+  library(Seurat)
+  library(future)
+  library(RColorBrewer)
+  library(remotes)
+  library(ggplot2)
+  library(SingleCellExperiment)
+})
+
+# load split seurat object
+seurat_obj_split_input <- readRDS("./seurat_obj_split.rds")
+print("input loaded")
+
+# specify output directory
+dir_out <- snakemake@config$integration_output_dir
+
+####################################
+## 01. Run integration
+####################################
+
+# RPCA integration on SCT transformed counts
+features_integration <- SelectIntegrationFeatures(object.list = seurat_object_list,
+                                                  nfeatures = 3000)
+seurat_object_list <- Seurat::PrepSCTIntegration(object.list = seurat_object_list,
+                                                 assay = "SCT",
+                                                 anchor.features = features_integration,
+                                                 verbose = TRUE
+)
+seurat_object_list <- lapply(seurat_object_list,
+                             FUN = Seurat::RunPCA,
+                             anchor.features = features_integration
+)
+
+
+reference_dataset <- which(names(seurat_object_list) == "batch17")
+
+
+integration_anchors <- FindIntegrationAnchors(object.list = seurat_object_list,
+                                              normalization.method = "SCT",
+                                              anchor.features = features_integration,
+                                              dims = seq_len(20),
+                                              reduction = "rpca",
+                                              k.anchor = 20,
+                                              reference = reference_dataset,
+                                              verbose = TRUE
+)
+
+seurat_object_list <- IntegrateData(anchorset = integration_anchors,
+                                    normalization.method = "SCT",
+                                    dims = seq_len(snakemake@config$n_dimred_integration),
+                                    verbose = TRUE
+)
+
+DefaultAssay(integrated_data) <- "integrated"
+
+seed <- 122334
+
+# Perform linear dimensional reduction by PCA on HVGs
+integrated_data <- RunPCA(object = integrated_data,
+                     features = VariableFeatures(integrated_data),
+                     seed.use = seed,
+                     npcs = 50)
+# Run UMAP
+integrated_data <- RunUMAP(integrated_data,  dims = 1:30, seed.use = seed)
+
+#Save seurat
[…]

### 2025-02-27 01_Atlas_generation/Integration/scripts/02_run_seuratv3.R

@@ -1,99 +1,0 @@
-####################################
-## 00. Load library and environment
-####################################
-
-# log
-log <- file(snakemake@log$log_file, open="wt")
-sink(log)
-sink(log, type="message")
-
-# image
-save.image(file = paste0(snakemake@log$log_object) )
-
-# load packages
-suppressPackageStartupMessages({
-  library(dplyr)
-  library(Seurat)
-  library(future)
-  library(RColorBrewer)
-  library(remotes)
-  library(ggplot2)
-  library(SingleCellExperiment)
-})
-
-# dir_out
-dir_out <- snakemake@config$integration_output_dir
-
-
-# load seurat object
-seurat_obj_split_input <- readRDS(snakemake@input[[1]])
-print("input loaded")
-
-
-
-####################################
-## 01. Run integration
-####################################
-
-rpca_integration = function(seurat_object_list, snakemake, features_integration, verbose = TRUE){
-
-  # RPCA integration on SCT transformed counts
-  features_integration <- SelectIntegrationFeatures(object.list = seurat_obj_split_input,
-                                                    nfeatures = 3000)
-  seurat_object_list <- Seurat::PrepSCTIntegration(object.list = seurat_object_list,
-                                                   assay = "SCT",
-                                                   anchor.features = features_integration,
-                                                   verbose = verbose
-  )
-  seurat_object_list <- lapply(seurat_object_list,
-                               FUN = Seurat::RunPCA,
-                               anchor.features = features_integration
-  )
-
-
-  reference_dataset <- which(names(seurat_object_list) == snakemake@config$reference_dataset)
-
-
-  integration_anchors <- FindIntegrationAnchors(object.list = seurat_object_list,
-                                                normalization.method = "SCT",
-                                                anchor.features = features_integration,
-                                                dims = seq_len(snakemake@config$n_dimred_integration),
-                                                reduction = "rpca",
-                                                k.anchor = 20,
-                                                reference = reference_dataset,
-                                                verbose = verbose
-  )
-  print(paste0('Length of features used for integration',
-               length(integration_anchors@anchor.features)))
-  print(paste0('Number of cells used for integration ',
-               dim(integration_anchors@anchors)[1]))
-  seurat_object_list <- IntegrateData(anchorset = integration_anchors,
-                                      normalization.method = "SCT",
[…]

### 2025-02-27 01_Atlas_generation/Integration_epithelial_only/01b_run_normalization_split.R

@@ -2,14 +2,6 @@
 ## 00. Load library and environment
 ####################################
 
-# log
-log <- file(snakemake@log$log_file, open="wt")
-sink(log)
-sink(log, type="message")
-
-# image
-save.image(file = paste0(snakemake@log$log_object) )
-
 # load packages
 suppressPackageStartupMessages({
   library(dplyr)
@@ -20,7 +12,7 @@ suppressPackageStartupMessages({
 
 
 # load seurat object
-seurat_obj <- readRDS(snakemake@input[[1]])
+seurat_obj <- readRDS(./seurat.rds)
 
 # Subset epithelial cells
 Idents(seurat_obj) <- seurat_obj@meta.data$AnnotationMain
@@ -39,10 +31,8 @@ seurat_obj <- subset(seurat_obj, idents = samples_integration)
 ## 01. Normalization split per batch
 ####################################
 
-regr.vars <- strsplit(snakemake@params[[1]], split='_')[[1]]
-
-# split by batch
-Idents(seurat_obj) <- seurat_obj@meta.data$sample
+# SCTransform and regress out percent.mt and sample
+regr.vars <- c("percent.mt","sample")
 seurat_obj <- SCTransform(seurat_obj,
                           vars.to.regress = regr.vars,
                           conserve.memory=FALSE,
@@ -50,8 +40,10 @@ seurat_obj <- SCTransform(seurat_obj,
                           vst.flavor = "v2",
                           return.only.var.genes=TRUE)
 
-seurat_obj_split <- SplitObject(seurat_obj, split.by = snakemake@params[[2]])
+# split by 10x capture batch
+seurat_obj_split <- SplitObject(seurat_obj, split.by = "batch")
 
-saveRDS(seurat_obj_split, snakemake@output[[1]])
+# save split seurat object
+saveRDS(seurat_obj_split, "./seurat_obj_split.rds")
 sessionInfo()
 date()

### 2025-02-27 01_Atlas_generation/Integration_epithelial_only/scripts/02_run_seuratv3.R

@@ -1,122 +1,0 @@
-####################################
-## 00. Load library and environment
-####################################
-
-# log
-log <- file(snakemake@log$log_file, open="wt")
-sink(log)
-sink(log, type="message")
-
-# image
-save.image(file = paste0(snakemake@log$log_object) )
-
-# load packages
-suppressPackageStartupMessages({
-  library(dplyr)
-  library(Seurat)
-  library(future)
-  library(RColorBrewer)
-  library(remotes)
-  library(ggplot2)
-  library(SingleCellExperiment)
-})
-
-# load seurat object
-seurat_obj_split_input <- readRDS(snakemake@input[[1]])
-print("input loaded")
-#####filter hvgs
-features_integration <- SelectIntegrationFeatures(object.list = seurat_obj_split_input,
-                                                    nfeatures = 3000)
-print("integration features loaded")
-
-##Gene lists
-#Haemoglobin genes
-hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes 
-#mitochondrial genes
-mt.genes <- grep("^MT-", features_integration, value=TRUE)#mitochondrial genes
-#ribosomal genes
-rb.genes <- grep(pattern = "^RP[SL]", features_integration, value=TRUE)
-##Core stress response genes#https://github.com/kieranrcampbell/scrnaseq-digestion-paper/blob/master/data/deliverables/coregene_df-FALSE-v3.csv
-stress.genes <- read.csv("/home/common/data/output/projects/ENDO/E039/A012/Tan22databases/coregene_stress-response.csv")$gene_symbol
-##Cell cycle genes #Giotti, B. et al. Assembly of a parts list of the human mitotic cell cycle machinery. J. Mol. Cell. Biol. 11, 703718 (2019).
-CellCycleGenes <- read.csv("/home/common/data/output/projects/ENDO/E039/A012/Tan22databases/cell_cycle.csv")$Genes
-
-print("gene lists loaded")
-
-#filter hvgs
-suffix='hvgs_filtered'
-features_integration <- Reduce(setdiff, list(features_integration, hb.genes,mt.genes,rb.genes))
-
-print("hvgs loaded")
-
-
-# dir_out
-dir_out <- snakemake@config$integration_output_dir
-
-
-####################################
-## 01. Run integration
-####################################
-
-rpca_integration = function(seurat_object_list, snakemake, features_integration, verbose = TRUE){
-
-  # RPCA integration on SCT transformed counts
[…]

### 2025-02-28 05a_DEG_analysis_muscat_Snakemake/config/profile/job_env.sh
deleted extra pipeline

@@ -1,2 +1,0 @@
-NAMESPACE=endo
-PIPELINE=endo

### 2025-02-28 05a_DEG_analysis_muscat_Snakemake/run_snakemake.sh

@@ -1,2 +1,0 @@
-#!/bin/bash
-snakemake --profile /endometriosis_endometrium_scRNA_atlas/05a_DEG_analysis_muscat/config/profile 

### 2025-02-28 05a_DEG_analysis_muscat_Snakemake/scripts/DGE_muscat_prolif_secretory.R

@@ -1,65 +1,0 @@
-####################################
-## 00. Load library and environment
-####################################
-
-
-suppressPackageStartupMessages({
-library(ggplot2)
-library(dplyr)
-library(limma)
-library(muscat)
-library(purrr)
-library(Seurat)
-library(scater)
-library(SingleCellExperiment)
-library(Matrix)
-})
-
-####################################
-## 01. Prepare Data
-####################################
-
-#load seurat object 
-data <- readRDS(snakemake@input[[1]])
-print("loaded data")
-
-#make SingleCellExperiment from RNA assay
-sce <- as.SingleCellExperiment(data, assay = "RNA")
-
-# remove undetected genes
-sce <- sce[rowSums(counts(sce) > 0) > 0, ]
-
-# remove lowly expressed genes (less than 10 counts)
-sce <- sce[rowSums(counts(sce) > 1) >= 10, ]
-
-##Data preparation
-sce$id <- paste0(sce$treatment, sce$sample)
-
-(sce <- prepSCE(sce,
-                kid = "Annotation", # subpopulation assignments
-                gid = "condition",  # group IDs (ENDO/non-ENDO)
-                sid = "sample",   # sample IDs 
-                drop = FALSE))  
-
-nk <- length(kids <- levels(sce$cluster_id))
-ns <- length(sids <- levels(sce$sample_id))
-names(kids) <- kids; names(sids) <- sids
-
-####################################
-## 02. Cell-level analysis: Mixed models
-####################################
-
-print("start DE testing:")
-       mm_dream <- mmDS(sce, method = "dream", coef = "group_idEndo",
-                 n_cells = 50, n_samples = 3, covs = "batch")
-
-saveRDS(mm_dream,"./mm_dream.rds")
-
-tbl_fil <- plyr::rbind.fill(mm_dream) %>%
-  dplyr::filter(p_adj.loc < 0.05) %>%
-  dplyr::arrange(p_adj.loc)
-
-write.csv(tbl_fil, "./mm_dream.csv")
-
-sessionInfo()
-date()

### 2025-02-28 05a_DEG_analysis_muscat_Snakemake/scripts/DGE_muscat_prolif_stringent.R

@@ -1,72 +1,0 @@
-####################################
-## 00. Load library and environment
-####################################
-
-
-suppressPackageStartupMessages({
-library(ggplot2)
-library(dplyr)
-library(limma)
-library(muscat)
-library(purrr)
-library(Seurat)
-library(scater)
-library(SingleCellExperiment)
-library(Matrix)
-})
-
-####################################
-## 01. Prepare Data
-####################################
-
-#load seurat object 
-data <- readRDS(snakemake@input[[1]])
-print("loaded data")
-
-#Filter for proliferative samples with stringent exclusion criteria
-Idents(data) <- data$DEG.analysis.Figure.3a.b
-data <- subset(data, idents = "TRUE")
-
-
-
-# List of annotations to loop through
-ANNOTATIONS <- c("AnnotationMain", "AnnotationRefined", "AnnotationUnited")
-
-# Loop through each annotation
-for (Annotation in ANNOTATIONS) {
-
-	#make SingleCellExperiment from RNA assay
-	sce <- as.SingleCellExperiment(data, assay = "RNA")
-
-	# remove undetected genes
-	sce <- sce[rowSums(counts(sce) > 0) > 0, ]
-
-	# remove lowly expressed genes (less than 10 counts)
-	sce <- sce[rowSums(counts(sce) > 1) >= 10, ]
-
-	##Data preparation
-	sce$id <- paste0(sce$treatment, sce$sample)
-
-	(sce <- prepSCE(sce,
-					kid = Annotation, # subpopulation assignments
-					gid = "condition",  # group IDs (ENDO/non-ENDO)
-					sid = "sample",   # sample IDs 
-					drop = FALSE))  
-
-	nk <- length(kids <- levels(sce$cluster_id))
-	ns <- length(sids <- levels(sce$sample_id))
-	names(kids) <- kids; names(sids) <- sids
-
-	####################################
-	## 02. Cell-level analysis: Mixed models
-	####################################
-
-	print("start DE testing:")
-		   mm_dream <- mmDS(sce, method = "dream", coef = "group_idEndo",
-					 n_cells = 50, n_samples = 3, covs = "batch")
-
-	saveRDS(mm_dream, paste0("./mm_dream",Annotation,".csv"))
-}
-
-sessionInfo()
-date()

### 2025-03-03 Figures/PaperFiguresextended.Rmd
Added additional code

@@
 
-DEGs
-E044A031
+##Figure 3a
+#ENDO/E044/A033/PaperFigures.Rmd
+#Clean up!! use DEG and GO table as basis for plots
+#Barplot unique genes per main cell type
+RNAlistRDS005_df %>% 
+  select(Symphony_Global_main5, gene, stage) %>%
+  distinct() %>%
+  ggplot(aes(x = stage, fill = factor(Symphony_Global_main5, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
+  geom_bar() +
+  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.3))
+ggsave(paste0(dir_out,"BarPlot_DEGs_main4.pdf"), 
+       plot = last_plot(), 
+       height = 4, width = 10)
+
+##Figure 3b
+#Clean up!! use DEG and GO table as basis for plots
+#ENDO/E044/A031/ExtractInterestingGenesFromGProfilerA031_RNA_AveExpr18_53202.Rmd
+gProfiler_df %>% 
+  #dplyr::select(p_value, term_size, query_size, intersection_size, term_id, term_name) %>% 
+  dplyr::filter(term_size > 25 & term_size < 750) %>% #larger terms are likely to be excessively broad and prone to false positives
+  distinct(cluster_id,analysis,assay,grade,term_name, .keep_all = TRUE) %>%
+  #filter(grade == "all")  %>%
+  filter(!grepl(pattern = "HPA|HP|CORUM|MIRNA", x = term_id)) %>%
+  dplyr::arrange(p_value) %>% 
+  #dplyr::slice_head(n = 20) %>% 
+  ggplot2::ggplot() + 
+  ggplot2::aes(y = reorder(term_name, -log(p_value)), x = -log(p_value), color = grade, size = intersection_size/term_size * 100) + #, shape = ordered_query
+  ggplot2::geom_point() + 
+  #ggplot2::scale_color_continuous(name = "-Log(adj. p-value)") + 
+  ggplot2::scale_size_continuous("Intersection size / Term size * 100 (%)") + 
+  ggplot2::theme_classic() + 
+  ggplot2::xlab("-Log(adj. p-value)") + 
+  ggplot2::ylab("enriched terms") + 
+  ggplot2::ggtitle("all cells")
+ggsave(paste0(dir_out, "GOterms_all_severe.pdf"),
+  plot = last_plot(),
+  width = 14, height = 14)
+
+##SFig5a
+
+#ENDO/E044/A031/ExtractInterestingGenesFromGProfilerA031_RNA_AveExpr18_53202.Rmd
+#Clean up!! use DEG and GO table as basis for plots
+
+#RNA assay DEGs from proliferative samples
+RNAlistRDS005_df <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv") %>%
+  filter(annotation_reference == "Tan") %>%
+  filter(AveExpr > 18.53202)
+##Exclude Haemoglobin genes, mitochondrial genes, #ribosomal genes
+hb.genes <- c("HBA1","HBA2","HBB","HBG2","HBG1","HBD","HBE1","HBZ")#haemoglobin genes, there are no hb genes in the DEG list
+mt.genes <- grep("^MT-", RNAlistRDS005_df$gene, value=TRUE)#mitochondrial genes, no genes
+rb.genes <- grep(pattern = "^RP[SL]", RNAlistRDS005_df$gene, value=TRUE)#ribosomal genes, 165 genes
+RNAlistRDS005_df <- RNAlistRDS005_df %>%
+  dplyr::filter(!gene %in% rb.genes) #Focus on protein coding genes
+
+#Mutate so each row contains only a single gene in the "gene" column
+gProfiler_df <- gProfiler_df %>%
+  mutate(gene = strsplit(intersection, ",")) %>%
+  unnest(gene) %>%
+  dplyr::rename(stage = grade) %>%
+  dplyr::filter(assay == "RNA")
+
+#Mutate so each row contains only a single gene in the "gene" column
+gProfiler_df <- gProfiler_df %>%
+  dplyr::filter(!is.na(DEG_category))
+table(gProfiler_df$term_name,gProfiler_df$cluster_id)
+
+table <- left_join(gProfiler_df, RNAlistRDS005_df[,c("gene","logFC","AveExpr","p_adj.loc","cluster_id","stage")], by = c("gene","cluster_id","stage"))
+
+#x = -log(p_adj.loc)
+table %>%
+  arrange(DEG_category) %>%
+  #dplyr::filter(term_name == "extracellular organelle") %>%
+  ggplot2::ggplot() + 
+  ggplot2::aes(y = reorder(gene, -log(logFC)), x = -log(p_adj.loc), color = stage, size = logFC) + 
+  ggplot2::geom_point() + 
+  ggplot2::scale_size_continuous("logFC") + 
+  ggplot2::scale_x_continuous(limits = c(1, 6)) +
+  ggplot2::theme_classic() +
+  ggplot2::facet_wrap(~term_name, scales = "free_y", ncol = 13) +
+  ggplot2::theme(strip.text = ggplot2::element_text(size = 8))
+ ggsave(filename = paste0(dir_out, 'DotPlot_DEGcategories_DEGs_logFC_pval.pdf'), width =25,height=12)
 

### 2025-03-03 07_Ligand_receptor_analysis/01_cellchat_prep.R

@@ -7,6 +7,14 @@ library(patchwork)
 library(Seurat)
 options(stringsAsFactors = FALSE)
 
+
+#dir_out <- "/home/common/data/output/projects/ENDO/E044/A087/"
+#dir.create(paste0(dir_out,"output/"), recursive = TRUE)
+#setwd(dir_out)
+# data <- /home/common/data/output/projects/ENDO/E044/A033/EndoAtlas_GEO_complete.rds
+# meta <- readRDS("/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas_meta.rds")
+data <- AddMetaData(data,meta)
+
 dir_out <- "CellChat/"
 setwd(dir_out)
 
@@ -18,18 +26,18 @@ setwd(dir_out)
 data  <- readRDS("ENDOatlas.rds")
 
 ##Filter Seurat for proliferative phase samples with stringet exclusion criteria and annotated cells
-samples <- unique(read.csv("DEG_samples.csv")$samples)
+DEGsamples <- unique(data@meta.data %>% filter(DEG.analysis..Figure.3a.b. == "TRUE") %>% pull(sample))
 Idents(data) <- data$sample
-data <- subset(data, idents = samples) 
-data <- subset(data, subset = Symphony_Refined_Final %in% levels(as.factor(data$Symphony_Refined_Final)))
+data <- subset(data, idents = DEGsamples) 
+data <- subset(data, subset = AnnotationRefined %in% levels(as.factor(data$AnnotationRefined)))
 
-#Filter out secretory specific cell types (no/low cell counts stop script)
+#Filter out secretory specific cell types (no/low cell counts stop script 02)
 secretory_celltypes <- c("dS3","dS4","fib C7 secretory","glandular early-secretory","mid-secretory","Prv_VSMC secretory","Prv_VSMC secretory_late","fib C7 secretory_late") 
-data <- subset(data, subset = Symphony_Refined_Final %in% setdiff(levels(as.factor(data$Symphony_Refined_Final)), secretory_celltypes))
+data <- subset(data, subset = AnnotationRefined %in% setdiff(levels(as.factor(data$AnnotationRefined)), secretory_celltypes))
 
 # Define the conditions for Case and Control
-Endo <- data@meta.data$sample %in% c(filter(data@meta.data, treatment == "Endo") %>% distinct(sample) %>% pull())
[…]

### 2025-03-04 Figures/Preprocessing_metadata.Rmd
removed redundant code

@@
 
-#Save color pals
-qual_col_pals <- brewer.pal.info[brewer.pal.info$category == "qual", ]
-col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
 ```
-
-Save epithelial object and upload to GEO.
-```{r}
-#Load monocl object
-data_monocl <- readRDS("/home/common/data/output/projects/ENDO/E044/A062/trajectory_day5-6-7_greenCircle/trajectory_timepoint0.rds")
-listData <- (data_monocl@colData@listData)
-listData <- listData["cell_ident"]
-
-EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))
-colnames(EndoAtlas_meta)
-
-########################
-# Extract cell.ident from listData
-cell_ids <- listData[["cell_ident"]]
-
-# Ensure cell_ids is not empty
-if (length(cell_ids) == 0) {
-  stop("cell.ident in listData is empty. Ensure it is populated before proceeding.")
-}
-
-# Filter and order EndoAtlas_meta by cell_ids
-filtered_meta <- EndoAtlas_meta[EndoAtlas_meta$cell_names %in% cell_ids, ]
-ordered_meta <- filtered_meta[match(cell_ids, filtered_meta$cell_names), ]
-
-# Verify alignment
-if (!all(ordered_meta$cell_names == cell_ids)) {
-  stop("Ordering of filtered EndoAtlas_meta does not match cell_ids.")
-}
-
-# Add columns from ordered_meta to listData
-columns_to_add <- c("sample", "EndometriosisStatus", "EndometriosisGrade", "CycleDay", 
-                    "ProgesteroneSerum", "MenstrualCyclePhase", "AnnotationMain", 
-                    "AnnotationRefined", "cell_names", "MenstrualCyclePhase_main")
-
-for (col in columns_to_add) {
-  if (col %in% colnames(ordered_meta)) {
-    listData[[col]] <- ordered_meta[[col]]
-  } else {
-    warning(paste("Column", col, "not found in EndoAtlas_meta. Skipping."))
-  }
-}
-
-# Update the main object
-data_monocl@colData@listData <- listData
-
-data_monocl@principal_graph_aux@listData[["UMAP"]][["pseudotime"]]
-
-saveRDS(data_monocl, paste0(dir_data, "epithelial_cells_monocle3.rds"))
-```

### 2025-03-04 07_Ligand_receptor_analysis/01_cellchat_prep.R

@@ -131,5 +131,5 @@ CellChat_visualization <- function(CONDITION){
       }
 }
 
-CellChat_visualization(Endo)
-CellChat_visualization(Control)
+CellChat_visualization("Endo")
+CellChat_visualization("Control")

### 2025-03-06 Figures/01_Generation_Figure1.R
cleaned up code and reformatted figure code

@@ -1,0 +1,89 @@
+# Aim : Code used for the generation of figure 1
+# Author: Lea Duempelmann
+
+## ---------------------------------- ##
+## Setup
+## ---------------------------------- ##
+
+# load librairies
+library(Seurat)
+library(dplyr)
+library(tibble)
+library(stringr)
+library(ggplot2)
+library(cowplot)
+library(ggbreak) 
+library(tidyr)
+library(patchwork)
+library(ComplexUpset)
+library(ggrepel)
+library(Polychrome)
+library(ComplexHeatmap)
+library(gplots)
+library(RColorBrewer)
+library(randomcoloR)
+
+## ---------------------------------- ##
+## Figure 1.c
+## ---------------------------------- ##
+
+#@Lea check these paths are correct 
+EndoAtlas <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas.rds")
+EndoAtlas_meta <- EndoAtlas@meta.data
+
+order <- EndoAtlas_meta %>%
+  select(AnnotationMain, AnnotationRefined) %>%
+  filter(!is.na(AnnotationMain)) %>%
+  distinct() %>%
+  arrange(AnnotationRefined) %>%
+  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
+  arrange(AnnotationMain)
+EndoAtlas$AnnotationRefined <- factor(x = EndoAtlas$AnnotationRefined, levels = order$AnnotationRefined)
+Fig1c <- DimPlot(EndoAtlas, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
+
+## ---------------------------------- ##
+## Figure 1.d
+## ---------------------------------- ##
+
+Fig1d_1 <- EndoAtlas_meta %>%
+  group_by(AnnotationMain) %>%
+  summarise(row_count = n()) %>% 
+  ggplot(aes(x = AnnotationMain, y = row_count)) +
+  geom_col() +
+  coord_flip()
+
+Fig1d_2 <- EndoAtlas_meta %>%
+  group_by(sample, AnnotationMain) %>%
+  summarise(row_count = n()) %>%
+  ggplot(aes(x = AnnotationMain, y = row_count)) +
+  geom_boxplot() +
+  scale_y_log10() +
+  coord_flip() +
[…]

### 2025-03-06 Figures/02_Generation_Figure2.R

@@ -1,0 +1,319 @@
+# Aim : Code used for the generation of figure 2
+# Author: Lea Duempelmann
+
+## ---------------------------------- ##
+## Setup
+## ---------------------------------- ##
+
+# load librairies
+library(Seurat)
+library(dplyr)
+library(tibble)
+library(stringr)
+library(ggplot2)
+library(cowplot)
+library(ggbreak) 
+library(tidyr)
+library(patchwork)
+library(ComplexUpset)
+library(ggrepel)
+library(Polychrome)
+library(ComplexHeatmap)
+library(gplots)
+library(RColorBrewer)
+library(randomcoloR)
+
+## ---------------------------------- ##
+## Figure 2.a
+## ---------------------------------- ##
+
+p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
+#Figure 2a
+biplot(p,
+       colby = 'MenstrualCyclePhase',
+       # ellipse config
+       ellipse = TRUE,
+       ellipseType = 't',
+       ellipseLevel = 0.90,
+       ellipseFill = TRUE,
+       ellipseAlpha = 1/4,
+       ellipseLineSize = 1.0,
+       labSize = 0,drawConnectors = FALSE,
+       legendPosition = 'top', legendLabSize = 16, legendIconSize = 8.0)
+
+## ---------------------------------- ##
+## Figure 2.b
+## ---------------------------------- ##
+
+EndoAtlas <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas.rds")
+EndoAtlas_meta <- EndoAtlas@meta.data
+
+#Figure 2b entire atlas
+Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
+Fig2b_1
+ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
+       plot = Fig2b_1,
+       width =20, height=9)
+
+#Figure 2b epithelial cells
+#@ Lea add how this object was generated EndoAtlas_epithelial_reintegrated
+Fig2b_2 <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
+
[…]

### 2025-03-06 Figures/03_Generation_Figure3.R

@@ -1,0 +1,157 @@
+# Aim : Code used for the generation of figure 3
+# Author: Lea Duempelmann
+
+## ---------------------------------- ##
+## Setup
+## ---------------------------------- ##
+
+# load librairies
+library(Seurat)
+library(dplyr)
+library(tibble)
+library(stringr)
+library(ggplot2)
+library(cowplot)
+library(ggbreak) 
+library(tidyr)
+library(patchwork)
+library(ComplexUpset)
+library(ggrepel)
+library(Polychrome)
+library(ComplexHeatmap)
+library(gplots)
+library(RColorBrewer)
+library(randomcoloR)
+
+## ---------------------------------- ##
+## Figure 3.a
+## ---------------------------------- ##
+
+##Figure 3a
+#ENDO/E044/A033/PaperFigures.Rmd
+#Clean up!! use DEG and GO table as basis for plots
+#Barplot unique genes per main cell type
+RNAlistRDS005_df %>% 
+  select(Symphony_Global_main5, gene, stage) %>%
+  distinct() %>%
+  ggplot(aes(x = stage, fill = factor(Symphony_Global_main5, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
+  geom_bar() +
+  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.3))
+ggsave(paste0(dir_out,"BarPlot_DEGs_main4.pdf"), 
+       plot = last_plot(), 
+       height = 4, width = 10)
+
+## ---------------------------------- ##
+## Figure 3.b
+## ---------------------------------- ##
+
+##Figure 3b
+#Clean up!! use DEG and GO table as basis for plots
+#ENDO/E044/A031/ExtractInterestingGenesFromGProfilerA031_RNA_AveExpr18_53202.Rmd
+gProfiler_df %>% 
+  #dplyr::select(p_value, term_size, query_size, intersection_size, term_id, term_name) %>% 
+  dplyr::filter(term_size > 25 & term_size < 750) %>% #larger terms are likely to be excessively broad and prone to false positives
+  distinct(cluster_id,analysis,assay,grade,term_name, .keep_all = TRUE) %>%
+  #filter(grade == "all")  %>%
+  filter(!grepl(pattern = "HPA|HP|CORUM|MIRNA", x = term_id)) %>%
+  dplyr::arrange(p_value) %>% 
+  #dplyr::slice_head(n = 20) %>% 
[…]

### 2025-03-06 Figures/05_Generation_Supplementary_Figure1.R

@@ -1,0 +1,119 @@
+## ---------------------------------- ##
+##  Setup
+## ---------------------------------- ##
+
+# load librairies
+library(ggplot2)
+library(rlang)
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(cowplot)
+library(gridExtra)
+library(patchwork)
+library(tibble)
+library(tinter)
+library(UpSetR)
+library(forcats)
+
+
+## ---------------------------------- ##
+##  Figure 1.a & b 
+## ---------------------------------- ##
+
+#Supplementary Data Figure 1a and 1b
+#CODE OK
+
+VlnPlot(EndoAtlas,
+        features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"),
+        group.by='EndometriosisStatus', ncol = 1, pt.size=0)
+ggsave(paste0(dir_out, "VlnPlot_GroupByEndometriosisStatus_nFeature_nCounts_RNA_SCT_percMT.pdf"),
+       plot = last_plot(),
+       width = 4, height = 13)
+
+#Percentage of main 5
+table <- as.data.frame(table(EndoAtlas_meta$AnnotationMain))
+table$Percentage <- round((table$Freq / sum(table$Freq)) * 100, 1)
+
+#Number of cells for CTL and Endo
+table(EndoAtlas_meta$EndometriosisStatus)
+
+#Number of patients for CTL and Endo
+EndoAtlas_meta %>% select(sample,EndometriosisStatus) %>%
+  distinct() %>%
+  group_by(EndometriosisStatus) %>%
+  summarize(number_of_rows = n())
+
+#mean number of cells per sample
+mean(table(EndoAtlas_meta$sample))
+
+#median count/feature
+median_summary <- EndoAtlas_meta %>%
+  select(nCount_RNA, nCount_SCT, nFeature_RNA, nFeature_SCT) %>%
+  summarize(
+    median_nCount_RNA = median(nCount_RNA),
+    median_nCount_SCT = median(nCount_SCT),
+    median_nFeature_RNA = median(nFeature_RNA),
+    median_nFeature_SCT = median(nFeature_SCT)
+  )
+
+## ---------------------------------- ##
+##  Figure 1.c
+## ---------------------------------- ##
+
+#Supplementary Data Figure 1c
+#CODE OK
+
+SFig1c <- FeaturePlot(EndoAtlas, features = "CycleDay", raster = FALSE)
+
[…]

### 2025-03-06 Figures/06_Generation_Supplementary_Figure2.R

@@ -1,0 +1,145 @@
+## ---------------------------------- ##
+##  Setup
+## ---------------------------------- ##
+
+# load librairies
+library(ggplot2)
+library(rlang)
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(cowplot)
+library(gridExtra)
+library(patchwork)
+library(tibble)
+library(tinter)
+library(UpSetR)
+library(forcats)
+
+# input path 
+input_path<- '../Tables' 
+  
+# source common functions 
+source('./common_functions.R')
+
+## ---------------------------------- ##
+##  Supplementary Data Figure 2
+## ---------------------------------- ##
+
+#CODE OK
+
+# Set SCT as default and remove RNA to save space
+DefaultAssay(EndoAtlas) <- "SCT"
+EndoAtlas@assays$RNA <- NULL
+gc()
+# Remove data without annotation (no NAs!)
+Idents(EndoAtlas) <- EndoAtlas$AnnotationMain
+EndoAtlas <- subset(EndoAtlas, subset = AnnotationMain %in% levels(as.factor(AnnotationMain$Symphony_Refined)))
+gc()
+
+### AnnotationMain markers
+genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2") #correct xxx
+pops_to_plot <- c("stromal","epithelial","myeloid","lymphocyte","endothelial")
+# DotPlot
+Seurat::DotPlot(
+  object = EndoAtlas,
+  scale = T,
+  features = genes_to_plot,
+  idents = pops_to_plot,
+) + theme(axis.text.x = element_text(angle = 90))
+# Save DotPlot
+ggsave(plot = last_plot(), "./SFig2_AnnotationMain.pdf", width = 10, height = 5)
+
+
+###Set identity to AnnotationRefined
+Idents(EndoAtlas) <- EndoAtlas$AnnotationRefined
+
+### DC markers (from Tan2022 Fig. 5a)
+genes_to_plot <- c("CD1C", "CLEC9A", "LAMP3")
+pops_to_plot <- c("mDC", "cDC1", "DC3", "cDC2", "pre-cDC2")
+# reorder clusters
[…]

### 2025-03-06 Figures/07_Generation_Supplementary_Figure3.R

@@ -1,0 +1,228 @@
+## ---------------------------------- ##
+##  Setup
+## ---------------------------------- ##
+
+# load librairies
+library(ggplot2)
+library(rlang)
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(cowplot)
+library(gridExtra)
+library(patchwork)
+library(tibble)
+library(tinter)
+library(UpSetR)
+library(forcats)
+
+
+## ---------------------------------- ##
+##  Supplementary figure 3.a
+## ---------------------------------- ##
+#Supplementary Figure 3a
+#CODE running, need to extend the PCA analysis folder
+#dir_out <- "/home/common/data/output/projects/ENDO/E044/A072/allcells_EndoAtlas_SCTassay/"
+p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
+
+#Supplementary Data Figure 3a
+biplot(p,
+       colby = 'EndometriosisStatus',
+       labSize = 0,drawConnectors = FALSE, 
+       hline = 0, vline = 0,
+       legendPosition = 'right')
+
+## ---------------------------------- ##
+##  Supplementary figure 3.b & e
+## ---------------------------------- ##
+
+#Supplementary Figure 3b and e
+#CODE running
+epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))
+
+#Menstrual Cycle Day
+SFig3b <- plot_cells(epithelial_cells_monocle3, color_cells_by = "CycleDay",
+                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
+                     graph_label_size=3, cell_size = 1)
+SFig3b
+ggsave(filename = paste0(dir_out, 'SFig3b.pdf'),
+       plot = SFig3b,
+       width =20, height=9)
+
+
+#Menstual Cycle Phase Wang
+SFig3e <- plot_cells(epithelial_cells_monocle3 , color_cells_by = "WangMenstrualCyclePhase",
+                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
+                     graph_label_size=3, cell_size = 1)
+SFig3e
+ggsave(filename = paste0(dir_out, 'SFig3e.pdf'),
+       plot = SFig3e,
[…]

### 2025-03-06 Figures/08_Generation_Supplementary_Figure4.R

@@ -1,0 +1,117 @@
+## ---------------------------------- ##
+##  Setup
+## ---------------------------------- ##
+
+# load librairies
+library(ggplot2)
+library(rlang)
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(cowplot)
+library(gridExtra)
+library(patchwork)
+library(tibble)
+library(tinter)
+library(UpSetR)
+library(forcats)
+
+
+## ---------------------------------- ##
+##  Figure 4.a
+## ---------------------------------- ##
+
+#Supplementary Data Figure 4a
+#CODE OK
+
+##Barplots of cell type frequency split by the main 5 cell types
+
+#Set colour palette
+P63 = createPalette(63, c("#ff0000", "#00ff00", "#0000ff")) 
+swatch(P63)
+names(P63) <- NULL
+
+
+#Make df with number of AnnotationRefined cells in each sample
+df <- EndoAtlas_meta%>%
+  filter(!is.na(AnnotationMain)) %>%
+  group_by(sample, AnnotationMain, AnnotationRefined, EndometriosisGrade, MenstrualCyclePhase, epithelial.pseudotime, EndometriosisStatus) %>%
+  summarise(row_count = n()) %>%
+  ungroup() 
+
+#Make bar plot
+p <- ggplot(df, aes(x = sample, y = row_count, fill = AnnotationRefined)) +
+  geom_bar(stat = 'identity', position = 'fill') +
+  scale_y_continuous(expand = c(0, 0)) +
+  scale_fill_manual(values = sample(P63)) +
+  xlab('Sample ID') +
+  theme(
+    axis.text.x = element_text(angle = 90, hjust = 1),
+    panel.grid.minor = element_blank(),
+    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
+  ) +
+  facet_grid(AnnotationMain ~ ., scales = "free_y")
+p
+#Save barplot
+ggsave(paste0(dir_out, "SFig4a_barplot.pdf"),
+       plot = p,
+       width = 15, height = 6)
+
+## ---------------------------------- ##
+##  Figure 4.b
+## ---------------------------------- ##
+
+#Supplementary Data Figure 4b
+#CODE OK
+
+##########################################################################################
[…]

### 2025-03-06 Figures/09_Generation_Supplementary_Figure5.R

@@ -1,0 +1,43 @@
+## ---------------------------------- ##
+##  Setup
+## ---------------------------------- ##
+
+# load librairies
+library(ggplot2)
+library(rlang)
+library(stringr)
+library(dplyr)
+library(tidyr)
+library(cowplot)
+library(gridExtra)
+library(patchwork)
+library(tibble)
+library(tinter)
+library(UpSetR)
+library(forcats)
+
+## ---------------------------------- ##
+##  Supplemental Figure 5.a
+## ---------------------------------- ##
+
+#Supplementary Data Figure 5a @Lea to do
+
+## ---------------------------------- ##
+##  Supplemental Figure 5.b-d
+## ---------------------------------- ##
+
+#Figure 3c and Supplementary Figure 5b
+gk1 <- rankNet(cellchat, mode = "comparison", stacked = T, do.stat = TRUE)
+gk1 
+ggsave(plot = gk1,
+       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12)
+
+
+#Load CellChat data
+cellchat <- readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
+
+#Supplementary Data Figure 5d TNF
+netVisual_bubble(cellchat, comparison = c(1, 2),signaling = c("TNF"), remove.isolate = FALSE) 
+ggsave(plot = last_plot(),
+       filename = paste0(dir_out, 'output_comparison_Refined/netVisual_bubble_TNF.pdf'), width =25,height=5)
+

### 2025-03-06 Figures/10_Generation_Supplementary_Figure6.R

@@ -26,19 +26,74 @@ source('./common_functions.R')
 ##  Figure 6.a
 ## ---------------------------------- ##
 
-## to do lea 
+# @Lea to check 
+#Supplementary Data Figure 6a
+# reshape df
+melted_meta <- melt(meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
+## create the bar plot for endo_EndometriosisGrade
+SFig6a <- ggplot(melted_meta, aes(x = variable, fill = endo_grade)) +
+  geom_bar(position = "stack", width = 0.7) +
+  facet_wrap(~ fct_rev(value), ncol = 2) +
+  labs(title = "Bar Plot of Samples by Endo EndometriosisGrade",
+       x = "Fold",
+       y = "Number of Samples") +
+  scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
+  theme_minimal() +
+  coord_flip()
 
 ## ---------------------------------- ##
 ##  Figure 6.b
 ## ---------------------------------- ##
 
-## to do lea 
+# @lea to check 
+#Supplementary Data Figure 6b
+SFig6b <- ggplot(melted_meta, aes(x = variable, fill = lane)) +
+  geom_bar(position = "stack", width = 0.7) +
+  facet_wrap(~ fct_rev(value), ncol = 2) +
+  labs(title = "Bar Plot of Samples by lane",
+       x = "Fold",
+       y = "Number of Samples") +
+  #scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
+  theme_minimal() +
+  coord_flip()
+SFig6b
 
 ## ---------------------------------- ##
 ##  Figure 6.c
 ## ---------------------------------- ##
 
-## to do lea 
+#@lea to check 
+#Supplementary Data Figure 6c
+
+metadata <- read.csv('/home/common/data/output/projects/ENDO/E042/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv')
+metadata$new_column <- NULL
+metadata$batch <- NULL
+
+
+# Remove 'names' column as it will be used as row labels
+row_names <- metadata$names
+metadata <- metadata[, -1]
+
+# Convert non-numeric columns to factors
[…]

### 2025-03-06 Figures/11_Generation_Supplementary_Figure7.R

@@ -22,19 +22,41 @@ input_path <- '../Tables'
 source('./common_functions.R')
 
 ## ---------------------------------- ##
-##  Figure 6.a
+##  Figure 7.a
 ## ---------------------------------- ##
 
-## TO do lea 
+##@Lea check 
+#LTs annotation for cycling fibroblasts
+data <- readRDS("/home/common/data/output/projects/ENDO/E031/A004/proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated.RDS")
+meta <- data@meta.data
+rm(data)
+gc()
+table(meta$Annotation, meta$EndometriosisStatus)
+
+#Make cell id compatible with new data set
+meta$cell_id <- paste(meta$sample, meta$cell.names, sep = '_')
+rownames(meta) <- meta$cell_id
+
+#Load entire dataset and add LTs annotation
+data_ds <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
+data_ds <- AddMetaData(data_ds,meta[,c("Annotation", "scds.cxds.score")])
+DimPlot(data_ds_new,group.by = "Annotation")
+
+#Highlight cycling fibroblasts
+Idents(data_ds) <- data_ds$Annotation
+table(data_ds$Annotation)
+cyclingfibs <- WhichCells(data_ds, idents = c("Cycling Fibroblasts"))
+DimPlot(data_ds, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
+
 
 ## ---------------------------------- ##
-##  Figure 6.b
+##  Figure 7.b
 ## ---------------------------------- ##
 
 # this figure represent a scheme and was done manually 
 
 ## ---------------------------------- ##
-##  Figure 6.c
+##  Figure 7.c
 ## ---------------------------------- ##
 
 #1. Get plot for evaluation
@@ -52,7 +74,7 @@ list_plot_evaluation_auc <- lapply(list_plot_evaluation_auc,function(x){x + ggti
 plot_grid(plotlist = c(list_plot_evaluation_auc),nrow = 1)
 
 ## ---------------------------------- ##
-##  Figure 6.d
+##  Figure 7.d
 ## ---------------------------------- ##
 
 validation_auc_cf <- read.csv(file.path(input_path,'Performance_cycling_fibroblasts_signature_confirmation.csv')) 
[…]

### 2025-03-06 05b_Functional_enrichment_analysis/DEG_gProfiler.Rmd
cleanup analyses code

@@
 
-dir_out <- "/home/common/data/output/projects/ENDO/E044/A031/"
+dir_out <- "/endometriosis_endometrium_scRNA_atlas/_Data/05b_Functional_enrichment_analysis/"
 setwd(dir_out)
+
+# define menstrual cycle phase 
+MenstrualCyclePhase = ""
+
+# define annotation level 
+Annotation = ""
+
 ```
-
 Determine inflection point based on unfiltered DEGs
 ```{r}
+
 ##Combine muscat output into one dataframe
-PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
+PATHS <- list.files(path = paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/",MenstrualCyclePhase,"_", Annotation), pattern = "mm_dream.rds", recursive = TRUE, full.names = TRUE)
 RNAlistRDS <- lapply(PATHS, readRDS)
@@
 ```
-
 summarize DEGs into df and plot
@@
 #RNA assay
-PATHS <- list.files(path = paste0("/home/common/data/output/projects/ENDO/E044/A030"), pattern = "res_mm_dream.rds", recursive = TRUE, full.names = TRUE)
+PATHS <- list.files(path = paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/",MenstrualCyclePhase,"_", Annotation), pattern = "mm_dream.rds", recursive = TRUE, full.names = TRUE)
 print(PATHS)
@@
 RNAlistRDS005 <- as.data.frame(do.call(rbind, RNAlistRDS005))
+
 #Check for cluster_ids that have or don't have DEGs with p_adj.loc < 0.05
@@
 setdiff(levels(factor(RNAlistRDSdf$cluster_id)),levels(factor(RNAlistRDS005$cluster_id)))
+
 #Add columns stage, clustering, annotation_reference
@@
 
-write.csv(RNAlistRDS005,"/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv")
-#RNAlistRDS005 <- read.csv("/home/common/data/output/projects/ENDO/E044/A031/RNAlistRDS005_unmerged.csv", row.names = 1)
+write.csv(RNAlistRDS005,file.path(dir_out,"RNAlistRDS005_unmerged.csv"))
 
 ```
-
 gProfiler based on RNA assay and per annotation with gene expression >18.53202

### 2025-03-06 05b_Functional_enrichment_analysis/DEG_gProfiler_example.R

@@ -1,0 +1,116 @@
+## ---------------------------------------- ##
+## Setup
+## ---------------------------------------- ##
+
+# load librairies
+library(gprofiler2)
+library(plyr)
+library(dplyr)
+library(ggplot2)
+library(tibble)
+library(cowplot)
+library(ggrepel)
+
+# define paths
+dir_out <- "/endometriosis_endometrium_scRNA_atlas/_Data/05b_Functional_enrichment_analysis/"
+path_degs_table <- ""
+setwd(dir_out)
+
+## ---------------------------------------- ##
+## Identify inflection point to filter out
+## lowly expressed genes 
+## ---------------------------------------- ##
+
+##Combine muscat output into one dataframe
+PATHS <- list.files(path = paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/",MenstrualCyclePhase,"_", Annotation), pattern = "mm_dream.rds", recursive = TRUE, full.names = TRUE)
+RNAlistRDS <- lapply(PATHS, readRDS)
+names(RNAlistRDS) <- print(PATHS)
+RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
+RNAlistRDS <- lapply(RNAlistRDS,plyr::rbind.fill)
+RNAlistRDSdf <- as.data.frame(do.call(rbind, RNAlistRDS))
+
+##exclude ribosomal genes (comment: does not change the inflection point)
+rb.genes_DEGs <- unique(grep(pattern = "^RP[SL]", RNAlistRDSdf$gene, value=TRUE)) #around 150 unique rb genes
+RNAlistRDSdf <- RNAlistRDSdf[!(RNAlistRDSdf$gene %in% rb.genes_DEGs), ]
+
+##Determine inflection point
+#Determine Inflection Point, No log2 scale, bw = 0.02
+plot(density(RNAlistRDSdf$AveExpr))
+# Create the density plot and store it in an object
+density_plot <- density(RNAlistRDSdf$AveExpr, bw = 0.02) #good values 0.025 or 0.02, 0.01 is too small
+plot(density_plot)
+# Calculate the second derivative of the density curve
+second_derivative <- diff(diff(density_plot$y)) / diff(density_plot$x)^2
+# Find where the second derivative crosses zero (inflection point)
+inflection_points <- density_plot$x[which(diff(sign(second_derivative)) != 0)]
+# Print or plot the inflection point(s)
[…]

### 2025-03-06 07_Ligand_receptor_analysis/01_cellchat_prep.R

@@ -8,13 +8,11 @@ library(Seurat)
 options(stringsAsFactors = FALSE)
 
 
-#dir_out <- "/home/common/data/output/projects/ENDO/E044/A087/"
-#dir.create(paste0(dir_out,"output/"), recursive = TRUE)
-#setwd(dir_out)
-# data <- /home/common/data/output/projects/ENDO/E044/A033/EndoAtlas_GEO_complete.rds
-# meta <- readRDS("/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas_meta.rds")
-data <- AddMetaData(data,meta)
+dir_out <- "/endometriosis_endometrium_scRNA_atlas/_Data/07_Ligand_receptor_analysis"
+dir.create(paste0(dir_out,"output/"), recursive = TRUE)
+setwd(dir_out)
 
+path_seurat_object <- "/endometriosis_endometrium_scRNA_atlas/_Data/ENDO_global.rds"
 dir_out <- "CellChat/"
 setwd(dir_out)
 
@@ -23,7 +21,7 @@ setwd(dir_out)
 ####################
 
 ##Load seurat object and add metadata
-data  <- readRDS("ENDOatlas.rds")
+data  <- readRDS(path_seurat_object)
 
 ##Filter Seurat for proliferative phase samples with stringet exclusion criteria and annotated cells
 DEGsamples <- unique(data@meta.data %>% filter(DEG.analysis..Figure.3a.b. == "TRUE") %>% pull(sample))

### 2025-03-06 07_Ligand_receptor_analysis/02_cellchat_comparison.R

@@ -6,7 +6,7 @@ library(CellChat)
 library(paFibroblastshwork)
 options(stringsAsFactors = FALSE)
 
-dir_out <- "A076/"
+dir_out <- '/endometriosis_endometrium_scRNA_atlas/_Data/07_Ligand_receptor_analysis'
 
 
 ####################################

### 2025-03-07 01_Atlas_generation/Integration/01b_run_normalization_split.R
Adapted paths

@@ -12,7 +12,7 @@ suppressPackageStartupMessages({
 
 
 # load seurat object
-seurat_obj <- readRDS(./seurat.rds)
+seurat_obj <- readRDS(../_Data/01_Atlas_generation/AlevinQC/seurat.rds)
 Idents(seurat_obj) <- seurat_obj@meta.data$sample
 
 ####################################
@@ -32,6 +32,6 @@ seurat_obj <- SCTransform(seurat_obj,
 seurat_obj_split <- SplitObject(seurat_obj, split.by = "batch")
 
 # save split seurat object
-saveRDS(seurat_obj_split, "./seurat_obj_split.rds")
+saveRDS(seurat_obj_split, "../_Data/01_Atlas_generation/Integration/seurat_obj_split.rds")
 sessionInfo()
 date()

### 2025-03-07 01_Atlas_generation/Integration/02_run_seuratv3.R

@@ -14,11 +14,8 @@ suppressPackageStartupMessages({
 })
 
 # load split seurat object
-seurat_obj_split_input <- readRDS("./seurat_obj_split.rds")
+seurat_obj_split_input <- readRDS("../_Data/01_Atlas_generation/Integration/seurat_obj_split.rds")
 print("input loaded")
-
-# specify output directory
-dir_out <- snakemake@config$integration_output_dir
 
 ####################################
 ## 01. Run integration
@@ -74,7 +71,7 @@ integrated_data <- RunUMAP(integrated_data,  dims = 1:30, seed.use = seed)
 #                                      dimreducs = c("pca", "umap"))
 
 # Save the filtered, normalized and integrated data
-saveRDS(integrated_data, "./seurat_integrated.rds")
+saveRDS(integrated_data, "../_Data/EndoAtlas.rds")
 print("saved seurat_integrated.rds")
 
 sessionInfo()

### 2025-03-07 01_Atlas_generation/Integration_epithelial_only/01b_run_normalization_split.R

@@ -12,7 +12,7 @@ suppressPackageStartupMessages({
 
 
 # load seurat object
-seurat_obj <- readRDS(./seurat.rds)
+seurat_obj <- readRDS("../_Data/EndoAtlas.rds")
 
 # Subset epithelial cells
 Idents(seurat_obj) <- seurat_obj@meta.data$AnnotationMain
@@ -44,6 +44,6 @@ seurat_obj <- SCTransform(seurat_obj,
 seurat_obj_split <- SplitObject(seurat_obj, split.by = "batch")
 
 # save split seurat object
-saveRDS(seurat_obj_split, "./seurat_obj_split.rds")
+saveRDS(seurat_obj_split, "../_Data/01_Atlas_generation/Integration_epithelial_only/seurat_obj_split.rds")
 sessionInfo()
 date()

### 2025-03-07 01_Atlas_generation/Integration_epithelial_only/02_run_seuratv3.R

@@ -14,7 +14,7 @@ suppressPackageStartupMessages({
 })
 
 # load split seurat object
-seurat_obj_split_input <- readRDS("./seurat_obj_split.rds")
+seurat_obj_split_input <- readRDS("../_Data/01_Atlas_generation/Integration_epithelial_only/seurat_obj_split.rds")
 print("input loaded")
 
 # specify output directory
@@ -69,12 +69,8 @@ integrated_data <- RunPCA(object = integrated_data,
 # Run UMAP
 integrated_data <- RunUMAP(integrated_data,  dims = 1:30, seed.use = seed)
 
-#Save seurat
-#integrated_data_diet <- Seurat::DietSeurat(integrated_data, assays = c("SCT","integrated"),
-#                                      dimreducs = c("pca", "umap"))
-
 # Save the filtered, normalized and integrated data
-saveRDS(integrated_data, "./seurat_integrated.rds")
+saveRDS(integrated_data, "../_Data/01_Atlas_generation/Integration_epithelial_only/epithelial_cells.rds")
 print("saved seurat_integrated.rds")
 
 sessionInfo()

### 2025-03-07 02_Cell_annotation/01_symphony.R

@@ -4,17 +4,19 @@
 
 # load packages
 suppressPackageStartupMessages({
-  source('/endometriosis_endometrium_scRNA_atlas/02_Cell_annotation/01_libs.R')# imports
+  source('../02_Cell_annotation/01_libs.R')# imports
 }) 
 
 #source utils
-source('/endometriosis_endometrium_scRNA_atlas/02_Cell_annotation/01_utils.R') # color definitions, symphony function and plotting functions
-
+source('../02_Cell_annotation/01_utils.R') # color definitions, symphony function and plotting functions
 
 
 ########################################################
 ##### 01. Run Symphony on entire EndoAtlas
 #######################################################
+
+#create first following symbolic link in Linux environment:
+#ln -s ../_Data/EndoAtlas.rds ../_Data/02_Cell_annotation/ENDO_global.rds
 
 #00 global annotation
 symphony(
@@ -28,7 +30,7 @@ symphony(
 #######################################################
 
 #Load integrated EndoAtlas
-data <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/ENDO_global.rds")
+data <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas.rds")
 Idents(data) <- data$AnnotationMain
 
 # Define the vector of annotations
@@ -38,7 +40,7 @@ for (annotation in ANNOTATIONMAIN) {
   # Subset data based on the current annotation
   AnnotationMain_subset <- subset(data, idents = annotation)
   # Save the subsetted data to an RDS file
-  saveRDS(AnnotationMain_subset, paste0("/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/", annotation, ".rds"))
+  saveRDS(AnnotationMain_subset, paste0("../_Data/02_Cell_annotation/ENDO_",annotation,".rds"))
 }
 
 

### 2025-03-07 02_Cell_annotation/01_utils.R

@@ -37,7 +37,7 @@ symphony <- function(CELLTYPEref,
   ######################################################## 
   
   #create directory
-  dir_out <- paste0("/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/",CELLTYPEref,"ref_",CELLTYPEquery,"query","/")
+  dir_out <- paste0("../_Data/02_Cell_annotation/",CELLTYPEref,"ref_",CELLTYPEquery,"query","/")
   dir.create(dir_out, recursive = TRUE)
   setwd(dir_out)
   
@@ -46,7 +46,7 @@ symphony <- function(CELLTYPEref,
   ########################################################
   
   #Load ref data set
-  cells_ref <- readRDS(paste0("/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/",CELLTYPEref,".rds"))
+  cells_ref <- readRDS(paste0("../_Data/02_Cell_annotation/Tan_",CELLTYPEref,".rds"))
   cells_ref@meta.data$dataset <- "ref"
   cells_ref <- NormalizeData(cells_ref, normalization.method = "LogNormalize", scale.factor = 10000, assay = "RNA")
   ref_exp_full <- cells_ref@assays$RNA@data
@@ -56,7 +56,7 @@ symphony <- function(CELLTYPEref,
   ref_metadata <- setnames(ref_metadata, old = c("subtypes", "phase"), new = c("cell_type", "batch"))
   
   #Load query data set
-  cells_query <- readRDS(paste0(dir_out,"/EndoAtlas_",CELLTYPEquery,".rds"))
+  cells_query <- readRDS(paste0(dir_out,"/ENDO_",CELLTYPEquery,".rds"))
   DefaultAssay(cells_query) <- "RNA"
   cells_query <- DietSeurat(cells_query, assay="RNA", dimreducs = "umap")
   cells_query@meta.data$dataset <- "query"
@@ -170,7 +170,7 @@ symphony <- function(CELLTYPEref,
   fig.size(3, 5)
   plotBasic(umap_labels_query, title = 'Reference', color.by = 'cell_type_pred_knn')
   
-  saveRDS(query$meta_data, paste0("prolif_secretory_",CELLTYPEquery,"_HarmonySymphony.RDS"))
+  saveRDS(query$meta_data, paste0("prolif_secretory_",CELLTYPEquery,"_Symphony.RDS"))
   ########################################################
   ##### 04. Visualization of mapping
   #########################################################

### 2025-03-07 02_Cell_annotation/02_MenstrualCycleSpecificCelltypes.R

@@ -13,7 +13,7 @@ library(ComplexUpset)
 library(ggrepel)
 library(PCAtools)
 library(scCustomize)
-dir_out <- "/endometriosis_endometrium_scRNA_atlas/_Data/02_Cell_annotation/"
+dir_out <- "../_Data/02_Cell_annotation/"
 dir.create(dir_out, recursive = TRUE)
 setwd(dir_out)
 
@@ -21,17 +21,17 @@ setwd(dir_out)
 #### 01. Load data and plot
 #######################
 ###Load data
-data <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/01_Atlas_generation/Integration/ENDOatlas.rds")
+data <- readRDS("../_Data/EndoAtlas.rds")
 
 ###Plotting for overview
 DimPlot_scCustom(data, reduction = "umap", group.by = "MenstrualCyclePhase")
-DimPlot_scCustom(data, reduction = "umap", group.by = "Symphony_Global_main5")
-DimPlot_scCustom(data, reduction = "umap", group.by = "Symphony_Refined")
+DimPlot_scCustom(data, reduction = "umap", group.by = "AnnotationMain")
+DimPlot_scCustom(data, reduction = "umap", group.by = "AnnotationRefined")
 DimPlot_scCustom(data, reduction = "umap", group.by = "sample", split.by = "MenstrualCyclePhase")
 #Highlight samples 65 and 130 (they are in the transition between periovulatory and secretory menstual cycle phase)
 Idents(data) <- data$sample
 cellshighlight <- WhichCells(data, idents = c("sample65","sample130"))
-DimPlot(data, label=T,reduction = "umap", group.by = "treatment", cells.highlight = cellshighlight, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
+DimPlot(data, label=T,reduction = "umap", group.by = "EndometriosisStatus", cells.highlight = cellshighlight, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
 
 ############################################################################
 ############################################################################
@@ -57,7 +57,7 @@ DimPlot_scCustom(data, reduction = "umap", split.by = "integrated_snn_res.1",gro
 ########################
 #### 03. Menstrual cycle specific mesenchymal cells
 #######################
-Idents(data) <- data$Symphony_Global_main5
+Idents(data) <- data$AnnotationMain
[…]

### 2025-03-07 03_PCA_analysis/PCAtools.R

@@ -9,8 +9,7 @@ library(Seurat)
 
 
 ##Set paths
-dir_ENDO <- "/endometriosis_endometrium_scRNA_atlas/"
-dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")
+dir_out <- "../_Data/03_PCA_analysis/")
 dir.create(dir_out, recursive = TRUE)
 
 
@@ -18,7 +17,7 @@ dir.create(dir_out, recursive = TRUE)
 ### Load Data
 ###############################################
 ##Load Data
-EndoAtlas <- readRDS(paste0(dir_ENDO, "_Data/EndoAtlas.rds"))
+EndoAtlas <- readRDS("../_Data/EndoAtlas.rds"))
 
 
 ###############################################

### 2025-03-07 04_Time_trajectory_analysis/trajectoryAnalysis_monocle3.r

@@ -14,8 +14,8 @@ library(dplyr)
 library(ggplot2)
 
 ##Set paths
-analysisFolder <- "endometriosis_endometrium_scRNA_atlas/_Data/04_Time_trajectory_analysis/"
-rawData <- "/endometriosis_endometrium_scRNA_atlas/_Data/01_Atlas_generation/Integration_epithelial_only/EndoAtlas_epithelial.rds"
+analysisFolder <- "../_Data/04_Time_trajectory_analysis/"
+rawData <- "../_Data/01_Atlas_generation/Integration_epithelial_only/epithelial_cells.rds"
 metadataFile <- ""
 subsetClusters <- c()  
 #-------------
@@ -27,25 +27,25 @@ data <- readRDS(rawData)
 
 if(metadataFile != ""){
   metadata <- readRDS(metadataFile)
-  data@meta.data$Symphony_Refined <- metadata[rownames(data@meta.data), "Symphony_Refined"]
-  data@meta.data$CyclePhase <- metadata[rownames(data@meta.data), "CyclePhase"]
+  data@meta.data$AnnotationRefined <- metadata[rownames(data@meta.data), "AnnotationRefined"]
+  data@meta.data$MenstrualCyclePhase <- metadata[rownames(data@meta.data), "MenstrualCyclePhase"]
   data@meta.data$CycleDay <- metadata[rownames(data@meta.data), "CycleDay"]
 }
 
 #Plots
 DefaultAssay(data) <- "SCT"
-Idents(data) <- data@meta.data$Symphony_Refined
+Idents(data) <- data@meta.data$AnnotationRefined
 
-pdf(paste0(analysisFolder, "/", out, "/UMAP_Symphony_Refined.pdf"))
-  DimPlot(data, reduction = "umap", label=TRUE, group.by="Symphony_Refined") + labs(title="Symphony_Refined")
+pdf(paste0(analysisFolder, "/", out, "/UMAP_AnnotationRefined.pdf"))
+  DimPlot(data, reduction = "umap", label=TRUE, group.by="AnnotationRefined") + labs(title="AnnotationRefined")
 dev.off()
 
 pdf(paste0(analysisFolder, "/", out, "/UMAP_CycleDay.pdf"))
   DimPlot(data, reduction = "umap", label=TRUE, group.by="CycleDay") + labs(title="CycleDay")
 dev.off()
 
-pdf(paste0(analysisFolder, "/", out, "/UMAP_CyclePhase.pdf"))
[…]

### 2025-03-07 05a_DEG_analysis_muscat/utils.R

@@ -3,7 +3,7 @@
 ####################################
 
 ##Set paths
-dir_out <- paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/")
+dir_out <- paste0("../_Data/05a_DEG_analysis_muscat/")
 dir.create(dir_out, recursive = TRUE, showWarnings = FALSE)
 
 
@@ -14,7 +14,7 @@ DEGanalysis_muscat <- function(MenstrualCyclePhase, Annotation) {
 ####################################
 
 ##load seurat object 
-data <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas.rds")
+data <- readRDS("../_Data/EndoAtlas.rds")
 print("loaded data")
 
 ##Subset samples by menstrual cycle phase
@@ -42,13 +42,13 @@ if (MenstrualCyclePhase == "Secretory") {
 }
 
 ##Rename required Annotation to 'Annotation'
-if (MenstrualCyclePhase == "AnnotationMain") {
+if (Annotation == "AnnotationMain") {
   data$Annotation <- data$AnnotationMain
 }
-if (MenstrualCyclePhase == "AnnotationRefined") {
+if (Annotation == "AnnotationRefined") {
   data$Annotation <- data$AnnotationRefined
 }
-if (MenstrualCyclePhase == "AnnotationUnited") {
+if (Annotation == "AnnotationUnited") {
   data$Annotation <- data$AnnotationUnited
 }
 

### 2025-03-07 05b_Functional_enrichment_analysis/DEG_gProfiler.R
removed redundant code

@@ -12,8 +12,7 @@ library(cowplot)
 library(ggrepel)
 
 # define paths
-dir_out <- "/endometriosis_endometrium_scRNA_atlas/_Data/05b_Functional_enrichment_analysis/"
-path_degs_table <- ""
+dir_out <- "../_Data/05b_Functional_enrichment_analysis/"
 setwd(dir_out)
 
 ## ---------------------------------------- ##
@@ -22,7 +21,7 @@ setwd(dir_out)
 ## ---------------------------------------- ##
 
 ##Combine muscat output into one dataframe
-PATHS <- list.files(path = paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/",MenstrualCyclePhase,"_", Annotation), pattern = "mm_dream.rds", recursive = TRUE, full.names = TRUE)
+PATHS <- list.files(path = paste0("/endometriosis_endometrium_scRNA_atlas/_Data/05a_DEG_analysis_muscat/Proliferative*/", pattern = "mm_dream.rds", recursive = TRUE, full.names = TRUE)
 RNAlistRDS <- lapply(PATHS, readRDS)
 names(RNAlistRDS) <- print(PATHS)
 RNAlistRDS <- unlist(RNAlistRDS, recursive = FALSE) #combine sub-lists
@@ -91,7 +90,7 @@ run_go_enrichment <- function(query){
 }
 
 go_enrichment_df <- lapply(degs_list,run_go_enrichment(degs_df))
-write.csv(go_enrichment_df,file = file.path(dir_out,'GO_enrichment_summary.csv'),quote=F,row.names=F) # @Lea to add the name of the table if saved 
+write.csv(go_enrichment_df,file = file.path(dir_out,'GO_enrichment_summary.csv'),quote=F,row.names=F)
 
 ##--------------------------------##
 ## Generate summary plot 
@@ -110,7 +109,7 @@ go_enrichment_df %>%
   ggplot2::ylab("enriched terms") + 
   ggplot2::ggtitle("all cells")
 
-ggsave(paste0(dir_out, "GOterms_all_severe.pdf"), # @Lea to add the name of the table if saved 
+ggsave(paste0(dir_out, "GO_enrichment_plot.pdf"),
        plot = last_plot(),
        width = 14, height = 14)
 

### 2025-03-07 07_Ligand_receptor_analysis/01_cellchat_prep.R
adapted paths

@@ -7,13 +7,9 @@ library(patchwork)
 library(Seurat)
 options(stringsAsFactors = FALSE)
 
-
-dir_out <- "/endometriosis_endometrium_scRNA_atlas/_Data/07_Ligand_receptor_analysis"
+path_seurat_object <- "../_Data/EndoAtlas.rds"
+dir_out <- "../_Data/07_Ligand_receptor_analysis/"
 dir.create(paste0(dir_out,"output/"), recursive = TRUE)
-setwd(dir_out)
-
-path_seurat_object <- "/endometriosis_endometrium_scRNA_atlas/_Data/ENDO_global.rds"
-dir_out <- "CellChat/"
 setwd(dir_out)
 
 ####################
@@ -24,7 +20,7 @@ setwd(dir_out)
 data  <- readRDS(path_seurat_object)
 
 ##Filter Seurat for proliferative phase samples with stringet exclusion criteria and annotated cells
-DEGsamples <- unique(data@meta.data %>% filter(DEG.analysis..Figure.3a.b. == "TRUE") %>% pull(sample))
+DEGsamples <- unique(data@meta.data %>% filter(DEG.analysis..Figure.3a.b == "TRUE") %>% pull(sample))
 Idents(data) <- data$sample
 data <- subset(data, idents = DEGsamples) 
 data <- subset(data, subset = AnnotationRefined %in% levels(as.factor(data$AnnotationRefined)))

### 2025-03-07 07_Ligand_receptor_analysis/02_cellchat_comparison.R

@@ -6,7 +6,7 @@ library(CellChat)
 library(paFibroblastshwork)
 options(stringsAsFactors = FALSE)
 
-dir_out <- '/endometriosis_endometrium_scRNA_atlas/_Data/07_Ligand_receptor_analysis'
+dir_out <- '../_Data/07_Ligand_receptor_analysis/'
 
 
 ####################################
@@ -70,7 +70,7 @@ gk1 <- rankNet(cellchat, mode = "comparison", stacked = T, do.stat = TRUE)
 gk2 <- rankNet(cellchat, mode = "comparison", stacked = F, do.stat = TRUE)
 gk1 + gk2
 ggsave(plot = gk1 + gk2,
-       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12)
+       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12) #For paper without previous cell type filtering
 
 ####################################
 #### Plot candidate pathways

### 2025-03-07 07_Ligand_receptor_analysis/02_cellchat_comparison.R
adapted paths and cleaned code

@@ -70,7 +70,7 @@ gk1 <- rankNet(cellchat, mode = "comparison", stacked = T, do.stat = TRUE)
 gk2 <- rankNet(cellchat, mode = "comparison", stacked = F, do.stat = TRUE)
 gk1 + gk2
 ggsave(plot = gk1 + gk2,
-       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12) #For paper without previous cell type filtering
+       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12)
 
 ####################################
 #### Plot candidate pathways

### 2025-03-07 Figures/01_Generation_Figure1.R

@@ -23,15 +23,15 @@ library(gplots)
 library(RColorBrewer)
 library(randomcoloR)
 
+dir_out <- "../_Data/Figures/"
+
 ## ---------------------------------- ##
 ## Figure 1.c
 ## ---------------------------------- ##
 
-#@Lea check these paths are correct 
-EndoAtlas <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas.rds")
-EndoAtlas_meta <- EndoAtlas@meta.data
+EndoAtlas <- readRDS("../_Data/EndoAtlas.rds")
 
-order <- EndoAtlas_meta %>%
+order <- EndoAtlas@meta.data %>%
   select(AnnotationMain, AnnotationRefined) %>%
   filter(!is.na(AnnotationMain)) %>%
   distinct() %>%
@@ -40,19 +40,25 @@ order <- EndoAtlas_meta %>%
   arrange(AnnotationMain)
 EndoAtlas$AnnotationRefined <- factor(x = EndoAtlas$AnnotationRefined, levels = order$AnnotationRefined)
 Fig1c <- DimPlot(EndoAtlas, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
+ggsave(filename = paste0(dir_out, 'Fig1c.pdf'),
+       plot = Fig1c,
+       width =20, height=9)
 
 ## ---------------------------------- ##
 ## Figure 1.d
 ## ---------------------------------- ##
 
-Fig1d_1 <- EndoAtlas_meta %>%
+Fig1d_1 <- EndoAtlas@meta.data %>%
   group_by(AnnotationMain) %>%
   summarise(row_count = n()) %>% 
   ggplot(aes(x = AnnotationMain, y = row_count)) +
   geom_col() +
   coord_flip()
+ggsave(paste0(dir_out, "Fig1d_1.pdf"),
+       plot = last_plot(),
+       width = 15, height = 6)
 
-Fig1d_2 <- EndoAtlas_meta %>%
+Fig1d_2 <- EndoAtlas@meta.data %>%
   group_by(sample, AnnotationMain) %>%
   summarise(row_count = n()) %>%
   ggplot(aes(x = AnnotationMain, y = row_count)) +
@@ -60,8 +66,12 @@ Fig1d_2 <- EndoAtlas_meta %>%
   scale_y_log10() +
   coord_flip() +
[…]

### 2025-03-07 Figures/02_Generation_Figure2.R

@@ -23,11 +23,13 @@ library(gplots)
 library(RColorBrewer)
 library(randomcoloR)
 
+dir_out <- "../_Data/Figures/"
+
 ## ---------------------------------- ##
 ## Figure 2.a
 ## ---------------------------------- ##
 
-p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
+p <- readRDS("../_Data/03_PCA_analysis/p_EndoAtlas.rds")
 #Figure 2a
 biplot(p,
        colby = 'MenstrualCyclePhase',
@@ -45,10 +47,9 @@ biplot(p,
 ## Figure 2.b
 ## ---------------------------------- ##
 
-EndoAtlas <- readRDS("/endometriosis_endometrium_scRNA_atlas/_Data/EndoAtlas.rds")
-EndoAtlas_meta <- EndoAtlas@meta.data
-
 #Figure 2b entire atlas
+#MenstrualCyclePhase
+EndoAtlas <- readRDS("../_Data/EndoAtlas.rds")
 Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
 Fig2b_1
 ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
@@ -56,29 +57,34 @@ ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
        width =20, height=9)
 
 #Figure 2b epithelial cells
-#@ Lea add how this object was generated EndoAtlas_epithelial_reintegrated
-Fig2b_2 <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
+#MenstrualCyclePhase
+epithelial_cells_monocle3 <- readRDS("../_Data/epithelial_cells_monocle3.rds")
+Fig2b_2 <- plot_cells(epithelial_cells_monocle3, color_cells_by = "MenstrualCyclePhase",
+                    label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
+                    graph_label_size=3, cell_size = 1)
+ggsave(filename = paste0(dir_out, 'Fig2b_2.pdf'),
+       plot = Fig2b_2,
+       width =12, height=9)
 
 ## ---------------------------------- ##
 ## Figure 2.c
 ## ---------------------------------- ##
 
-#Figure 2c
[…]

### 2025-03-07 Figures/03_Generation_Figure3.R

@@ -23,6 +23,9 @@ library(gplots)
 library(RColorBrewer)
 library(randomcoloR)
 
+dir_out <- "../_Data/Figures/"
+
+
 ## ---------------------------------- ##
 ## Figure 3.a
 ## ---------------------------------- ##
@@ -74,16 +77,13 @@ ggsave(paste0(dir_out, "GOterms_all_severe.pdf"),
 ## Figure 3.c
 ## ---------------------------------- ##
 
-CellChat
-E044A076
-#cellchat <-readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
-dir_out <- "/home/common/data/output/projects/ENDO/E044/A076/"
+cellchat <- readRDS("../_Data/07_Ligand_receptor_analysis/cellchat_CntrlvsEndo.rds")
+#without previous cell filtering
 
-#Figure 3c and Supplementary Figure 5b
 gk1 <- rankNet(cellchat, mode = "comparison", stacked = T, do.stat = TRUE)
 gk1 
 ggsave(plot = gk1,
-       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12)
+       filename = paste0(dir_out, 'output_comparison_Refined/rankNet.pdf'), width =10,height=12) 
 
 ## ---------------------------------- ##
 ## Figure 3.d
@@ -119,8 +119,9 @@ dev.off()
 ## ---------------------------------- ##
 ## Figure 3.e
 ## ---------------------------------- ##
+
 #Load CellChat data
-cellchat <- readRDS(paste0(dir_out,"cellchat_CntrlvsEndo_Symphony_Refined_Final.rds"))
+cellchat <- readRDS("../_Data/07_Ligand_receptor_analysis/cellchat_CntrlvsEndo.rds")
 
 #Figure 3e ICAM
 netVisual_bubble(cellchat, comparison = c(1, 2),signaling = c("ICAM"), remove.isolate = FALSE) 
@@ -132,6 +133,9 @@ ggsave(plot = last_plot(),
 ## Figure 3.f
 ## ---------------------------------- ##
 
+#Load CellChat data
+cellchat <- readRDS("../_Data/07_Ligand_receptor_analysis/cellchat_CntrlvsEndo.rds")
+
 #Figure 3f CXCL
 #Load data
[…]

### 2025-03-07 Figures/04_Generation_Figure4.R

@@ -21,7 +21,7 @@ library(tinter)
 input_path <- '../Tables' 
 
 # source common functions 
-source('./common_functions.R')
+source('../Figures/common_functions.R')
 
 ## ---------------------------------- ##
 ##  Figure 4.b

### 2025-03-08 Figures/PaperFiguresextended.Rmd
updated figure code

@@
 #Figure 2b epithelial cells
-Fig2b_2 <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
+Fig2b_2 <- DimPlot(epithelial_cells_monocle3, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
 Fig2b_2

### 2025-03-08 Figures/03_Generation_Figure3.R

@@ -22,22 +22,21 @@ library(ComplexHeatmap)
 library(gplots)
 library(RColorBrewer)
 library(randomcoloR)
+library(CellChat)
 
 dir_out <- "../_Data/Figures/"
-
 
 ## ---------------------------------- ##
 ## Figure 3.a
 ## ---------------------------------- ##
 
-##Figure 3a
-#ENDO/E044/A033/PaperFigures.Rmd
-#Clean up!! use DEG and GO table as basis for plots
 #Barplot unique genes per main cell type
-RNAlistRDS005_df %>% 
-  select(Symphony_Global_main5, gene, stage) %>%
+SupTable4 <- read.csv("./Tables/SupplementaryTable4.csv") 
+SupTable4$AnnotationMain <- gsub("myeloid|lymphoid", "immune", SupTable4$AnnotationMain)
+SupTable4 %>% 
+  select(AnnotationMain, gene, stage) %>%
   distinct() %>%
-  ggplot(aes(x = stage, fill = factor(Symphony_Global_main5, levels = c("immune", "endothelial", "epithelial", "stromal")))) +
+  ggplot(aes(x = stage, fill = factor(AnnotationMain, levels = c("immune", "endothelial", "epithelial", "mesenchymal")))) +
   geom_bar() +
   theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.3))
 ggsave(paste0(dir_out,"BarPlot_DEGs_main4.pdf"), 
@@ -48,27 +47,26 @@ ggsave(paste0(dir_out,"BarPlot_DEGs_main4.pdf"),
 ## Figure 3.b
 ## ---------------------------------- ##
 
-##Figure 3b
-#Clean up!! use DEG and GO table as basis for plots
-#ENDO/E044/A031/ExtractInterestingGenesFromGProfilerA031_RNA_AveExpr18_53202.Rmd
-gProfiler_df %>% 
+#Read in SupplementaryTable5
+SupTable5 <- read.table("./Tables/SupplementaryTable5.txt", sep = "\t", header = TRUE)
+
+#Filter and plot GO terms
+SupTable5 %>% 
   #dplyr::select(p_value, term_size, query_size, intersection_size, term_id, term_name) %>% 
   dplyr::filter(term_size > 25 & term_size < 750) %>% #larger terms are likely to be excessively broad and prone to false positives
[…]

### 2025-03-08 Figures/05_Generation_Supplementary_Figure1.R

@@ -16,14 +16,18 @@ library(tinter)
 library(UpSetR)
 library(forcats)
 
+# set dir_out
+dir_out <- "../_Data/Figures/"
 
 ## ---------------------------------- ##
 ##  Figure 1.a & b 
 ## ---------------------------------- ##
 
-#Supplementary Data Figure 1a and 1b
-#CODE OK
+#Load data
+EndoAtlas <- readRDS("../_Data/EndoAtlas.rds")
+EndoAtlas_meta <- EndoAtlas@meta.data
 
+#Plot and save
 VlnPlot(EndoAtlas,
         features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"),
         group.by='EndometriosisStatus', ncol = 1, pt.size=0)
@@ -58,56 +62,59 @@ median_summary <- EndoAtlas_meta %>%
   )
 
 ## ---------------------------------- ##
-##  Figure 1.c
+##  Supplementary Data Figure 1.c
 ## ---------------------------------- ##
 
-#Supplementary Data Figure 1c
-#CODE OK
+#Load data
+EndoAtlas <- readRDS("../_Data/EndoAtlas.rds")
 
+#Plot and save
 SFig1c <- FeaturePlot(EndoAtlas, features = "CycleDay", raster = FALSE)
+ggsave(filename = paste0(dir_out, 'SFig1c.pdf'),
+       plot = SFig1c,
+       width =20, height=9)
 
 ## ---------------------------------- ##
-##  Figure 1.d
+##  Supplementary Data Figure 1.d
 ## ---------------------------------- ##
 
-#Supplementary Data Figure 1d
-#CODE OK
+#Load data
+EndoAtlas <- readRDS("../_Data/EndoAtlas.rds")
 
+#Plot and save
 SFig1d <- FeaturePlot(EndoAtlas, features = "ProgesteroneSerum", raster = FALSE)
-
+ggsave(filename = paste0(dir_out, 'SFig1d.pdf'),
+       plot = SFig1d,
+       width =20, height=9)
 
 ## ---------------------------------- ##
-##  Figure 1.e
+##  Supplementary Data Figure 1.e
 ## ---------------------------------- ##
 
-dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")
-
+#Load data
 p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas_DEGsamples.rds"))
 
-#Figure 1e
+#Plot and save
[…]

### 2025-03-08 Figures/06_Generation_Supplementary_Figure2.R

@@ -22,33 +22,38 @@ input_path<- '../Tables'
 # source common functions 
 source('./common_functions.R')
 
+# set dir_out
+dir_out <- "../_Data/Figures/"
+
 ## ---------------------------------- ##
 ##  Supplementary Data Figure 2
 ## ---------------------------------- ##
 
-#CODE OK
+#Load data
+EndoAtlas <- readRDS("../_Data/EndoAtlas.rds")
 
 # Set SCT as default and remove RNA to save space
 DefaultAssay(EndoAtlas) <- "SCT"
 EndoAtlas@assays$RNA <- NULL
 gc()
+
 # Remove data without annotation (no NAs!)
 Idents(EndoAtlas) <- EndoAtlas$AnnotationMain
-EndoAtlas <- subset(EndoAtlas, subset = AnnotationMain %in% levels(as.factor(AnnotationMain$Symphony_Refined)))
+EndoAtlas <- subset(EndoAtlas, subset = AnnotationMain %in% levels(as.factor(EndoAtlas$AnnotationMain)))
 gc()
 
 ### AnnotationMain markers
 genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2") #correct xxx
 pops_to_plot <- c("stromal","epithelial","myeloid","lymphocyte","endothelial")
 # DotPlot
-Seurat::DotPlot(
+p_main <- Seurat::DotPlot(
   object = EndoAtlas,
   scale = T,
   features = genes_to_plot,
   idents = pops_to_plot,
 ) + theme(axis.text.x = element_text(angle = 90))
 # Save DotPlot
-ggsave(plot = last_plot(), "./SFig2_AnnotationMain.pdf", width = 10, height = 5)
+ggsave(plot = p_main, "../_Data/Figures/SFig2_AnnotationMain.pdf", width = 10, height = 5)
 
 
 ###Set identity to AnnotationRefined
@@ -62,14 +67,14 @@ pops_seurat <- subset(EndoAtlas, idents = pops_to_plot)
 pops_seurat@active.ident <- factor(pops_seurat@active.ident, 
                                    levels=pops_to_plot)
 # DotPlot
-p <- Seurat::DotPlot(
+p_DC <- Seurat::DotPlot(
[…]

### 2025-03-08 Figures/09_Generation_Supplementary_Figure5.R

@@ -4,6 +4,7 @@
 
 # load librairies
 library(ggplot2)
+library(CellChat)
 library(rlang)
 library(stringr)
 library(dplyr)
@@ -16,28 +17,95 @@ library(tinter)
 library(UpSetR)
 library(forcats)
 
+dir_out <- "../_Data/Figures/"
+
 ## ---------------------------------- ##
 ##  Supplemental Figure 5.a
 ## ---------------------------------- ##
 
-#Supplementary Data Figure 5a @Lea to do
+SupTable4 <- read.csv("../Tables/SupplementaryTable4.csv") 
+SupTable4$stage <- gsub("-", "_", SupTable4$stage)
+SupTable5 <- read.table("../Tables/SupplementaryTable5.txt", sep = "\t", header = TRUE)
+
+#Mutate so each row contains only a single gene in the "gene" column
+SupTable5 <- SupTable5 %>%
+  mutate(gene = strsplit(gene.intersection.between.DEGs.and.terms, ",")) %>%
+  unnest(gene) %>%
+  dplyr::rename(stage = grade) 
+
+#join tables
+dfFig5a <- left_join(SupTable5, SupTable4[,c("gene","logFC","AveExpr","p_adj.loc","cluster_id","stage")], by = c("gene","cluster_id","stage"))
+
+#Save vector with term names of interest
+term_name_oi <- c("Immune System","VEGFA-VEGFR2 signaling","focal adhesion","Focal adhesion","extracellular space","Chemokine signaling pathway","NF-kappa B signaling pathway","TGF-beta receptor signaling")
+
+#plot and save
+dfFig5a  %>%
+  dplyr::filter(term_name %in% term_name_oi) %>%
+  ggplot2::ggplot() + 
+  ggplot2::aes(y = reorder(gene, -log(logFC)), x = -log(p_adj.loc), color = stage, size = logFC) + 
+  ggplot2::geom_point() + 
+  ggplot2::scale_size_continuous("logFC") + 
+  ggplot2::scale_x_continuous(limits = c(1, 6)) +
+  ggplot2::theme_classic() +
+  ggplot2::facet_wrap(~term_name, scales = "free_y", ncol = 13) +
+  ggplot2::theme(strip.text = ggplot2::element_text(size = 8))
+ggsave(filename = paste0(dir_out, 'DotPlot_DEGcategories_DEGs_logFC_pval.pdf'), width =25,height=12)
[…]

### 2025-03-08 Figures/10_Generation_Supplementary_Figure6.R

@@ -15,22 +15,28 @@ library(tibble)
 library(tinter)
 library(UpSetR)
 library(forcats)
+library(reshape2)
 
 # input path 
-input_path<- '../Tables' 
-  
+input_path <- '../Tables/' 
+
+#output path
+dir_out <- "../_Data/Figures/"
+
 # source common functions 
 source('./common_functions.R')
 
 ## ---------------------------------- ##
-##  Figure 6.a
+##  Supplementary Data Figure 6.a
 ## ---------------------------------- ##
 
-# @Lea to check 
-#Supplementary Data Figure 6a
+#Load data
+ScaiNET_meta <- read.csv(file.path(input_path,'ScaiNET_metadata_prolif_folds.csv.csv')) 
+
 # reshape df
-melted_meta <- melt(meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
-## create the bar plot for endo_EndometriosisGrade
+melted_meta <- melt(ScaiNET_meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
+
+# barplot and save
 SFig6a <- ggplot(melted_meta, aes(x = variable, fill = endo_grade)) +
   geom_bar(position = "stack", width = 0.7) +
   facet_wrap(~ fct_rev(value), ncol = 2) +
@@ -40,48 +46,49 @@ SFig6a <- ggplot(melted_meta, aes(x = variable, fill = endo_grade)) +
   scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
   theme_minimal() +
   coord_flip()
+ggsave(filename = paste0(dir_out, 'SFig6a.pdf'),
+       plot = SFig6a,
+       width =12, height=7)
 
 ## ---------------------------------- ##
-##  Figure 6.b
+##  Supplementary Data Figure 6.b
 ## ---------------------------------- ##
+#Load data
+ScaiNET_meta <- read.csv(file.path(input_path,'ScaiNET_metadata_prolif_folds.csv.csv')) 
 
-# @lea to check 
-#Supplementary Data Figure 6b
+# reshape df
+melted_meta <- melt(ScaiNET_meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
+
+# barplot and save
[…]

### 2025-03-08 Figures/11_Generation_Supplementary_Figure7.R

@@ -22,41 +22,28 @@ input_path <- '../Tables'
 source('./common_functions.R')
 
 ## ---------------------------------- ##
-##  Figure 7.a
+##  Supplementary Data Figure 7.a
 ## ---------------------------------- ##
 
-##@Lea check 
-#LTs annotation for cycling fibroblasts
-data <- readRDS("/home/common/data/output/projects/ENDO/E031/A004/proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated.RDS")
-meta <- data@meta.data
-rm(data)
-gc()
-table(meta$Annotation, meta$EndometriosisStatus)
-
-#Make cell id compatible with new data set
-meta$cell_id <- paste(meta$sample, meta$cell.names, sep = '_')
-rownames(meta) <- meta$cell_id
-
-#Load entire dataset and add LTs annotation
-data_ds <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
-data_ds <- AddMetaData(data_ds,meta[,c("Annotation", "scds.cxds.score")])
-DimPlot(data_ds_new,group.by = "Annotation")
+#Load data
+EndoAtlas <- readRDS("../_Data/EndoAtlas.rds")
 
 #Highlight cycling fibroblasts
-Idents(data_ds) <- data_ds$Annotation
-table(data_ds$Annotation)
-cyclingfibs <- WhichCells(data_ds, idents = c("Cycling Fibroblasts"))
-DimPlot(data_ds, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
-
+Idents(EndoAtlas) <- EndoAtlas$cycling.fibroblast
+cyclingfibs <- WhichCells(EndoAtlas, idents = "TRUE")
+SFig7a <- DimPlot(data_ds, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
+ggsave(filename = paste0(dir_out, '../_Data/Figures/SFig7a.pdf'),
+       plot = SFig7a,
+       width =20, height=9)
 
 ## ---------------------------------- ##
-##  Figure 7.b
+##  Supplementary Data Figure 7.b
 ## ---------------------------------- ##
 
 # this figure represent a scheme and was done manually 
 
 ## ---------------------------------- ##
-##  Figure 7.c
[…]

### 2025-03-09 02_Cell_annotation/01_symphony.R
Added description on how I downloaded and converted the Tan 2022 ref files

@@ -10,9 +10,38 @@ suppressPackageStartupMessages({
 #source utils
 source('../02_Cell_annotation/01_utils.R') # color definitions, symphony function and plotting functions
 
+########################################################
+##### 01. Download data from Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5) as annotation reference
+#######################################################
+
+####### Download data in the command line
+#cd ../_Data/02_Cell_annotation/
+#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_global.h5ad
+#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_endothelial.h5ad
+#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_epithelial.h5ad
+#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_lymphocyte.h5ad
+#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_myeloid.h5ad
+#wget --no-check-certificate https://singlecell.jax.org/datasets/endometriosis/h5ad/endo-2022_stromal.h5ad
+
+####### Convert .h5ad files to .h5seurat files and save as .rds
+library(SeuratDisk)
+
+# List of cell types
+cell_types <- c("global", "lymphocyte", "epithelial", "endothelial", "myeloid", "stromal")
+
+# Convert each file to h5seurat format and save as RDS
+for (cell_type in cell_types) {
+  # Convert to h5seurat
+  Convert(paste0("endo-2022_", cell_type, ".h5ad"), dest = "h5seurat", assay = "RNA", overwrite = TRUE)
+  
+  # Load data and save as RDS
+  data <- LoadH5Seurat(paste0("endo-2022_", cell_type, ".h5seurat"), assays = "RNA")
+  saveRDS(data, paste0("endo-2022_", cell_type, ".rds"))
+}
+
 
 ########################################################
-##### 01. Run Symphony on entire EndoAtlas
+##### 02. Run Symphony on entire EndoAtlas
 #######################################################
 
 #create first following symbolic link in Linux environment:
@@ -26,7 +55,7 @@ symphony(
 
 
 ########################################################
-##### 02. Save EndoAtlas split into the 5 main cell types
+##### 03. Save EndoAtlas split into the 5 main cell types
 #######################################################
 
[…]

### 2025-03-09 Figures/06_Generation_Supplementary_Figure2.R
removed unnecessary comment

@@ -38,7 +38,7 @@ EndoAtlas <- subset(EndoAtlas, subset = AnnotationMain %in% levels(as.factor(Ann
 gc()
 
 ### AnnotationMain markers
-genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2") #correct xxx
+genes_to_plot <- c("EPCAM","CDH1","PDGFRA","COL1A1", "VWF","PECAM1","PTPRC","CD68","CD14","CD2")
 pops_to_plot <- c("stromal","epithelial","myeloid","lymphocyte","endothelial")
 # DotPlot
 Seurat::DotPlot(

### 2025-03-10 Figures/10_Generation_Supplementary_Figure6.R
changed referencing of scainet to scaivision, corrected minor bugs and improved axis labelling

@@ -16,6 +16,7 @@ library(tinter)
 library(UpSetR)
 library(forcats)
 library(reshape2)
+library(ComplexHeatmap)
 
 # input path 
 input_path <- '../Tables/' 
@@ -31,18 +32,18 @@ source('./common_functions.R')
 ## ---------------------------------- ##
 
 #Load data
-ScaiNET_meta <- read.csv(file.path(input_path,'ScaiNET_metadata_prolif_folds.csv.csv')) 
+ScaiVision_meta <- read.csv(file.path(input_path,'ScaiVision_metadata_prolif_folds.csv')) 
 
 # reshape df
-melted_meta <- melt(ScaiNET_meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
+melted_meta <- melt(ScaiVision_meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
 
 # barplot and save
 SFig6a <- ggplot(melted_meta, aes(x = variable, fill = endo_grade)) +
   geom_bar(position = "stack", width = 0.7) +
   facet_wrap(~ fct_rev(value), ncol = 2) +
-  labs(title = "Bar Plot of Samples by Endo EndometriosisGrade",
+  labs(title = "Bar Plot of Samples by Endometriosis Grade",
        x = "Fold",
-       y = "Number of Samples") +
+       y = "Number of Samples",fill="endometriosis status") +
   scale_fill_manual(values = c("severe" = "red", "mild" = "orange", "CTL" = "blue")) +
   theme_minimal() +
   coord_flip()
@@ -54,18 +55,19 @@ ggsave(filename = paste0(dir_out, 'SFig6a.pdf'),
 ##  Supplementary Data Figure 6.b
 ## ---------------------------------- ##
 #Load data
-ScaiNET_meta <- read.csv(file.path(input_path,'ScaiNET_metadata_prolif_folds.csv.csv')) 
+ScaiVision_meta <- read.csv(file.path(input_path,'ScaiVision_metadata_prolif_folds.csv')) 
 
 # reshape df
-melted_meta <- melt(ScaiNET_meta, id.vars = c("lane", "endo_grade"), measure.vars = c("fold1", "fold2", "fold3", "fold4", "fold5"))
[…]

### 2025-03-10 Figures/11_Generation_Supplementary_Figure7.R

@@ -31,7 +31,7 @@ EndoAtlas <- readRDS("../_Data/EndoAtlas.rds")
 #Highlight cycling fibroblasts
 Idents(EndoAtlas) <- EndoAtlas$cycling.fibroblast
 cyclingfibs <- WhichCells(EndoAtlas, idents = "TRUE")
-SFig7a <- DimPlot(data_ds, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
+SFig7a <- DimPlot(EndoAtlas, reduction = "umap", cells.highlight = cyclingfibs, cols.highlight = c("darkred"), cols= "grey", raster = FALSE)
 ggsave(filename = paste0(dir_out, '../_Data/Figures/SFig7a.pdf'),
        plot = SFig7a,
        width =20, height=9)
